### 2016-11-21 ms/valley_crossing.tex
initial code and writeup

%\documentclass[rmp,twocolumn]{revtex4}
\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{color}
\usepackage{breqn}
\usepackage{graphicx}
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{enumerate}

% Code syntax highlighting
\usepackage{listings}
\lstloadlanguages{C++,Python}

\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
\newcommand{\REF}[1]{ref.~\citep{#1}}

\newcommand{\comment}[1]{{\color{red}#1}}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.42}

\DeclareMathOperator*{\argmin}{arg\,min}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Genetic draft and the evolution of ``irreducible complexity"}
%\author{Taylor~Kessinger}
%\author{Jeremy~Van~Cleve}
%\affiliation{Department of Biology, University of Kentucky}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

Living systems are characterized by complex adaptations, at least some of which have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
Such adaptations have been termed ``irreducibly complex", and the process by which they evolve is known as ``crossing a fitness valley".
Previous efforts have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate of individual mutations.
We investigate the rate at which ``irreducibly complexity" evolves in these rapidly adapting populations, including both asexual and sexual populations.
We confirm that, as in neutrally evolving populations, recombination can aid in the crossing of deep fitness valleys but find that its effects are mitigated.
\end{abstract}

\maketitle

\section*{Introduction}

The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
This appears to be how Darwin thought most complex systems, such as the vertebrate eye, evolved (Darwin 1859), and this assumption underlies much of the field of adaptive dynamics.
However, many adaptations exist that do not appear to have formed by this process.
One class of such adaptations has been termed ``irreducibly complex" because, in their current forms, they cease to function when one or more of their parts is removed.
Muller (1918) described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other, giving the appearance of irreducibility.
An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a lower fitness mutational ``valley" in order to reach a higher fitness ``peak".
We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation (Trotter et al. 2013).

The rate of valley crossing in asexual populations is increasingly well understood (Weissman 2009).
If the number of mutants produced every generation is high, the arisal of a complex adaptation that is destined to fix may be guaranteed.
Otherwise, wild type individuals may give rise to neutral or deleterious intermediate mutants, which drift to fixation.
A third possibility is for wild type individuals to birth mutant ``bubbles", or transient subpopulations.
These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation.
These models were successfully extended to sexually reproducing populations by Weissman et al. (2010), which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.

Previous studies of valley crossing focused exclusively on populations where the background fitness variation is small enough to be negligible.
In these populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift.
In a drafting asexual population, the fate of an allele depends sharply on its genetic background.
Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
In sexual populations, much the same description applies.
The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ (Neher and Shraiman 2011) or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within a small block length (Neher et al. 2013).

Drift and draft are fundamentally different forces.
It is not possible to rescale a drafting population, for example by defining a reduced ``effective population size".
Drafting populations do not admit a diffusion approximation, and they exhibit qualitatively different genealogies and site frequency spectra relative to a drifting population.
In fact, the possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
It is reasonable to suspect that draft might likewise have profoundly different effects on the evolution of complex adaptations, a notion that was explored in Neher and Shraiman (2011).
We investigate this possibility using analytical approaches and forward simulations, confirming that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.

\end{document}

### 2016-11-21 writeup/writeup.tex
initial code and writeup

\documentclass[10pt]{revtex4}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{mathtools}

\begin{document}

The dynamics of complex adaptations, such as the probability that a particular complex adaptation will appear and its time to fixation, are increasingly well understood.
However, traditional results only apply in populations where genetic drift, not genetic draft, is the major force that shapes neutral variation.
As these are radically different forces (they result in completely different coalescent dynamics, and notably genetic draft does not allow for a diffusion approximation), we expect complex adaptations to behave differently in each situation.
Neher and Hallatschek (2013) suggested that the determining factor is the compound parameter $N\sigma$, where $\sigma$ is the standard deviation in fitness and $N$ is the population size.
When this product is much less than one, drift is more important.
When it is much greater than one, draft is more important.
We begin by outlining the results in the neutral (drift) case, as summarized by Weissman et al. (2009) and Weissman et al. (2010).
We then move on to discuss Neher et al. (2009), which summarized the dynamics of genetic draft in asexual populations, and Neher et al. (2013), which extended these results to sexual populations using a scaling argument.

Consider the case of a complex adaptation that requires two mutations, the first of which is neutral or deleterious and the second of which is beneficial, in the drift regime.
There are several ways for the complex adaptation to appear (after which it must drift to enough individuals to sweep through the population).
First, both mutations can appear in one individual (deterministic), which requires $N\mu^2 \sim 1$, with $\mu$ the mutation rate.
Otherwise, the intermediate mutation must appear and drift to an appreciable frequency, at which point the second mutation can arise on a background containing that intermediate mutation.
What constitutes an ``appreciable frequency" will depend on the population size $N$, the benefit $s$ associated with the adaptation, and the fitness penalty $\delta$ associated with the intermediate.
For small populations or small fitness effects (where $\delta < 1/N$), the intermediate is essentially neutral and can drift to more or less the entire population (neutral or sequential fixation).
Otherwise, the intermediate will appear and persist for short times, forming ``bubbles" that transiently persist at low frequencies (tunneling).
The size of these bubbles is determined by the fitness detriment and advantage of the intermediate and the adaptation, respectively (Weissman refers to different regimes herein as neutral semi-deterministic, neutral stochastic, and deleterious tunneling, respectively).
A final possibility is for the two mutations to arise on different genetic backgrounds, which recombine.
This can speed up the rate of valley crossing somewhat, provided that $s > r$, roughly speaking.
If this is true, the rate at which beneficial adaptations establish is higher than or at least comparable to the rate at which they are broken up by recombination.

We will outline this in a little more detail now.
Note that in the asexual case, the rate of valley crossing can be obtained by considering the establishment probability of the complex adaptation, which will depend (recursively) on the dynamics of each intermediate.

\subsection{Valley crossing in asexual populations}

We will briefly discuss the different possible parameter regimes and the valley crossing times that result therefrom.
We first consider the regime where $N\mu$ is small.
In this case, mutations are fairly rare: there will generally only be one intermediate mutation lineage at a time, and the first successful double mutant subpopulation that arises within this lineage determines the fixation probability.
The important number in this regime is the weight $W(t) = \int_0^t n(t^\prime) dt^\prime$, with $n(t)$ the number of mutant individuals at time $t$.
We also define the total weight $W = \int_0^T n(t)dt$, with $T$ the duration over which an intermediate ``bubble" persists.

When a beneficial intermediate arises within such a ``bubble" and later sweeps to fixation, this is called \emph{tunneling}. The beneficial intermediate then fixes with probability $\frac{1-\exp{-s}}{1-\exp{-Ns}} \approx s$. Hence $Ws\mu$ gives the fixation probability.

From diffusion theory, a neutral mutation will generally persist for about $T$ generations and reach a maximum frequency of $T$ individuals, with probability $1/T$.
The weight is therefore of order $T^2$.
Hence $T > 1/\sqrt{s\mu}$ is the condition for tunneling to be likely to occur, which implies that the fixation probability is roughly $p \sim \sqrt{s\mu}$. This is referred to as \emph{neutral tunneling}.

If the lineage is not neutral ($\delta > 0$), the dynamics are not necessarily much different.
Provided that $\delta < \sqrt{s\mu}$, the mutation will likely give rise to a beneficial subpopulation that is destined to fix well before it grows to order $1/\delta$ (at sizes smaller than this, its dynamics are effectively neutral).
Note that this differs from the condition for conventional neutrality, $\delta < 1/N$.
On the other hand, if $\delta > \sqrt{s\mu}$, then the mutant lineage's size is effectively capped at $1\delta$.
The total weight will be of order $1/\delta^2$, and the fixation probability is $p \sim s\mu/delta$. This is the \emph{deleterious tunneling} regime.

All of this assumes that the mutant subpopulation, while remaining small compared to $N$, can give rise to a beneficial mutant destined to fix: for effectively neutral mutations, this implies $1/\sqrt{s\mu} < N$.
If this condition does not obtain, then intermediate mutants can drift to fixation (with probability $1/N$) before the beneficial mutant arises and fixes.
This is the \emph{sequential fixation} regime, in which $p \sim 1/N$.
For deleterious mutations, the same rules apply, except that the condition is $1/\delta > N$ instead of $1/\sqrt{s\mu} > N$.

Finally, the preceding analysis has assumed that both intermediate and double mutants are sufficiently rare that only one such lineage exists at any given time. This assumption can be relaxed, though the results are not necessarily enlightening. The major effect is that as the number of mutations in the population increases, they will start to interfere with one another during expansion.

\subsection{More rigorous analysis}

The preceding argument has relied on estimating the average value of the weight $W$.
The true distribution of bubble sizes will determine the probability that a particular mutant $k$ is successful, however, via

\begin{equation}
p_k = \int_0^\infty dw P(W = w) (1-e^{u_k p_{k+1}}) = 1-\mathbb{E}(e^{u_k p_{k+1}}).
\end{equation}

This has the form of a Laplace transform, with

\begin{equation}
p_k = 1-\varphi(\mu_k p_{k+1})
\end{equation}
and
\begin{equation}
\varphi(y) = \mathbb{E}(e^{-yW}).
\end{equation}
Calculating $\varphi$ is difficult because $W$ is not a Markov random variable. 
However, we can instead consider the variable $n(t), W(t)$ and compute the transform $\Phi (x,y,t) = \mathbb{E}(e^{-xn-yW})$.
Evaluating it at $x=0$ will average over all values of $n$ and return the Laplace transform $\phi$ for the weight.
We proceed by considering the growth of a single mutant lineage as a branching process.
The time evolution of $p(n,w)$, the probability that there are $n$ individuals and a total weight $w$ at a particular time $t$, is given by the master equation
\begin{equation}
p_{t+dt}(n,w) = (n+1)p_t(n+1,w)dt + (n-1)(1-\delta)p_t(n-1,w)dt + (1-(2-\delta)ndt)p_t(n,w-ndt).
\end{equation}
Using the relation $p(n,w-ndt) = p(n,w)-ndt\partial_w p_t(n,w)$ allows us to rewrite this as the partial differential equation
\begin{equation}
\partial_t p_t (n,w) = (n+1)p_t(n+1,w) + (n-1)(1-\delta) p_t(n-1,w) - n(2-\delta)p_t(n,w)-n\partial_w p_t(n,w).
\end{equation}
The transform $\Phi(x,y,t) = \sum_n^\infty \int_{-\infty}^\infty dw p_t(n,w) e^{-xn-yw}$.
Differentiating both sides and substituting the above yields a PDE for $\Phi$, which can be arranged into
\begin{equation}
\partial_t \Phi = -e^x - (1-\delta e^{-x} + (2-\delta) + y)\partial_x \phi .
\end{equation}
This can be solved using the characteristic equation $\frac{dx}{dt} = e^x + (1-\delta)e^{-x} - 2+\delta-y$: the requirement that the lineage starts at $t=0$ with one individual means $p_0(n,w) = \delta_{1,n} \delta (w)$, with $\delta$ the Kronecker and Dirac deltas respectively.
The boundary condition for $\Phi$ becomes $\Phi(x,y,0) = e^{-x}$. Solving for $\Phi$, then for $\phi(y,t) = \Phi(0,y,t)$, and finally for $\varphi = \lim_{t\to \infty} \phi (y,t)$ (the Laplace transform for the total bubble size) yields:
\begin{equation}
\varphi(y) = \frac{2-\delta+y-\sqrt{(2-\delta+y)^2 - 4(1-\delta)}}{2(1-\delta)}.
\end{equation}
This expression allows us to derive the ``weight" for each individual mutant bubble.
We have assumed that the population size is large enough that transient bubble sizes don't ``butt up" against the population size. This assumption can be relaxed but is (to my mind) not particularly informative.

\subsection{Valley crossing in sexual populations}

In sexual populations, recombination can affect the valley crossing rate in one of two basic ways.
First, recombination can increase the total number of double mutants that appear by causing deleterious single mutants to appear on the same genetic background.
Second, recombination can slow the expansion of beneficial double mutants by forcing them to recombine with (mostly wild type) individuals.
We will briefly consider some basic parameter regimes, then explore the analytics.

First, suppose $r \gg s, \delta$.
In this limit, the single mutants will be present at roughly linkage equilibrium: when their frequencies $x_a, x_b$ satisfy $(s+2\delta)x_ax_b - \delta(x_a+x_b) > 1/N$, the double mutant will begin to spread deterministically in the population (selection, the left side, will overwhelm drift, the right side).
If selection against single mutants is strong, i.e., $\delta^2/s \gg \max (\mu, 1/N)$, the rate at which a population reaches this threshold will be sharply reduced.
In linkage equilibrium, the frequencies will be distributed roughly as
\begin{equation}
P(x_a,x_b) \sim \exp(-N(\delta(x_a+x+b) - sx_ax_b).
\end{equation}
The population must drift along the ridge where $x_a = x_b$, which will take approximately $\tau \sim \exp(N\delta^2/s)$ generations: this is because the population must drift to the unstable saddle point at $x_a = x_b \approx \delta/s$ and then ride selection to the stable equilibrium at $x_a = x_b = 1$, and the exponent roughly sets the probability of making it to the saddle point from $x_a = x_b \approx 0$.

Second, suppose $r \ll s$.
A successful double mutant grows initially at a rate $s - r = \bar{s}$, which roughly sets the fixation probability.
If $N\mu$ is large enough that single mutants are produced constantly, then the appearance of the single mutant is essentially deterministic, as in the asexual case.
If not, then we must consider the branching process dynamics, but the wait time is dominated by the waiting time to the first successful single mutant (i.e., the first one destined for fixation), which is distributed as $\tau \approx 2N\mu/p$, with $p$ the success probability.
$p$ can be decomposed into a probability that the mutation is successful due to successive mutations (something like tunneling, almost identical to the asexual case) or due to recombination between another nascent mutant subpopulation.

We consider only the sexual case.
During the time $T$ over which a bubble of genotype $Ab$ persists, a number of individuals $\approx N\mu T$ $aB$ individuals arise.
The success probability is therefore related to $N\mu T \int P_{r,T^\prime} \mathrm{prob} (T^\prime) dT^\prime$, where the last term corresponds to the probability density $\mathrm{prob}(T^\prime)$ of wait times for the $aB$ lineage, times the probability $P_{r,T^\prime}$ of success contingent on $T$.
The number of $AB$ lineages produced by recombination is of order $r/N \int n_{Ab}(t) n_{aB}(t) dt$, and the lineages reach sizes of $T$ and $T^\prime$ respectively, persisting for $T^\prime$ generations: so the total integral is simply $TT^{\prime 2}$.
The total success probability $P_{r,T,T^\prime}$ therefore grows as $N\mu TT{^\prime 2}$ until it saturates near $1$ at $T^\prime \sim \sqrt{Nr\bar{s}}$.

The preceding analysis has assumed implicitly that $\delta$ is somewhat small.
If this assumption is relaxed, then $T$ and $T^\prime$ are roughly bounded by $1/\delta$, as in the asexual case.
Interestingly, this causes the valley crossing time to be reduced by a much greater degree in the sexual case than in the asexual case, meaning that recombination facilitates the crossing of deep valleys.

\subsection{Valley crossing when $N\sigma \gg 1$}

The past sections have assumed that the distribution of fitnesses in the population is roughly equal, so that the primary force governing the behavior of neutral alleles is genetic drift.
In this situation, standard diffusion results (including fixation probabilities) apply, and there is essentially no effect arising from genetic backgrounds of differential fitness.
The condition $N\sigma \ll 1$ is sufficient, where $\sigma$ is the standard deviation in fitness.

If, on the other hand, we have $N\sigma \gg 1$, then the fate of any individual mutation will depend sharply on the genetic background on which it arises.
Mutations that arise on unfit backgrounds may be doomed to extinction, and mutations that arise on fit backgrounds may be destined for fixation even if their are themselves deleterious.
If fitness variation is due to the effects of many weak loci, the population takes the form of a Gaussian traveling ``wave", and the mean fitness $\bar{x}$ advances due to selection at a rate $\dot{\bar{x}} = \sigma^2$.
This advance is determined by the dynamics of the nose, where deviations from Gaussianity become important (due to the discrete number of individuals).
Below some cutoff point $\Theta$, the establishment probability is essentially zero: above this cutoff point it starts to increase roughly as $P_\mathrm{fix} \sim x+s-r-\bar{x}$, so that the genetic background fitness $x$ may dominate. This may appear to increase markedly the probability of a successful double mutant, but the size of mutant bubbles will tend to be smaller due to the advancing mean fitness $\bar{x}$.

We will focus solely on the valley crossing problem here.
As in the sexual case, the birth of a double mutant can be interpreted as a Poisson process, so that
\begin{equation}
P(T) = e^{-\mu \int_0^T (n_1(t) + n(2)t) dt}
\end{equation}
gives the probability that a double mutant has not yet arisen.
The time integrated bubble size (weight) then provides the number of chances for a successful double mutant to arise.
This must be obtained by considering the dynamics of bubbles that arise on any of the genetic backgrounds $x$ in the population.
As argued above, this is effectively negligible below a certain cutoff.
The calculation still needs to be filled in, but the Laplace transform for the bubble size $\Phi(z) = \mathbb{E}(e^-zw)$ is given by
\begin{equation}
\Phi(z) =
\begin{cases}
ze^{r\sigma^{-1}\sqrt{-2\log z}} &|s| \ll re^{-r\sigma^{-1}\sqrt{-2\log z}} \\
\frac{z}{|s|} &|s| \gg re^{-r\sigma^{-1}\sqrt{-2\log z}}.
\end{cases}
\end{equation}
It is worth noting that $r$ here is the total recombination rate between two arbitrary loci, not necessarily between the two loci of interest.
In the draft dominated regime, we can use the Laplace transform to compute the weight distribution, which scales as $P(w)\sim 1/w^2 \sqrt{2\log w}$, compared to $P(w) \sim w^{-3/2}$ in the drift case.
As expected, draft decreases the bubble size distribution but lessens the dependence on $\delta$.
Some more precise analytics will be needed here.

\subsection{Valley crossing when $N\sigma \gg 1$ in the high recombination limit}

At sufficiently high recombination rates $r$ between loci, the concept of those loci segregating within a common fitness wave will start to break down.
In fact high enough recombination can effectively cause different parts of the genome to evolve neutrally even though the rate of adaptation is high.
This is tantamount to a return to drift dominance.
The relevant parameter turns out to be $N\sigma_b$, where $\sigma_b$ is the proportion of $\sigma$ that segregates within an ``effectively asexual" block: such a block will be large enough that recombination effectively ``cancels out" the amplification of fit variants due to selection.

We proceed by identifying the block length.
If $N\sigma \gg 1$, the fittest individuals are the only ones whose lineages stand a good chance of surviving.
They are roughly $x_c = \sigma \sqrt{2 \log N\sigma}$ ahead of the mean, and their lineages will take approximately $\sigma^{-1} \sqrt{2 \log N\sigma}$ generations to dominate the population.
Thus, two random individuals had a common ancestor roughly $\sigma^{-1} \sqrt{2 \log N\sigma}$ generations ago with probability of order $1$.
On the other hand, if $N\sigma \ll 1$, coalescence is dominated by genetic drift and takes approximately $N$ generations. So the mean pair coalescence time is given by
\begin{equation}
\left< T_2 \right> \approx
\begin{cases}
N & \mathrm{if~} N\sigma \ll 1, \\
c\sigma^{-1}\sqrt{2 log N\sigma} & \mathrm{if~} N\sigma \gg 1.
\end{cases}
\end{equation}

Let $\xi$ be the characteristic distance over which loci share most of their history, which is the same as the appropriate block length.
$\xi$ will decrease due to recombination:
\begin{equation}
\xi (t) = \frac{L}{1 + L\rho t} \approx \frac{1}{\rho t}.
\end{equation}
A block of the genome will harbor a fraction $\sigma_\xi^2$ of the total fitness variance, assuming that fitness variation is distributed roughly evenly throughout the genome.
For a block of length $\xi$, this is
\begin{equation}
\sigma^2_\xi = \sigma^2 \frac{\xi}{L}.
\end{equation}
The relevant $\sigma_\xi$ is the amount of fitness variation $\sigma_b$ that segregates in a block of length $\xi_b$ that is unlikely to be broken up during the coalescence timescale.

If fitness variation in the block is substantial (i.e., $N\sigma_b \gg 1$), coalescence in this part of the genome will occur in $\left< T_2 \right> = c \sqrt{2\log N\sigma_b}/\sigma_b$ generations.
Hence
\begin{equation}
\xi_b = \frac{\sigma_b}{c\rho\sqrt{2\log N\sigma_b}},
\end{equation}
or equivalently
\begin{equation}
\sigma_b = \frac{\sigma^2}{L\rho c \sqrt{2 \log N\sigma_b}}
\end{equation}
and
\begin{equation}
\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}.
\end{equation}

If, on the other hand, coalescence in this block is \emph{not} primarily due to exponential amplification of fit lineages but rather due to drift processes, then we recover $\xi_b \sim (N\rho)^{-1}$, as expected under genetic drift.
$\xi$ sets the characteristic distance over which linkage disequilibrium in the genome decays exponentially.
Within blocks shorter than $\xi$, most loci tend to share most of their genetic history, and LD remains elevated.
Sexual populations thereby behave quite similarly to asexual ones, with $\sigma$ being replaced by $\sigma_b$.
When $N\sigma_b \ll 1$, the Kingman limit applies, and when $N\sigma_b \gg 1$, the BSC arises in the genetic block under consideration.

We might therefore immediately consider two kinds of complex adaptation: one where the two loci occupy part of the same ``fitness wave" (because the distance between them is shorter than $\xi_b$) and one where they do not (and the recombination rate between them is arbitrarily large).
We can immediately make two predictions, which still need a little bit of theory to back them up.
In the first case, the valley crossing dynamics should be comparable to the ``singular traveling wave" case, with $\sigma_b$ substituted instead of $\sigma$.
In the second case, we probably need to do a little bit more thinking.
The traveling wave model should still set the fixation probability for each individual mutation, but then we need to consider a model similar to the $r > s$ limit in Weissman (2010) in order to obtain the full probability and wait time for the complex adaptation.

\end{document}

### 2016-11-21 src/simulate_population.py

@@ -4,7 +4,6 @@ simulate_population.py
 Author: Taylor Kessinger
 Date: September 24, 2016
 Description: Simulation of valley crossing using FFPopSim.
-    A number of modifications to FFPopSim still TODO.
     Also TODO: direct calculation of Nsigma_b.
 '''
 import sys
@@ -13,30 +12,52 @@ import FFPopSim as h
 import random as rand
 import matplotlib.pyplot as plt
 from time import time
+import argparse
+import math
+import cPickle as pickle
 
 t0 = time()
 
+parser = argparse.ArgumentParser()
+parser.add_argument('--N', type=int, default=10000)
+parser.add_argument('--s', type=float, default=0.1)
+parser.add_argument('--delta', type=float, default=0.1)
+parser.add_argument('--rho', type=float, default=0.0)
+parser.add_argument('--sigma', type=float, default=0.1)
+parser.add_argument('--mu', type=float, default=0.001)
+parser.add_argument('--numsamples',type=int,default=10)
+parser.add_argument('--runno', type=int, default=0)
+parser.add_argument('--name', type=str, default='test')
+
+args=parser.parse_args()
+
 L = 300 #number of loci. 1000 is about as many as we're likely to need.
-N = 10000 #population size. recall that Nsigma is the main parameter.
-mu = 0 #with all_polymorphic, mutation rate is irrelevant anyway
-sigma = 0.05 #fixed fitness variance
-rho = 0.001 #outcrossing rate
-cx = 0.001 #per-locus crossover rate
+N = args.N #population size. recall that Nsigma is the main parameter.
+mu = args.mu #per site mutation rate for focal mutations
+sigma = args.sigma #fitness variance
+rho = args.rho #per generation outcrossing rate
+cx = 0 #per locus crossover rate
 #unfortunately, nonzero values of rho and cx increase the runtime somewhat
 
-fitness_check = N/2 #for debugging: number of generations to let pass before checking the gaussianity of the fitness distribution
+s = args.s
+delta = args.delta
 
-mutant_locs = np.array([int(L*0.35), int(L*0.65)]) #locations of the focal mutations: this should avoid "edge" effects
-deleterious_effects = [-0.001,-0.001] #fitness costs of single mutants
-beneficial_effect = 0.001 #fitness advantage of double mutant
[…]

### 2016-11-21 submit/valley_crossing_submit.py
submit script

@@ -1,0 +1,53 @@
+#!/usr/bin/env python
+
+# Set your minimum acceptable walltime, format: day-hours:minutes:seconds
+#SBATCH --time=0-00:30:00
+
+# Set name of job shown in squeue
+#SBATCH --job-name src/branching_cluster.py
+
+# Request CPU resources
+#SBATCH --ntasks=1
+#SBATCH --ntasks-per-node=1
+#SBATCH --cpus-per-task=1
+'''
+valley_crossing_submit.py
+Author: Taylor Kessinger
+Date: November 19, 2016
+Description: Submit script for valley crossing simulations.
+'''
+
+import os
+import shutil
+
+s_range = {}
+alpha_range = {}
+numruns = {}
+numsims = {}
+sigma_range['161120'] = [0.00001,0.00003,0.0001,0.0003,0.001,0.003,0.01,0.03,0.1] #range of sigma values
+rho_range['161120'] = [0] #range of recombination rates
+N_range['161120'] = [10000] #range of population size values
+delta_range['161120'] = [0.00001,0.00003,0.0001,0.0003,0.001,0.003,0.01,0.03,0.1] #range of deleterious intermediate coefficients
+s_range['161120'] = [0.00001,0.00003,0.0001,0.0003,0.001,0.003,0.01,0.03,0.1] #range of selection coefficients for the adaptation
+#we will need to think carefully about which ``slices" of parameter space we can explore.
+numruns['161120'] = 1
+numsamples['161120'] = 10
+
+name = '161120'
+
+sigma_range, rho_range, N_range, delta_range, s_range, numruns, numsamples = sigma_range[name], rho_range[name], N_range[name], delta_range[name], s_range[name], numruns[name], numsamples[name]
+
+if not os.path.exists('output/valley_crossing_sims_'+name):
+    os.makedirs('output/valley_crossing_sims_'+name)
+
+shutil.copy('src/simulate_population.py', 'output/valley_crossing_sims_'+name)
+
+for s in s_range:
+    for N in N_range:
+        for runno in range(numruns):
+            call_list = ['sbatch', 'src/valley_crossing.py', '--name', name, '--s', str(s), '--N', str(N), '--numsamples', str(numsamples),
+             '--runno', str(runno)]
+            call = ' '.join(call_list)
+            print call
+            #sp.call(call)
+            os.system(call)

### 2016-11-22 FFPopSim/doc/cpp/mainpage.cpp
FFPopSim modifications

@@ -1,0 +1,54 @@
+/**
+ * @file mainpage.cpp
+ * @brief Main page for the documentation.
+ * @author Richard Neher, Boris Shraiman, Fabio Zanini
+ * @version 
+ * @date 2012-04-19
+ *
+ * @mainpage FFPopSim Documentation
+ *
+ * @image html quote.png
+ *
+ * FFPopSim is a C++/Python library used to simulate evolution of genome populations. This is the home page of the C++ documentation.
+ *
+ * Two kinds of simulations are supported, depending on their genome size:
+ * - few loci (up to ~20): the *lowd classes and functions, mainly `haploid_lowd` and `hypercube_lowd`.
+ *   The full space of possible genomes is generated and simulations can be performed with any population size.
+ *
+ * - many loci (> 20, tested up to several tens of thousands): the *highd classes and functions, mainly `haploid_highd` and `hypercube_highd`.
+ *   Only the observed genotypes are monitored. If a new genotype is created either by mutation or by recombination, a more resources are required.
+ *   Population sizes up to \f$10^6\f$ can be modeled this way. Furthermore, genealogies of single loci can be tracked, and trees can be exported
+ *   to standard formats at the end.
+ *
+ * In addition, FFPopSim includes a specialized subclass of `haploid_highd` used for simulating evolution of the Human Immunodeficiency Virus (HIV),
+ * `hivpopulation`. Viruses have a fixed genome size of 10000, and default constructor values are offered to start reasonable simulations quickly.
+ * Furthermore, helper I/O routine are included.
+ *
+ * @section tests-examples Tests and Examples
+ *
+ * Examples and tests are found in the `tests` folder. They include a few functions that show that the Fast Fourier Transform-based algorithm for
+ * recombination yields the same result as the more expensive naive algorithm.
+ *
+ * @section compiling Compiling and Linking against FFPopSim
+ *
+ * To build the C++ part of FFPopSim, refer to the INSTALL file.
+ * Building your own programs using FFPopSim is then a two-step process, @em compiling @em and @em linking @em.
+ * For the compile step, you need to import the header(s) you are using, like in the example above in this page. The FFPopSim headers include recursively all sublibraries such as GSL. Moreover, the compiler must be told where those headers are located; a typical compile call using g++ is the following:
+ * \verbatim
+ g++ -c -I../pkg/include/ -o example_lowd.o example_lowd.cpp
+ \endverbatim
+ *
[…]

### 2016-11-22 FFPopSim/doc/python/source/conf.py

@@ -1,0 +1,251 @@
+# -*- coding: utf-8 -*-
+#
+# FFPopSim documentation build configuration file, created by
+# sphinx-quickstart2 on Wed Aug 22 09:37:23 2012.
+#
+# This file is execfile()d with the current directory set to its containing dir.
+#
+# Note that not all possible configuration values are present in this
+# autogenerated file.
+#
+# All configuration values have a default; values that are commented out
+# serve to show the default.
+
+import sys, os
+
+# If extensions (or modules to document with autodoc) are in another directory,
+# add these directories to sys.path here. If the directory is relative to the
+# documentation root, use os.path.abspath to make it absolute, like shown here.
+sys.path.insert(0, os.path.abspath('../../../pkg/python'))
+sys.path.append(os.path.abspath('../sphinxext'))
+
+# -- General configuration -----------------------------------------------------
+
+# If your documentation needs a minimal Sphinx version, state it here.
+#needs_sphinx = '1.0'
+
+# Add any Sphinx extension module names here, as strings. They can be extensions
+# coming with Sphinx (named 'sphinx.ext.*') or your custom ones.
+extensions = ['sphinx.ext.autodoc',
+              'sphinx.ext.autosummary',
+              'sphinx.ext.todo',
+              'sphinx.ext.mathjax',
+              'sphinx.ext.ifconfig',
+              'ipython_console_highlighting',
+             ]
+
+# Add any paths that contain templates here, relative to this directory.
+templates_path = ['_templates']
+
+# The suffix of source filenames.
+source_suffix = '.rst'
+
+# The encoding of source files.
+#source_encoding = 'utf-8-sig'
+
+# The master toctree document.
+master_doc = 'index'
+
+# General information about the project.
+project = u'FFPopSim'
+copyright = u'2012-2013, Fabio Zanini, Richard Neher'
+
+# The version info for the project you're documenting, acts as replacement for
+# |version| and |release|, also used in various other places throughout the
+# built documents.
+#
+# The short X.Y version.
+version = '2.0'
+# The full version, including alpha/beta/rc tags.
+release = '2.0'
+
+# The language for content autogenerated by Sphinx. Refer to documentation
+# for a list of supported languages.
+#language = None
+
+# There are two options for replacing |today|: either, you set today to some
+# non-false value, then it is used:
+#today = ''
+# Else, today_fmt is used as the format for a strftime call.
+#today_fmt = '%B %d, %Y'
+
[…]

### 2016-11-22 FFPopSim/examples/benefits_of_sex.py

@@ -1,0 +1,88 @@
+'''
+author:     Richard Neher, Fabio Zanini
+date:       11/07/12
+content:    Compare the speed of evolution at different recombination rates.
+            Sexual populations reach higher fitnesses than asexual ones.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import argparse
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as ffpop
+
+
+# parse the command line arguments
+parser = argparse.ArgumentParser(description="compare the speed of evolution at different recombination rates")
+parser.add_argument('--pop', default=10000, type=float, help='Population size (N)')
+parser.add_argument('--mut', default=0.0001,type=float, help='mutation rate')
+parser.add_argument('--sel', default=0.01,type=float, help='effect size of beneficial mutations')
+parser.add_argument('--Ttraj', default=1000,type=int, help='Length of trajectory in generations')
+parser.add_argument('--dt',  default=10,type =int, help='time increments of trajectory')
+params=parser.parse_args()
+
+# set up the population
+L = 256                                         # number of loci
+pop=ffpop.haploid_highd(L)                      # produce an instance of haploid_highd
+pop.recombination_model = ffpop.CROSSOVERS      # specify crossover recombination
+pop.crossover_rate = 2.0 / L                    # on average 2 crossovers per recombination event
+pop.set_trait_additive(np.ones(L)*params.sel)   # set additive contribution of each locus to s
+pop.mutation_rate = params.mut                  # set mutation rate
+outcrossing_rates = np.linspace(0,.2, 6)        # outcrossing rates to be simulated
+
+# loop over the outcrossing rates
+for r in outcrossing_rates:
+    print "\nEvolving a population with outcrossing rate",r
+    pop.outcrossing_rate = r                    # set outcrossing rates
+    pop.set_wildtype(params.pop)                # initialize a wildtype population of size params.pop
+    
+    # simulate for params.Ttraj generations and record the mean fitness,
+    # fitness variance, participation ratio and the number of clones
+    pfit = pop.get_fitness_statistics()
+    popstat = []
+    pop.status()
[…]

### 2016-11-22 FFPopSim/examples/condensation.py

@@ -1,0 +1,88 @@
+'''
+author:     Richard Neher, Fabio Zanini
+date:       11/07/12
+content:    Genotype condensation is driven by epistasis.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import time
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as ffpop
+
+
+# parse the command line arguments
+import argparse
+parser = argparse.ArgumentParser(description="Simulate a population on a mixed additive/epistatic fitness function")
+parser.add_argument('--pop', default=10000, type=float, help='Population size (N)')
+parser.add_argument('--rec',  default=0,type=float, help='out-crossing rate (r)')
+parser.add_argument('--sigma', default=0.05,type=float, help='Sigma')
+parser.add_argument('--hsq', default=0,type=float, help='heritability')
+parser.add_argument('--Ttraj', default=200,type=int, help='Length of trajectory in generations')
+parser.add_argument('--dt',  default=1,type =int, help='time increments of trajectory')
+params=parser.parse_args()
+
+# set up the population
+L = 1000                                            # number of loci
+pop = ffpop.haploid_highd(L)                        # create an instance of the class
+pop.outcrossing_rate = params.rec                   # outcrossing rate
+pop.recombination_model = ffpop.CROSSOVERS          # recombination model
+
+# set random epistatic fitness lanscape
+pop.set_random_epistasis(params.sigma * np.sqrt(1 - params.hsq))
+
+# set heritability
+if (params.hsq > 0):
+    pop.set_trait_additive(np.ones(L) * params.sigma * sqrt(params.hsp / L))
+
+
+# initialize population in linkage equilibrium with all
+# allele frequencies at 0.5
+pop.set_allele_frequencies(np.ones(L) * 0.5, params.pop)
+
+# evolve and store statistics along the way
+popstat = []
+t1 = time.time()
+print "\nEvolve a population to illustrate the effect of epistasis on population structure"
+pop.status()
+print
+for gen in range(params.dt,params.Ttraj, params.dt):
+    if gen%(params.Ttraj/10)==0: print gen, "out of", params.Ttraj, "generations"
+    # append current statistics to the list
[…]

### 2016-11-22 FFPopSim/examples/decay_of_LD_highd.py

@@ -1,0 +1,76 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini and Richard Neher
+date:       23/08/12
+content:    Example of haploid_highd on linkage relaxation via recombination
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 10000                           # population size
+L = 100                             # number of loci
+mu = 0.0                            # no new mutations
+r = 0.1/L                           # crossover rate
+
+# set up population
+pop = h.haploid_highd(L)                # produce an instance of haploid_highd
+pop.carrying_capacity = N               # set the steady-state population size
+pop.recombination_model = h.CROSSOVERS  # specify crossover recombination
+pop.outcrossing_rate = 1                # obligate sexual
+pop.crossover_rate = r                  # assign the crossover rate per locus
+pop.mutation_rate = mu                  # assign the mutation rate per locus
+
+# initialize the population with
+#    - N/2 individuals with genotype  00..00 or --..--
+#    - N/2 with the opposite genotype 11..11 or ++..++
+pop.set_genotypes([np.zeros(L,dtype='int'),
+                   np.ones(L,dtype='int')],
+                  [N/2, N/2])
+
+# locus pairs for which LD is to be tracked
+locus_pairs = [[0,10],
+               [0,20],
+               [40,50],
+               [10,60]]
+
+print "\nTrack LD and compare to deterministic expectations\n"
+pop.status()
+
+# get initial LD
+LD_trajectories = [[pop.get_LD(l1,l2) for l1,l2 in locus_pairs]]
+tp = [pop.generation]
+
+# evolve with accuracy of 5 generations and save LD along the way
+for ii in xrange(50):
+
+    pop.evolve(5)
+
+    # get LD and time
+    LD_trajectories.append([pop.get_LD(l1,l2) for l1,l2 in locus_pairs])
+    tp.append(pop.generation)
+    
+LD_trajectories = np.array(LD_trajectories)
+tp = np.array(tp)
+
+# plot the LD trajectories and compare to exponential decay
[…]

### 2016-11-22 FFPopSim/examples/decay_of_LD_lowd.py

@@ -1,0 +1,63 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       23/08/12
+content:    Example of haploid_lowd on linkage relaxation via recombination
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 500000                          # Population size
+L = 4                               # number of loci
+mu = 0.0                            # no new mutations
+r = 0.01                            # recombination rate
+
+### set up
+pop = h.haploid_lowd(L)             # produce an instance of haploid_lowd with L loci
+pop.carrying_capacity = N           # set the steady-state population size
+pop.set_recombination_rates(r)      # assign the recombination rate
+pop.set_mutation_rates(mu)          # assign the mutation rate
+
+# initialize the population with
+#    - N/2 individuals with genotype 0, that is ----
+#    - N/2 with the opposite genotype, that is ++++
+pop.set_genotypes([0, 2**L - 1],
+                  [N/2, N/2])
+
+print "\nTrack LD and compare to deterministic expectations\n"
+
+pop.status()
+
+# get initial LD
+LD_trajectories = [[pop.generation, pop.get_LD(0,1), pop.get_LD(0,2), pop.get_LD(0,3)]]
+
+# evolve with accuracy of 5 generations and save LD along the way
+for ii in xrange(50):
+
+    pop.evolve(5)
+    
+    # get LD and time
+    LD_trajectories.append([pop.generation, pop.get_LD(0,1), pop.get_LD(0,2), pop.get_LD(0,3)])
+
+LD_trajectories=np.array(LD_trajectories)
+
+# plot the LD trajectories and compare to exponential decay
+cols = ['r', 'b', 'g', 'm', 'c']
+for ii in xrange(LD_trajectories.shape[1]-1):
+    plt.plot(LD_trajectories[:,0], LD_trajectories[:,ii+1], color=cols[ii], label=r'$D_{0'+str(ii+1)+'}$')
+    plt.plot(LD_trajectories[:,0], 0.25 * np.exp(-LD_trajectories[:,0] * r * (ii+1)), ls='--', color=cols[ii])
+
+plt.legend()
+plt.xlabel('Time [generations]')
+plt.ylabel('LD $D_{ij}$')
+
+plt.ion()
+plt.show()

### 2016-11-22 FFPopSim/examples/drift_vs_draft.py

@@ -1,0 +1,81 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       23/08/12
+content:    Example of haploid_highd showing how neutral alleles are affected
+            by linked selective sweeps
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import FFPopSim as h
+
+
+# specify parameters
+L = 256                                           # simulate 256 loci
+
+# set up population
+pop = h.haploid_highd(L)                        # produce an instance of haploid_highd with L loci
+pop.carrying_capacity = 50000                   # set the average population size to 50000
+pop.outcrossing_rate = 1                        # make the species obligate outcrossing
+pop.crossover_rate = 0.02 / pop.L               # set the crossover rate of the segment to 2 centimorgans
+pop.mutation_rate = 0.1 / pop.carrying_capacity # per locus mutation rate equal to 0.1/N
+
+# set fitness landscape
+selection_coefficients = 0.0*np.ones(pop.L)     # most loci are neutral
+m = 10
+selection_coefficients[::m] = -0.1              # every m-th locus is strongly deleterious
+pop.set_trait_additive(selection_coefficients)  # trait 0 is by default fitness
+
+# initialize the population in linkage equilibrium with the specified allele frequencies
+initial_allele_frequencies = 0.5*np.ones(pop.L)  # define some initial allele frequencies as 1/2
+initial_allele_frequencies[::m] = 0.0            # set a subset of alleles to frequency 0
+pop.set_allele_frequencies(initial_allele_frequencies, pop.carrying_capacity)
+
+# evolve for 2000 generations and track the allele frequencies
+maxgen = 2000
+allele_frequencies = [pop.get_allele_frequencies()]
+tp = [pop.generation]
+while pop.generation < maxgen:
+    pop.evolve(10)
+
+    # save allele frequencies and time
+    allele_frequencies.append(pop.get_allele_frequencies()) 
+    tp.append(pop.generation)
+
+    # every 200 generations, make one of the deleterious mutations beneficial
+    if (pop.generation % 200 == 0):
[…]

### 2016-11-22 FFPopSim/examples/escape.py

@@ -1,0 +1,97 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini, Richard Neher
+date:       28/05/12
+content:    Example of immune escape in HIV using a few effective loci and
+            haploid_lowd
+
+Note: this example shows some alternative Python syntaxes from most other
+example scripts, such as the following:
+
+    - if __name__ == '__main__'
+    - 1<<L
+    - fig.add_subplot(1, 1, 1)
+    - '{:04d}'.format(...)
+    - line[0].set_dashes
+    - bbox_to_anchor = (...)
+
+They are mainly cosmetic changes for easy scripts such as this one, but might
+become relevant for more complex programs. Please refer to the Python
+documentation for more information on those topics.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import FFPopSim as h
+
+
+# specify parameters
+L = 4                           # number of loci
+N = 1e10                        # population size
+mu = 1e-5                       # mutation rate
+r = 1e-4                        # recombination rate
+f = [0.3, 0.2, 0.1, 0.05]       # fitness (additive effects)
+
+
+
+# Script
+if __name__ == '__main__':
+
+    pop = h.haploid_lowd(L)     # produce an instance of haploid_lowd with L loci
+    pop.set_genotypes([0],[N])  # start with a wildtype population
+    
+    pop.set_recombination_rates(r)  # set recombination rate
+    pop.set_mutation_rates(mu)      # set mutation rate
+    pop.set_fitness_additive(f)     # set fitness landscape
+
+    # evolve while tracking genotype frequencies
+    times = []
+    genotype_frequencies = []
+    while (pop.get_genotype_frequency(0b1111) < 0.99) and (pop.generation < 1e7):
+        pop.evolve()
+
+        # get genotype frequencies and time
+        genotype_frequencies.append(pop.get_genotype_frequencies())
+        times.append(pop.generation)
+
+    genotype_frequencies = np.asarray(genotype_frequencies)
+
+    # check what genotypes reach high frequencies
+    ind = []
+    for i in xrange(1<<L):
[…]

### 2016-11-22 FFPopSim/examples/fitness_wave.py

@@ -1,0 +1,98 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini, Richard Neher
+date:       07/09/12
+content:    Example of tracking the fitness distribution wave along the
+            simulation.
+            
+Note: This example also shows subclassing of FFPopSim classes. In this case, the
+syntax construct if __name__ == '__main__' is actually useful to import the
+subclass in other scripts.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import FFPopSim as h
+
+
+# subclass haploid_lowd
+class haploid_lowd_track(h.haploid_lowd):
+    '''This class tracks the fitness distribution'''
+
+    def __init__(self, *args, **kwargs):
+        '''Initialize the class instance'''
+        super(haploid_lowd_track, self).__init__(*args, **kwargs)
+
+        # prepare the tracking list
+        self.fitness_wave = []
+
+
+    def evolve(self, *args, **kwargs):
+        '''Evolve the population and keep track of the fitness histogram'''
+        ret = super(haploid_lowd_track, self).evolve(*args, **kwargs)
+        self.fitness_wave.append((self.generation,
+                                  self.get_fitness_histogram(density=True,
+                                                             n_sample=50000,
+                                                             bins=30)))
+        return ret
+
+
+    def plot_tracks(self):
+        '''Plot the tracked histograms'''
+        l = len(self.fitness_wave)
+        colors = cm.jet([int(255.0 * i / l) for i in xrange(l)])
+
+        # plot each histogram
+        for i, (g, dist) in enumerate(self.fitness_wave):
+            x = dist[1]
+            x = 0.5 * (x[:-1] + x[1:])
+            y = dist[0] + 1e-5
+            plt.plot(x, y, c=colors[i], lw=2, label='gen '+str(g))
+        
+        plt.xlabel('Fitness')
+        plt.legend(loc=1)
+        plt.yscale('log')
+        plt.ylim(1e-1)
+        plt.title('Fitness distribution over time')
+
+        plt.ion()
+        plt.show()
+
+
+
+
+# specify parameters
+N = 500000                          # Population size
+L = 12                              # number of loci
+mu = 1e-5                           # mutation rate
+r = 0.01                            # recombination rate
+
+
+# script
+if __name__ == '__main__':
[…]

### 2016-11-22 FFPopSim/examples/genetic_drift.py

@@ -1,0 +1,61 @@
+'''
+author:     Richard Neher, Fabio Zanini
+date:       11/07/12
+content:    Example on genetic drift using haploid_highd
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+from matplotlib import pyplot as plt
+import matplotlib.cm as cm
+import FFPopSim as h
+
+
+# specify parameters
+L = 256                                         # simulate 256 loci
+
+# set up population
+pop = h.haploid_highd(L)                        # produce an instance of haploid_highd with L loci
+pop.carrying_capacity = 50000                   # set the average population size to 50000
+pop.outcrossing_rate = 1                        # make the species obligate outcrossing
+pop.crossover_rate = 0.02 / pop.L               # set the crossover rate of the segment to 2 centimorgans
+pop.mutation_rate = 0.1 / pop.carrying_capacity # per locus mutation rate equal to 0.1/N
+
+
+# initialize the population in linkage equilibrium with the specified allele frequencies
+initial_allele_frequencies = 0.5*np.ones(pop.L) # define some initial allele frequencies as 1/2
+pop.set_allele_frequencies(initial_allele_frequencies, pop.carrying_capacity)
+
+# evolve for 2000 generations and track the allele frequencies
+maxgen = 2000
+allele_frequencies = [pop.get_allele_frequencies()] 
+tp = [pop.generation]
+
+print "Illustrate genetic drift on allele frequency trajectories."
+pop.status()    #print status message
+while pop.generation < maxgen:
+    if (pop.generation%(maxgen/10)==0): print pop.generation,"out of",maxgen, "generations"
+    pop.evolve(10)
+
+    # save allele frequencies and time
+    allele_frequencies.append(pop.get_allele_frequencies())
+    tp.append(pop.generation)
+
+# convert to an array to enable slicing
+allele_frequencies = np.array(allele_frequencies)
+
+# plot the result
+plt.figure()
+for locus in xrange(5,pop.L,50):                # plot a few neutral trajectories
+    plt.plot(tp, allele_frequencies[:,locus], c=cm.cool(locus), lw=2)
+
+plt.title('Genetic Drift')
+plt.xlabel('Time [generations]')
+plt.ylabel('Allele frequencies')
+
+plt.ion()
+plt.show()
+

### 2016-11-22 FFPopSim/examples/hiv_fitness.py

@@ -1,0 +1,56 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini, Richard Neher
+date:       25/04/12
+content:    Example of the hivpopulation subclass. This example is similar to
+            fitness_wave.py, but exploits the specialized subclass hivpopulation
+            for simulating HIV populations.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 10000                       # population size
+adaptive_fraction = 0.01        # fraction of beneficial/adaptive sites
+effect_size_adap = 0.03         # mean selection coefficient of beneficial alleles
+
+
+# script
+if __name__ == '__main__':
+
+    pop = h.hivpopulation(N)    # create population with default parameters
+    
+    # set random replication/fitness landscape
+    pop.set_replication_landscape(adaptive_fraction=adaptive_fraction,
+                                  effect_size_adaptive=effect_size_adap)
+    
+    # prepare figure
+    fig = plt.figure()
+    ax = fig.add_subplot(1,1,1)
+    colors = ['b', 'r', 'g', 'cyan']
+    
+    # evolve and plot histograms
+    x0 = pop.get_fitness_statistics().mean
+    for i in xrange(4):
+        pop.evolve(250)
+        h = pop.get_fitness_histogram()
+        x = h[1][:-1] - x0
+        y = h[0]
+        w = (x[1:] - x[:-1]).mean()
+        ax.bar(x, y,width=w,
+               color=colors[i%len(colors)], alpha=0.8,
+               label=str(pop.generation))
+
+    ax.set_xlabel('Fitness (relative to founder)')
+    ax.set_title('Fitness distribution of the population')
+    ax.legend(loc=2)
+
+    plt.ion()
+    fig.show()

### 2016-11-22 FFPopSim/examples/mutation_selection_balance_highd.py

@@ -1,0 +1,98 @@
+'''
+author:     Richard Neher, Fabio Zanini
+date:       11/07/12
+content:    Example on the steady state distribution of allele frequency in a 
+            balance between mutation and genetic drift using haploid_highd.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 500                             # population size
+L = 64                              # number of loci
+s = np.linspace(-2 ,2, L) / N       # additive selection coefficients for L loci, scaled to N
+mu = 0.5 / N                        # mutation rate, scaled to N
+r  = 50.0 / N / L                   # recombination rate for each interval between loci
+
+# set up population
+pop = h.haploid_highd(L)            # produce an instance of haploid_highd with L loci
+
+# set and additive fitness function. Note that FFPopSim models fitness landscape
+# in a +/- rather than 0/1 basis, hence the factor 1/2
+pop.set_fitness_additive(0.5 * s)
+
+pop.mutation_rate = mu                   # mutation rate
+pop.recombination_model = h.CROSSOVERS   # recombination model
+pop.outcrossing_rate = 1                 # obligate sexual
+pop.crossover_rate = r                   # crossover rate
+
+# initialize population in linkage equilibrium with frequencies 0.5 and size N
+pop.carrying_capacity = N           # set the population size
+pop.set_allele_frequencies(0.5 * np.ones(L), N)
+
+print "Evolve for >> N generations and compare allele frequency distributions \nto expectations from diffusion theory."
+pop.status()
+pop.evolve(10 * N)                      # evolve for 10N generations to equilibrate
+
+# evolve and sample allele_frequencies
+nsamples = 10000
+allele_frequencies = np.zeros((nsamples,L))
+for ii in range(nsamples):
+    pop.evolve(0.1 * N)                 # N / 10 generations between successive samples
+
+    # print output every 100 generations
+    if (ii % 100 == 0):
[…]

### 2016-11-22 FFPopSim/examples/mutation_selection_balance_lowd.py

@@ -1,0 +1,81 @@
+'''
+author:     Richard Neher, Fabio Zanini
+date:       11/07/12
+content:    Example on the steady state distribution of allele frequency in a 
+            balance between mutation and genetic drift using haploid_lowd.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 500                             # population size
+L = 4                               # number of loci
+s = np.linspace(-0.2 ,0.7, L) / N   # additive selection coefficients for L loci, scaled to N
+mu = 0.4 / N                        # mutation rate, scaled to N
+r = 5.0 / N                         # recombination rate for each interval between loci. 
+
+# set up population
+pop = h.haploid_lowd(L)             # produce an instance of haploid_lowd with L loci
+
+# set and additive fitness function. Note that FFPopSim models fitness landscape
+# in a +/- rather than 0/1 basis, hence the factor 1/2
+pop.set_fitness_additive(0.5 * s) 
+
+pop.set_mutation_rates(mu)          # mutation rate
+pop.set_recombination_rates(r)      # recombination rate (CROSSOVERS model by default)
+
+# initialize the population with N wildtype individuals, that is ----
+pop.carrying_capacity = N           # set the population size
+pop.set_genotypes([0], [N])
+
+print "Evolve for >> N generations and compare allele frequency distributions \nto expectations from diffusion theory."
+pop.status()
+pop.evolve(10 * N)                  # run for 10N generations to equilibrate
+
+# evolve and sample allele frequencies
+nsamples = 10000      
+allele_frequencies = np.zeros((nsamples,L))
+for ii in range(nsamples):
+    pop.evolve(0.1 * N)             # N / 10 generations between successive samples
+
+    # get allele frequencies
+    allele_frequencies[ii,:] = pop.get_allele_frequencies()
+
+
+# prepare allele frequency histogram
+af_bins = np.linspace(0,1,26)                   # bins for histogram
[…]

### 2016-11-22 FFPopSim/examples/neutral_LD_highd.py

@@ -1,0 +1,75 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       23/08/12
+content:    Example of haploid_highd demonstrating the balance between
+            recombination and genetic drift in a finite population. The example
+            plots the average r^2 (squared correlation coefficient) for loci at
+            different distances along the chromosome.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+N = 500                            # population size
+L = 1000                           # number of loci
+mu = 0.1 / N                       # mutation rate
+r = 10.0 / L / N                   # crossover rate
+
+# set up population
+pop = h.haploid_highd(L)             # produce an instance of haploid_lowd with L loci
+pop.carrying_capacity = N            # set the steady-state population size
+pop.outcrossing_rate=1               # obligate sexual
+pop.crossover_rate = r               # crossover rate per locus
+pop.mutation_rate=mu                 # mutation rate per locus
+
+# initialize the population with
+#    - N/2 individuals with genotypes 00000000...0000
+#    - N/2 with the opposite genotype 11111111...1111
+pop.set_genotypes([np.zeros(L), np.ones(L)],[N/2, N/2])
+
+# locus pairs for which LD is to be tracked
+ld_points = np.arange(5,L-1,100)
+locus_pairs = [ [L/2, l1] for l1 in ld_points]
+
+print "Evolve for >> N generations and measure correlations between loci, aka LD."
+pop.status()
+pop.evolve(10 * N)                      # evolve for 10N generations to equilibrate
+
+pop.evolve(4*N)                     # evolve for 4N to equilibrate
+
+# evolve the population and track linkage disequilibrium (LD)
+nsamples = 10000      
+LD=[]
+rsq = []
+for ii in range(nsamples):
+    pop.evolve(0.1 * N)             # N / 10 generations between successive samples
+
+    if (ii%100==0):
+        print ii, "out of", nsamples, "samples"
+
+    # get allele frequencies
[…]

### 2016-11-22 FFPopSim/examples/speed_highd.py

@@ -1,0 +1,156 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       24/08/12
+content:    Example of haploid_lowd on the algorithm runtime complexity.
+            The same series of simulations are performed:
+                - for a human-like population,
+                - for an HIV-like (viral) population,
+            in this order.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+import time
+
+
+# 1. specify parameters for a human population, per MB of DNA
+r = 1e-8                       # crossover rate
+mu = 1e-8                      # mutation rate
+G = 100                        # generations
+NS = 1000
+exec_time_MB = {}
+Llist_MB = [1e5, 3e5, 1e6]
+Nlist = [100, 300, 1000, 3000, 10000, 30000]
+
+for L in Llist_MB:
+    exec_time = []
+    for N in Nlist:
+        # set up population
+        pop = h.haploid_highd(L)    # produce an instance of haploid_lowd with L loci
+        pop.carrying_capacity = N   # set the population size
+
+        pop.recombination_model = h.CROSSOVERS  # set recombination model
+        pop.outcrossing_rate = 1.0              # obligate sexual
+        pop.crossover_rate = r                  # crossover rates
+        pop.mutation_rate = mu                  # mutation rate (per locus)
+        
+        pop.set_wildtype(N)                     # set a wildtype population of size N
+
+        pop.evolve(1.0 / (L * (mu + r)))        # evolve until equilibrium
+        
+        # run for G generations to measure execution time    
+        t1=time.time()
+        pop.evolve(G)
+        t2=time.time()
+        print L, "bases, human like, population size:", N,
+        print "time required for", G, "generations:",
+        print round(t2 - t1, 3), "seconds"
+        
+        # set additive fitness landscape and repeat
+        selection_coefficients = np.zeros(L)
+        for locus in xrange(0, int(L), int(L/NS)):
+            selection_coefficients[locus] = 0.01 * ((np.random.rand()-0.98)>0)
+        pop.set_fitness_additive(selection_coefficients)
+
[…]

### 2016-11-22 FFPopSim/examples/speed_lowd.py

@@ -1,0 +1,150 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       24/08/12
+content:    Example of haploid_lowd on the algorithm runtime complexity.
+            The same series of simulations are performed:
+                - with multiple crossovers, O(3^L),
+                - with at most one crossover, O(L 2^L),
+                - without recombination, O(L 2^L),
+            in this order.
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+import time
+
+
+# specify parameters
+N = 1e12                        # population size
+Lmax_general = 13               # maximal number of loci
+Lmax_single_xo = 17             # maximal number of loci
+r = 0.01                        # recombination rate
+mu = 0.001                      # mutation rate
+G = 100                         # generations
+
+# 1. multiple crossovers
+print "\ngeneral recombination"
+exec_time = []
+for L in range(2,Lmax_general+1):
+    print L,"loci, maximum",Lmax_general,
+    # set up population
+    pop = h.haploid_lowd(L)     # produce an instance of haploid_lowd with L loci
+    pop.carrying_capacity = N   # set the population size
+
+    # set and additive fitness function with random coefficients. 
+    # Note that FFPopSim models fitness landscapes in the -1/+1 basis
+    pop.set_fitness_additive(0.01 * np.random.randn(L))
+
+    pop.set_recombination_rates(r)  # recombination rates
+    pop.set_mutation_rates(mu)      # mutation rate
+    
+    # initialize the population with N individuals in linkage equilibrium
+    pop.set_allele_frequencies(0.2 * np.ones(L), N)
+
+    # run for G generations to measure execution time
+    t1=time.time()
+    pop.evolve(G)
+    t2=time.time()
+
+    print "time required for",G,"generations:",round(t2-t1,3),'s'
+    exec_time.append([L, t2-t1])            # store the execution time
+exec_time=np.array(exec_time)
+
+
+# 2. single crossover recombination
+print "\nsingle crossover"
+exec_time_single_xo = []
[…]

### 2016-11-22 FFPopSim/examples/valley.py

@@ -1,0 +1,116 @@
+# vim: fdm=indent
+'''
+author:     Richard Neher, Fabio Zanini
+date:       25/05/12
+content:    Find the time for valley crossing
+'''
+# Import modules (setting the path should not be necessary when the module is
+# installed in the PYTHONPATH)
+import sys
+sys.path.insert(0, '../pkg/python')
+
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+
+# specify parameters
+simulate = False                # use extant data or repeat the simulation?
+L = 4                           # number of loci
+N = 1e10                        # population size
+s1 = 1e-5                       # fitness advantage of wildtype (half)
+s2 = 0.01                       # fitness advantage of quadruple mutant (half)
+
+# recombination rates to check out
+rs = np.logspace(-4,-3,10).tolist() + \
+     [0.00125, 0.0015, 0.00175, 0.002,\
+      0.00225, 0.0025, 0.00275, 0.003,\
+      0.0031, 0.0032, 0.0033, 0.0034, \
+      0.0035, 0.00375, 0.004, 0.005, \
+      0.0075, 0.01]
+
+# mutation rates to check out
+mutation_rates=[1e-7, 1e-6, 1e-5]
+
+
+# script
+if __name__ == '__main__':
+
+    # prepare figure
+    fig = plt.figure()
+    ax = fig.add_subplot(1,1,1)
+    lstyles = [[1,0],
+               [2, 4, 5, 4],
+               [7, 4, 7, 4]]
+    colors=['b', 'g', 'r', 'm', 'c']
+    
+    # cycle over mutation rates
+    for k, mu in enumerate(mutation_rates):
+
+        # simulate the population? ...
+        if simulate:
+
+            # prepare result vectors
+            times = np.zeros_like(rs)
+            dtimes = np.zeros_like(rs)
+
+            for i, r in enumerate(rs):    
+                ttmp = []
+                for j in xrange(50):
+        
+                    # set up population
+                    pop = h.haploid_lowd(L)         # produce an instance of haploid_lowd with L loci
+
+                    pop.set_genotypes([0], [N])     # set a wildtype population of size N
+
+                    pop.set_recombination_rates(r)  # recombination rate
+                    pop.set_mutation_rates(mu)      # mutation rate
+
+                    # set the fitness valley:
+                    #    - intermediates have fitness 0
+                    #    - wildtype has fitness 2 s1
[…]

### 2016-11-22 FFPopSim/profile/profile.cpp

@@ -1,0 +1,68 @@
+/**
+ * @file highd.cpp
+ * @brief Profile functions for the simulation library.
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-04-20
+ */
+/* Include directives */
+#include <boost/progress.hpp>
+#include "hivpopulation.h"
+
+/* Defines */
+#define HIGHD_BADARG -1354341
+#define NOTHING 1e-10
+
+/* Be verbose? */
+#define PROFILE_VERBOSE 1
+
+/* Declarations */
+int hiv_profile();
+
+
+/* MAIN */
+int main(int argc, char **argv){
+	boost::progress_timer timer;
+
+	int status= 0;
+	if (argc > 1) {
+		cout<<"Usage: "<<argv[0]<<endl;
+		status = 1;
+	} else {
+                for(int i=0; i < 5; i++)
+		status += hiv_profile();
+	}
+	cout<<"Number of errors: "<<status<<endl;
+	return status;
+}
+
+
+int hiv_profile() {
+	int N = 1000, err=0;
+	ifstream model;
+
+	for(size_t i=0; i < 1; i++) {
+		if(PROFILE_VERBOSE) cerr<<"Population n. "<<i<<"...";
+		model.open("../tests/hiv_model.dat", ifstream::in);
+		hivpopulation pop(N, 1, 2e-5, 1e-3, 1e-3);
+	
+		if(PROFILE_VERBOSE) cerr<<"Reading model coefficients...";
+		pop.read_replication_coefficients(model);
+		if(PROFILE_VERBOSE) cerr<<"read!"<<endl;
+	
+		for(size_t j = 0; j < 5; j++) {
+			pop.get_max_fitness();
+			err += pop.evolve(100);
+		}
+		if(PROFILE_VERBOSE) cerr<<"done"<<endl;
+
+		cout<<"Fitnesses: ";
+		for(size_t k=0; k!= 5; k++)	{
+			cout<<pop.get_fitness(k)<<"  ";		
+		}
+		cout<<endl;
+		model.close();
+	}
+
+	return err;
+}

### 2016-11-22 FFPopSim/setup.py

@@ -1,0 +1,105 @@
+#!/usr/bin/env python
+'''
+author:     Richard Neher, Fabio Zanini
+date:       23/08/12
+license:    GPL3
+content:    Distutils setup script for the Python bindings of FFPopSim.
+
+            If your compiler has problems finding GSL and/or BOOST and you are
+            sure they are installed on your system, check the section
+            PLATFORM-DEPENDENT OPTIONS below.
+
+            *Note*: this file is called by the Makefile with the command
+
+                python setup.py build_ext
+            
+            to build the C++/Python extension. It can, however, also be called
+            directly including with other commands such as
+
+                python setup.py install
+
+            to install the Python bindings of FFPopSim on the system. Note that
+            calling this file directly does not clean the 'build' folder.
+
+Static building: copy your GSL includes and static lib files into
+
+    static/<architechture>/
+
+following the directory structure (i.e. include/gsl and /lib). Then, set the
+'platform' variable here below accordingly and build.
+'''
+from distutils.core import setup, Extension
+from numpy import distutils as npdis
+import sys
+import platform as pfm
+
+############################################################################
+#									   #
+#			PLATFORM-DEPENDENT OPTIONS			   #
+#									   #
+############################################################################
+# Please add your include folders to the following list, where the compiler
+# can find GSL, BOOST, and Python 2.7 headers.
+includes = ['/usr/include', '/usr/local/include', '/opt/local/include']
+
+# Please add your shared library folders to the following list, where the linker
+# can find Python 2.7 (GLS is static in this branch)
+library_dirs = ['/ebio/ag-neher/share/programs/lib']
+
+############################################################################
+#                !!  DO NOT EDIT BELOW THIS LINE  !!                       #
+############################################################################
+VERSION = '2.0'
+SRCDIR = 'src'
+PYBDIR = SRCDIR+'/python'
+
+# Find the current platform and Python interpreter 64-bitness
+# Mac OSX
+if pfm.system().lower() == 'darwin':
+    if sys.maxsize > 2**32:
+        platform = 'mac_64'
+    else:
+        platform = 'mac_32'
+# Linux
+else:
+    if pfm.architecture()[0][:2] == '64':
+        platform = 'linux_64'
+    else:
+        platform = 'linux_32'
+print 'The current platform is recognized as: '+platform+''
+print 'If this is not correct, please modify setup.py manually.'
[…]

### 2016-11-22 FFPopSim/src/ffpopsim_generic.h

@@ -1,0 +1,119 @@
+/**
+ * @file popgen.h
+ * @brief Header file with the classes and types provided with the library. 
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2010-10-27
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#ifndef FFPOPGEN_GENERIC_H_
+#define FFPOPGEN_GENERIC_H_
+
+#include <time.h>
+#include <cmath>
+#include <vector>
+#include <bitset>
+#include <string>
+#include <sstream>
+#include <iostream>
+#include <fstream>
+#include <iomanip>
+#include <gsl/gsl_sf.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_histogram.h>
+#include <gsl/gsl_histogram2d.h>
+#include <gsl/gsl_matrix.h>
+#include <gsl/gsl_vector.h>
+#include <gsl/gsl_eigen.h>
+#include <boost/dynamic_bitset.hpp>
+
+#define MIN(a,b) (a<b)?a:b
+#define MAX(a,b) (a>b)?a:b
+#define RNG gsl_rng_taus2		//choose the random number generator algorithm, see http://www.gnu.org/software/gsl/manual/html_node/Random-number-generator-algorithms.html
+
+#define FREE_RECOMBINATION 1
+#define CROSSOVERS 2
+#define SINGLE_CROSSOVER 3
+
+using namespace std;
+
+/**
+ * @brief Pairs of an index and a value
+ */
+struct index_value_pair_t {
+	size_t index;
+	double val;
+	index_value_pair_t(int index_in=0, double val_in=0) : index(index_in), val(val_in) {};
+};
[…]

### 2016-11-22 FFPopSim/src/ffpopsim_highd.h

@@ -1,0 +1,609 @@
+/**
+ * @file popgen_highd.h
+ * @brief Header file for high-dimensional simulations
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-04-19
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ *
+ * HP_VERBOSE: degree of verbosity of haploid_highd. Levels:
+ * - 0: no messages
+ * - 1: most messages (enter/exit function)
+ * - 2: all messages
+ */
+#ifndef FFPOPSIM_HIGHD_H_
+#define FFPOPSIM_HIGHD_H_
+#include "ffpopsim_generic.h"
+
+#define HCF_MEMERR -131545
+#define HCF_BADARG -131546
+#define HCF_VERBOSE 0
+#define WORDLENGTH 28 	//length used to chop bitsets into words
+
+using namespace std;
+
+
+/**
+ * @brief Trait coefficient for a set of loci.
+ *
+ * This struct is used in the hypercube_highd class for saving trait coefficients.
+ * See also hypercube_highd::add_coefficient.
+ *
+ * Note: words and bits are deprecated.
+ */
+struct coeff_t {
+	int order;
+	double value;
+	int *loci;
+	coeff_t(double value_in, vector <int> loci_in){
+		value=value_in;
+		order=loci_in.size();
+		loci=new int [order];
+		for (int i=0; i<order; i++) loci[i]=loci_in[i];
+	}
+};
+
+/**
+ * @brief Trait coefficient for a single locus.
+ *
+ * Note: word and bit are deprecated. For this reason, this class is actually equivalent
[…]

### 2016-11-22 FFPopSim/src/ffpopsim_lowd.h

@@ -1,0 +1,257 @@
+/**
+ * @file popgen_lowd.h
+ * @brief Header file for low-dimensional simulations
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-04-19
+ *
+ * Copyright (c) 2012-2013, Richard Neher,Fabio Zanini
+ * All rights reserved.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#ifndef FFPOPSIM_LOWD_H_
+#define FFPOPSIM_LOWD_H_
+#include "ffpopsim_generic.h"
+
+#define HC_MEMERR -131545		//memory error code
+#define HC_BADARG -131546		//bad argument error code
+#define HC_VERBOSE 0			//debugging: if set to one, each function prints out a message into the error stream
+#define HC_FUNC 1			//hypercube_lowd.func is up-to-date
+#define HC_COEFF -1			//hypercube_lowd.coeff is up-to-date
+#define HC_FUNC_EQ_COEFF 0		//hypercube_lowd.func equal hypercube_lowd.coeff
+
+using namespace std;
+
+/**
+ * @brief Binary hypercube_lowd used in low-dimensional simulations.
+ *
+ * This class is a generic object that can be used to represent various things, e.g.
+ * - the fitness landscape or any other phenotypic landscape;
+ * - the genotype frequencies of a population with a genome of size L.
+ *
+ * If you are planning to model a whole population evolving on the hypercube_lowd, see the class haploid_lowd.
+ *
+ * Notes on scalability:
+ * - The number of genotypes to store increases as \f$2^L\f$, where L is the number of sites
+ * - The number of recombination intermediates to be stored increases as \f$3^L\f$, this class can thus only be used for \f$L\f$ up to 20 or so.
[…]

### 2016-11-22 FFPopSim/src/haploid_highd.cpp

@@ -1,0 +1,2118 @@
+// vim: tabstop=8:softtabstop=8:shiftwidth=8:noexpandtab
+/*
+ * haploid_highd.cpp
+ *
+ *  Created on: Aug 28, 2008
+ *      Author: neher
+ *    Modified: Dec 13, 2011
+ *   Committer: Fabio Zanini
+ *
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include <math.h>
+#include "ffpopsim_highd.h"
+
+/* Initialize the number of instances to zero */
+size_t haploid_highd::number_of_instances = 0;
+
+/**
+ * @brief Default constructor
+ *
+ * The population objects are initialized, but neither the random generator is initialized nor the memory
+ * for traits, gametes etc. is allocated. Please call set_up() before using the class.
+ *
+ * Note: The sequence is assumed to be linear (not circular). You can change this by hand if you wish so.
+ */
+haploid_highd::haploid_highd(int L_in, int rng_seed, int n_o_traits, bool all_polymorphic_in, bool most_polymorphic_in) {
+	if (L_in < 1 or n_o_traits < 1) {
+		if(HP_VERBOSE) cerr <<"haploid_highd::haploid_highd(): Bad Arguments! Both L and the number of traits must be larger or equal one."<<endl;
+		throw HP_BADARG;
+	}
+
+	// Check that all_polymorphic is not used with more than one phenotypic trait
+	if  (n_o_traits > 1 and all_polymorphic_in) {
[…]

### 2016-11-22 FFPopSim/src/haploid_lowd.cpp

@@ -1,0 +1,1131 @@
+/*
+ * haploid_lowd.cpp
+ *
+ *  Created on: Jan 27, 2010
+ *  Author: Richard Neher
+ *  Modified by: Fabio Zanini
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "ffpopsim_lowd.h"
+
+/* Initialize the number of instances to zero */
+size_t haploid_lowd::number_of_instances=0;
+
+/**
+ * @brief Default constructor
+ *
+ * @param L_in number of loci (at least 1)
+ * @param rngseed seed for the random number generator. If this is zero, time(NULL)+getpid() is used.
+ */
+haploid_lowd::haploid_lowd(int L_in, int rng_seed) {
+	if (L_in <1) {
+		cerr <<"haploid_lowd::haploid_lowd(): Bad Arguments! L must be larger or equal one."<<endl;
+		throw HG_BADARG;
+	}
+
+	// Set attributes
+	number_of_loci = L_in;
+	population_size = carrying_capacity = 0;
+	recombination_model = FREE_RECOMBINATION;
+	mem = false;
+	outcrossing_rate = 1.0;
+	circular = false;
+	generation = 0;
+	long_time_generation = 0;
+
+	//In case no seed is provided, get one from the OS
+	seed = rng_seed ? rng_seed : get_random_seed();
+
+	// Note: we should clean up the mess made by allocate_mem(). This requires more fine-grained
+	// control than we currently have.
+	int err = allocate_mem();
+	if(err)	throw err;
+	
+	number_of_instances++;
+}
+
+/**
+ * @brief Default destructor
[…]

### 2016-11-22 FFPopSim/src/hivgene.cpp

@@ -1,0 +1,41 @@
+/**
+ * @file hivgene.cpp
+ * @brief Implementation of an HIV gene
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-05-18
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "hivpopulation.h"
+
+/**
+ * @brief Default constructor for an HIV gene
+ *
+ * @param start_in first site of the gene in the HIV genome
+ * @param end_in last site + 1 of the gene in the HIV genome
+ */
+hivgene::hivgene(unsigned int start_in, unsigned int end_in) {
+		if((start_in>=HIVGENOME) || (end_in>HIVGENOME) || (end_in<=start_in))
+			throw (int)HIVPOP_BADARG;
+		else {
+			start = start_in;
+			end = end_in;
+		}
+}

### 2016-11-22 FFPopSim/src/hivpopulation.cpp

@@ -1,0 +1,192 @@
+/**
+ * @file hivpopulation.cpp
+ * @brief Implementation of an HIV population with drug treatment.
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-04-23
+ *
+ * Copyright (c) 2012-2013, Richard Neher,  Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "hivpopulation.h"
+
+/**
+ * @brief Construct a HIV population with certain parameters
+ *
+ * @param N_in number of viral particles 
+ * @param rng_seed seed for the random number generator. If this is 0, time(NULL)+getpid() is used.
+ * @param mutrate mutation rate in events / generation / site
+ * @param coinfection_rate probability of coinfection of the same cell by two viral particles in events / generation
+ * @param crossover_rate probability of template switching during coinfection in events / site
+ *
+ * *Note*: the genome length is 10000 (see HIVGENOME).
+ * *Note*: exceptions are propagated from the base class constructor (haploid_highd).
+ */
+hivpopulation::hivpopulation(int N_in, int rng_seed, double mutation_rate_in, double coinfection_rate_in, double crossover_rate_in) : haploid_highd(HIVGENOME, rng_seed, 2), env(ENV_START, ENV_END), treatment(0) {
+
+	outcrossing_rate = coinfection_rate_in;
+	mutation_rate = mutation_rate_in;
[…]

### 2016-11-22 FFPopSim/src/hivpopulation.h

@@ -1,0 +1,105 @@
+/**
+ * @file hivpopulation.h
+ * @brief Header file for a typical HIV population (subclass of haploid_highd)
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-04-23
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#ifndef HIVPOPULATION_H_
+#define HIVPOPULATION_H_
+
+#include <string>
+#include <fstream>
+#include <sstream>
+#include <iomanip>
+#include <boost/algorithm/string.hpp>
+
+#include "ffpopsim_highd.h"
+
+#define HIVPOP_VERBOSE 0
+#define HIVPOP_BADARG -1354341
+#define NOTHING 1e-10
+#define HIVGENOME 10000
+
+// HIV genes
+#define ENV_START 7000
+#define ENV_END 8000
+
+
+/**
+ * @brief HIV gene.
+ *
+ * Attributes:
+ * - start: the starting position of the gene
+ * - stop: the (last position + 1) of the gene
+ */
+struct hivgene {
+	unsigned int start;
+	unsigned int end;
+	hivgene(unsigned int start_in=0, unsigned int end_in=HIVGENOME);
+};
+
+/**
+ * @brief HIV population with facultative drug treatment
+ *
+ * This class exemplifies the haploid_highd base class. It mainly adds one trait,
+ * "treatment", which is the same for all individuals and represents the presence
+ * or absence of drug treatment (in a continuous manner, \f$0 \leq \f$ treatment
+ * \f$\leq 1\f$).
+ *
+ * The replication capacity in absence of drug is encoded in the first trait. The
[…]

### 2016-11-22 FFPopSim/src/hypercube_highd.cpp

@@ -1,0 +1,292 @@
+/*
+ *  hypercube_highd.cpp
+ *  Haploids
+ *
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "ffpopsim_highd.h"
+
+hypercube_highd::hypercube_highd()
+{
+mem=false;
+if (HCF_VERBOSE) cerr<<"hypercube_highd::hypercube_highd(): constructing...!\n";
+}
+
+//constructor
+hypercube_highd::hypercube_highd(int dim_in, int s) {
+	if (HCF_VERBOSE) cerr<<"hypercube_highd::hypercube_highd(): constructing...!\n";
+	set_up(dim_in, s);
+}
+
+//set up and allocate memory, perform basic consistency checks on input.
+int hypercube_highd::set_up(int dim_in, int s) {
+	if (dim_in>0) {
+		if (HCF_VERBOSE) cerr<<"hypercube_highd::set_up(): setting up...!";
+                // FIXME: we should really go for a global rng seeder for the whole library
+		if (s==0) s = time(NULL);
+		seed = s;
+		if (HCF_VERBOSE) cerr<<"done.\n";
+
+                dim = dim_in;
+                mem = false;
+                hcube_allocated = false;
+                hypercube_mean = 0;
+                epistatic_std = 0;
+
+		return allocate_mem();
+	} else {
+		cerr <<"hypercube_highd: need positive dimension!\n";
+		return HCF_BADARG;
+	}
+}
+
+//destructor
+hypercube_highd::~hypercube_highd() {
+	if (HCF_VERBOSE) cerr<<"hypercube_highd::~hypercube_highd(): destructing...!\n";
[…]

### 2016-11-22 FFPopSim/src/hypercube_lowd.cpp

@@ -1,0 +1,565 @@
+/*
+ *  hypercube_lowd.cpp
+ *
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "ffpopsim_lowd.h"
+
+//default constructor
+hypercube_lowd::hypercube_lowd() {
+	mem=false;
+	if (HC_VERBOSE) cerr<<"hypercube_lowd::hypercube_lowd(): constructing...!\n";
+}
+
+//constructor
+hypercube_lowd::hypercube_lowd(int dim_in, int s) {
+	if (HC_VERBOSE) cerr<<"hypercube_lowd::hypercube_lowd(): constructing...!\n";
+	set_up(dim_in, s);
+}
+
+//check consistency of input dimension and call allocation routine, seed for the rng is provided (s)
+int hypercube_lowd::set_up(int dim_in, int s) {
+	if (dim_in>0 && (unsigned int)dim_in <= 8*sizeof(int)) {
+		if (HC_VERBOSE) cerr<<"hypercube_lowd::set_up(): setting up...!";
+		dim=dim_in;
+		mem=false;
+		if (s==0) seed=time(NULL);
+		else seed=s;
+		if (HC_VERBOSE) cerr<<"done.\n";
+		return allocate_mem();
+	} else {
+		cerr <<"hypercube_lowd: got: "<<dim_in<<", need positive and <32 dimension!\n";
+		return HC_BADARG;
+	}
+}
+
+//destructor
+hypercube_lowd::~hypercube_lowd() {
[…]

### 2016-11-22 FFPopSim/src/multiLocusGenealogy.cpp

@@ -1,0 +1,82 @@
+/*
+ * multi_locus_genealogy.cpp
+ *
+ *  Created on: Oct 14, 2012
+ *      Author: richard
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+
+#include "ffpopsim_highd.h"
+
+/**
+ * @brief Default constructor
+ */
+multi_locus_genealogy::multi_locus_genealogy() {
+}
+
+/**
+ * @brief Default destructor
+ */
+multi_locus_genealogy::~multi_locus_genealogy() {
+}
+
+/**
+ * @brief Start tracking a new locus
+ *
+ * @param new_locus locus to be tracked
+ */
+void multi_locus_genealogy::track_locus(int new_locus) {
+	loci.push_back(new_locus);
+	rooted_tree temp_tree;
+	trees.push_back(temp_tree);
+	vector <node_t> temp_generation;
+	newGenerations.push_back(temp_generation);
+}
+
+/**
+ * @brief Add a new generation to all trees.
+ *
+ * @param baseline the new generation to be added
+ */
+void multi_locus_genealogy::add_generation(double baseline) {
+	for (unsigned int locusIndex = 0; locusIndex<trees.size(); locusIndex++){
+		trees[locusIndex].add_generation(newGenerations[locusIndex], baseline);
+	}
+}
+
+/**
+ * @brief: allocates memory for a sufficient number of slots in the genealogy
+ *
+ * @params: the total number of slots required (should be population.size() in most cases)
+ */
+int multi_locus_genealogy::extend_storage(int n) {
[…]

### 2016-11-22 FFPopSim/src/python/FFPopSim.py

@@ -1,0 +1,3549 @@
+# This file was automatically generated by SWIG (http://www.swig.org).
+# Version 2.0.12
+#
+# Do not make changes to this file unless you know what you are doing--modify
+# the SWIG interface file instead.
+
+
+
+
+"""
+C++/Python library for population genetics.
+
+This library offers *two* simulation packages for population genetics: one for
+low-dimensional simulations (up to ~15 loci) and one for high-dimensional ones.
+
+Each package is based on a big class that represents a population:
+
+   - ``haploid_lowd`` for low-dimensional populations
+   - ``haploid_highd`` for high-dimensional simulations
+
+A simple example routine is the following::
+
+    #####################################
+    #   EXAMPLE SCRIPT                  #
+    #####################################
+    import numpy as np
+    import matplotlib.pyplot as plt
+    import FFPopSim as h
+    
+    c = h.haploid_lowd(4)
+    c.set_allele_frequencies([0,0.3,0.6,0.9], N=1000) 
+    c.evolve(100)
+    c.plot_diversity_histogram()
+    plt.show()
+    #####################################
+
+which evolves a population with 4 loci for 100 generations starting from fixed
+allele frequencies, under neutral conditions, and plots the diversity
+histogram afterwards.
+
+For more usage examples, please consult the ``tests`` and ``examples`` folders. 
+
+"""
+
+
+from sys import version_info
+if version_info >= (3,0,0):
+    new_instancemethod = lambda func, inst, cls: _FFPopSim.SWIG_PyInstanceMethod_New(func)
+else:
+    from new import instancemethod as new_instancemethod
+if version_info >= (2,6,0):
+    def swig_import_helper():
+        from os.path import dirname
+        import imp
+        fp = None
+        try:
+            fp, pathname, description = imp.find_module('_FFPopSim', [dirname(__file__)])
+        except ImportError:
+            import _FFPopSim
+            return _FFPopSim
+        if fp is not None:
+            try:
+                _mod = imp.load_module('_FFPopSim', fp, pathname, description)
+            finally:
[…]

### 2016-11-22 FFPopSim/src/rootedTree.cpp

@@ -1,0 +1,782 @@
+/*
+ * tree.cpp
+ *
+ *  Created on: Oct 14, 2012
+ *      Author: richard
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+
+#include "ffpopsim_highd.h"
+/*
+ * this overloads the ostream operator to output keys of nodes and edges
+ */
+std::ostream& operator<< ( std::ostream& os, const tree_key_t& key )
+  {
+    os <<"age: "<< key.age << " index "
+      << key.index;
+    return os;
+  }
+
+
+rooted_tree::rooted_tree() {
+	reset();
+}
+rooted_tree::~rooted_tree() {
+}
+
+/*
+ * @brief deletes everything and reinserts a root and an MRCA
+ */
+void rooted_tree::reset(){
+	node_t root_node;
+	node_t mrca_node;
+ 	edge_t to_root;
+
+ 	nodes.clear();
+ 	edges.clear();
+ 	leafs.clear();
+
+ 	//the root node will never be touched, the MRCA moves up with the tree
+	root.age=-3;
+	root.index=0;
+	MRCA.age=-2;
+	MRCA.index=0;
+
+	root_node.own_key = root;
+	root_node.parent_node = root;
+	root_node.clone_size=-1;
+	root_node.number_of_offspring=-1;
+	root_node.crossover[0]=0;
+	root_node.crossover[1]=RT_VERYLARGE;
+	root_node.child_edges.clear();
+	root_node.child_edges.push_back(MRCA);
+
+	mrca_node.own_key = MRCA;
+	mrca_node.parent_node = root;
+	mrca_node.clone_size=-1;
[…]

### 2016-11-22 FFPopSim/src/sample.cpp

@@ -1,0 +1,162 @@
+/*
+ * sample.cpp
+ *
+ *  Created on: Aug 28, 2008
+ *
+ * Copyright (c) 2012-2013, Richard Neher, Fabio Zanini
+ * All rights reserved.
+ *
+ * This file is part of FFPopSim.
+ *
+ * FFPopSim is free software: you can redistribute it and/or modify
+ * it under the terms of the GNU General Public License as published by
+ * the Free Software Foundation, either version 3 of the License, or
+ * (at your option) any later version.
+ *
+ * FFPopSim is distributed in the hope that it will be useful,
+ * but WITHOUT ANY WARRANTY; without even the implied warranty of
+ * MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the
+ * GNU General Public License for more details.
+ *
+ * You should have received a copy of the GNU General Public License
+ * along with FFPopSim. If not, see <http://www.gnu.org/licenses/>.
+ */
+#include "ffpopsim_generic.h"
+
+sample::sample() {
+	mem_dis=false;
+	mem_values=false;
+	with_range=false;
+	bins=0;
+	number_of_values=0;
+	values=NULL;
+}
+
+sample::~sample() {
+	if (mem_dis)
+		gsl_histogram_free(distribution);
+	if (mem_values)
+		delete [] values;
+}
+
+int sample::set_up(int n)
+{
+	if (n<1)
+	{
+		cerr <<"sample::set_up(): number of values has to be greater than zero! Got: "<<bins<<endl;
+		return SAMPLE_ERROR;
+	}
+	else
+	{
+		number_of_values=n;
+		values=new double [number_of_values];
+		mem_values=true;
+	}
+	return 0;
+}
+
+int sample::set_distribution(int bins_in)
+{
+	if (bins_in<1)
+	{
+		cerr <<"sample::set_distribution(): number of bins has to be greater than zero! Got: "<<bins<<endl;
+		return SAMPLE_ERROR;
+	}
+	else
+	{
+		bins=bins_in;
+		distribution=gsl_histogram_alloc(bins);
+	}
+	return 0;
+}
+
+int sample::calc_mean()
+{
+	double v;
+	if (mem_values==false)
+	{
+		cerr <<"sample::calc_mean(): Set values first!"<<endl;
+		return SAMPLE_ERROR;
+	}
+	mean=0;
[…]

### 2016-11-22 FFPopSim/tests/highd.cpp

@@ -1,0 +1,437 @@
+/**
+ * @file highd.cpp
+ * @brief Tests for the high-dimensional simulation library.
+ * @author Richard Neher, Boris Shraiman, Fabio Zanini
+ * @version 
+ * @date 2012-04-20
+ */
+
+/* Include directives */
+#include "ffpopsim_highd.h"
+#define HIGHD_BADARG -1354341
+#define NOTHING 1e-10
+
+/* Be verbose? */
+#define HIGHD_VERBOSE 1
+
+/* Test generic library access */
+int library_access() {
+	index_value_pair_t ivp;
+	ivp.index = 4;
+	ivp.val = 4.5;
+	if(HIGHD_VERBOSE) {
+		cerr<<"index_value_pair_t: index = "<<ivp.index;
+		cerr<<", val = "<<ivp.val<<endl;
+	}
+
+	coeff_single_locus_t csl(ivp.val, ivp.index);
+	if(HIGHD_VERBOSE) {
+		cerr<<"coeff_single_locus_t: locus = "<<csl.locus;
+		cerr<<", value = "<<csl.value<<endl;
+	}
+	return 0;
+}
+
+
+/* Test sample initialization */
+int sample_initialize() {
+	int N = 5;
+
+	sample sam;
+	sam.set_up(N);
+
+	for(int i=0; i< N; i++)
+		sam.values[i] = i + 3;
+	if(HIGHD_VERBOSE)
+		cerr<<"Mean of sample: "<<sam.calc_mean()<<endl;
+	return 0;
+}
+
+/* Test hypercube initialization */
+int hc_initialize() {
+	hypercube_highd hc;
+	int status = hc.set_up(4,3);
+	if(HIGHD_VERBOSE){
+		cerr<<"First constructor: Memory allocation = "<<status<<endl;
+		cerr<<"First constructor: Dimension: "<<hc.get_dim()<<endl;
+	}
+	hypercube_highd hc2(4,3);
+	if(HIGHD_VERBOSE){
+		cerr<<"Second constructor: Dimension: "<<hc2.get_dim()<<endl;
+	}
+	return 0;
+}
+
+/* Test setting coefficients on hypercube */
+int hc_setting() {
+	int L = 1000;
+
+	hypercube_highd hc(L,3);
+
+	double values[] = {0.1, 0.043, 1.3, -4.3};
+	int myints[] = {16,2,77,29};
+	vector<int> loci;
+	for(int i=0; i<4; i++) {
[…]

### 2016-11-22 FFPopSim/tests/lowd.cpp

@@ -1,0 +1,190 @@
+/**
+ * @file lowd.cpp
+ * @brief Tests for the low-dimensional simulation library.
+ * @author Richard Neher, Boris Shraiman, Fabio Zanini
+ * @version 
+ * @date 2012-04-20
+ */
+/* Include directives */
+#include <string>
+#include "ffpopsim_lowd.h"
+#define LOWD_BADARG -1354341
+#define NOTHING 1e-10
+
+/* Be verbose? */
+#define LOWD_VERBOSE 1
+
+/* Test generic library access */
+int library_access() {
+	index_value_pair_t ivp;
+	ivp.index = 4;
+	ivp.val = 4.5;
+
+	if(LOWD_VERBOSE) {
+		cerr<<"index_value_pair: index = "<<ivp.index;
+		cerr<<", val = "<<ivp.val<<endl;
+	}
+	return 0;
+}
+
+
+/* Test sample initialization */
+int sample_initialize() {
+	int N = 5;
+
+	sample sam;
+	sam.set_up(N);
+
+	for(int i=0; i< N; i++)
+		sam.values[i] = i + 3;
+	if(LOWD_VERBOSE)
+		cerr<<"Mean of sample: "<<sam.calc_mean()<<endl;
+	return 0;
+}
+
+/* Test hypercube_lowd initialization */
+int hc_initialize() {
+	hypercube_lowd hc;
+	int status = hc.set_up(4,3);
+	if(LOWD_VERBOSE){
+		cerr<<"Memory allocation = "<<status<<endl;
+		cerr<<"Dimension: "<<hc.get_dim()<<endl;
+	}
+	return 0;
+}
+
+/* Test setting coefficients on hypercube_lowd */
+int hc_setting() {
+	int L = 2;
+	double additive[L];
+
+	for(int i=0; i<L;i++) {
+		additive[i] = 3;
+	}
+
+	hypercube_lowd hc(3, L);
+	hc.reset();
+	hc.additive(additive);
+	hc.fft_coeff_to_func();
+	
+	if(LOWD_VERBOSE){
+		cerr<<"Func values: ";
+		for(int gt=0; gt < (1<<L); gt++)
+			cerr<<hc.get_func(gt)<<" ";
+		cerr<<endl;
+
+		cerr<<"Coeff values: ";
+		for(int gt=0; gt < (1<<L); gt++)
+			cerr<<hc.get_coeff(gt)<<" ";
+		cerr<<endl;
+	}
+	return 0;
+}
+
[…]

### 2016-11-22 FFPopSim/tests/python_highd.py

@@ -1,0 +1,71 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       22/08/12
+content:    Test script for the python bindings (haploid_highd)
+'''
+
+# Import module
+import sys
+sys.path.append('../pkg/python')
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+from Bio import Phylo
+
+
+# Globals
+L = 1000    # number of loci
+N = 300    # population size
+
+
+# Construct class
+pop = h.haploid_highd(L, all_polymorphic=False)
+
+# Start tracking genealogy of two loci
+pop.track_locus_genealogy([3, 60])
+
+# Test fitness landscapes
+rep = np.zeros(L)
+rep[np.random.random(L) > 0.5] = -0.1
+pop.set_trait_additive(rep)
+
+# Show the additive part of the fitness landscape
+print pop.get_trait_additive()
+
+# Test population initialization
+pop.track_locus_genealogy([3,6])
+pop.set_wildtype(N)
+#pop.set_allele_frequencies([0.3] * L, N)
+pop.mutation_rate = 1e-5
+pop.outcrossing_rate = 1e-2
+pop.crossover_rate = 1e-3
+
+# Test allele frequency readout
+print np.max(pop.get_allele_frequency(4))
+
+# Test evolution
+from time import time as ti
+t0 = ti()
+pop.evolve(30)
+t1 = ti()
+print 'Time for evolving population for 30 generations: {:1.1f} s'.format(t1-t0)
+
+## Write genotypes
+#pop.write_genotypes('test.txt', 100)
+#pop.write_genotypes_compressed('test.npz', 100)
+
+## Plot histograms
+#plt.ion()
+#pop.plot_fitness_histogram()
+#pop.plot_divergence_histogram(color='r')
+#pop.plot_diversity_histogram(color='g')
+
+# Look at the genealogy
+print pop.genealogy
+tree = pop.genealogy.get_tree(3)
+subtree = tree.create_subtree_from_keys([tree.leafs[1], tree.leafs[2]])
+
+# Convert the tree into Biopython format and plot it
+treeBio = tree.to_Biopython_tree()
+Phylo.draw(treeBio)

### 2016-11-22 FFPopSim/tests/python_hiv.py

@@ -1,0 +1,54 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       25/04/12
+content:    Test script for the python bindings
+'''
+
+# Import module
+import sys
+sys.path.append('../pkg/python')
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+# Construct class
+pop = h.hivpopulation(1000)
+
+# Test I/O fitness landscapes
+pop.set_replication_landscape(lethal_fraction=0.05,
+                            number_valleys=0)
+pop.read_replication_coefficients('hiv_model.dat')
+rep = pop.get_replication_additive()
+rep[np.random.random(10000) > 0.5] = -0.1
+pop.set_replication_additive(rep)
+
+# Show the additive part of the fitness landscape
+print pop.get_trait_additive()
+
+# Test population initialization
+pop.set_allele_frequencies([0.3] * h.HIVGENOME, 1000)
+
+# Test allele frequency readout
+print np.max(pop.get_allele_frequency(4))
+
+# Test evolution
+from time import time as ti
+t0 = ti()
+pop.evolve(30)
+t1 = ti()
+print 'Time for evolving HIV for 30 generations: {:1.1f} s'.format(t1-t0)
+
+# Write genotypes
+pop.write_genotypes('test.txt', 100)
+pop.write_genotypes_compressed('test.npz', 100)
+
+# Plot histograms
+plt.ion()
+pop.plot_fitness_histogram()
+pop.plot_divergence_histogram(color='r')
+pop.plot_diversity_histogram(color='g')
+
+# Test treatment changes
+pop.treatment = 0.4
+print pop.treatment

### 2016-11-22 FFPopSim/tests/python_lowd.py

@@ -1,0 +1,58 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       14/05/12
+content:    Test script for the python bindings to the low-dimensional
+            simulation
+'''
+
+# Import module
+import sys
+sys.path.append('../pkg/python')
+import numpy as np
+import matplotlib.pyplot as plt
+import FFPopSim as h
+
+# Construct class
+pop = h.haploid_lowd(4)
+
+# Test initialization
+#pop.set_allele_frequencies([0,0.3,0.6,0.9], 1000)
+pop.set_genotypes([1,2],[400,800])
+
+# Test setting the recombination/mutation rates
+pop.set_recombination_rates([0.01, 0.03, 0.02], h.SINGLE_CROSSOVER)
+pop.set_mutation_rates([0.003,0.002,0.004,0.005],
+                       [0.006,0.004,0.008,0.010])
+
+# Test getting the mutation rate
+print pop.get_mutation_rates(direction=0)
+print pop.get_mutation_rates(direction=1)
+
+# Test setting / getting fitness
+pop.set_fitness_additive([0.02,0.03,0.04,0.02])
+pop.get_fitnesses()
+
+# Test allele frequency readout
+print pop.get_allele_frequencies()
+
+# Test evolution
+gens = 100
+from time import time as ti
+t0 = ti()
+pop.evolve(gens)
+t1 = ti()
+print 'Time for evolving the population for '+str(gens)+' generations: {:1.1f} s'.format(t1-t0)
+
+# Print population size
+print pop.N
+
+# Test divergence / diversity statistics
+print pop.get_divergence_statistics()
+print pop.get_diversity_statistics()
+
+# Plot histograms
+plt.ion()
+pop.plot_fitness_histogram()
+pop.plot_divergence_histogram(color='r')
+pop.plot_diversity_histogram(color='g')

### 2016-11-22 FFPopSim/tests/recombination_lowd.cpp

@@ -1,0 +1,377 @@
+/**
+ * @file recombination_lowd.cpp
+ * @brief Test recombination routine for haploid_lowd
+ * @author Richard Neher, Fabio Zanini
+ * @version 
+ * @date 2012-08-28
+ *
+ * Tests for the recombination routine in haploid_lowd are performed,
+ * for the general case and for single crossover.
+ */
+/* Include directives */
+#include <string>
+#include "ffpopsim_lowd.h"
+#define LOWD_BADARG -1354341
+#define NOTHING 1e-10
+
+/* Be verbose? */
+#define LOWD_VERBOSE 1
+
+/* Declaration */
+class haploid_lowd_test : public haploid_lowd {
+public:
+	// construction / destruction
+	haploid_lowd_test(int L=1, int rngseed=0) : haploid_lowd(L, rngseed){};
+	~haploid_lowd_test();
+
+	//testing
+	int test_recombinant_distribution();
+	int test_recombination(double *rec_rates);
+	int mutation_drift_equilibrium(double** mutrates);
+	int test_single_crossover_set_rates();
+};
+
+
+/* Implementation */
+/**
+ * @brief Default destructor
+ */
+haploid_lowd_test::~haploid_lowd_test() {}
+
+/**
+ * @brief Test the recombination routine using Fourier transforms
+ *
+ * @returns zero if both routines agree, -1 otherwise
+ *
+ * Debugging routine: calculates the distribution of recombinants explicitly and
+ * compares the result to the recombinant distribution obtained via fourier transform
+ */
+int haploid_lowd_test::test_recombinant_distribution(){
+	double *test_rec;
+	double dev=0;
+	//allocate memory for the recombinant distribution calculated step-by-step
+	test_rec=new double [(1<<number_of_loci)];
+	int mother, father;
+	//now calculate the recombinant distribution from pairs of parents.
+	int gt1, gt2, rec_pattern, inherited;
+
+	cout <<"test_recombinant_distribution(): start..."<<endl;
+
+	if (recombination_model == FREE_RECOMBINATION){
+		//calculate recombinants the efficient way
+		calculate_recombinants_free();
+		for (gt1=0; gt1<(1<<number_of_loci); gt1++){	//target genotype
[…]

### 2016-11-22 FFPopSim/tests/test_genealogy.cpp

@@ -1,0 +1,311 @@
+/**
+ * @file test_genealogies.cpp
+ * @brief Tests for the genealogies in the high-dimensional simulation library.
+ * @author Richard Neher, Boris Shraiman, Fabio Zanini
+ * @version 
+ * @date 2012-04-20
+ */
+
+/* Include directives */
+#include "ffpopsim_highd.h"
+#define HIGHD_BADARG -1354341
+#define NOTHING 1e-10
+
+/* Be verbose? */
+#define HIGHD_VERBOSE 1
+
+// test neutral evolution
+int kingman_coalescent() {
+	int L = 1000;
+	int N = 100;
+	int nbins = 20;
+	haploid_highd pop(L);
+
+	pop.set_mutation_rate(1e-3);
+	pop.outcrossing_rate = 1e-1;
+	pop.crossover_rate = 1e-2;
+	pop.recombination_model = CROSSOVERS;
+	vector <int> gen_loci;
+	gen_loci.push_back(100);
+	gen_loci.push_back(500);
+	gen_loci.push_back(900);
+	pop.track_locus_genealogy(gen_loci);
+
+	pop.set_wildtype(N);		// start with a population of the right size
+
+	stat_t fitstat;
+	gsl_histogram *SFS = gsl_histogram_alloc(nbins);
+	gsl_histogram_set_ranges_uniform(SFS,0,1);
+	int nrun=500,T[4];
+	T[0]=0;T[1]=0; T[2]=0; T[3]=0;
+	pop.evolve(4*N);
+	for (int n=0; n<nrun; n++){
+		cout <<n<<" out of "<<nrun<<endl;
+		pop.evolve(N);
+
+		vector <int> clones;
+		vector <tree_key_t> clone_keys;
+		rooted_tree subtree;
+		tree_key_t temp;
+		for (unsigned int genlocus=0; genlocus<gen_loci.size(); genlocus++){
+			T[0]+=pop.get_generation() - pop.genealogy.trees[genlocus].get_MRCA().age-1;
+			pop.genealogy.trees[genlocus].SFS(SFS);
+
+			for (int cs = 2; cs<5; cs++){
+				clones.clear();
+				clone_keys.clear();
+				pop.random_clones(cs,&clones);
+				for (vector <int>::iterator clone=clones.begin(); clone!=clones.end(); clone++){
+					temp.age=pop.get_generation()-1;
[…]

### 2017-05-25 writeup/writeup.tex
writeup updates--expanding some areas, trying to unify

@@
 We then move on to discuss Neher et al.
-(2009), which summarized the dynamics of genetic draft in asexual populations, and Neher et al.
+(2011), which summarized the dynamics of genetic draft in asexual populations, and Neher et al.
 (2013), which extended these results to sexual populations using a scaling argument.
@@
 \end{equation}
-This can be solved using the characteristic equation $\frac{dx}{dt} = e^x + (1-\delta)e^{-x} - 2+\delta-y$: the requirement that the lineage starts at $t=0$ with one individual means $p_0(n,w) = \delta_{1,n} \delta (w)$, with $\delta$ the Kronecker and Dirac deltas respectively.
+This can be solved using the characteristic equation $\frac{dx}{dt} = e^x + (1-\delta)e^{-x} - 2+\delta-y$.
+It is found that $\Phi$ depends on $x$ and $t$ via $\frac{e^{-x}-a_{-}}{a_{+}-e^{-x}} \exp(-(1-\delta)(a_{+} - a_{-})t)$, where $a_{\pm}$ are given by
+\begin{equation}
+a_{\pm}(y) = \frac{2-\delta+y \pm \sqrt{(2-\delta+y)^2 -4(1-\delta)}}{2(1-\delta)}.
+\end{equation}
+Note that the requirement that the lineage start at $t=0$ with one individual means $p_0(n,w) = \delta_{1,n} \delta (w)$, with $\delta$ the Kronecker and Dirac deltas respectively.
 The boundary condition for $\Phi$ becomes $\Phi(x,y,0) = e^{-x}$.
@@
 This expression allows us to derive the ``weight" for each individual mutant bubble.
-We have assumed that the population size is large enough that transient bubble sizes don't ``butt up" against the population size.
-This assumption can be relaxed but is (to my mind) not particularly informative.
+We need to take the inverse Laplace transform of $\varphi (y)$ to do so.
+It is
+\begin{equation}
+P(w) = \frac{\exp(-(2-\delta)w)}{w\sqrt{1-\delta}}I_1(2w\sqrt{1-\delta}),
+\end{equation}
+where $I_1$ is a modified Bessel function of the first kind.
+Expanding out the modified Bessel function in the limit of $w \gg 1$ yields
+\begin{equation}
+P(w) \approx \frac{\exp(-(2-\delta-2\sqrt{1-\delta})w)}{w^{3/2}\sqrt{4\pi(1-\delta)^3}} \times (1-\frac{3}{16w\sqrt{1-\delta}} + O(1/w^2)).
+\end{equation}
+This has the behavior $P(w) \sim w^{-3/2}$ until dropping off at $w \sim 1/\delta^2$.
 
@@
 In this limit, the single mutants will be present at roughly linkage equilibrium: when their frequencies $x_a, x_b$ satisfy $(s+2\delta)x_ax_b - \delta(x_a+x_b) > 1/N$, the double mutant will begin to spread deterministically in the population (selection, the left side, will overwhelm drift, the right side).
+Thus, double mutants will spread deterministically while single mutants are at low frequency only if $\delta \gg s$.
 If selection against single mutants is strong, i.e., $\delta^2/s \gg \max (\mu, 1/N)$, the rate at which a population reaches this threshold will be sharply reduced.
-In linkage equilibrium, the frequencies will be distributed roughly as
+There are essentially two stable equilibria, one where all the population is wild-type and one where the population is all double mutant, with an unstable saddle point at $x_a = x_b \approx \delta/s$.
+
+If the population is large ($N \gg \delta^2/s$), then in linkage equilibrium, the frequencies will be distributed roughly as
 \begin{equation}
@@
 \end{equation}
-The population must drift along the ridge where $x_a = x_b$, which will take approximately $\tau \sim \exp(N\delta^2/s)$ generations: this is because the population must drift to the unstable saddle point at $x_a = x_b \approx \delta/s$ and then ride selection to the stable equilibrium at $x_a = x_b = 1$, and the exponent roughly sets the probability of making it to the saddle point from $x_a = x_b \approx 0$.
+In frequency space, $P$ peaks at fixation for either the wild type or double mutant and has a saddle point at $\delta/s$, as previously mentioned.
+If $\delta \ll s$, the population must drift along the ridge where $x_a = x_b$, which will take approximately $\log \tau \sim N\delta^2/s$ generations, because the probability of the population being near the saddle point is lower than that of being near the wild-type equilibrium by a factor of $\exp (-N\delta^2/s)$.
+The population must drift to the unstable saddle point at $x_a = x_b \approx \delta/s$ and then ride selection to the stable equilibrium at $x_a = x_b = 1$, and the exponent roughly sets the probability of making it to the saddle point from $x_a = x_b \approx 0$.
+This reasoning applies provided $r \gg s \gg \delta \gg \max (\sqrt{\mu s},\sqrt{s/N}$.
+Conversely, if $\delta \gg s$, what is necessary is for at least one mutant to drift close to fixation, which takes order $\log \tau = N\delta$ generations.
+Finally, if $\delta \gg r$, selection keeps the population in linkage equilibrium, and a lucky double mutant must drift up to high frequency while getting lucky enough to avoid recombination.
 
-Second, suppose $r \ll s$.
-A successful double mutant grows initially at a rate $s - r = \bar{s}$, which roughly sets the fixation probability.
+Now, suppose $r \ll s$.
+A successful double mutant grows initially at a rate $s - r = \tilde{s}$, which roughly sets the fixation probability.
 If $N\mu$ is large enough that single mutants are produced constantly, then the appearance of the single mutant is essentially deterministic, as in the asexual case.
 If not, then we must consider the branching process dynamics, but the wait time is dominated by the waiting time to the first successful single mutant (i.e., the first one destined for fixation), which is distributed as $\tau \approx 2N\mu/p$, with $p$ the success probability.
-$p$ can be decomposed into a probability that the mutation is successful due to successive mutations (something like tunneling, almost identical to the asexual case) or due to recombination between another nascent mutant subpopulation.
 
-We consider only the sexual case.
+Suppose a mutant lineage arises and drifts for $T$ generations.
+The probability that it will give rise to a successful double mutant and sweep can be expressed as
+\begin{equation}
+p \approx \int P_T \mathrm{prob}(T)dT.
+\end{equation}
+For times $t \ll \min(N, 1/\delta)$ the mutant is unlikely to reach a high frequency, and the expected number of individuals $\mathrm{E}(n(t)) = e^{-\delta t}$ is small ($\sim 1$) until approximately $t \sim 1/\delta$, when selection starts to ``cut off" the lineage size (as with asexual populations).
+In general the size of $n(t)$, as well as the number of mutants or recombinants produced by a lineage or in a given time interval, will have exponential tails, so a good approximation is to consider only the drift times and let the other variables be more or less fixed.
+Moreover, as selection ``cuts off" the lineage size and drift time around $1/\delta$, we can simply assume the drift time is limited by that.
+
+$P_T$ can be decomposed into a probability that the mutation is successful due to successive mutations (something like tunneling, almost identical to the asexual case) or due to recombination between another nascent mutant subpopulation.
+The asexual population route is almost identical to the previous section, with the exception that $\tilde{s}$, not $s$, sets the growth rate and hence fixation probability of a double mutant.
+
+We now zero in on the sexual path $P_{r,T}$.
 During the time $T$ over which a bubble of genotype $Ab$ persists, a number of individuals $\approx N\mu T$ $aB$ individuals arise.
-The success probability is therefore related to $N\mu T \int P_{r,T^\prime} \mathrm{prob} (T^\prime) dT^\prime$, where the last term corresponds to the probability density $\mathrm{prob}(T^\prime)$ of wait times for the $aB$ lineage, times the probability $P_{r,T^\prime}$ of success contingent on $T$.
+The success probability is therefore given by
+\begin{equation}
+P_{r,T} \approx N\mu T \int P_{r,T^\prime} \mathrm{prob} (T^\prime) dT^\prime,
+\end{equation} where the last term corresponds to the probability density $\mathrm{prob}(T^\prime)$ of wait times for the $aB$ lineage, times the probability $P_{r,T^\prime}$ of success contingent on $T$.
 The number of $AB$ lineages produced by recombination is of order $r/N \int n_{Ab}(t) n_{aB}(t) dt$, and the lineages reach sizes of $T$ and $T^\prime$ respectively, persisting for $T^\prime$ generations: so the total integral is simply $TT^{\prime 2}$.
-The total success probability $P_{r,T,T^\prime}$ therefore grows as $N\mu TT{^\prime 2}$ until it saturates near $1$ at $T^\prime \sim \sqrt{Nr\bar{s}}$.
+The total success probability $P_{r,T,T^\prime}$ therefore grows as $N\mu TT{^\prime 2}$ until it saturates near $1$ at $T^\prime \sim \sqrt{Nr\tilde{s}}$.
+The success probability becomes
+
+\begin{equation}
+P_{r,T} \sim
+\begin{cases}
+T^3 \mu r \tilde{s} & T \ll (N/r\tilde{s})^{1/3}, \\
+T^3/2 \mu \sqrt{Nr \tilde{s}} & (N/r\tilde{s})^{1/3} \ll T \ll (N\mu^2r\tilde{s})^{-1/3}, \\
+1 & T \gg (N\mu^2r\tilde{s})^{-1/3}.
+\end{cases}
+\end{equation}
 
@@
 Below some cutoff point $\Theta$, the establishment probability is essentially zero: above this cutoff point it starts to increase roughly as $P_\mathrm{fix} \sim x+s-r-\bar{x}$, so that the genetic background fitness $x$ may dominate.
-This may appear to increase markedly the probability of a successful double mutant, but the size of mutant bubbles will tend to be smaller due to the advancing mean fitness $\bar{x}$.
+This may appear to increase markedly the probability of a successful double mutant, but the size of mutant bubbles will tend to be smaller due to the advancing mean fitness $\bar{x} = \sigma^2t$ (i.e., the mean fitness is time dependent).
 
 We will focus solely on the valley crossing problem here.
-As in the sexual case, the birth of a double mutant can be interpreted as a Poisson process, so that
+The theory is rather similar to the drift case: consider the weight of mutant bubbles, then what happens to individuals within those bubbles.
+The culmination is, essentially, to show that the distribution of weights $P(w)$ scales not as $w^{-3/2}$, as in the drift case, but as $w^{-2}/\sqrt{2 \log w}$, i.e., smaller mutant bubbles.
+What have hitherto been missing from this theory are clear analytics for the contribution of the parameters $s, \delta, r, \sigma$ outside of whether $N\sigma \gg 1$ or not.
+(In this section, $s$ will be used to mean the opposite of $\delta$ for a nascent mutant bubble.
+In general the notation needs cleaning up.)
+Furthermore the only type of recombination explicitly covered here is the ``communal" recombination model, where individuals essentially recombine with the entire population: this can be shown to behave similarly to ``free" recombination (e.g., rapid template switching in HIV recombination).
+A discussion of how crossovers work will be covered in the next section.
+
+As in the drift case, the birth of a double mutant can be interpreted as a Poisson process, so that
 \begin{equation}
-P(T) = e^{-\mu \int_0^T (n_1(t) + n(2)t) dt}
+P(T) = e^{-\mu \int_0^T (n_1(t) + n_2(t)) dt}
 \end{equation}
@@
 This must be obtained by considering the dynamics of bubbles that arise on any of the genetic backgrounds $x$ in the population.
-As argued above, this is effectively negligible below a certain cutoff.
-The calculation still needs to be filled in, but the Laplace transform for the bubble size $\Phi(z) = \mathbb{E}(e^-zw)$ is given by
+We will determine this now.
+Let $x$ be the fitness of the lineage, $k$ be the number of mutants at the previous time step, $n$ be the current number, and $T$ and $t$ be the current and previous times.
+The fate of a mutant lineage is governed by the master equation
+\begin{align*}
+-\partial_t p(n,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(n,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(n,T|k,t,x) \\
+& +kp(n,T|k-1,t,x) \\
+&+rk\sum_{n^\prime} \int_{x^\prime} K_{x x^\prime} p(n-n^\prime , T|k-1, t, x) p(n^\prime,T|t,t,x^\prime).
+\end{align*}
+Here, $K$ represents a recombination ``kernel", i.e., a function that assigns to each $x$ an offspring fitness $x^\prime$ with some probability.
+The first term corresponds to the probability that nothing happens, the second to the duplication of one of the $k$ individuals (as usual, the death rate is normalized to $1$), the third to one of the $k$ individuals dying, and the last to outcrossing, yielding an individual with background fitness $x^\prime$.
+
+We can use generating functions to remove the convolution over $n$. Define $G(\lambda, T|k,t,x) = \sum_n \lambda^n p(n,T|k,t,x)$. Then
+\begin{align*}
+-\partial_t G(\lambda,T|k,t,x) = &-k(2+x-\bar{x}+s+r)G(\lambda,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)G(\lambda,T|k,t,x) \\
+& + kG(\lambda,T|k-1,t,x) \\
+& + rk\int_{x^\prime} K_{x x^\prime} G(\lambda, T|k-1, t, x) G(\lambda,T|t,t,x^\prime).
+\end{align*}
+From this we can use the fact that all $k$ individuals are independent, i.e., $G(\lambda,t|k,t,x) = G^k(\lambda,T|t,x)$.
+The right hand side becomes $-\partial_t G(\lambda,T|k,t,x) = -kG^{k-1}(\lambda,T|t,x)\partial_t G(\lambda,T|t,x)$. Dividing out $kG^{k-1}(\lambda,T|t,x)$ yields
+\begin{align*}
+-\partial_t G(\lambda,T|k,t,x) = &-(2+x-\bar{x}+s+r)G(\lambda,T|t,x) \\
+& + (1+x-\bar{x}+s)G^2(\lambda,T|t,x) \\
+& + 1 \\
+& + r\int_{x^\prime} K_{x x^\prime} G(\lambda,T|t,t,x^\prime).
+\end{align*}
+(The last term is justified by the fact that $G(\lambda,T|k-1,t,x)$ doesn't depend on $x^\prime$.)
+As usual, the boundary condition is $G(\lambda,T|T,x) = \lambda$.
+Substitutiong $\phi = 1 - G$ yields
+\begin{align*}
+\partial_t \phi(\lambda,T|k,t,x) = &-(x-\bar{x}+s-r)G(\lambda,T|t,x) \\
+& + (1+x-\bar{x}+s)\phi^2(\lambda,T|t,x) \\
+& - r\int_{x^\prime} K_{x x^\prime} \phi(\lambda,T|t,t,x^\prime).
+\end{align*}
+The boundary condition is now $\phi(\lambda,T|T,x) = 1-\lambda$.
+In the weak selection limit, the non-constant parts of the $\phi^2$ prefactor are small, so we have
+\begin{align*}
+-\partial_t \phi(\lambda,T|k,t,x) = &(x-\bar{x}+s-r)G(\lambda,T|t,x) \\
+& - \phi^2(\lambda,T|t,x) \\
+& + r\int_{x^\prime} K_{x x^\prime} \phi(\lambda,T|t,t,x^\prime).
+\end{align*}
+Using $\bar{x} = vt = \sigma^2 t$, we can rescale via $\psi(T,t,x) = \phi(\lambda,T,t,x)/(\sigma\epsilon)$, with $\epsilon = 1-\lambda$.
+Similarly rescaling $\tilde{r} = r/\sigma$, $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, and $\tau = \sigma(T-t)$ yields
+\begin{align*}
+\partial_t \psi(\tau,\chi) = &\tilde{r}\int K_{\chi \chi^\prime} \psi (\tau, \chi^\prime) \\
+& + (\chi + \tau + \tilde{s} - \tilde{r}) \psi (\tau,\chi) \\
+& - \epsilon \psi^2 (\tau,chi).
+\end{align*}
+Integrating this over the fitness distribution $P(\chi,\tau)$ would yield the scaled generating function $\Phi(\tau) = \int d\chi P(\chi,tau)\psi(\tau,\chi)$.
+We can proceed by leaving the bulk of the equation in the form of $\phi$, since ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
+Here $w$ represents (as before) the weight of a mutant bubble and evolves according to
+\begin{align*}
+-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(w,T|k,t,x) \\
+& +kp(w,T|k-1,t,x) \\
+&+rk \int dw^\prime \int_{x^\prime} K_{x x^\prime} p(w-w^\prime , T|k-1, t, x) p(w^\prime,T|t,t,x^\prime).
+\end{align*}
+We still have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$.
+Rescaling variables as above and introducing $\theta = \chi+\tau+\tilde{s}+z-\tilde{r}$ and $\hat{\phi} = (1-\hat{p})/\sigma)$ yields
 \begin{equation}
-\Phi(z) =
+\partial_t \hat{\phi}(\tau,z,\chi) = z + \tilde{r} \Phi(\tau, z) + \theta \hat{\phi}(\tau,z,\chi) - \hat{\phi}^2(\tau,z,\chi).
+\end{equation}
+The boundary condition is given by $\hat{\phi}(0,z,\chi) = 0$, since the ``weight" is zero at time zero.
+We can solve this by sending $\tau$ to infinity, assuming that tunneling is rare and requires many mutant ``bubbles" to occur.
+
+We will need to figure out what happens to mutations in the ``bulk" of the wave versus the ``nose".
+The above expression for $\Phi$ yields a differential equation
+\begin{equation}
+\partial_\tau \Phi(\tau) = \partial_\tau \int d\chi P(\chi,\tau)\psi(\tau,\chi) = \tilde{s} \Phi(\tau) - \epsilon \int d\chi P(\tau,\chi) \psi^2 (\tau,\chi).
+\end{equation}
+Note that, for beneficial mutations, the two right hand terms balance each other over long time scales, but for neutral and deleterious mutations, no such steady state exists.
+For the ``communal" recombination model used in Neher and Shraiman (2011) and (2010), the recombination contribution in the ODE for $\psi$ simplifies to $\tilde{r} \Phi(\tau)$.
+
+We proceed in the limits of large positive and negative growth rate $\beta = \chi+\tau+\tilde{s}-\tilde{r}$.
+For $\beta < 0$, only the first term in the equation for $\psi$ matters, as $psi < 1/\epsilon$: for $\beta > 0$, only the second one matters.
+Accordingly,
+\begin{equation}
+\psi(\tau,\chi) =
 \begin{cases}
-ze^{r\sigma^{-1}\sqrt{-2\log z}} &|s| \ll re^{-r\sigma^{-1}\sqrt{-2\log z}} \\
-\frac{z}{|s|} &|s| \gg re^{-r\sigma^{-1}\sqrt{-2\log z}}.
+\tilde{r} e^{\beta^2/2} \int^\tau_0 d\tau^\prime \Phi(\tau^\prime)e^{-\beta^2/2} & \beta < \beta_c (\tau) \\
+\beta/\epsilon & \beta > \beta_c (\tau).
+\end{cases}
+\end{equation}
+
+(Still to-do: write up $\beta_c$ and explain how the crossover occurs.
+Some of the following calculations will still need cleaning up.)
+
+Correspondingly,
+\begin{equation}
+\hat{\phi}(z,\chi) =
+\begin{cases}
+e^{\beta^2/2}(z+\tilde{r}\Phi(z)) & \beta < \beta_c (\tau) \\
+\beta & \beta > \beta_c (\tau).
+\end{cases}
+\end{equation}
+
+The jump from this condition to the Laplace transform for the bubble size $\Phi(z) = \mathbb{E}(e^{-zw})$ needs to be filled in, but the transform in the low $\tilde{s}$ (neutral) or high $\tilde{r}$ limit is given by
+\begin{equation}
+\Phi(z) = z\frac{e^{\tilde{r}\beta_c - 1}}{\tilde{r}} = 
+\begin{cases}
+z\sqrt{-2\log z} & \tilde{r} \beta_c \ll 1 \\
+z e^{\tilde{r}\beta_c}/\tilde{r} & \beta_c \gg 1.
 \end{cases}
@@
 It is worth noting that $r$ here is the total recombination rate between two arbitrary loci, not necessarily between the two loci of interest.
-In the draft dominated regime, we can use the Laplace transform to compute the weight distribution, which scales as $P(w)\sim 1/w^2 \sqrt{2\log w}$, compared to $P(w) \sim w^{-3/2}$ in the drift case.
+
+The inverse Laplace transform is then given by
+\begin{equation}
+P(w) = \oint \frac{dz}{2\pi i} e^{zw} \bar{\Phi}.
+\end{equation}
+The low-$r$ regime above yields
+\begin{equation}
+P(w) = \oint \frac{dz}{2\pi i} e^{zw}z\sqrt{\log z^2} \sim \frac{1}{w^2\sqrt{\log w^2}}.
+\end{equation}
+The weight distribution scales as $P(w)\sim 1/w^2 \sqrt{2\log w}$, compared to $P(w) \sim w^{-3/2}$ in the drift case.
 As expected, draft decreases the bubble size distribution but lessens the dependence on $\delta$.
 Some more precise analytics will be needed here.
+In particular, note that, looking at the high and low $\tilde{r}$ expressions for $\Phi$, it seems the weight will barely depend at all on $\sigma$, which seems... wrong.
 
@@
 
-At sufficiently high recombination rates $r$ between loci, the concept of those loci segregating within a common fitness wave will start to break down.
+At sufficiently high recombination rates $r$ between loci, or (alternately) in a large enough genome with crossovers, the concept of those loci segregating within a common fitness wave will start to break down.
 In fact high enough recombination can effectively cause different parts of the genome to evolve neutrally even though the rate of adaptation is high.

### 2017-09-01 src/bubble_draft.py
updated simulation and error checking scripts

@@ -1,0 +1,96 @@
+import sys
+sys.path.append('/home/tkessinger/.local/lib/python3.5/site-packages')
+
+import numpy as np
+import fwdpy11
+import fwdpy11.model_params
+import fwdpy11.fitness
+import fwdpy11.wright_fisher
+
+# plotting stuff
+import matplotlib.pyplot as plt
+import seaborn as sns
+import powerlaw as pl
+
+class BubbleRecorder(object):
+    def __init__(self,ngens,N):
+        #An index variable referring to where we will next store data
+        self.i = 0
+        self.weights = dict()
+        
+    def __call__(self,pop):
+        for mut in range(len(pop.mutations)):
+            # infinite sites model so each mutation has unique position
+            mkey = pop.mutations[mut].pos
+            # if mutation neutral (neutral ones draft) 
+            if pop.mutations[mut].neutral:
+                # bubble is open
+                if pop.mcounts[mut] > 0:
+                    # update bubble size of currently segregating mutation
+                    if mkey in self.weights:
+                        self.weights[mkey] += pop.mcounts[mut]
+                    # new segregating mutation; open bubble
+                    else:
+                        self.weights[mkey] = pop.mcounts[mut]
+                    
+def evolve_draft(ngens, N, s, mu, r, burn = 500, seed = 42):
+    """
+    Evolve drafting region and record bubble sizes
+    """
+
+    pop = fwdpy11.SlocusPop(N)
+    rng=fwdpy11.GSLrng(seed)
+    
+    params = fwdpy11.model_params.SlocusParams(
+        nregions = [fwdpy11.Region(0, 1, 1)],
+        sregions = [fwdpy11.ExpS(0, 1, 1, s, 1.0)],
+        recregions = [fwdpy11.Region(0, 1, 1)],
+        gvalue = fwdpy11.fitness.SlocusAdditive(2.0),
+        demography = np.array([N]*ngens, dtype=np.uint32),
+        rates=(mu, mu, r)
+        )
+    recorder = BubbleRecorder(ngens, N)
+    
+    fwdpy11.wright_fisher.evolve(rng,pop,params,recorder)
+
+    return pop, np.array(list(recorder.weights.values()))
+
+
+# plot using the 'powerlaw' package and show fit
[…]

### 2017-09-01 src/bubble_sizes_multi.py

@@ -72,7 +72,7 @@ pop = h.haploid_highd(L, number_of_traits=2, number_of_mutant_loci = 1, most_pol
 pop.carrying_capacity = N
 pop.mutation_rate = nothing
 #pop._set_mutant_loci([mutant_loc])
-pop._set_mutant_loci(mutant_loc)
+pop._set_mutant_loci(np.int32(mutant_loc))
 
 pop.recombination_model = h.FREE_RECOMBINATION #recombination model, can be either FREE_RECOMBINATION or CROSSOVERS
 pop.outcrossing_rate = rho
@@ -91,8 +91,8 @@ run_time = args.runtime
 #run_time = np.max([1e4,10*N]) #number of generations to simulate
 
 num_attempts = 0 #attempts at crossing the valley, i.e., number of deleterious or neutral single mutants
-mutation_in_play = False #flag to see whether the mutations are currently segregating
-weights = []
+mutation_in_play = [False for x in range(mut_loci)] #flag to see whether the mutations are currently segregating
+weights = [[] for x in range(mut_loci)]
 bubble_openings = []
 bubble_closings = []
 
@@ -110,11 +110,12 @@ selection_coefficients[mutant_loc]=1e-10
 pop.set_trait_additive(selection_coefficients,0)
 pop.set_trait_additive(np.zeros(L),1)
 
+
 #pop.trait_weights[0] = 0
 #print pop.trait_weights
 
-for loc in mutant_loc:
-    pop.add_trait_coefficient(deleterious_effects[0],loc,1)
+#for loc in mutant_loc:
+pop.add_trait_coefficient(deleterious_effects[0],np.int32(mutant_loc),1)
 #mutant_loc might need to be a list
 afreqs = []
 
@@ -127,26 +128,27 @@ if not os.path.exists('output/bubble_size_sims_'+args.name):
 
 vars = []
 
+
 while pop.generation < run_time:
 #    if pop.get_allele_frequency(mutant_loc) == 0:
 #        pop.flip_single_locus(mutant_loc)
     #print pop.generation
     pfit = pop.get_trait_statistics(0)
-    #pfit1 = pop.get_trait_statistics(1)
-    #fitvar = pop.get_fitness_statistics().variance
+    pfit1 = pop.get_trait_statistics(1)
+    fitvar = pop.get_fitness_statistics().variance
     #vars.append(pfit.variance)
[…]

### 2017-09-01 src/plot_bubbles.py

@@ -7,8 +7,9 @@ Description: Glob and plot bubble size distribution.
 '''
 import sys
 sys.path.append('/home/tkessinger/tools/FFPopSim/pkg/python/')
+sys.path.append('/home/tkessinger/.local/lib/python3.5/site-packages')
 import numpy as np
-import FFPopSim as h
+#import FFPopSim as h
 import random as rand
 import matplotlib.pyplot as plt
 from time import time
@@ -16,12 +17,13 @@ import argparse
 import math
 import cPickle as pickle
 import glob
+import powerlaw as pl
 
 import seaborn as sns
 sns.set_context('paper', font_scale=1.5)
 sns.set_style('white')
 
-name = '4'
+name = '10'
 
 numruns = {}
 sigma_range = {}
@@ -63,37 +65,100 @@ if name == '4':
     delta_range[name] = [0.0, 1e-2]
     run_time[name] = 1000000
     numruns[name] = 5
+    
+if name == '5':
+
+    N_range[name] = [10000]
+    sigma_range[name] = [2e-8, 2e-1]
+    #delta_range[name] = [0, 1e-3, 1e-4, 1e-5]
+    delta_range[name] = [0.0, 1e-2]
+    run_time[name] = 50000
+    numruns[name] = 10
+
+if name == '6':
+
+    s_mean_range = {}
+    N_range[name] = [10000]
+    sigma_range[name] = [1e-6, 1e-2]
+    #delta_range[name] = [0, 1e-3, 1e-4, 1e-5]
+    delta_range[name] = [0.0, 1e-2]
+    run_time[name] = 50000
+    numruns[name] = 10
+    
+if name == '7':
+
+    s_mean_range = {}
+    N_range[name] = [50000]
+    sigma_range[name] = [1e-6, 1e-2]
+    #delta_range[name] = [0, 1e-3, 1e-4, 1e-5]
+    delta_range[name] = [0.0, 1e-2]
+    run_time[name] = 200000
+    numruns[name] = 10
+    
+if name == '8':
+
+    N_range[name] = [100000]
+    sigma_range[name] = [2e-8, 2e-1]
+    delta_range[name] = [0.0, 1e-2]
+    run_time[name] = 1000000
+    numruns[name] = 10
+
+if name == '9':
+
+    N_range[name] = [100000]
+    sigma_range[name] = [2e-8, 2e-1]
+    delta_range[name] = [0.0, 1e-2]
+    run_time[name] = 1000000
[…]

### 2017-09-01 src/simulate_population.py

@@ -8,6 +8,7 @@ Description: Simulation of valley crossing using FFPopSim.
 '''
 import sys
 sys.path.append('/home/tkessinger/tools/FFPopSim/pkg/python/')
+sys.path.append('/home/tkessinger/Documents/draft_valley_crossing/FFPopSim/pkg/python/')
 import numpy as np
 import FFPopSim as h
 import random as rand
@@ -32,6 +33,7 @@ parser.add_argument('--numsamples',type=int,default=10)
 parser.add_argument('--runno', type=int, default=0)
 parser.add_argument('--runtime', type=int, default=100000)
 parser.add_argument('--name', type=str, default='test')
+parser.add_argument('--fixed_s', type=bool, default=False)
 
 args=parser.parse_args()
 
@@ -71,7 +73,11 @@ pop.crossover_rate = cx
 
 pop.set_wildtype(N)
 
-s_mean = 1e-6 #average effect of beneficial mutations: this turns out to be mostly irrelevant due to sigma, above
+if args.fixed_s:
+    s_mean = sigma
+    sigma_hist = []
+else:
+    s_mean = 1e-2 #average effect of beneficial mutations: this turns out to be mostly irrelevant due to sigma, above
 base_burn_time = 0.1*N #number of generations to wait for population to equilibrate: ideally this should be of order T_MRCA
 burn_time = base_burn_time
 burned_in = False #flag for whether the population has equilibrated
@@ -111,7 +117,10 @@ pop.add_trait_coefficient(beneficial_effect, mutant_locs,1) #set the epistatic e
 
 afreqs = []
 
-prefix = 'N_'+str(N)+'_sigma_'+str(sigma)+'_mu_'+str(mu)+'_rho_'+str(rho)+'_s_'+str(s)+'_delta_'+str(delta)
+if args.fixed_s:
+    prefix = 'N_'+str(N)+'_smean_'+str(sigma)+'_mu_'+str(mu)+'_rho_'+str(rho)+'_s_'+str(s)+'_delta_'+str(delta)
+else:
+    prefix = 'N_'+str(N)+'_sigma_'+str(sigma)+'_mu_'+str(mu)+'_rho_'+str(rho)+'_s_'+str(s)+'_delta_'+str(delta)
 
 if not os.path.exists('output/valley_crossing_sims_'+args.name):
     os.makedirs('output/valley_crossing_sims_'+args.name)
[…]

### 2017-09-01 writeup/writeup.tex
cleaner version of weight distribution calcs

@@
 -\partial_t p(n,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(n,T|k,t,x) \\
-& + k(1+x-\bar{x}+s)p(n,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(n,T|k+1,t,x) \\
 & +kp(n,T|k-1,t,x) \\
@@
 
-We can use generating functions to remove the convolution over $n$. Define $G(\lambda, T|k,t,x) = \sum_n \lambda^n p(n,T|k,t,x)$. Then
+We can use generating functions to remove the convolution over $n$. Define $\hat{p}(\lambda, T|k,t,x) = \sum_n \lambda^n p(n,T|k,t,x)$. Then
 \begin{align*}
--\partial_t G(\lambda,T|k,t,x) = &-k(2+x-\bar{x}+s+r)G(\lambda,T|k,t,x) \\
-& + k(1+x-\bar{x}+s)G(\lambda,T|k,t,x) \\
-& + kG(\lambda,T|k-1,t,x) \\
-& + rk\int_{x^\prime} K_{x x^\prime} G(\lambda, T|k-1, t, x) G(\lambda,T|t,t,x^\prime).
+-\partial_t \hat{p}(\lambda,T|k,t,x) = &-k(2+x-\bar{x}+s+r)\hat{p}(\lambda,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)\hat{p}(\lambda,T|k+1,t,x) \\
+& + k\hat{p}(\lambda,T|k-1,t,x) \\
+& + rk\int_{x^\prime} K_{x x^\prime} \hat{p}(\lambda, T|k-1, t, x) \hat{p}(\lambda,T|t,t,x^\prime).
 \end{align*}
-From this we can use the fact that all $k$ individuals are independent, i.e., $G(\lambda,t|k,t,x) = G^k(\lambda,T|t,x)$.
-The right hand side becomes $-\partial_t G(\lambda,T|k,t,x) = -kG^{k-1}(\lambda,T|t,x)\partial_t G(\lambda,T|t,x)$. Dividing out $kG^{k-1}(\lambda,T|t,x)$ yields
+From this we can use the fact that all $k$ individuals are independent, i.e., $\hat{p}(\lambda,t|k,t,x) = \hat{p}^k(\lambda,T|t,x)$.
+The right hand side becomes $-\partial_t \hat{p}(\lambda,T|k,t,x) = -k\hat{p}^{k-1}(\lambda,T|t,x)\partial_t \hat{p}(\lambda,T|t,x)$.
+Dividing out $k\hat{p}^{k-1}(\lambda,T|t,x)$ yields
 \begin{align*}
--\partial_t G(\lambda,T|k,t,x) = &-(2+x-\bar{x}+s+r)G(\lambda,T|t,x) \\
-& + (1+x-\bar{x}+s)G^2(\lambda,T|t,x) \\
+-\partial_t \hat{p}(\lambda,T|k,t,x) = &-(2+x-\bar{x}+s+r)\hat{p}(\lambda,T|t,x) \\
+& + (1+x-\bar{x}+s)\hat{p}^2(\lambda,T|t,x) \\
 & + 1 \\
-& + r\int_{x^\prime} K_{x x^\prime} G(\lambda,T|t,t,x^\prime).
+& + r\int_{x^\prime} K_{x x^\prime} \hat{p}(\lambda,T|t,t,x^\prime).
 \end{align*}
-(The last term is justified by the fact that $G(\lambda,T|k-1,t,x)$ doesn't depend on $x^\prime$.)
-As usual, the boundary condition is $G(\lambda,T|T,x) = \lambda$.
-Substitutiong $\phi = 1 - G$ yields
+(The last term is justified by the fact that $\hat{p}(\lambda,T|k-1,t,x)$ doesn't depend on $x^\prime$.)
+As usual, the boundary condition is $\hat{p}(\lambda,T|T,x) = \lambda$, corresponding to one individual at time $0$.
+Substitutiong $\phi = 1 - \hat{p}$ yields
 \begin{align*}
-\partial_t \phi(\lambda,T|k,t,x) = &-(x-\bar{x}+s-r)G(\lambda,T|t,x) \\
+\partial_t \phi(\lambda,T|k,t,x) = &-(x-\bar{x}+s-r)\phi(\lambda,T|t,x) \\
 & + (1+x-\bar{x}+s)\phi^2(\lambda,T|t,x) \\
@@
 & + (\chi + \tau + \tilde{s} - \tilde{r}) \psi (\tau,\chi) \\
-& - \epsilon \psi^2 (\tau,chi).
+& - \epsilon \psi^2 (\tau,\chi).
 \end{align*}
-Integrating this over the fitness distribution $P(\chi,\tau)$ would yield the scaled generating function $\Phi(\tau) = \int d\chi P(\chi,tau)\psi(\tau,\chi)$.
+Integrating this over the fitness distribution $P(\chi,\tau)$ would yield the scaled generating function $\Phi(\tau) = \int d\chi P(\chi,\tau)\psi(\tau,\chi)$.
+(Note that integrating the recombination term over $P(\chi,\tau)$, in the free recombination model, yields simply $\tilde{r}\Phi(\tau)$.)
+
+Similar reasoning will allow us to derive the behavior of the weight distribution.
 We can proceed by leaving the bulk of the equation in the form of $\phi$, since ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
@@
 -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(w,T|k,t,x) \\
-& + k(1+x-\bar{x}+s)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
 & +kp(w,T|k-1,t,x) \\
@@
 \end{align*}
-We still have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$.
-Rescaling variables as above and introducing $\theta = \chi+\tau+\tilde{s}+z-\tilde{r}$ and $\hat{\phi} = (1-\hat{p})/\sigma)$ yields
+We still have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$, by the fact that each of the $k$ copies of the mutant allele is independent of the other.
+Differentiating $\hat{p}$ with respect to $w$ pulls out a factor of $-z$, but otherwise the behavior is very similar to the previous case.
+Rescaling variables as above and introducing $\theta = \chi+\tau+\tilde{s}+z-\tilde{r}$ and $\phi = (1-\hat{p}(z,T|1,t,x))/\sigma)$ yields
 \begin{equation}
-\partial_t \hat{\phi}(\tau,z,\chi) = z + \tilde{r} \Phi(\tau, z) + \theta \hat{\phi}(\tau,z,\chi) - \hat{\phi}^2(\tau,z,\chi).
+\partial_\tau \phi(\tau,z,\chi) = z + \tilde{r} \Phi(\tau, z) + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi),
 \end{equation}
-The boundary condition is given by $\hat{\phi}(0,z,\chi) = 0$, since the ``weight" is zero at time zero.
-We can solve this by sending $\tau$ to infinity, assuming that tunneling is rare and requires many mutant ``bubbles" to occur.
+where $\Phi(\tau, z) = \int d\chi P(\chi, \tau) \phi(\tau, z, \chi)$ is, as before, the rescaled version of $\phi$ when integrating over the entire fitness distribution $P(\chi, \tau)$.
+(In the communal recombination model, the contribution of the recombination term simply becomes $\tilde{r}\Phi$, since the fitness of an individual is, in that model, determined by simply sampling randomly from the population.)
 
-We will need to figure out what happens to mutations in the ``bulk" of the wave versus the ``nose".
-The above expression for $\Phi$ yields a differential equation
+We can send $\tau$ to infinity and look for steady state solutions, since all we are interested in is the size of a single, short lived bubble (i.e., there is only ever one bubble at a time).
+For large values of $\theta$, the dominant terms are $\theta \phi(\tau,z,\chi)$ and $-\phi^2(\tau,z,\chi)$.
+Setting their sum equal to zero (steady state solution) and solving for $\phi$ yields $\phi(z,\chi) \approx \theta$.
+
+For small values of $\theta$, the remaining terms are more important.
+We have $\partial_\tau \phi = z + \tilde{r} \Phi(z) + \theta \phi(z, \chi)$.
+The relevant ``integrating factor" here is $\mu(\theta) = e^{\int -\theta d\theta} = e^{-\theta^2/2}$ (the use of $\theta$ rather than $\tau$ is justified by the fact that they depend linearly on each other).
+We therefore obtain, in the low $\theta$ limit,
+
 \begin{equation}
-\partial_\tau \Phi(\tau) = \partial_\tau \int d\chi P(\chi,\tau)\psi(\tau,\chi) = \tilde{s} \Phi(\tau) - \epsilon \int d\chi P(\tau,\chi) \psi^2 (\tau,\chi).
+\phi(z,\chi) = \frac{\int_0^\tau (z+\tilde{r}\Phi(\tau^\prime)) e^{-\theta^\prime^2/2} d\tau^\prime}{e^{-\theta^2/2}}.
 \end{equation}
-Note that, for beneficial mutations, the two right hand terms balance each other over long time scales, but for neutral and deleterious mutations, no such steady state exists.
-For the ``communal" recombination model used in Neher and Shraiman (2011) and (2010), the recombination contribution in the ODE for $\psi$ simplifies to $\tilde{r} \Phi(\tau)$.
 
-We proceed in the limits of large positive and negative growth rate $\beta = \chi+\tau+\tilde{s}-\tilde{r}$.
-For $\beta < 0$, only the first term in the equation for $\psi$ matters, as $psi < 1/\epsilon$: for $\beta > 0$, only the second one matters.
-Accordingly,
+The $z$ term in the integrand simply becomes $z\sqrt{2\pi} \sim z$.
+The second term needs to be treated with slightly more care.
+The major contribution to the integral will come from a well defined, large $\tau^\prime$, corresponding to the maximum of $\exp(-\theta^\prime^2/2 + \log \Phi(\tau^\prime))$ at $\tau^\prime = \tau - \theta$.
+$\Phi(\tau)$ changes slowly and can be pulled out of the integral, yielding another Gaussian integral.
+We therefore have $\phi(z,\chi) \approx (z+\tilde{r}\Phi(z))e^{\theta^2}/2$, or, summarizing:
+
 \begin{equation}
-\psi(\tau,\chi) =
+\phi(z,\chi) \approx
 \begin{cases}
-\tilde{r} e^{\beta^2/2} \int^\tau_0 d\tau^\prime \Phi(\tau^\prime)e^{-\beta^2/2} & \beta < \beta_c (\tau) \\
-\beta/\epsilon & \beta > \beta_c (\tau).
+(z+\tilde{r}\Phi(z))e^{\theta^2}/2 & \theta \ll \Theta_c, \\
+\theta & \theta \gg \Theta_c.
 \end{cases}
@@
 
-(Still to-do: write up $\beta_c$ and explain how the crossover occurs.
-Some of the following calculations will still need cleaning up.)
+The crossover between these two solutions occurs roughly at some critical $\Theta_c$ given by setting them equal to each other:
 
-Correspondingly,
 \begin{equation}
-\hat{\phi}(z,\chi) =
+(z+\tilde{r}\Phi(z))e^{\Theta_c^2/2} = \Theta_c.
+\end{equation}
+
+We would still like to find $\Phi(z)$. The solvability condition given for $\Phi(z)$ by integrating the equation for $\partial_\tau \phi(z,\chi)$ over the fitness distribution is
+
+\begin{equation}
+0 = z + \tilde{s}\Phi(z) - \int \frac{d\chi}{\sqrt{2\pi}}e^{-\chi^2/2}\phi^2(z,\chi).
+\end{equation}
+
+For small $\tilde{r}$, the integral on the right is dominated by the crossover point $\Theta_c$.
+Using the above form of $\phi(z,\chi)$ allows us to solve the integral: it becomes $ze^{-\tilde{r}\Theta_c} + \tilde{r}e^{-\tilde{r}\Theta_c}\Phi(z)$.
+We thus have
+
+\begin{equation}
+\Phi(z) = z\frac{1-e^{-\tilde{r}\Theta_c}}{\tilde{r}e^{-\tilde{r}\Theta_c} - \tilde{s}}.
+\end{equation}
+
+Using the crossover condition gives us an expression for $\Theta_c$:
+
+\begin{equation}
+\Theta_c e^{-\Theta_c^2/2} = z\frac{\tilde{r}-\tilde{s}}{\tilde{r}e^{-\tilde{r}\Theta_c} - \tilde{s}}.
+\end{equation}
+
+If $\tilde{s}$ is more important than $\tilde{r}e^{-\tilde{r}\Theta_c}$, we have $\Theta_c = \sqrt{-2\log(ze^{\tilde{r}\Theta_c}/\Theta_c}$, which means
+
+\begin{equation}
+\Phi(z) = z\frac{e^{\tilde{r}\Theta_c - 1}}{\tilde{r}} =
 \begin{cases}
-e^{\beta^2/2}(z+\tilde{r}\Phi(z)) & \beta < \beta_c (\tau) \\
-\beta & \beta > \beta_c (\tau).
+z\sqrt{-2\log z} & \tilde{r}\Theta_c \ll 1 \\
+ze^{\tilde{r}\Theta_c}/\tilde{r} & \tilde{r}\Theta_c \gg 1.
 \end{cases}
@@
 
-The jump from this condition to the Laplace transform for the bubble size $\Phi(z) = \mathbb{E}(e^{-zw})$ needs to be filled in, but the transform in the low $\tilde{s}$ (neutral) or high $\tilde{r}$ limit is given by
+If the opposite is true, we have instead
+
 \begin{equation}
-\Phi(z) = z\frac{e^{\tilde{r}\beta_c - 1}}{\tilde{r}} = 
+\Phi(z) = \frac{z}{\tilde{s}}(1-e^{\tilde{r}\Theta_c}) =
 \begin{cases}
-z\sqrt{-2\log z} & \tilde{r} \beta_c \ll 1 \\
-z e^{\tilde{r}\beta_c}/\tilde{r} & \beta_c \gg 1.
+\frac{z\tilde{r}\sqrt{-2\log z}}{\tilde{s}} & \tilde{r}\Theta_c \ll 1 \\
+\frac{z}{\tilde{s}} & \tilde{r}\Theta_c \gg 1.
 \end{cases}
 \end{equation}
-It is worth noting that $r$ here is the total recombination rate between two arbitrary loci, not necessarily between the two loci of interest.
 
-The inverse Laplace transform is then given by
+Note the dependence on $z\sqrt{-2\log z}$ in the low $\tilde{r}$ limit.
+We can now compute the weight distribution, having finally obtained $\Phi(z)$, the Laplace transform thereof.
+We have
+
 \begin{equation}
-P(w) = \oint \frac{dz}{2\pi i} e^{zw} \bar{\Phi}.
+P(w) = \oint_C \frac{dz}{2\pi i}e^{zw}\bar{\Phi}(z).
 \end{equation}
-The low-$r$ regime above yields
+
+Computing this integral with either of the $\tilde{r}\Theta_c \ll 1$ expressions will yield
+
 \begin{equation}
-P(w) = \oint \frac{dz}{2\pi i} e^{zw}z\sqrt{\log z^2} \sim \frac{1}{w^2\sqrt{\log w^2}}.
+P(w) \sim \oint_C \frac{dz}{2\pi i}e^{zw}z \sqrt{-2 \log z} = \int^\infty_0 \frac{dy y e^{-yw}}{\sqrt{\log y^{-2}}} \sim \frac{w^{-2}}{\log w^2}.
 \end{equation}
+
+%The boundary condition is given by $\hat{\phi}(0,z,\chi) = 0$, since the ``weight" is zero at time zero.
+%We can solve this by sending $\tau$ to infinity, assuming that tunneling is rare and requires many mutant ``bubbles" to occur.
+%
+%We will need to figure out what happens to mutations in the ``bulk" of the wave versus the ``nose".
+%The above expression for $\Phi$ yields a differential equation
+%\begin{equation}
+%\partial_\tau \Phi(\tau) = \partial_\tau \int d\chi P(\chi,\tau)\psi(\tau,\chi) = \tilde{s} \Phi(\tau) - \epsilon \int d\chi P(\tau,\chi) \psi^2 (\tau,\chi).
+%\end{equation}
+%Note that, for beneficial mutations, the two right hand terms balance each other over long time scales, but for neutral and deleterious mutations, no such steady state exists.
+%For the ``communal" recombination model used in Neher and Shraiman (2011) and (2010), the recombination contribution in the ODE for $\psi$ simplifies to $\tilde{r} \Phi(\tau)$.
+%
+%We proceed in the limits of large positive and negative growth rate $\beta = \chi+\tau+\tilde{s}-\tilde{r}$.
+%For $\beta < 0$, only the first term in the equation for $\psi$ matters, as $psi < 1/\epsilon$: for $\beta > 0$, only the second one matters.
+%Accordingly,
+%\begin{equation}
+%\psi(\tau,\chi) =
+%\begin{cases}
+%\tilde{r} e^{\beta^2/2} \int^\tau_0 d\tau^\prime \Phi(\tau^\prime)e^{-\beta^2/2} & \beta < \beta_c (\tau) \\
+%\beta/\epsilon & \beta > \beta_c (\tau).
+%\end{cases}
+%\end{equation}
+%
+%(Still to-do: write up $\beta_c$ and explain how the crossover occurs.
+%Some of the following calculations will still need cleaning up.)
+%
+%Correspondingly,
+%\begin{equation}
+%\hat{\phi}(z,\chi) =
+%\begin{cases}
+%e^{\beta^2/2}(z+\tilde{r}\Phi(z)) & \beta < \beta_c (\tau) \\
+%\beta & \beta > \beta_c (\tau).
+%\end{cases}
+%\end{equation}
+%
+%The jump from this condition to the Laplace transform for the bubble size $\Phi(z) = \mathbb{E}(e^{-zw})$ needs to be filled in, but the transform in the low $\tilde{s}$ (neutral) or high $\tilde{r}$ limit is given by
+%\begin{equation}
+%\Phi(z) = z\frac{e^{\tilde{r}\beta_c - 1}}{\tilde{r}} = 
+%\begin{cases}
+%z\sqrt{-2\log z} & \tilde{r} \beta_c \ll 1 \\
+%z e^{\tilde{r}\beta_c}/\tilde{r} & \beta_c \gg 1.
+%\end{cases}
+%\end{equation}
+%It is worth noting that $r$ here is the total recombination rate between two arbitrary loci, not necessarily between the two loci of interest.
+%
+%The inverse Laplace transform is then given by
+%\begin{equation}
+%P(w) = \oint \frac{dz}{2\pi i} e^{zw} \bar{\Phi}.
+%\end{equation}
+%The low-$r$ regime above yields
+%\begin{equation}
+%P(w) = \oint \frac{dz}{2\pi i} e^{zw}z\sqrt{\log z^2} \sim \frac{1}{w^2\sqrt{\log w^2}}.
+%\end{equation}
 The weight distribution scales as $P(w)\sim 1/w^2 \sqrt{2\log w}$, compared to $P(w) \sim w^{-3/2}$ in the drift case.
 As expected, draft decreases the bubble size distribution but lessens the dependence on $\delta$.
-Some more precise analytics will be needed here.
-In particular, note that, looking at the high and low $\tilde{r}$ expressions for $\Phi$, it seems the weight will barely depend at all on $\sigma$, which seems... wrong.
+The unfortunate effect of this calculation is that we have explicitly \emph{not} used the assumption of no recombination.
+That might simplify the analysis substantially and is the next thing to check.
+
+(Following is an aborted attempt to study the behavior of an allele directly using the generating function and an explicitly time dependent advancing wave. It is included for completeness.)
+Let's proceed using the assumption of no recombination, but that the effect of mutations is small enough (or the number of loci large enough) for the fitness distribution to remain roughly Gaussian.
+We have, as before,
+
+\begin{equation}
+p(n,t) = dt(n-1)(1+s+x-\bar{x})p(n-1,t-dt) + dt(n+1)p(n+1,t-dt) + (1-ndt(2+s+x-\bar{x}))p(n,t-dt).
+\end{equation}
+
+Rearranging yields
+\begin{equation}
+p(n,t) - p(n,t-dt) = dt(n-1)(1+s+x-\bar{x})p(n-1,t-dt) + dt(n+1)p(n+1,t-dt) - ndt(2+s+x-\bar{x})p(n,t-dt).
+\end{equation}
+
+The limit $dt \to 0$ (and dividing both sides by $dt$) then yields the master equation
+
+\begin{equation}
+\partial_t p_t(n) = (n-1)(1+s+x-\bar{x})p_t(n-1) + (n+1)p_t(n+1) - n(2+s+x-\bar{x})p_t(n).
+\end{equation}
+
+Define $G(\lambda,t) = \sum_n \lambda^n p_t(n)$. Then
+
+\begin{equation}
+\partial_t G(\lambda,t) = (1+s+x-\bar{x})\sum_n (n-1) \lambda^n p_t(n-1) + \sum_n (n+1) \lambda^n p_t(n+1) - (2+s+x-\bar{x})\sum_n n \lambda^n p_t(n).
+
+\end{equation}
+
+The sum in the first right hand side term can be rewritten as $\sum_n (n-1) \lambda^n p_t(n-1) = \sum_n n\lambda^{n+1} p_t(n)$ by relabeling elements, which is just $\lambda^2 \sum_n n\lambda^{n-1} p_t(n)$, or $\lambda^2 \partial_\lambda G$.
+The second becomes $\sum_n (n+1) \lambda^n p_t(n+1) = \sum_n n \lambda^{n-1} p_n$, which is just $\partial_\lambda G$.
+The third becomes $\sum_n n \lambda^n p_t(n) = \lambda \sum_n n \lambda^{n-1} p_n$, which is just $\lambda \partial_\lambda G$.
+Thus
+
+\begin{equation}
+\partial_t G(\lambda,t) = \left( (1+s+x-\bar{x})\lambda^2 - (2+s+x-\bar{x})\lambda + 1)\partial_\lambda G(\lambda, t).
+\end{equation}
+
+If there were no $\bar{x}$ term, this could easily be solved via the method of characteristics, but unfortunately $\bar{x} = \bar{x}(t) = vt = \sigma^2 t$, which is time dependent.
+We will obviously have to restrict our analysis to the case $1 + s + x > \bar{x}$, or else the lineage will (on average) shrink and any long time solution will end in extinction.
+Define the surface
+
+\begin{equation}
+a(\lambda,t)G_\lambda + b(\lambda,t)G_t + c(\lambda,t)G = f(\lambda,t,G).
+\end{equation}
+
+The normal to this integral surface is given by
+\begin{equation}
+\vec{n} = \vec{\nabla} F = G_\lambda \vec{i} + G_t \vec{j} - \vec{k}.
+\end{equation}
+So the PDE is simply $(a(\lambda,t),b(\lambda,t),f(\lambda,t,G)) \cdot \vec{n} = 0$, which is another way of saying that the PDE's vector field $\vec{v} = a(\lambda,t)\vec{i} + b(\lambda,t)\vec{j} + f(\lambda,t,G)\vec{k}$ is normal to the integral surface at every point.
+This means $\vec{v}$ can be written as a curve along the integral surface $F$.
+We can parametrize it using the dummy variable $\tau$ and write
+\begin{equation}
+\vec{p}(\tau ) = \lambda(\tau) \vec{i} + t(\tau) \vec{j} + G(\tau ) \vec{k},
+\end{equation}
+or
+\begin{equation}
+\vec{p}^\prime = \frac{d\vec{p}}{d\tau} = \frac{d\lambda}{d\tau} \vec{i} + \frac{dt}{d\tau} \vec{j} + \frac{dG}{d\tau} \vec{k},
+\end{equation}
+which is normal to $\vec{n}$ and hence proportional to $\vec{v}$.
+If the proportionality constant is $\kappa$, we have $\frac{d\lambda}{d\tau} = \kappa a(\lambda,t)$, $\frac{dt}{d\tau} = \kappa b(\lambda,t)$, and $\frac{dG}{d\tau} = \kappa f(\lambda,t,G)$, which gives us simply
+\begin{equation}
+\frac{d\lambda}{a} = \frac{dt}{b} = \frac{dG}{f}.
+\end{equation}
+In this case, we have $a(\lambda, t) = (1+s+x-\bar{x})\lambda^2 - (2+s+x-\bar{x})\lambda + 1)$, $b(\lambda, t) = -1$, and $c(\lambda, t) = f(\lambda,t,G) = 0$.
+To avoid a divide by zero error, we can simply flip the above equation.
+We end up with
+\begin{equation}
+\frac{d\lambda}{dt} = -((1+s+x-\bar{x})\lambda^2 - (2+s+x-\bar{x})\lambda + 1) = P(s,x,\lambda, t), \frac{dG}{dt} = 0,
+\end{equation}
+with the so called ``Cauchy data" (initial/boundary conditions) given above.
+It remains to solve these ODEs.
+It is worth noting that $P$ can be factored into
+\begin{equation}
+P(s,x,\lambda, t) = ((1+s+x-\bar{x})\lambda - 1)(\lambda - 1).
+\end{equation}
+The first equation we need to solve becomes:
+\begin{equation}
+\int \frac{d\lambda}{P(s,x,\lambda, t)} = \int dt
+\end{equation}
+Decomposing $P$ and using partial fractions allows us to write
+\begin{equation}
+\int \frac{d\lambda}{P(s,x,\lambda, t)} = \int d\lambda\left(\frac{A}{\lambda - \frac{1}{\alpha}} + \frac{B}{\lambda - 1}),
+\end{equation}
+with $\alpha = s+x-\bar{x}$, $A = \frac{1}{\frac{1}{\alpha} - 1}$, $B = -A = \frac{-1}{\frac{1}{\alpha} - 1}$.
+Integrating the first term on the left yields $\frac{1}{\frac{1}{\alpha}-1}\log(\lambda - \frac{1}{\alpha})$, and the second term yields $-\frac{1}{\frac{1}{\alpha}-1}\log(\lambda - 1)$.
+We thus have $\frac{1}{\frac{1}{\alpha}-1}\log(\frac{\lambda-\frac{1}{\alpha}}{\lambda-1}) + C_1$.
+The right hand side becomes of course simply $t$.
+Next, the equation $\frac{dG}{dt} = 0$ yields simply that $G = C_2$, or that $G$ depends on $t$ only through its dependence on $\lambda$.
+At this point, if $\alpha$ were simply a constant (in the case of no fitness variance, there would be no need to consider $x$ at all, and $\alpha$ would equal $s$), we could simply proceed to choose a clever value for $C_1$ and $C_2$ and solve for $\lambda(t)$.
+The Cauchy data $G(\lambda,0) = \lambda$ and $G(1,t) = 1$ would then guide us in determining the form of $t$.
+However, $\alpha$ is not a constant: it is hiding a nasty time dependence (via $\bar{x} = \sigma^2 t$).
+There might be a way around this, but it is not obvious.
 

### 2017-09-01 writeup/writeup.tex
compile typos fixed hopefully

@@
 \begin{equation}
-\phi(z,\chi) = \frac{\int_0^\tau (z+\tilde{r}\Phi(\tau^\prime)) e^{-\theta^\prime^2/2} d\tau^\prime}{e^{-\theta^2/2}}.
+\phi(z,\chi) = \frac{\int_0^\tau (z+\tilde{r}\Phi(\tau^\prime)) e^{-{\theta'}^2/2} d\tau^\prime}{e^{-\theta^2/2}}.
 \end{equation}
@@
 The second term needs to be treated with slightly more care.
-The major contribution to the integral will come from a well defined, large $\tau^\prime$, corresponding to the maximum of $\exp(-\theta^\prime^2/2 + \log \Phi(\tau^\prime))$ at $\tau^\prime = \tau - \theta$.
+The major contribution to the integral will come from a well defined, large $\tau^\prime$, corresponding to the maximum of $\exp(-\theta'^2/2 + \log \Phi(\tau^\prime))$ at $\tau^\prime = \tau - \theta$.
 $\Phi(\tau)$ changes slowly and can be pulled out of the integral, yielding another Gaussian integral.
@@
 \partial_t G(\lambda,t) = (1+s+x-\bar{x})\sum_n (n-1) \lambda^n p_t(n-1) + \sum_n (n+1) \lambda^n p_t(n+1) - (2+s+x-\bar{x})\sum_n n \lambda^n p_t(n).
-
 \end{equation}
@@
 \begin{equation}
-\partial_t G(\lambda,t) = \left( (1+s+x-\bar{x})\lambda^2 - (2+s+x-\bar{x})\lambda + 1)\partial_\lambda G(\lambda, t).
+\partial_t G(\lambda,t) = \left( (1+s+x-\bar{x})\lambda^2 - (2+s+x-\bar{x})\lambda + 1 \right)\partial_\lambda G(\lambda, t).
 \end{equation}
@@
 \begin{equation}
-\int \frac{d\lambda}{P(s,x,\lambda, t)} = \int d\lambda\left(\frac{A}{\lambda - \frac{1}{\alpha}} + \frac{B}{\lambda - 1}),
+\int \frac{d\lambda}{P(s,x,\lambda, t)} = \int d\lambda\left(\frac{A}{\lambda - \frac{1}{\alpha}} + \frac{B}{\lambda - 1} \right),
 \end{equation}

### 2017-09-01 fwdpy11/bubble_draft.py

@@ -95,3 +95,4 @@ def plot_weight_dist(weights, kde = True, hist = False, bins = 20, loc = 3):
 ### ipython code: this produces a very good fit to the drift case of -alpha = -3/2
 # %time pop, weights = evolve_draft(20000, 10000, 0.01, 0.001, 0.001, seed=314)
 # plot_weight_powerlaw(weights)
+

### 2017-09-15 valley_crossing.tex
sketching some heuristics, filling in theory

--- 2016-11-21 ms/valley_crossing.tex
@@
 
+\section*{Model}
+
+Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
+We focus on two loci, which are initially fixed for alleles $a$ and $b$.
+These alleles mutate to $A$ and $B$ at rates $\mu_a$ and $\mu_b$, respectively (we neglect back mutations).
+Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
+For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$: we discuss the effects of relaxing this assumption in the appendix.
+We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate at which the mean fitness advances.
+Organisms cross over at rate $\rho$ per site per generation.
+Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.
+
+In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with Weissman et al.
+(2009): when $\rho > 0$, the rate agrees with Weissman et al.
+(2010).
+This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
+We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
+By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
+We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
+We sketch some heuristics in this limit before providing a more formal analysis.
+
+First, we consider the case where $\rho = 0$ and focus on what Weissman et al. (2009) refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
+As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
+In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
+If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
+We refer to $w$ as the ``weight" of the bubble.
+The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
+Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
+The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
+This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.
+
+In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
+In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
+Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
+This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
+However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
+This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
+(Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
+Can start with equation 3 of Neher and Shraiman (2009) and drop the recombination term.
+This is tantamount to a ``mean field" approximation but might still work out okay.
+Other possible approach is to use something like Desai and Fisher's (2007) expression for the fixation probability and rescale it to a traveling wave model.
+Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)
+
+We now turn to the case where $\rho > 0$.
+Because we are explicitly dealing with crossovers, the results of Neher, Shraiman, and Kessinger (2013) are useful here.
+We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
+The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
+The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
+If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
+If not, then the value of $N\sigma_b$ becomes the relevant factor.
+$N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
+On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
+
 \end{document}

### 2017-09-26 ms/valley_crossing.tex
bibliography, manuscript additions

@@
 The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
-This appears to be how Darwin thought most complex systems, such as the vertebrate eye, evolved (Darwin 1859), and this assumption underlies much of the field of adaptive dynamics.
+This appears to be how Darwin thought most complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies much of the field of adaptive dynamics.
 However, many adaptations exist that do not appear to have formed by this process.
 One class of such adaptations has been termed ``irreducibly complex" because, in their current forms, they cease to function when one or more of their parts is removed.
-Muller (1918) described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other, giving the appearance of irreducibility.
+\citet{muller_1918} described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other, giving the appearance of irreducibility.
 An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a lower fitness mutational ``valley" in order to reach a higher fitness ``peak".
-We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation (Trotter et al. 2013).
+We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation \citep{trotter_2014}.
 
-The rate of valley crossing in asexual populations is increasingly well understood (Weissman 2009).
+The rate of valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
 If the number of mutants produced every generation is high, the arisal of a complex adaptation that is destined to fix may be guaranteed.
@@
 These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation.
-These models were successfully extended to sexually reproducing populations by Weissman et al.
-(2010), which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
+These models were successfully extended to sexually reproducing populations by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
 Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
@@
 In sexual populations, much the same description applies.
-The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ (Neher and Shraiman 2011) or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within a small block length (Neher et al.
-2013).
+The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013} or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within a small block length \citep{neher_kessinger_2013}.
 
@@
 In fact, the possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
-It is reasonable to suspect that draft might likewise have profoundly different effects on the evolution of complex adaptations, a notion that was explored in Neher and Shraiman (2011).
+It is reasonable to suspect that draft might likewise have profoundly different effects on the evolution of complex adaptations, a notion that was explored in \citet{neher_shraiman_2011}.
 We investigate this possibility using analytical approaches and forward simulations, confirming that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
@@
 
+\section*{Model}
+
+Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
+We focus on two loci, which are initially fixed for alleles $a$ and $b$.
+These alleles mutate to $A$ and $B$ at rates $\mu_a$ and $\mu_b$, respectively (we neglect back mutations).
+Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
+For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$: we discuss the effects of relaxing this assumption in the appendix.
+We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate at which the mean fitness advances.
+Organisms cross over at rate $\rho$ per site per generation.
+Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.
+
+In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with \citet{weissman_2009}: when $\rho > 0$, the rate agrees with \citet{weissman_2010}.
+This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
+We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
+By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
+We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
+We sketch some heuristics in this limit before providing a more formal analysis.
+
+First, we consider the case where $\rho = 0$ and focus on what \citet{weissman_2009} refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
+As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
+In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
+If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
+We refer to $w$ as the ``weight" of the bubble.
+The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
+Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
+The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
+This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.
+
+In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
+In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
+Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
+This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
+However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
+This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
+(Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
+Can start with equation 3 of \citet{neher_shraiman_2011} and drop the recombination term, which is what I'm doing now.
+This is tantamount to a ``mean field" approximation but might still work out okay.
+Other possible approach is to use something like the expression of  \citet{desai_fisher_2007} for the fixation probability and rescale it to a traveling wave model.
+Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)
+
+One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+That is to say, the total rate of valley crossing will be given by $\int P(x) p_{\mathrm{fix}}(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
+One cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
+This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
+We proceed to derive both $p_{\mathrm{fix}}(x,s)$ and $P(w)$ in the zero recombination limit.
+
+The establishment probability $p(x,s)$ obeys the PDE
+\begin{equation}
+v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
+\end{equation}
+To see this, consider the generic expression for the fixation probability
+\begin{equation}
+p(X,t-dt) = p(X,t)-dt(D+B(X,t))p(X,t) + dt(B(X,t)(2p(X,t)-p(X,t)^2)
+\end{equation}
+\citep{barton_1995}. Taking the limit as $dt \to 0$ and inserting $D = 1, B = 1 + X -\bar{X}(t)+s$ yields a PDE for $p(X,t)$.
+Defining $x = X - vt$, we then have $\partial_t p(X-vt) = -v\partial_x p(x)$.
+
+This is a Bernoulli equation and is readily solved by substituting $u(x,s) = 1/p(x,s)$. We then have that $\partial_x u(x,s) = -\frac{\partial_x p(x,s)}{p(x,s)^2}$, so that
+\begin{equation}
+\partial_x u(x,s) = -\frac{(x+s)}{v}u(x,s) - \frac{1+x+s}{v}.
+\end{equation}
+The use of the integrating factor $e^{\int \frac{x+s}{v} dx} = Ce^{\frac{(x+s)^2}{2v}}$ yields
+\begin{equation}
+Ce^{\frac{(x+s)^2}{2v}} (\partial_x u + \frac{x+s}{v}u) = -Ce^{\frac{(x+s)^2}{2v}}\frac{1+x+s}{v}.
+\end{equation}
+The left side is simply $\partial_x(Ce^{\frac{(x+s)^2}{2v}}u)$, so (after dividing $C$ out) we can integrate both sides, then multiply by $e^{\frac{-(x+s)^2}{2v}}$ to obtain
+\begin{equation}
+u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}} - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
+\end{equation}
+with C a constant.
+(Need to think about how to eliminate the constant: what are the boundary conditions here?)
+
+Next, we seek the weight distribution as a function of $x$.
+Let $p(w,T|k,t,x)$ be the probability that a bubble has attained a weight of $w$ at time $T$, given that there were $k$ individuals in the lineage at time $t$.
+Ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
+Here $w$ represents (as before) the weight of a mutant bubble and evolves according to
+\begin{align*}
+-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
+& +kp(w,T|k-1,t,x). \\
+\end{align*}
+We have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$, by the assumption that the behavior of each of the $k$ copies of the mutant allele is independent of the other.
+Recall that, in the traveling wave approximation, $\bar{x} = vt = \sigma^2 t$.
+The notation here can be cleaned up by defining $\phi = 1 - \hat{p}$, as well as $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, $\tau = \sigma(T-t)$, and $\theta = \chi+\tau+\tilde{s}+z$.
+Rewriting the preceding in terms of $\phi$ (note that differentiating $p$ with respect to $w$ simply pulls out a factor of $z$) yields
+\begin{equation}
+\partial_\tau \phi(\tau,z,\chi) = z + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi).
+\end{equation}
+Unfortunately $\theta$ depends on $\tau$, so we will have to be careful here.
+This is a Riccati equation.
+Let $\phi_0(\tau,z,\chi)$ be a particular solution for $\phi$: then the general solution becomes
+\begin{equation}
+\phi(\tau,z,\chi) = \phi_0(\tau,z,\chi) + \frac{\Phi(\tau,z,\chi)}{C + \int \Phi(\tau,z,\chi) d\tau},
+\end{equation}
+where $\Phi(\tau,z,\chi) = e^{\int (-2\phi_0 + \theta ) d\tau}$.
+It remains to find a particular solution of $\phi$.
+We can ``guess" a steady state solution, in which case $0 = \phi^2 - \theta \phi - z = 0$: this corresponds to $\phi_0 = \frac{\theta \pm \sqrt{\theta^2 + 4z}}{2}$.
+Computing the exponent in the integral (and changing integration variables from $\tau$ to $\theta$) yields
+\begin{equation}
+\Phi (\tau, z, \chi) = e^{-\int \sqrt{\theta^2 + 4z} d\theta} = e^{\theta\sqrt{\theta^2+4z} + \frac{4z}{2}\log (\theta + \sqrt{\theta^2 + 4z})}.
+\end{equation}
+(Still need to solve this and carefully consider the integral boundary.)
+
+We now turn to the case where $\rho > 0$.
+Because we are explicitly dealing with crossovers, the results of \citet{neher_kessinger_2013} are useful here.
+We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
+The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
+The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
+If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
+If not, then the value of $N\sigma_b$ becomes the relevant factor.
+$N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
+On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
+
+\bibliographystyle{plainnat}
+\bibliography{bib}
+
 \end{document}

### 2017-10-05 ms/valley_crossing.tex
thoughts on how to compute fixation probability

@@
 \begin{equation}
-u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}} - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
+u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}}) - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
 \end{equation}
-with C a constant.
-(Need to think about how to eliminate the constant: what are the boundary conditions here?)
+with $C$ a constant. Of course, $p = 1/u$.
+
+Unfortunately it is not clear how to get rid of $C$, as there is no obvious ``initial condition".
+The most we are likely to know about $p$ is that $p$ approaches $0$ in the limit of low $x$, but behaves $\sim s$ in the limit of high $x$ before saturating at $1$ (however, there are unlikely to be any individuals in the population at such high $x$).
+One possibility is to impose an \emph{ersatz} ``solvability condition": surely the fixation probability of a neutral mutation averaged over the entire population must equal $1/N$. We would then have
+\begin{equation}
+\int_{-\infty}^{\infty} P(x)p(x,0)dx = \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{e^{-x^2/2\sigma^2}}{-e^{-x^2/2\sigma^2} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - 1 + Ce^{-x^2/2\sigma^2}}dx = \frac{1}{N}.
+\end{equation}
+Multiplying the top and bottom of the integrand by $e^{x^2/2\sigma^2}$ would clean this up, yielding
+\begin{equation}
+\frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{1}{-\frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - e^{x^2/2\sigma^2} + C}dx = \frac{1}{N}.
+\end{equation}
+Unfortunately this integral has an imaginary error function in the denominator, so there is not likely to be a closed form expression: we could explore it numerically and see where it peaks, however.
+
+Another idea would be to consider an alternative way of solving the differential equation for $p$.
+Return to
+\begin{equation}
+v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
+\end{equation}
+Suppose we make the \emph{ansatz} that $x+s \ll 1$, and define $\chi = x/\sigma$, $\tilde{s} = s/\sigma$, and $\tilde{p}(\chi,\tilde{s}) = p(x,s)/\sigma$. Then
+\begin{equation}
+\partial_\chi \tilde{p}(\chi,\tilde{s}) = (\chi + \tilde{s})\tilde{p}(\chi,\tilde{s}) - \tilde{p}(\chi,\tilde{s})^2.
+\end{equation}
+Now define $q(\chi,\tilde{s})$ such that $\tilde{w}(\chi,\tilde{s}) = \partial_x \log q(\chi,\tilde{s})$.
+Then
+\begin{equation}
+\partial_\chi^2 q(\chi,\tilde{s}) - (\chi+\tilde{s})\partial_\chi q(\chi, \tilde{s}) = 0.
+\end{equation}
+Defining $\vartheta = \chi + \tilde{s}$ and $q(\vartheta) = e^{\vartheta^2/4}h(\vartheta)$ yields
+\begin{equation}
+\partial_\vartheta^2 h(\vartheta) - (\frac{1}{2} + \frac{\theta^2}{4})h(\vartheta) = 0,
+\end{equation}
+which is the parabolic cylinder equation.
+We then need to find the form of $q$ that has the correct asymptotics.
+(This does have a solution.)
 

### 2017-11-10 writeup/weights.tex
brief writeup of weights equation

\documentclass[10pt]{revtex4}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{mathtools}

\begin{document}

We are interested in deriving the distribution of weights $P(w)$ \emph{as a function of the background fitness $x$}.
To do this, we begin with the master equation governing the evolution of the weight distribution in the case of recombination.
Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $k$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
We have
\begin{align*}
-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(w,T|k,t,x) \\
& + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
& +kp(w,T|k-1,t,x) \\
&+rk \int dw\prime \int_{x\prime} K_{x x\prime} p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x\prime).
\end{align*}

The first term is the probability flux out of state $(w,t)$.
The second corresponds to birth, in which case the final state $(w,t)$ is reached with probability $p(w,T|k+1,t,x)$.
The third corresponds to death, in which case the final state $(w,t)$ is reached with probability $p(w,T|k-1,t,x)$.
The last term corresponds to removing an individual with fitness $x$ and generating an individual with fitness $x\prime$, summed over all the ways the $k-1$ individuals with fitness $x$ and the one with fitness $x\prime$ can yield a bubble of $w$ individuals at time $T$.
The ``kernel" depends on the recombination model.
In Richard's analysis the ``free recombination" kernel is used, where offspring fitness is simply generated randomly by sampling from the entire population: $K_{xx\prime} = \frac{1}{\sqrt{2\pi \sigma^2}} \exp \left[ -\frac{(x\prime - x(t))^2}{2\sigma^2} \right]$.

Since we are interested in the weights, we need to consider not the generating function (which would make the analysis a little easier, as we could simply follow Ben's analysis almost line for line) but rather the Laplace transform $\hat{p}(z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
We can simplify the resulting equation using $\hat{p}(z,T|k,t,x) = \hat{p}^k(z,T|1,t,x)$ by the \emph{ansatz} that all the lineages are separate, and also by using $\partial_t \hat{p}^k = k\hat{p}^{k-1} \partial_t \hat{p}$.
This yields

\begin{align*}
-k\hat{p}^{k-1} \partial_t\hat{p} + kz\hat{p}^k = &-k(2+x-\bar{x}+s+r)\hat{p}^k \\
& + k(1+x-\bar{x}+s)\hat{p}^{k+1} \\
& +k\hat{p}^{k-1} \\
&+rk \times \mathrm{recombination~term}.
\end{align*}

Collecting terms and dividing out by $k$ and $\hat{p}^{k-1}$ yields

\begin{align*}
-\partial_t\hat{p} = &-z\hat{p}\\
&-(2+x-\bar{x}+s+r)\hat{p} \\
&+ (1+x-\bar{x}+s)\hat{p}^2 \\
& + 1\\
&+r \times \mathrm{recombination~term}.
\end{align*}

Removing the $1$ is done by replacing $\hat{p}$ with $1-\phi$ (or equivalently $\phi = 1- \hat{p}$, yielding (after some algebra):
\begin{align*}\partial_t \phi = z - (x-\bar{x} + s - r+z)\phi + (1+x-\bar{x} + s)\phi^2 - r \times \mathrm{recombination~term},
\end{align*}
where the recombination term has become $\int dx K_{xx\prime} \phi$.
With the ``free recombination" kernel, that term simply because $\Phi$, the scaled fixation probability after averaging over the entire population.
Switching variables by defining $\tilde{r} = r/\sigma$, $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, $\tau = \sigma(T-t)$, and $\theta = \chi+\tau+\tilde{s}+z-\tilde{r}$, and making the approximation that the prefactor of $\phi^2$ is about $1$, yields
\begin{equation}
\partial_\tau \phi(\tau,z,\chi) = z + \tilde{r} \Phi(\tau, z) + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi),
\end{equation}
where $\Phi(\tau, z) = \int d\chi P(\chi, \tau) \phi(\tau, z, \chi)$ is, as before, the rescaled version of $\phi$ when integrating over the entire fitness distribution $P(\chi, \tau)$.

We can send $\tau$ to infinity and look for steady state solutions, since all we are interested in is the size of a single, short lived bubble (i.e., there is only ever one bubble at a time).
For large values of $\theta$, the dominant terms are $\theta \phi(\tau,z,\chi)$ and $-\phi^2(\tau,z,\chi)$.
Setting their sum equal to zero (steady state solution) and solving for $\phi$ yields $\phi(z,\chi) \approx \theta$.

For small values of $\theta$, the remaining terms are more important.
We have $\partial_\tau \phi = z + \tilde{r} \Phi(z) + \theta \phi(z, \chi)$.
The relevant ``integrating factor" here is $\mu(\theta) = e^{\int -\theta d\theta} = e^{-\theta^2/2}$ (the use of $\theta$ rather than $\tau$ is justified by the fact that they depend linearly on each other).
We therefore obtain, in the low $\theta$ limit,

\begin{equation}
\phi(z,\chi) = \frac{\int_0^\tau (z+\tilde{r}\Phi(\tau\prime)) e^{-\theta\prime^2/2} d\tau\prime}{e^{-\theta^2/2}}.
\end{equation}

The $z$ term in the integrand simply becomes $z\sqrt{2\pi} \sim z$.
The second term needs to be treated with slightly more care.
The major contribution to the integral will come from a well defined, large $\tau^\prime$, corresponding to the maximum of $\exp(-\theta^\prime^2/2 + \log \Phi(\tau^\prime))$ at $\tau^\prime = \tau - \theta$.
$\Phi(\tau)$ changes slowly and can be pulled out of the integral, yielding another Gaussian integral.
We therefore have $\phi(z,\chi) \approx (z+\tilde{r}\Phi(z))e^{\theta^2}/2$, or, summarizing:

\begin{equation}
\phi(z,\chi) \approx
\begin{cases}
(z+\tilde{r}\Phi(z))e^{\theta^2}/2 & \theta \ll \Theta_c, \\
\theta & \theta \gg \Theta_c.
\end{cases}
\end{equation}

The crossover between these two solutions occurs roughly at some critical $\Theta_c$ given by setting them equal to each other:

\begin{equation}
(z+\tilde{r}\Phi(z))e^{\Theta_c^2/2} = \Theta_c.
\end{equation}

For what happens from here, consult the writeup.

This is where the fly in the ointment arises. Much of that analysis depends on a nonzero $\tilde{r}$.
If we send it to zero, we instead have

\begin{equation}
\phi(z,\chi) \approx
\begin{cases}
ze^{\theta^2}/2 & \theta \ll \Theta_c, \\
\theta & \theta \gg \Theta_c.
\end{cases}
\end{equation}

This leaves us with no way to solve for $\Phi$, but anyway $\Phi$ is not what we need: we would like to retain the dependence on $\chi$ explicitly, to get the background dependent behavior of $\phi$.
Trying to get $\Theta_c$ by itself yields $\Theta_c^2 = \log 2\Theta_c - \log z$, which there might well be some recursive way to approximate.
Reducing it to $e^{\Theta_c}/\Theta_c = 2/z$ might help: some variant of the Lambert $W$ function could be used (via $-\Theta_c e^{-\Theta_c} = -z/2$, or $\Theta_c = -W(-z/2)$: I'd have to check and see if this behaves well).
The bubble size distribution would then be recovered via $P(w) = \oint_C \frac{dz}{2\pi i} e^{zw} \phi(z,\chi)$, where we would have to choose a suitable curve.
This might yield a nonsense answer, though, in the event that we have a singular perturbation.
(It is however not obvious that the perturbation needs to be singular.
In the fixation probability case, the singular perturbation arises from the fact that extinction is assured unless a mutation can jump onto higher fitness backgrounds.
In the bubble size distribution case, we assume that extinction is assured anyway.)

We can also try returning to the branching process equation and replacing the recombination term with a mutation distribution:

\begin{align*}
-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(B+D+U)p(w,T|k,t,x) \\
& + kBp(w,T|k+1,t,x) \\
& +kDp(w,T|k-1,t,x) \\
&+kU \int dw\prime \int ds\prime \rho(s\prime) p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x+s\prime).
\end{align*}

We have left $B$ and $D$ as the generic birth and death rates, with $U$ the mutation rate.
The integral is now a sum over all the ways an individual with fitness $x$ can mutate into one with fitness $x+s$, removing an individual with fitness $x$ and adding one with fitness $x+s$.
Accounting for the mutation term will be somewhat difficult, but we can simplify immediately by using $\rho(s) = \delta(s)$. Then

\begin{align*}
-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(B+D+U)p(w,T|k,t,x) \\
& + kBp(w,T|k+1,t,x) \\
& +kDp(w,T|k-1,t,x) \\
&+kU \int dw^\prime p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x+s),
\end{align*}

i.e., the mutation distribution ``collapses" onto a single value of $s$.
Now we need to be careful, because we are about to take the Laplace transform of the whole thing. As before, we have

\begin{align*}
-k\hat{p}^{k-1} \partial_t\hat{p} + kz\hat{p}^k = &-k(B+D+U) \hat{p}^k \\
& + kB\hat{p}^{k+1} \\
& +kD\hat{p}^{k-1} \\
&+Uk \int dw\prime \int dz p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x+s) e^{-zw}.
\end{align*}

In the last term, we could interchange the order of integration to get $\int dz e^{-zw} \int dw^\prime p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x+s)$, but then it is not totally obvious how to get rid of the convolution over $w\prime$. I still need to think about this.

\end{document}

### 2017-12-03 writeup/weights.tex
updated weights equation writeup

@@
 The second term needs to be treated with slightly more care.
-The major contribution to the integral will come from a well defined, large $\tau^\prime$, corresponding to the maximum of $\exp(-\theta^\prime^2/2 + \log \Phi(\tau^\prime))$ at $\tau^\prime = \tau - \theta$.
+The major contribution to the integral will come from a well defined, large $\tau\prime$, corresponding to the maximum of $\exp(-\theta\prime^2/2 + \log \Phi(\tau^\prime))$ at $\tau\prime = \tau - \theta$.
 $\Phi(\tau)$ changes slowly and can be pulled out of the integral, yielding another Gaussian integral.
@@
 
+I have thought about this.
+Let's go back to that original differential equation arising from the master equation equation:
+\begin{align*}
+-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s+r)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
+& +kp(w,T|k-1,t,x) \\
+&+rk \int dw\prime \int_{x\prime} K_{x x\prime} p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x\prime).
+\end{align*}
+The last term is the tricky bit, but it is important to remember that $K_{xx\prime}$ ``sends" $x$ to $x\prime$.
+The two terms in the inner integral refer to having one lineage that has reached weight $w - w\prime$ at time $T$, given that at time $t$ there were $k-1$ individuals, and a lineage with weight $w\prime$ at time $T$, given that at time $t$ a recombination event sent an individual in the original lineage  (carrying whatever focal mutation) into a new lineage of size $1$.
+We can now briefly reconsider what happens when we apply the Laplace transform (applying the integral operator $\int_0^\infty dw e^{-zw}$ to both sides--note that Neher and Shraiman (2011) actually has a typo here, as it uses $dz$ instead) to both sides.
+
+We have that $\int_0^\infty dw e^{-zw} \partial_t p(w,T|k,t,x) = \partial_t \int_0^\infty dw e^{-zw} p(w,T|k,t,x) = \partial_t \hat{p}(z,T|k,t,x) = \partial_t \hat{p}(z,T|1,t,x)^k = k\hat{p}(z,T|1,t,x)^k \partial_t p(z,T|1,t,x)$.
+Unfortunately we cannot simply exchange the order of integration in the second term, but we can use the Laplace transform identity $\mathcal{L}(df(t)/dt) = sF(s) - f(0)$, with $F(s)$ the transform of $f(t)$: this becomes $z\hat{p}(z,T|k,t,x) - \delta_{t,T}$, under the argument that the weight is zero at the initial timepoint but nonzero thereafter, provided the lineage doesn't immediately go extinct.
+So our preceding analysis holds.
+The last term remains tricky. It is not obvious what $\int dw e^{-zw} \int dw\prime \int_{x\prime} K_{x x\prime} p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x\prime)$.
+But it is instructive to note that the transform can basically be split into two parts, a contribution corresponding to $w\prime$ and another corresponding to $w - w\prime$, via $e^{-zw} = e^{-z(w-w\prime)}e^{-zw\prime}$.
+This causes the first term inside the integral to become $\hat{p}(z,T|k-1,t,x)$ and the second to become $\hat{p}(z,T|1,t,x\prime)$, consistent with our earlier argument.
+The fact that $K_{xx\prime}$ in the free recombination case is just a Gaussian centered around $\bar{x}$ is what causes that term to become $\Phi$.
+
+Now we need to think about what changes in the mutation case.
+Essentially, just as $K_{xx\prime}$ sends $x$ to $x\prime$, so too the distribution of mutational effects $\rho(s)$ sends $x$ to $x+s$.
+So the differential equation governing the evolution of $p$ becomes
+
+\begin{align*}
+-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s+U)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x})p(w,T|k+1,t,x) \\
+& +kp(w,T|k-1,t,x) \\
+&+Uk \int_0^w dw\prime \int_0^\infty ds \rho(s) p(w-w\prime , T|k-1, t, x) p(w\prime,T|1,t,x+s).
+\end{align*}
+
+This is starting to look like the differential equation of Good et al. (2012) for the fixation probability, though it took some thought to get there. By the same argument, the Laplace transform yields
+
+\begin{align*}
+-k\hat{p}(z,T|1,t,x)^{k-1} \partial_t\hat{p}(z,T|1,t,x) + kz\hat{p}(z,T|1,t,x)^k = &-k(2+x-\bar{x}+U)\hat{p}(z,T|1,t,x)^k \\
+& + k(1+x-\bar{x})\hat{p}(z,T|1,t,x)^{k+1} \\
+& +k\hat{p}(1,T|k,t,x)^{k-1} \\
+& +Uk \int_0^\infty ds \rho(s) \hat{p}(z,T|1, t, x)^{k-1} p(z,T|1,t,x+s).
+\end{align*}
+
+Note that the first term in the integral can be pulled out, as it doesn't depend on $s$.
+Collecting terms and dividing out by $k$ and $\hat{p}^{k-1}$ yields
+
+\begin{align*}
+-\partial_t\hat{p}(z,T|t,x) = &-z\hat{p}(z,T|t,x)\\
+&-(2+x-\bar{x}+U)\hat{p}(z,T|t,x) \\
+&+ (1+x-\bar{x})\hat{p}(z,T|t,x)^2 \\
+& + 1\\
+& +U \int_0^\infty ds \rho(s) \hat{p}(z,T|t, x+s).
+\end{align*}
+
+Swapping in $phi(z,T|t,x) = 1-\hat{p}(z,T|t,x)$ yields
+
+\begin{align*}\partial_t \phi(z,T|t,x) = z - (x-\bar{x} + s - r+z)\phi(z,T|t,x) + (1+x-\bar{x} + s)\phi(z,T|t,x)^2 - U\int_0^\infty ds \rho(s) \hat{p}(z,T|t,x_s),
+\end{align*}
+
+Switching variables as before and assuming $x-\bar{x} \ll 1$ yields
+
+\begin{equation}
+\partial_\tau \phi(\tau,z,\chi) = z + \tilde{U} \int_0^\infty d\tilde{s} \rho(\tilde{s}) \phi(\tau, z, \chi+\tilde{s}) + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi).
+\end{equation}
+
+This is ultimately what we will need to solve.
+Using some tricks similar to Good et al.
+(2012) to get rid of the integral will hopefully be fruitful, though I still think the $U \to 0$ limit should be tractable (it just might be that asymptotic matching does not work).
+
 \end{document}

### 2017-12-04 src/plt_asex_crossings.py
script for plotting asexual valley crossing times

@@ -2,11 +2,12 @@
 '''
 plt_asex_crossings.py
 Author: Taylor Kessinger
-Date: June 16 20, 2017
+Date: June 16, 2017
 Description: Glob and plot valley crossing times as a function of input variables.
 '''
 import sys
 sys.path.append('/home/tkessinger/tools/FFPopSim/pkg/python/')
+sys.path.append('/home/tkessinger/Documents/draft_valley_crossing/FFPopSim/pkg/python/')
 import numpy as np
 import FFPopSim as h
 import random as rand
@@ -21,13 +22,22 @@ import seaborn as sns
 sns.set_context('paper', font_scale=1.5)
 sns.set_style('white')
 
-name = 'asex_crossings_1'
+#recently used plots, possibly paper worthy:
+#in all cases N*mu ~= 0.3
+#var_sigma_6: variable sigma, N = 10**7
+#var_sigma_7: variable sigma, N = 10**6
+#var_sigma_8: variable sigma, N = 10**5
+
+
+name = 'var_sigma_6'
 
 numruns = {}
 sigma_range = {}
 N_range = {}
 s_range = {}
 delta_range = {}
+mu_range = {}
+rho_range = {}
 
 run_time = {}
 
@@ -40,42 +50,579 @@ if name == 'asex_crossings_1':
     run_time[name] = 1000000
     numruns[name] = 5
     rho = 0.0
+    graphtype = 'sigma_delta'
     
+if name == 'asex_crossings_2':
+
+    N_range[name] = [100000]
+    sigma_range[name] = [1e-8,1e-6,1e-4,1e-2]
+    delta_range[name] = [0.0, 1e-3]
+    s_range[name] = [0.01]
+    run_time[name] = 1000000
+    numruns[name] = 5
+    rho_range[name] = [0]
+    fixed_s = False
+    graphtype = 'sigma_delta'
+
+if name == 'asex_crossings_2_fixed_s':
+
+    N_range[name] = [100000]
+    sigma_range[name] = [5e-8,5e-6,5e-4,5e-2]
+    delta_range[name] = [0.0, 1e-3]
+    s_range[name] = [0.01]
+    run_time[name] = 1000000
+    numruns[name] = 5
+    rho_range[name] = [0]
+    fixed_s = True
+    graphtype = 'sigma_delta'
+
+if name == 'asex_weissman_2009_1':
[…]

### 2018-03-02 ms/valley_crossing.tex
updated ms and bibliography

@@
 \usepackage{graphicx}
-%\usepackage{caption}
-%\usepackage{subcaption}
+\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup{compatibility=false}
+
 \usepackage{amsmath}
@@
 \usepackage{enumerate}
+%\usepackage{subfig}
 
@@
 \title{Genetic draft and the evolution of ``irreducible complexity"}
-%\author{Taylor~Kessinger}
-%\author{Jeremy~Van~Cleve}
+\author{Taylor~Kessinger}
+\author{Jeremy~Van~Cleve}
 %\affiliation{Department of Biology, University of Kentucky}
@@
 However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate of individual mutations.
-We investigate the rate at which ``irreducibly complexity" evolves in these rapidly adapting populations, including both asexual and sexual populations.
-We confirm that, as in neutrally evolving populations, recombination can aid in the crossing of deep fitness valleys but find that its effects are mitigated.
+We investigate the rate at which irreducible complexity evolves in these rapidly adapting populations, focusing on asexual populations.
+We confirm that rapid adaptation overall increases the time required to cross a valley but can make it easier for deeper valleys to be crossed.
 \end{abstract}
@@
 
-The rate of valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
+Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
 If the number of mutants produced every generation is high, the arisal of a complex adaptation that is destined to fix may be guaranteed.
-Otherwise, wild type individuals may give rise to neutral or deleterious intermediate mutants, which drift to fixation.
+In small populations, wild type individuals may give rise to neutral or deleterious intermediate mutants, which drift to fixation.
 A third possibility is for wild type individuals to birth mutant ``bubbles", or transient subpopulations.
-These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation.
-These models were successfully extended to sexually reproducing populations by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
+These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
+These phenomena were successfully extended to sexually reproducing populations by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
 Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
@@
 
-Previous studies of valley crossing focused exclusively on populations where the background fitness variation is small enough to be negligible.
+Previous studies of valley crossing focused primarily on populations where the background fitness variation is small enough to be negligible.
 In these populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
@@
 In sexual populations, much the same description applies.
-The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013} or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within a small block length \citep{neher_kessinger_2013}.
+The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013} or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within an effectively asexual block length \citep{neher_kessinger_2013}.
 
 Drift and draft are fundamentally different forces.
-It is not possible to rescale a drafting population, for example by defining a reduced ``effective population size".
+It is not possible to understand a drafting population in terms of a drifting one, for example by defining a reduced ``effective population size".
 Drafting populations do not admit a diffusion approximation, and they exhibit qualitatively different genealogies and site frequency spectra relative to a drifting population.
 In fact, the possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
-It is reasonable to suspect that draft might likewise have profoundly different effects on the evolution of complex adaptations, a notion that was explored in \citet{neher_shraiman_2011}.
-We investigate this possibility using analytical approaches and forward simulations, confirming that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
+Draft likewise has profoundly different effects on the evolution of complex adaptations.
+\citet{neher_shraiman_2011} explored the dynamics of stochastic tunneling in rapidly adapting sexual populations.
+The primary mode of recombination considered therein is ``free" recombination, in which parents essentially shuffle the entirety of their genomes during reproduction.
+We extend this to asexual populations and comment on alternate modes of recombination, such as crossovers.
+Using analytical approaches and forward simulations, we confirm that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
 These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate at which the mean fitness advances.
-Organisms cross over at rate $\rho$ per site per generation.
+Let $x$ be the background fitness of a lineage.
+The distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
 Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.
 
-In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with \citet{weissman_2009}: when $\rho > 0$, the rate agrees with \citet{weissman_2010}.
-This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
-We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
-By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
-We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
-We sketch some heuristics in this limit before providing a more formal analysis.
+In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a selective sweep to occur.
+At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
+We begin by zeroing in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
+Once this is understood, extending our results to incorporate sequential fixation or the arisal of a complex adaptation in a single generation is straightforward.
 
+There are essentially three components to this process.
+First, a single mutant lineage must appear, which occurs with rate $N\mu$.
+During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm fix}$.
+Finally, the complex adaptation must sweep to fixation.
+The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
+In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble and $T_1$ is the sweep time.
+We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.
+
+We refer to single mutant lineages as ``bubbles".
+Let $n(t)$ be the total number of single mutant individuals in the bubble, and suppose it appears at time $t_0$ and goes extinct at time $T$.
+Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, representing the total number of chances there are for a successful double mutant to appear.
+Most of the time, a bubble will simply arise and go extinct.
+However, with probability $1-e^{-\mu p_{\mathrm fix} W}$, the bubble will give rise to a successful double mutant.
+Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm fix} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
+
+One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+That is to say, one cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
+Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
+As a result, $W$ will depend on $x$ and $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and $s$.
+These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
+The expected valley crossing time then becomes $1/N\mu\Phi$.
+
+We begin by considering the dynamics of $p_{\mathrm fix}$ before moving on to consider the distribution of bubble sizes.
+Let $p = p_{\mathrm fix}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
+In the absence of any perturbations, the evolution of $p$ can be described by a birth-death process, yielding
+\begin{equation}
+-\partial_t p(x,s,t) = Bp(x,s,t) - (B+D)p(x,s,t)^2
+\end{equation}
+\citep{barton_1995, good_2012}, with $B$ and $D$ the birth and death rates respectively.
+The only dependence of $p$ on $t$ is via the time-changing mean fitness $vt$.
+Defining $X = x - vt$, we then have $\partial_t p(x-vt) = -v\partial_X p(X)$.
+This can be rewritten as a wave equation
+\begin{equation}
+v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2.
+\end{equation}
+This is a Bernoulli equation and is readily solved, but the only appropriate boundary condition yields $p = 0$ everywhere.
+This is intuitive: the inexorable advance of the mean fitness implies that all lineages are ultimately doomed to extinction \emph{unless} they can increase their own fitness either via recombination or mutation.
+We thus need to amend the model to include a mutation term.
+Let $U$ be the mutation rate of the genetic background and $\epsilon$ be the fitness effect of a mutation, distributed as $\rho(\epsilon)$.
+Then we have
+\begin{equation}
+v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon,
+\end{equation}
+where $\epsilon$ is now the fitness effect of a mutation in the background distribution, $\rho(\epsilon)$ is the distribution of mutational effects, and $U$ is the mutation rate of the background.
+Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
+\begin{equation}
+p(X,s) =
+\begin{cases}
+0 &\mathrm{~if~} X+s < 0, \\
+X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
+X+s &\mathrm{~if~} X+s > X_c.
+\end{cases}
+\end{equation}
+(If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
+Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
+The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
+Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
+Then $X_c$ is determined by solving
+\begin{equation}
+2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}.
+\end{equation}
+
+We now consider the behavior of single mutant bubbles.
+Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm fix}$.
+We proceed to derive an equation for the time evolution of the Laplace transform.
+
+Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
+We have
+\begin{align*}
+-(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
+& + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
+& +kp(w,T|k-1,t,x).
+\end{align*}
+Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
+Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
+We therefore have
+\begin{equation}
+-\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
+\end{equation}
+Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
+\begin{equation}
+\partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
+\end{equation}
+where $\tilde{\phi}$ = $\phi/\sigma$.
+We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
+In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
+However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
+To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
+Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
+\begin{equation}
+\partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
+\end{equation}
+Using matched asymptotics allows us to write
+\begin{equation}
+\tilde{\phi} = 
+\begin{cases}
+\tilde{z} &\mathrm{~if~} \theta < 0, \\
+\theta &\mathrm{~if~} \theta > 0.
+\end{cases}
+\end{equation}
+Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
+Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
+
+%\begin{figure}
+\begin{figure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\end{subfigure}
+\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes.
+In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$.
+Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges.
+Each box represents about $100$ simulation runs.}
+\label{fig:deterministic}
+\end{figure}
+
+\section*{Methods}
+
+To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package.
+The simulation code is available from the authors upon request or via \texttt{GitHub}.
+We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
+For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model, where any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
+Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation.
+In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
+
+We set the ``focal" loci (the loci at which the epistatic alleles segregate) to be at positions $L/4$ and $3L/4$.
+Simulations are allowed to proceed for an ``equilibration" time of $N/10$ generations to allow sufficient genetic diversity to be introduced.
+During this time, the mutation rate $\mu$ for the focal loci is set to zero.
+After equilibration, we set a per-site mutation rate $\mu$ for the focal loci, with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$, consistent with our analytical model.
+When the double mutant reaches frequency $0.5$, we consider it to have fixed, then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background, and allow the population to equilibrate for another $N/10$ generations.
+This offers some assurance that independent valley crossing trials are indeed independent despite occurring in the same simulation run.
+
+\section*{Results}
+
+\begin{figure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/seq_fix.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/neut_tunnel.pdf}
+\end{subfigure}
+\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes.
+In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^4$ and $\mu = 3\times 10^{-5}$, and on the right, $N = 10^5$ and $\mu = 3\times 10^{-6}$.
+Labeling is as in figure \ref{fig:deterministic}.}
+\label{fig:tunneling}
+\end{figure}
+
+Here we present the key results of our simulation runs.
+We first studied the effect of increasing $\sigma$ in what \citet{weissman_2009} describes as the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes.
+Figures \ref{fig:deterministic} and \ref{fig:tunneling} show the expected result: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
+It is tempting to interpret this as a reduced $N_e$ due to a decrease in genetic diversity and the strength of selection.
+The underlying dynamics are more complicated, however.
+As $\sigma$ increases, the effect of the genetic background becomes more significant in determining the fate of an allele than the allele's fitness effect is.
+The frequency of alleles is buffeted by genetic draft, so alleles behave as though they are more neutral than they really are.
+This depresses the fixation probability of the complex adaptation, which appears to account for the majority of the change in fixation time.
+In the case of stochastic tunneling, the overall size distribution of bubble sizes is reduced, as well: the distribution $P(w)$ scales not as $w^{-3/2}$ but as $w^{-2}$ \citep{neher_shraiman_2011}, so there are fewer chances for the complex adaptation to arise and fix.
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{Figures/var_sigma_delta.pdf}
+\caption{Valley crossing times for varied $\sigma$ and $\delta$.
+In these simulations, $N = 10^5$, $\mu = 3 \times 10^{-6}$, and $s = 10^{-2}$.
+Note the lack of dependence on $\delta$ at high $\sigma$.}
+\label{fig:sigma_delta}
+\end{figure}
+
+On the other hand, the reduced effectiveness of selection should generally be helpful in crossing deeper valleys, by mitigating the deleterious effect of single mutants.
+This can be seen in figure \ref{fig:sigma_delta}.
+At high values of $\sigma$, the fitness effect of the deleterious intermediate \emph{almost does not matter}, meaning that very deep valleys can be crossed.
+This supports the view that traversing rugged fitness landscapes by valley crossing may be the rule, not the exception, in rapidly adapting populations.
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{Figures/bands.pdf}
+\caption{Valley crossing times for fixed $N\sigma$, with $N$ allowed to vary up or down by a factor of $2$ (and $\sigma$ adjusted accordingly).
+Error bars are quartile ranges.
+Here, $\mu = 3 \times 10^{-6}$, $\delta = 10^{-3}$, and $s = 10^{-2}$.}
+\label{fig:bands}
+\end{figure}
+
+Finally, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
+Figure \ref{fig:bands} suggests that, at high values of $\sigma$, the value of $N\sigma$ is more important than $N$ itself in determining the valley crossing rate.
+The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
+
+%In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with \citet{weissman_2009}: when $\rho > 0$, the rate agrees with \citet{weissman_2010}.
+%This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
+%We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
+%By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
+%We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
+%We sketch some heuristics in this limit before providing a more formal analysis.
+
+\iffalse
 First, we consider the case where $\rho = 0$ and focus on what \citet{weissman_2009} refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
@@
 We then need to find the form of $q$ that has the correct asymptotics.
-(This does have a solution.)
+This does have a solution:
+\begin{equation}
+q(\chi) = \int_0^\infty d\lambda e^{\vartheta \lambda - \lambda^2/2}/\lambda.
+\end{equation}
 
@@
 On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
+\fi
 

### 2018-03-02 src/simulate_population.py
numerical solution of bubble size ODE

@@ -37,7 +37,7 @@ parser.add_argument('--fixed_s', type=bool, default=False)
 
 args=parser.parse_args()
 
-L = 20 #number of loci. 1000 is about as many as we're likely to need.
+L = 200 #number of loci. 1000 is about as many as we're likely to need.
 N = args.N #population size. recall that Nsigma is the main parameter.
 mu = args.mu #per site mutation rate for focal mutations
 sigma = args.sigma #fitness variance
@@ -125,6 +125,8 @@ else:
 if not os.path.exists('output/valley_crossing_sims_'+args.name):
     os.makedirs('output/valley_crossing_sims_'+args.name)
 
+total_number_of_mutations = 0
+
 
 while pop.generation < run_time:
     pfit = pop.get_trait_statistics(0)
@@ -143,6 +145,7 @@ while pop.generation < run_time:
         pop.mutation_rate = mu #set the focal loci to start mutating
         burned_in = True
     if burned_in == True:
+        total_number_of_mutations += pop.number_of_mutations[-1]
         for di, del_locus in enumerate(mutant_locs):
             if pop.get_allele_frequency(del_locus) != 0 and mutations_in_play[di] == False:
             #if either mutant has appeared
@@ -198,6 +201,7 @@ while pop.generation < run_time:
     if not pop.generation%(10**2) and pop.generation > base_burn_time and args.fixed_s:
         sigma_hist.append([np.sqrt(pop.get_trait_statistics(0).variance), np.sqrt(pop.get_fitness_statistics().variance)])
     if not pop.generation%(10**4):
+        print pop.generation
         with open('output/valley_crossing_sims_'+args.name+'/'+prefix+'_dm_weights_'+str(args.runno)+'.pkl', 'w') as dm_weights_file:
             pickle.dump(dm_weights, dm_weights_file)
         with open('output/valley_crossing_sims_'+args.name+'/'+prefix+'_weights_'+str(args.runno)+'.pkl', 'w') as weights_file:
@@ -212,6 +216,8 @@ while pop.generation < run_time:
             pickle.dump(bubble_closings, closings_file)        
         with open('output/valley_crossing_sims_'+args.name+'/'+prefix+'_dm_successes_'+str(args.runno)+'.pkl', 'w') as dm_successes_file:
[…]

### 2018-03-04 fwdpy11/.rendered.bubble_recorder.cpp
updates to mathematica file and python weight dist

@@ -1,19 +1,5 @@
 // clang-format off
-<%
-#import fwdpy11 so we can find its C++ headers
-import fwdpy11 as fp11
-#add fwdpy11 header locations to the include path
-cfg['include_dirs'] = [ fp11.get_includes(), fp11.get_fwdpp_includes() ]
-#On OS X using clang, there is more work to do.  Using gcc on OS X
-#gets rid of these requirements. The specifics sadly depend on how
-#you initially built fwdpy11, and what is below assumes you used
-#the provided setup.py + OS X + clang:
-#cfg['compiler_args'].extend(['-stdlib=libc++','-mmacosx-version-min=10.7'])
-#cfg['linker_args']=['-stdlib=libc++','-mmacosx-version-min=10.7']
-setup_pybind11(cfg)
-#An alternative to the above is to add the first line to CPPFLAGS
-#and the second to LDFLAGS when compiling a plugin on OS X using clang.
-%>
+
 // clang-format on
 
 #include <pybind11/pybind11.h>

### 2018-03-04 fwdpy11/bubble_draft.py

@@ -4,25 +4,25 @@ import fwdpy11.model_params
 import fwdpy11.fitness
 import fwdpy11.wright_fisher
 
+# plotting stuff
+import matplotlib.pyplot as plt
+import powerlaw as pl
+
 # pybind11 C++ recorder
 import cppimport
 cppimport.set_quiet(False)
 bubble_recorder = cppimport.imp("bubble_recorder")
 
-# plotting stuff
-import matplotlib.pyplot as plt
-import seaborn as sns
-import powerlaw as pl
 
 class BubbleRecorder(object):
     def __init__(self):
         self.weights = dict()
-        
-    def __call__(self,pop):
+
+    def __call__(self, pop):
         for mut in range(len(pop.mutations)):
             # infinite sites model so each mutation has unique position
             mkey = pop.mutations[mut].pos
-            # if mutation neutral (neutral ones draft) 
+            # if mutation neutral (neutral ones draft)
             if pop.mutations[mut].neutral:
                 # bubble is open
                 if pop.mcounts[mut] > 0:
@@ -32,51 +32,53 @@ class BubbleRecorder(object):
                     # new segregating mutation; open bubble
                     else:
                         self.weights[mkey] = pop.mcounts[mut]
-                    
-def evolve_draft(ngens, N, s, mu, r, burn = 500, seed = 42):
+
+
+def evolve_draft(ngens, N, s, mu, r, burn=500, seed=42):
     """
     Evolve drafting region and record bubble sizes
     """
 
     pop = fwdpy11.SlocusPop(N)
-    rng=fwdpy11.GSLrng(seed)
-    
+    rng = fwdpy11.GSLrng(seed)
+
     params = fwdpy11.model_params.SlocusParams(
-        nregions = [fwdpy11.Region(0, 1, 1)],
-        sregions = [fwdpy11.ExpS(0, 1, 1, s, 1.0)],
-        recregions = [fwdpy11.Region(0, 1, 1)],
-        gvalue = fwdpy11.fitness.SlocusAdditive(2.0),
-        demography = np.array([N]*ngens, dtype=np.uint32),
+        nregions=[fwdpy11.Region(0, 1, 1)],
+        sregions=[fwdpy11.ExpS(0, 1, 1, s, 1.0)],
+        recregions=[fwdpy11.Region(0, 1, 1)],
[…]

### 2018-03-19 fwdpy11/bubble_batchrun.py
fixed append issue

@@ -8,42 +8,46 @@ import subprocess
 import itertools as it
 import numpy as np
 
-ngens = 50000
-burn = 30000
-mu = 1e-4
-r = 1e-5
+ngens = 10000
+theta = 0.4
+r = 1e-6
 reps = 10
 seed = 42
 
 pars = ['N', 's']
-parvals =  [[1000, 5000, 10000, 50000], list(np.linspace(0.002,0.1,50))]
+parvals = [[1000, 5000, 10000, 50000], list(np.linspace(0.002, 0.1, 50))]
 
 parsets = list(it.product(*parvals))
 npars = len(pars)
 nsets = len(parsets)
 
-outputbase = 'bubble_sim_data_14-03-2018'
+outputbase = 'bubble_sim_data_18-03-2018'
 
 for p in range(0,nsets):
-    pset = parsets[p]
+    pset = dict(zip(pars, parsets[p]))
     
     simstr = 'bubble_sim.py '
     # add parameter values as command line options
     simstr += '--file={} '.format(outputbase + '.pkl')
-    simstr += '--mu={} --r={} --ngens={} --burn={} --reps={} --seed={} '.format(mu,
-                                                                                r,
-                                                                                ngens,
-                                                                                burn,
-                                                                                reps,
-                                                                                seed)
-    simstr += ' '.join(['--{}={}'.format(p[0], p[1]) for p in zip(pars, pset)])
+    simstr += (
+        '--N={} --s={} --mu={} --r={} --ngens={} --burn={} --reps={} --seed={} '
+        .format(pset['N'],
+                pset['s'],
+                theta/(4*pset['N']),
+                r,
+                ngens,
+                5*pset['N'],
+                reps,
+                seed))
 
-    cmdstr = 'sbatch --job-name=\'bubble_sim\' --output={}_{}.out --wrap=\'{}\''.format(outputbase,
-                                                                                        str(p+1).zfill(len(str(nsets))),
-                                                                                        simstr)
+    cmdstr = (
+        'sbatch --job-name=\'bubble_sim\' --output={}_{}.out --wrap=\'{}\''
+        .format(outputbase,
+                str(p+1).zfill(len(str(nsets))),
+                simstr))
 
-    # print(cmdstr)
-    cmd = shlex.split(cmdstr)
+    print(cmdstr)
+    # cmd = shlex.split(cmdstr)
 
-    proc = subprocess.Popen(cmd)
-    proc.wait()
+    # proc = subprocess.Popen(cmd)
+    # proc.wait()

### 2018-03-19 fwdpy11/bubble_sim.py

@@ -45,7 +45,7 @@ def evolve_draft(ngens, N, s, mu, r, burn=500, reps=10, seed=42):
         fwdpy11.wright_fisher.evolve(rng, pop, params)
 
         # recorded runs
-        pdict.update(demography=np.array([N]*(ngens-burn), dtype=np.uint32))
+        pdict.update(demography=np.array([N]*ngens, dtype=np.uint32))
         recorder = bubble_recorder.BubbleRecorder()
         fwdpy11.wright_fisher.evolve(rng, pop, params, recorder)
 
@@ -69,7 +69,7 @@ def main():
     parser.add_argument("--s", type=float, default=0.001, help="selection coefficient")
     parser.add_argument("--mu", type=float, default=0.0001, help="mutation rate")
     parser.add_argument("--r", type=float, default=0.00001, help="recombination rate")
-    parser.add_argument("--ngens", type=int, default=10000, help="number of generations")
+    parser.add_argument("--ngens", type=int, default=10000, help="number of generations (after burnin)")
     parser.add_argument("--burn", type=int, default=5000, help="burnin generations")
     parser.add_argument("--reps", type=int, default=10, help="number of replicates")
     parser.add_argument("--seed", type=int, default=42, help="random number seed")
@@ -88,7 +88,7 @@ def main():
                                                      args.reps,
                                                      args.seed)
 
-    with open(args.file, 'wb') as f:
+    with open(args.file, 'ab') as f:
         fcntl.flock(f, fcntl.LOCK_EX)
         pickle.dump(
             {'pars': dict(args._get_kwargs()),

### 2018-03-19 ms/valley_crossing.tex

@@
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
-During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm fix}$.
+During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
 Finally, the complex adaptation must sweep to fixation.
@@
 Most of the time, a bubble will simply arise and go extinct.
-However, with probability $1-e^{-\mu p_{\mathrm fix} W}$, the bubble will give rise to a successful double mutant.
-Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm fix} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
-
+However, with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
+Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
 One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
@@
 
-We begin by considering the dynamics of $p_{\mathrm fix}$ before moving on to consider the distribution of bubble sizes.
+We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
 Let $p = p_{\mathrm fix}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
@@
 We now consider the behavior of single mutant bubbles.
-Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm fix}$.
+Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
 We proceed to derive an equation for the time evolution of the Laplace transform.
@@
 We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
-For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model, where any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
-Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation.
+For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model.
+Any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
+Hence, the population experiences a constant influx of beneficial mutations at the background loci.
+Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
 In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
+We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.
 
@@
 This offers some assurance that independent valley crossing trials are indeed independent despite occurring in the same simulation run.
+
+We partitioned our simulations as follows.
+First, we considered the process of sequential fixation, in which the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
+In the drift limit, this obtains when $N \ll \mathrm{min}(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for drift to inhibit it.
+Next, we considered the stochastic tunneling regime.
+Tunneling becomes relevant when $N$ is too large for sequential fixation but also $\ll 1/\mu$, meaning that bubbles are unlikely to overlap and, if they do overlap, are unlikely to interfere with each other.
+By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, meaning that double mutants are produced at a high rate (semi-deterministic tunneling).
 
@@
 The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
+
+\section*{Discussion}
+
+We have used simulation methods to examine the rate at which rapidly adapting populations cross fitness valleys.
+Though the overall time required is generally higher than in slowly adapting (drifting) populations, there are two important caveats.
+First, the effect of the deleterious intermediate is much smaller.
+Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is likely to be much longer in rapidly adapting populations.
+This means that the presence of a fitness valley is not as steep of an impediment to the evolution of a complex adaptation for a drafting population as it is for a drifting one.
+One may compare our results to those of \citet{ochs_2015}, which found that valley crossing in drifting populations is slowed by intermediate population sizes.
+
+We have focused exclusively on asexual populations, but extensions to sexual populations are possible.
+The relevant parameter is likely to be the product of $N$ and the proportion of the fitness standard deviation segregating in a small, effectively asexual block, $\sigma_b$ \citep{neher_kessinger_2013}.
+If both $N\sigma$ and $N\sigma_b$ are large, and the focal loci segregate in the same block, then the analysis presented here should determine the crossing time.
+If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the distance between the focal loci is much larger than the block length $\xi_b$, then recombination between the focal loci is likely to be effectively ``free".
+In that case, the communal recombination analysis of \citet{neher_shraiman_2011} and \citet{neher_shraiman_2010} is likely to be important.
+The rate at which complex adaptations arise and fix will depend sensitively not on the distribution of background mutation effects but on the dynamics of recombination.
 

### 2018-03-19 fwdpy11/bubble_batchrun.py
new sims

@@ -9,6 +9,7 @@ import itertools as it
 import numpy as np
 
 ngens = 10000
+burnx = 5
 theta = 0.4
 r = 1e-6
 reps = 10
@@ -25,7 +26,7 @@ outputbase = 'bubble_sim_data_18-03-2018'
 
 for p in range(0,nsets):
     pset = dict(zip(pars, parsets[p]))
-    
+
     simstr = 'bubble_sim.py '
     # add parameter values as command line options
     simstr += '--file={} '.format(outputbase + '.pkl')
@@ -36,7 +37,7 @@ for p in range(0,nsets):
                 theta/(4*pset['N']),
                 r,
                 ngens,
-                5*pset['N'],
+                burnx*pset['N'],
                 reps,
                 seed))
 
@@ -46,8 +47,8 @@ for p in range(0,nsets):
                 str(p+1).zfill(len(str(nsets))),
                 simstr))
 
-    print(cmdstr)
-    # cmd = shlex.split(cmdstr)
+    # print(cmdstr)
+    cmd = shlex.split(cmdstr)
 
-    # proc = subprocess.Popen(cmd)
-    # proc.wait()
+    proc = subprocess.Popen(cmd)
+    proc.wait()

### 2018-03-19 fwdpy11/bubble_sim.py

@@ -76,8 +76,6 @@ def main():
     parser.add_argument("--file", default='data.pkl', help="output file")
 
     args = parser.parse_args()
-    if args.burn > args.ngens:
-        parser.error('burn must be < ngens')
 
     pops, weights, alpha, means, vars = evolve_draft(args.ngens,
                                                      args.N,

### 2018-03-19 fwdpy11/bubble_sim.py
added timing elements

@@ -9,6 +9,7 @@ import powerlaw as pl
 import pickle
 import fcntl
 from argparse import ArgumentParser
+from datetime import datetime
 
 # pybind11 C++ recorder
 import cppimport
@@ -77,6 +78,8 @@ def main():
 
     args = parser.parse_args()
 
+    start = datetime.now()
+    print (':: sim started at ' + str(start))
     pops, weights, alpha, means, vars = evolve_draft(args.ngens,
                                                      args.N,
                                                      args.s,
@@ -95,6 +98,9 @@ def main():
              'alpha': alpha,
              'means': means,
              'vars': vars}, f)
+
+    end = datetime.now
+    print(':: sim finished at ' + str(end) + ' and took ' + str(end-start))
 
 
 if __name__ == "__main__":

### 2018-03-21 fwdpy11/bubble_batchrun.py
more mutation for bubbles

@@ -10,7 +10,7 @@ import numpy as np
 
 ngens = 10000
 burnx = 5
-theta = 0.4
+theta = 4.0
 r = 1e-6
 reps = 10
 seed = 42
@@ -22,9 +22,9 @@ parsets = list(it.product(*parvals))
 npars = len(pars)
 nsets = len(parsets)
 
-outputbase = 'bubble_sim_data_18-03-2018'
+outputbase = 'bubble_sim_data_21-03-2018'
 
-for p in range(0,nsets):
+for p in range(0, nsets):
     pset = dict(zip(pars, parsets[p]))
 
     simstr = 'bubble_sim.py '

### 2018-03-21 fwdpy11/bubble_sim.py

@@ -79,7 +79,8 @@ def main():
     args = parser.parse_args()
 
     start = datetime.now()
-    print (':: sim started at ' + str(start))
+    print (':: sim started at '
+           + start.isoformat(sep=' ', timespec='seconds'))
     pops, weights, alpha, means, vars = evolve_draft(args.ngens,
                                                      args.N,
                                                      args.s,
@@ -100,7 +101,9 @@ def main():
              'vars': vars}, f)
 
     end = datetime.now
-    print(':: sim finished at ' + str(end) + ' and took ' + str(end-start))
+    print(':: sim finished at '
+          + end.isoformat(sep=' ', timespec='seconds')
+          + ' and took ' + str(end-start))
 
 
 if __name__ == "__main__":

### 2018-03-27 fwdpy11/bubble_sim.py
typo

@@ -100,7 +100,7 @@ def main():
              'means': means,
              'vars': vars}, f)
 
-    end = datetime.now
+    end = datetime.now()
     print(':: sim finished at '
           + end.isoformat(sep=' ', timespec='seconds')
           + ' and took ' + str(end-start))

### 2018-03-27 ms/valley_crossing.tex

@@
 Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
-In sexual populations, much the same description applies.
-The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013} or, in a sexual population, the proportion of fitness variation $\sigma_b$ that segregates within an effectively asexual block length \citep{neher_kessinger_2013}.
+The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013}.
+In sexual populations, the accuracy of this description depends on the recombination model.
+When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
+Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
+When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
+If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively.
 
@@
 We extend this to asexual populations and comment on alternate modes of recombination, such as crossovers.
-Using analytical approaches and forward simulations, we confirm that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
+
+The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic without the ``smoothing" effect provided by recombination.
+We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are very difficult to obtain.
+We confirm that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
+In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is more likely to favor crossing a fitness valley.
 These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
-Let $p = p_{\mathrm fix}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
+Let $p = p_{\mathrm{fix}}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
 In the absence of any perturbations, the evolution of $p$ can be described by a birth-death process, yielding
@@
 We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
-For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model.
-Any time a locus becomes monomorphic, a mutation is injected into a random individual in the population.
-Hence, the population experiences a constant influx of beneficial mutations at the background loci.
-Background fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
+For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
+In this way, the population experiences a constant influx of beneficial mutations at the background loci.
+These beneficial mutations' fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
 In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
@@
 After equilibration, we set a per-site mutation rate $\mu$ for the focal loci, with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$, consistent with our analytical model.
-When the double mutant reaches frequency $0.5$, we consider it to have fixed, then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background, and allow the population to equilibrate for another $N/10$ generations.
+When the double mutant reaches frequency $0.5$, we consider it to have fixed.
+We then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background, and allow the population to equilibrate for another $N/10$ generations.
 This offers some assurance that independent valley crossing trials are indeed independent despite occurring in the same simulation run.
@@
 
-Finally, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
+In addition, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
 Figure \ref{fig:bands} suggests that, at high values of $\sigma$, the value of $N\sigma$ is more important than $N$ itself in determining the valley crossing rate.
 The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
+In a slowly adapting population, $N$ determines the transition from sequential fixation to tunneling: if bubbles reach a maximum size that is of order $N$, they start to noticeably decrease the mean fitness, making fixation easier.
+When $N\sigma \gg 1$, bubbles almost never reach a large enough size for this condition to obtain.
+
+Finally, we examine the ratio $\tau_{\mathrm{sweep}}/\tau_{\mathrm{valley}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+Though this ratio will generally depend on $\delta$, as well as the selection coefficient $s_v$ of the complex adaptation and $s_u$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
+For several combinations of parameter values, we find that the ratio is higher for large $\sigma$ than for low $\sigma$.
+This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
+These results contrast with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
 

### 2018-04-03 ms/valley_crossing.tex
changes to manuscript, discussion of ratio data

@@
 One class of such adaptations has been termed ``irreducibly complex" because, in their current forms, they cease to function when one or more of their parts is removed.
-\citet{muller_1918} described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other, giving the appearance of irreducibility.
-An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a lower fitness mutational ``valley" in order to reach a higher fitness ``peak".
+\citet{muller_1918} described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other as parts of a system.
+When any such component of a system is removed, the system ceases to function, giving the impression of ``irreducibility".
+An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
 We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation \citep{trotter_2014}.
@@
 These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
-These phenomena were successfully extended to sexually reproducing populations by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
+Different modes of valley crossing were compared to simple sweep models by \citet{ochs_2015}, which found that valley crossing is disfavored at intermediate population sizes (where tunneling is the major mode of crossing) but can be favored in either large or small populations.
+Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
 Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
@@
 
-Previous studies of valley crossing focused primarily on populations where the background fitness variation is small enough to be negligible.
-In these populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
+Previous studies of valley crossing focused primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
+This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
+In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
+Such alleles can be effectively modeled by a diffusion approximation, and the ancestry of a population is described well by the Kingman coalescent.
 When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift.
-In a drafting asexual population, the fate of an allele depends sharply on its genetic background.
+Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum total of hitchhiking effects due to the genetic background.
+Draft is fundamentally different from drift: for example, drafting populations exhibit genealogies that are better described by an alternative coalescent process such as the Bolthausen-Sznitman coalescent or another Beta coalescent, and allele frequencies experience large jumps that cannot be encapsulated by a diffusion approximation.
+In addition, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency 1--or, equivalently, a depression in intermediate frequency alleles, which are swept out quickly.
 Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
 The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013}.
+
 In sexual populations, the accuracy of this description depends on the recombination model.
@@
 When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
-If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively.
+If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.
 
-Drift and draft are fundamentally different forces.
-It is not possible to understand a drafting population in terms of a drifting one, for example by defining a reduced ``effective population size".
-Drafting populations do not admit a diffusion approximation, and they exhibit qualitatively different genealogies and site frequency spectra relative to a drifting population.
-In fact, the possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
+It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbes of interest, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller.
+The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
 Draft likewise has profoundly different effects on the evolution of complex adaptations.
@@
 
-The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic without the ``smoothing" effect provided by recombination.
-We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are very difficult to obtain.
+The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic without the smoothing effect provided by recombination.
+We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
 We confirm that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
-In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is more likely to favor crossing a fitness valley.
+In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is often more likely to make the first choice than a drifting population.
 These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$: we discuss the effects of relaxing this assumption in the appendix.
-We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate at which the mean fitness advances.
+We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
 Let $x$ be the background fitness of a lineage.
@@
 
-There are essentially three components to this process.
+There are essentially three components to the process of stochastic tunneling.
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
@@
 Let $n(t)$ be the total number of single mutant individuals in the bubble, and suppose it appears at time $t_0$ and goes extinct at time $T$.
-Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, representing the total number of chances there are for a successful double mutant to appear.
+Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, or the total number of chances there are for a successful double mutant to appear.
 Most of the time, a bubble will simply arise and go extinct.
@@
 We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
+To do so, we follow the method of \citet{good_2012}.
 Let $p = p_{\mathrm{fix}}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
@@
 \end{equation}
-\citep{barton_1995, good_2012}, with $B$ and $D$ the birth and death rates respectively.
+\citep{barton_1995}, with $B$ and $D$ the birth and death rates respectively.
 The only dependence of $p$ on $t$ is via the time-changing mean fitness $vt$.
@@
 \begin{equation}
-v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon,
+v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon.
 \end{equation}
-where $\epsilon$ is now the fitness effect of a mutation in the background distribution, $\rho(\epsilon)$ is the distribution of mutational effects, and $U$ is the mutation rate of the background.
 Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
@@
 
+Finally, we compared the dynamics of valley crossing versus sweeping mutations.
+We performed simulations of a sweeping focal beneficial mutation at position $L/2$, with the background loci kept polymorphic as in the valley crossing simulations.
+We recorded the time for a beneficial mutation to arise and reach frequency $0.5$, considering it to have fixed at that point.
+We then computed the ratio of the sweep time to the valley crossing time for an equivalent complex adaptation.
+
 \section*{Results}
@@
 
-Finally, we examine the ratio $\tau_{\mathrm{sweep}}/\tau_{\mathrm{valley}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+A value greater than $1$ means that valley crossing takes much longer than a simple beneficial sweep does.
 Though this ratio will generally depend on $\delta$, as well as the selection coefficient $s_v$ of the complex adaptation and $s_u$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
-For several combinations of parameter values, we find that the ratio is higher for large $\sigma$ than for low $\sigma$.
+For several combinations of parameter values, we find that the ratio $\alpha$ is higher for large $\sigma$ than for low $\sigma$.
+This can be seen in figure \ref{fig:alpha_ratio}, in which we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
 This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
-These results contrast with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
+These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
+They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is the same: in their simulations, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult.
 

### 2018-04-25 ms/valley_crossing.tex
adding refs

@@
 We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
-We confirm that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations, but that much deeper valleys can be crossed.
-In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is often more likely to make the first choice than a drifting population.
+We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
+However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
+In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is often more likely to follow the first path than a drifting population, even when the fitness valley is rather deep.
 These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
-For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$: we discuss the effects of relaxing this assumption in the appendix.
+For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$.
 We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
@@
 
-There are essentially three components to the process of stochastic tunneling.
+There are three components to the process of stochastic tunneling.
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
@@
 Most of the time, a bubble will simply arise and go extinct.
-However, with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
+However, a double mutant appearing a fixing can be modeled as a Poisson process: with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
 Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
 One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
-That is to say, one cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
+That is to say, it is not possible to simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
 Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
@@
 
+(Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
 We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
-To do so, we follow the method of \citet{good_2012}.
-Let $p = p_{\mathrm{fix}}(x,s,t)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
-In the absence of any perturbations, the evolution of $p$ can be described by a birth-death process, yielding
+To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
+Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
+Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
+These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
+Then the fixation probability turns out to be governed by the integro-differential equation
 \begin{equation}
--\partial_t p(x,s,t) = Bp(x,s,t) - (B+D)p(x,s,t)^2
+v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
 \end{equation}
-\citep{barton_1995}, with $B$ and $D$ the birth and death rates respectively.
-The only dependence of $p$ on $t$ is via the time-changing mean fitness $vt$.
-Defining $X = x - vt$, we then have $\partial_t p(x-vt) = -v\partial_X p(X)$.
-This can be rewritten as a wave equation
-\begin{equation}
-v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2.
-\end{equation}
-This is a Bernoulli equation and is readily solved, but the only appropriate boundary condition yields $p = 0$ everywhere.
-This is intuitive: the inexorable advance of the mean fitness implies that all lineages are ultimately doomed to extinction \emph{unless} they can increase their own fitness either via recombination or mutation.
-We thus need to amend the model to include a mutation term.
-Let $U$ be the mutation rate of the genetic background and $\epsilon$ be the fitness effect of a mutation, distributed as $\rho(\epsilon)$.
-Then we have
-\begin{equation}
-v \partial_X p(X,s) = (X+s)p(X,s) - (1+X+s)p(X,s)^2 + U\int_0^\infty \rho(\epsilon) \left[ p(X+\epsilon,s)-p(X,s)\right] d\epsilon.
-\end{equation}
+where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.
+The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
+This should be intuitive, as lineages in a large population are doomed to extinction as the mean fitness inexorably advances, unless they an increase their fitness via recombination or mutation.
 Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
@@
 
-To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package.
+To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package that implements a modified Wright-Fisher model.
 The simulation code is available from the authors upon request or via \texttt{GitHub}.
+In \texttt{FFPopSim}, individuals are organized by ``clones" (sets of individuals with the same genome): each individual gives rise to a Poisson distributed number of offspring, with the mean dependent on the relative fitness.
+The mean offspring number is further adjusted to keep the population roughly at a pre-specified carrying capacity.
 We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
-For the genetic background, we make use of an \emph{ersatz} ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
+Genetic loci are partitioned into two groups, the ``focal" loci (at which the epistatic alleles will segregate) and the ``background" loci (which are responsible for the underlying fitness distribution).
+For the background loci, we make use of an \emph{ersatz} ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
 In this way, the population experiences a constant influx of beneficial mutations at the background loci.
 These beneficial mutations' fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
-In this way, $U$ and $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
+In this way, the background mutation rate $U$ and average fitness effect $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
 We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.
@@
 During this time, the mutation rate $\mu$ for the focal loci is set to zero.
-After equilibration, we set a per-site mutation rate $\mu$ for the focal loci, with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$, consistent with our analytical model.
+After equilibration, we set a per-site mutation rate $\mu$ for the focal loci (which controls both forward and backward mutations), with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$, consistent with our analytical model.
 When the double mutant reaches frequency $0.5$, we consider it to have fixed.
@@
 Next, we considered the stochastic tunneling regime.
-Tunneling becomes relevant when $N$ is too large for sequential fixation but also $\ll 1/\mu$, meaning that bubbles are unlikely to overlap and, if they do overlap, are unlikely to interfere with each other.
-By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, meaning that double mutants are produced at a high rate (semi-deterministic tunneling).
+Tunneling becomes relevant when $N$ is too large for sequential fixation, i.e., the tunneling probability becomes more significant than the probability that the intermediate will fix: this mandates $N \gtrsim 1/\sqrt{\mu s}$.
+It also requires that $\ll 1/\mu$, meaning that only one mutant lineage is likely to be extant at any given time and, if they do co-occur, are unlikely to interfere with each other or substantially affect the mean fitness.
+This happens to be the necessary condition for a birth-death process to be an appropriate model.
+By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, reaching mutation-selection balance at low frequency and providing many opportunities for a lucky double mutant to appear (so-called ``semi-deterministic tunneling").
+At still higher population sizes ($\gtrsim s/\mu^2$), multiple double mutants are likely to appear every generation, and their dynamics can be modeled as deterministic.
 
 Finally, we compared the dynamics of valley crossing versus sweeping mutations.
-We performed simulations of a sweeping focal beneficial mutation at position $L/2$, with the background loci kept polymorphic as in the valley crossing simulations.
+We performed simulations with a sweeping focal beneficial mutation, with selection coefficient $s_{\mathrm{sweep}}$, at position $L/2$, with the background loci kept polymorphic as in the valley crossing simulations.
 We recorded the time for a beneficial mutation to arise and reach frequency $0.5$, considering it to have fixed at that point.
-We then computed the ratio of the sweep time to the valley crossing time for an equivalent complex adaptation.
+We then computed the ratio of the sweep time to the valley crossing time for an equivalent complex adaptation, with valley depth $\delta$ and fitness advantage $s_{\mathrm{valley}}$.
+To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation favors valley crossing over simple sweeps.
 
@@
 At high values of $\sigma$, the fitness effect of the deleterious intermediate \emph{almost does not matter}, meaning that very deep valleys can be crossed.
-This supports the view that traversing rugged fitness landscapes by valley crossing may be the rule, not the exception, in rapidly adapting populations.
+This supports the view that traversing rugged fitness landscapes by valley crossing may be quite common in rapidly adapting populations.
 
@@
 
+\begin{figure}
+\begin{subfigure}[b]{0.2\textwidth}
+\includegraphics[width=\textwidth]{Figures/tau_valley_-6.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.2\textwidth}
+\includegraphics[width=\textwidth]{Figures/tau_valley_-2.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.2\textwidth}
+\includegraphics[width=\textwidth]{Figures/alpha_-6.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.2\textwidth}
+\includegraphics[width=\textwidth]{Figures/alpha_-2.pdf}
+\end{subfigure}
+\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
+Left two plots: valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
+Right two plots: the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.
+Deeper valleys in smaller populations are disfavored relative to simple sweeps.
+But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar).}
+\label{fig:ratios}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{Figures/alpha_ratio.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$. Genetic draft is most helpful for crossing deep valleys at large population sizes.}
+\label{fig:alpha_ratio}
+\end{figure}
+
 Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
@@
 Though this ratio will generally depend on $\delta$, as well as the selection coefficient $s_v$ of the complex adaptation and $s_u$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
-For several combinations of parameter values, we find that the ratio $\alpha$ is higher for large $\sigma$ than for low $\sigma$.
-This can be seen in figure \ref{fig:alpha_ratio}, in which we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
+In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
 This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.

### 2018-05-04 ms/valley_crossing.tex
extending the discussion

@@
 where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.
+
+We zero in on the ``infinitesimal" model, where mutations are of small effect compared to $\sigma$.
+In that case, the fixation probability is governed by the equation
+
+\begin{equation}
+v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
+\end{equation}
+
 The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
@@
 
+The dynamics by which rugged fitness landscapes are traversed has received significant attention in the literature in recent years.
+Valley crossing has previously been shown to be favored in small populations, where the effects of selection against intermediates are mitigated, and in large populations, where multiple mutants are much more likely to appear \citep{weissman_2009, ochs_2015}.
+Low amounts of recombination also appear to be helpful \citep{weissman_2010}.
+An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
+Like population subdivision, genetic draft favors the appearance of complex adaptations by forcing selection to act on entire genotypes, which may be carrying useful but provisionally harmful deleterious intermediate mutations.
+This can be contrasted with effects such as high migration rates and high recombination rates \citep{neher_shraiman_2009}, which force selection to act on individual alleles rather than genotypes.
+
+Two further intuitions can be drawn from this argument.
+First, there is a sort of duality between the evolution of complex adaptations and the evolution of cooperation, which can be seen as a complex behavioral adaptation.
+It is well known that higher migration rates between demes, which cause organisms to be less likely to interact with close kin, disfavor the evolution of cooperation: likewise, when migration rates are low, cooperation can be favored \citep{van_cleve_2015} (cite some other support here).
+Evolutionary forces which strengthen existing associations between loci are more likely to lead to such complex phenotypes: evolutionary forces which weaken these associations hinder their evolution.
+This reasoning applies whether the loci in question appear in the same individual (as in the case of sign epistasis) or in different individuals (as in the case of social evolution).
+These effects can even interact: cooperation can function as an additional evolutionary force that favors valley crossing \citep{Obolski:Lewin-Epstein:2017} (cite some additional support, possible reference to evolutionary capacitance--there is some work on army ants that has similar implications).
+The interplay between cooperation and valley crossing is an area that needs further study, as it may shed light on the evolution of cooperation within large microbial populations such as (cite some examples here).
+
+Second, the relationship between genetic draft and valley crossing has implications for Wright's ``shifting balance theory".
+Population subdivision followed by colonization of a new fitness peak may be an important aspect of the evolutionary process, but it is not obvious \emph{how} important \citep{coyne_barton_turelli_2000}.
+However, populations which experience both subdivision \emph{and} rapid adaptation are ideal candidates in which the shifting balance theory may apply.
+Such populations include HIV, in which host-specific adaptations evolve quickly (cite some support) and in which deleterious mutations are known to hitchhike to high frequency \citep{zanini_2013, zanini_2015}.
+Our work also shines light on the likely pathway through which multi-drug resistance evolves in HIV.
+In recent years of the HIV pandemic, resistance has generally been slower to evolve \citep{feder_2015}, in general because more and stronger drugs are used.
+There are two possible explanations for this.
+One option is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
+The other is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.
+Both factors are likely to play a role.
+However, since HIV is a population in which draft, not drift, is the dominant factor and hence population size is less important in determining the evolutionary dynamics, our work suggests that the second is likely to be more significant compared to other populations.
+
 %In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with \citet{weissman_2009}: when $\rho > 0$, the rate agrees with \citet{weissman_2010}.

### 2018-05-08 ms/valley_crossing.tex
minor edits

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Genetic draft and the evolution of ``irreducible complexity"}
+\title{Genetic draft and valley crossing}
 \author{Taylor~Kessinger}
@@
 However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
-In addition, we confirm that, given the option of crossing a fitness valley or simply allowing a beneficial mutation to sweep, a rapidly adapting population is often more likely to follow the first path than a drifting population, even when the fitness valley is rather deep.
+In addition, we confirm that in rapidly adapting population, adaptive events are more likely to be complex adaptations than they are in drifting, neutrally evolving populations: genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
 These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$.
-We further assume that the large number of loci of weak effect contribute to a background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
+We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
 Let $x$ be the background fitness of a lineage.
@@
 
-In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a selective sweep to occur.
+In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a double mutant to sweep through the population.
 At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
@@
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
-During the lifetime of this mutant lineage, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
+If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
+We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a ``bubble": the time integrated number of individuals $W$ is the ``weight".
+During the lifetime of a bubble, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
 Finally, the complex adaptation must sweep to fixation.
 The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
-In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble and $T_1$ is the sweep time.
+In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
 We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.
 
-We refer to single mutant lineages as ``bubbles".
-Let $n(t)$ be the total number of single mutant individuals in the bubble, and suppose it appears at time $t_0$ and goes extinct at time $T$.
-Then $W = \int_{t_0}^T n(t)dt$ is the ``weight" of the bubble, or the total number of chances there are for a successful double mutant to appear.
 Most of the time, a bubble will simply arise and go extinct.
-However, a double mutant appearing a fixing can be modeled as a Poisson process: with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
+If mutation and fixation are sufficiently rare events, then a double mutant fixing can be modeled as a Poisson process: with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
 Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
@@
 
-To test our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package that implements a modified Wright-Fisher model.
-The simulation code is available from the authors upon request or via \texttt{GitHub}.
+To analyze our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package that implements a modified Wright-Fisher model.
+The simulation code is available via \texttt{GitHub}.
 In \texttt{FFPopSim}, individuals are organized by ``clones" (sets of individuals with the same genome): each individual gives rise to a Poisson distributed number of offspring, with the mean dependent on the relative fitness.
@@
 We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
-Genetic loci are partitioned into two groups, the ``focal" loci (at which the epistatic alleles will segregate) and the ``background" loci (which are responsible for the underlying fitness distribution).
-For the background loci, we make use of an \emph{ersatz} ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
+Genetic loci are partitioned into two groups, the two ``focal" loci (at which the epistatic alleles will segregate) and ``background" loci (which are responsible for the underlying fitness distribution).
+For the background loci, we make use of a modified ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
 In this way, the population experiences a constant influx of beneficial mutations at the background loci.
-These beneficial mutations' fitness effects are drawn from an exponential distribution, and the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation: they are multipled by the current variance and divided by $\sigma^2$.
+These mutations' fitness effects are drawn from an exponential distribution; the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation (they are multipled by the current variance and divided by $\sigma^2$).
 In this way, the background mutation rate $U$ and average fitness effect $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
@@
 When the double mutant reaches frequency $0.5$, we consider it to have fixed.
-We then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background, and allow the population to equilibrate for another $N/10$ generations.
+We then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background by drawing anew from an exponential distribution, and allow the population to equilibrate for another $N/10$ generations.
 This offers some assurance that independent valley crossing trials are indeed independent despite occurring in the same simulation run.
@@
 We partitioned our simulations as follows.
-First, we considered the process of sequential fixation, in which the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
-In the drift limit, this obtains when $N \ll \mathrm{min}(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for drift to inhibit it.
+We first studied the effect of increasing $\sigma$ in what \citet{weissman_2009} describes as the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes.
+In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
+In the drift limit, sequential fixation applies when $N \ll \mathrm{min}(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
 Next, we considered the stochastic tunneling regime.
@@
 We recorded the time for a beneficial mutation to arise and reach frequency $0.5$, considering it to have fixed at that point.
-We then computed the ratio of the sweep time to the valley crossing time for an equivalent complex adaptation, with valley depth $\delta$ and fitness advantage $s_{\mathrm{valley}}$.
+We then computed the ratio of the sweep time to the valley crossing time for a comparable complex adaptation, with valley depth $\delta$ and larger fitness advantage $s_{\mathrm{valley}}$.
 To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation favors valley crossing over simple sweeps.
@@
 Here we present the key results of our simulation runs.
-We first studied the effect of increasing $\sigma$ in what \citet{weissman_2009} describes as the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes.
-Figures \ref{fig:deterministic} and \ref{fig:tunneling} show the expected result: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
+In general, we expect valley crossing in rapidly adapting populations to require more time, as bubbles are doomed to extinction more quickly and beneficial lineages are less likely to fix, especially when valleys are shallow.
+Figures \ref{fig:deterministic} and \ref{fig:tunneling} confirm these expectations: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
 It is tempting to interpret this as a reduced $N_e$ due to a decrease in genetic diversity and the strength of selection.
@@
 At high values of $\sigma$, the fitness effect of the deleterious intermediate \emph{almost does not matter}, meaning that very deep valleys can be crossed.
-This supports the view that traversing rugged fitness landscapes by valley crossing may be quite common in rapidly adapting populations.
+This supports the view that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations.
 
+\iffalse
 \begin{figure}
@@
 When $N\sigma \gg 1$, bubbles almost never reach a large enough size for this condition to obtain.
+\fi
 
@@
 \caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
+In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
 Left two plots: valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
@@
 Deeper valleys in smaller populations are disfavored relative to simple sweeps.
-But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar).}
+But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar).
+In both plots, $N = 10^5$ is a fairly hard cutoff beyond which tunneling becomes semi-deterministic, as $N\mu > 1$ means that, on average, more than one mutant lineage is seeded every generation.}
 \label{fig:ratios}
@@
 Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
-A value greater than $1$ means that valley crossing takes much longer than a simple beneficial sweep does.
-Though this ratio will generally depend on $\delta$, as well as the selection coefficient $s_v$ of the complex adaptation and $s_u$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
+A higher value of the ratio $\alpha$ means that valley crossing takes much longer than a simple beneficial sweep does.
+Though the precise value of this ratio will generally depend on $\delta$, as well as the selection coefficient $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
+In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
 For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
 In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
-This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
-These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
-They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is the same: in their simulations, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult.
 
@@
 This means that the presence of a fitness valley is not as steep of an impediment to the evolution of a complex adaptation for a drafting population as it is for a drifting one.
-One may compare our results to those of \citet{ochs_2015}, which found that valley crossing in drifting populations is slowed by intermediate population sizes.
+This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
+These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
+They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is the same: in their simulations, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult.
 

### 2018-05-13 ms/valley_crossing.tex
re

@@
 
+We begin by considering the dynamics of $p_{\mathrm{fix}}(x,s)$ before moving on to consider the distribution of bubble sizes.
+We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
+In this case, the background beneficial mutation rate $U_b$ and mean fitness effect $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
+Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
+(note: cite Hallatschek (2013) and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
+\begin{equation}
+v\partial_x p(x,s) = x p(x,s) + D\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
+\end{equation}
+in which we have assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
+The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
+
+In order to compute $\Phi$ we must next consider the behavior of mutant lineages.
+Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
+\citet{neher_shraiman_2011} showed that $\phi$ follows the equation
+\begin{equation}
+\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - sigma^2 t + \delta)\phi^2,
+\end{equation}
+but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
+Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu p_{\mathrm{fix}}(x)$), means that a simple closed form expression for $\Phi$ is not forthcoming.
+We therefore focus on simulation methods to study the dynamics of valley crossing.
+
+\iffalse
 (Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
@@
 Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
+\fi
 
@@
 \begin{figure}
-\begin{subfigure}[b]{0.2\textwidth}
+\begin{subfigure}[b]{0.25\textwidth}
 \includegraphics[width=\textwidth]{Figures/tau_valley_-6.pdf}
 \end{subfigure}
-\begin{subfigure}[b]{0.2\textwidth}
+\begin{subfigure}[b]{0.25\textwidth}
 \includegraphics[width=\textwidth]{Figures/tau_valley_-2.pdf}
 \end{subfigure}
-\begin{subfigure}[b]{0.2\textwidth}
-\includegraphics[width=\textwidth]{Figures/alpha_-6.pdf}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/alpha_ratios.pdf}
 \end{subfigure}
-\begin{subfigure}[b]{0.2\textwidth}
-\includegraphics[width=\textwidth]{Figures/alpha_-2.pdf}
-\end{subfigure}
-\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
+\caption{(note: combine left two figures so they have only one colorbar--this will be easy) Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
 In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{Figures/alpha_ratio.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$. Genetic draft is most helpful for crossing deep valleys at large population sizes.}
+\includegraphics[width=0.4\textwidth]{Figures/compare_ratios.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$.
+Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes.
+But when semi-deterministic tunneling becomes more important than stochastic tunneling, it can slow down the crossing rate.}
 \label{fig:alpha_ratio}
@@
 In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
+In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
+\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutnats at a low but constant rate.
+This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
 

### 2018-05-24 ms/valley_crossing.tex
adding sources, filling in gaps

@@
 Such alleles can be effectively modeled by a diffusion approximation, and the ancestry of a population is described well by the Kingman coalescent.
-When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift.
+When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
 Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum total of hitchhiking effects due to the genetic background.
-Draft is fundamentally different from drift: for example, drafting populations exhibit genealogies that are better described by an alternative coalescent process such as the Bolthausen-Sznitman coalescent or another Beta coalescent, and allele frequencies experience large jumps that cannot be encapsulated by a diffusion approximation.
-In addition, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency 1--or, equivalently, a depression in intermediate frequency alleles, which are swept out quickly.
+Draft is fundamentally different from drift: for example, drafting populations exhibit genealogies that are better described by an alternative coalescent process such as the Bolthausen-Sznitman coalescent or another Beta coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}, and allele frequencies experience large jumps that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}.
+In addition, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency 1--or, equivalently, a depression in intermediate frequency alleles, which are swept out quickly \citep{kosheleva_2013, neher_shraiman_2011}.
 Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
@@
 
-It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbes of interest, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller.
-The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly.
+It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbes of interest, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller \citep{masel_2011}.
+The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
 Draft likewise has profoundly different effects on the evolution of complex adaptations.
@@
 Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
-(note: cite Hallatschek (2013) and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
+\citet{hallatschek_2011} and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
 \begin{equation}
@@
 \begin{equation}
-\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - sigma^2 t + \delta)\phi^2,
+\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - \sigma^2 t + \delta)\phi^2,
 \end{equation}
@@
 \begin{figure}
-\begin{subfigure}[b]{0.25\textwidth}
-\includegraphics[width=\textwidth]{Figures/tau_valley_-6.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.25\textwidth}
-\includegraphics[width=\textwidth]{Figures/tau_valley_-2.pdf}
+\begin{subfigure}[b]{0.43\textwidth}
+\includegraphics[width=\textwidth]{Figures/tau_compare.pdf}
 \end{subfigure}
@@
 \end{subfigure}
-\caption{(note: combine left two figures so they have only one colorbar--this will be easy) Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
+\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
 In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
@@
 In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
-\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutnats at a low but constant rate.
+\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
+At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
 This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
@@
 Though the overall time required is generally higher than in slowly adapting (drifting) populations, there are two important caveats.
-First, the effect of the deleterious intermediate is much smaller.
+First, the effect of the deleterious intermediate is much smaller: for $\delta \lesssim \sigma$, the precise value of $\delta$ almost does not affect the crossing rate.
 Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is likely to be much longer in rapidly adapting populations.
@@
 This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
-These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem (competition between a sweeping beneficial mutant and a valley crossing adaptation) in a slowly evolving population.
-They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is the same: in their simulations, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult.
+These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem in a slowly evolving population.
+They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is similar: in their model, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult especially for populations of intermediate size.
 
@@
 If both $N\sigma$ and $N\sigma_b$ are large, and the focal loci segregate in the same block, then the analysis presented here should determine the crossing time.
-If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the distance between the focal loci is much larger than the block length $\xi_b$, then recombination between the focal loci is likely to be effectively ``free".
-In that case, the communal recombination analysis of \citet{neher_shraiman_2011} and \citet{neher_shraiman_2010} is likely to be important.
-The rate at which complex adaptations arise and fix will depend sensitively not on the distribution of background mutation effects but on the dynamics of recombination.
+If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the distance between the focal loci is much larger than the block length $\xi_b$, then recombination between the focal loci can be modeled as ``free", with individuals effectively shuffling their entire genomes every generation.
+In that case, the communal recombination analysis of \citet{neher_shraiman_2011} and \citet{neher_shraiman_2010} is likely to be important: the relationship between $\sigma$ and the recombination rate $\rho$ determines the crossing rate.
+If $\sigma$ is large, then draft dominates the crossing rate: if $\rho$ is large, then drift dominates it.
 
@@
 Valley crossing has previously been shown to be favored in small populations, where the effects of selection against intermediates are mitigated, and in large populations, where multiple mutants are much more likely to appear \citep{weissman_2009, ochs_2015}.
-Low amounts of recombination also appear to be helpful \citep{weissman_2010}.
+Low amounts of recombination also appear to be helpful \citep{weissman_2010}: at high values of the recombination rate, double mutant individuals outcross with the wild type, producing more deleterious intermediates and retarding the growth of the double mutant.
 An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
@@
 First, there is a sort of duality between the evolution of complex adaptations and the evolution of cooperation, which can be seen as a complex behavioral adaptation.
-It is well known that higher migration rates between demes, which cause organisms to be less likely to interact with close kin, disfavor the evolution of cooperation: likewise, when migration rates are low, cooperation can be favored \citep{van_cleve_2015} (cite some other support here).
+In social evolution theory, high migration rates between demes, which cause organisms to be less likely to interact with close kin, disfavor the evolution of cooperation: likewise, when migration rates are low, cooperation can be favored \citep{van_cleve_2015}.
 Evolutionary forces which strengthen existing associations between loci are more likely to lead to such complex phenotypes: evolutionary forces which weaken these associations hinder their evolution.
 This reasoning applies whether the loci in question appear in the same individual (as in the case of sign epistasis) or in different individuals (as in the case of social evolution).
-These effects can even interact: cooperation can function as an additional evolutionary force that favors valley crossing \citep{Obolski:Lewin-Epstein:2017} (cite some additional support, possible reference to evolutionary capacitance--there is some work on army ants that has similar implications).
-The interplay between cooperation and valley crossing is an area that needs further study, as it may shed light on the evolution of cooperation within large microbial populations such as (cite some examples here).
+These effects can even interact: cooperation can function as an additional evolutionary force that favors valley crossing \citep{Obolski:Lewin-Epstein:2017} and can in some cases be a valley crossing adaptation in itself \citep{van_cleve_2013}.
+The interplay between cooperation and valley crossing is an area that needs further study, as it may shed light on the evolution of cooperation within large microbial populations such as yeast \citep{gore_2009, gore_2013} and bacterial biofilms \citep{rainey_2003, van_gestel_2014}.
 
@@
 However, populations which experience both subdivision \emph{and} rapid adaptation are ideal candidates in which the shifting balance theory may apply.
-Such populations include HIV, in which host-specific adaptations evolve quickly (cite some support) and in which deleterious mutations are known to hitchhike to high frequency \citep{zanini_2013, zanini_2015}.
+Such populations include HIV, in which host-specific adaptations evolve quickly \citep{zhang_1997, wain_2007, dapp_2017, theys_2018} and in which deleterious mutations are known to hitchhike to high frequency \citep{zanini_2013, zanini_2015}.
 Our work also shines light on the likely pathway through which multi-drug resistance evolves in HIV.
@@
 There are two possible explanations for this.
-One option is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
+One is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
 The other is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.

### 2018-05-29 ms/valley_crossing.tex
adding refs still...

@@
 
-\iffalse
-(Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
-We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
-To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
-Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
-Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
-These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
-Then the fixation probability turns out to be governed by the integro-differential equation
-\begin{equation}
-v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
-\end{equation}
-where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.
+% \iffalse
+% (Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
+% We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
+% To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
+% Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
+% Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
+% These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
+% Then the fixation probability turns out to be governed by the integro-differential equation
+% \begin{equation}
+% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
+% \end{equation}
+% where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.
 
-We zero in on the ``infinitesimal" model, where mutations are of small effect compared to $\sigma$.
-In that case, the fixation probability is governed by the equation
+% We zero in on the ``infinitesimal" model, where mutations are of small effect compared to $\sigma$.
+% In that case, the fixation probability is governed by the equation
 
-\begin{equation}
-v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
-\end{equation}
+% \begin{equation}
+% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
+% \end{equation}
 
-The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
-This should be intuitive, as lineages in a large population are doomed to extinction as the mean fitness inexorably advances, unless they an increase their fitness via recombination or mutation.
-Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
-\begin{equation}
-p(X,s) =
-\begin{cases}
-0 &\mathrm{~if~} X+s < 0, \\
-X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
-X+s &\mathrm{~if~} X+s > X_c.
-\end{cases}
-\end{equation}
-(If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
-Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
-The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
-Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
-Then $X_c$ is determined by solving
-\begin{equation}
-2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}.
-\end{equation}
+% The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
+% This should be intuitive, as lineages in a large population are doomed to extinction as the mean fitness inexorably advances, unless they an increase their fitness via recombination or mutation.
+% Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
+% \begin{equation}
+% p(X,s) =
+% \begin{cases}
+% 0 &\mathrm{~if~} X+s < 0, \\
+% X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
+% X+s &\mathrm{~if~} X+s > X_c.
+% \end{cases}
+% \end{equation}
+% (If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
+% Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
+% The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
+% Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
+% Then $X_c$ is determined by solving
+% \begin{equation}
+% 2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}.
+% \end{equation}
 
-We now consider the behavior of single mutant bubbles.
-Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
-We proceed to derive an equation for the time evolution of the Laplace transform.
+% We now consider the behavior of single mutant bubbles.
+% Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
+% We proceed to derive an equation for the time evolution of the Laplace transform.
 
-Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
-We have
-\begin{align*}
--(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
-& + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
-& +kp(w,T|k-1,t,x).
-\end{align*}
-Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
-Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
-We therefore have
-\begin{equation}
--\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
-\end{equation}
-Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
-\begin{equation}
-\partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
-\end{equation}
-where $\tilde{\phi}$ = $\phi/\sigma$.
-We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
-In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
-However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
-To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
-Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
-\begin{equation}
-\partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
-\end{equation}
-Using matched asymptotics allows us to write
-\begin{equation}
-\tilde{\phi} = 
-\begin{cases}
-\tilde{z} &\mathrm{~if~} \theta < 0, \\
-\theta &\mathrm{~if~} \theta > 0.
-\end{cases}
-\end{equation}
-Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
-Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
-\fi
+% Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
+% We have
+% \begin{align*}
+% -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
+% & + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
+% & +kp(w,T|k-1,t,x).
+% \end{align*}
+% Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
+% Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
+% We therefore have
+% \begin{equation}
+% -\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
+% \end{equation}
+% Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
+% \begin{equation}
+% \partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
+% \end{equation}
+% where $\tilde{\phi}$ = $\phi/\sigma$.
+% We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
+% In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
+% However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
+% To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
+% Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
+% \begin{equation}
+% \partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
+% \end{equation}
+% Using matched asymptotics allows us to write
+% \begin{equation}
+% \tilde{\phi} = 
+% \begin{cases}
+% \tilde{z} &\mathrm{~if~} \theta < 0, \\
+% \theta &\mathrm{~if~} \theta > 0.
+% \end{cases}
+% \end{equation}
+% Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
+% Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
+% \fi
 
@@
 
-\iffalse
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{Figures/bands.pdf}
-\caption{Valley crossing times for fixed $N\sigma$, with $N$ allowed to vary up or down by a factor of $2$ (and $\sigma$ adjusted accordingly).
+% \iffalse
+% \begin{figure}
+% \includegraphics[width=0.4\textwidth]{Figures/bands.pdf}
+% \caption{Valley crossing times for fixed $N\sigma$, with $N$ allowed to vary up or down by a factor of $2$ (and $\sigma$ adjusted accordingly).
 Error bars are quartile ranges.
 Here, $\mu = 3 \times 10^{-6}$, $\delta = 10^{-3}$, and $s = 10^{-2}$.}
-\label{fig:bands}
-\end{figure}
+% \label{fig:bands}
+% \end{figure}
 
-In addition, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
-Figure \ref{fig:bands} suggests that, at high values of $\sigma$, the value of $N\sigma$ is more important than $N$ itself in determining the valley crossing rate.
-The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
-In a slowly adapting population, $N$ determines the transition from sequential fixation to tunneling: if bubbles reach a maximum size that is of order $N$, they start to noticeably decrease the mean fitness, making fixation easier.
-When $N\sigma \gg 1$, bubbles almost never reach a large enough size for this condition to obtain.
-\fi
+% In addition, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
+% Figure \ref{fig:bands} suggests that, at high values of $\sigma$, the value of $N\sigma$ is more important than $N$ itself in determining the valley crossing rate.
+% The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
+% In a slowly adapting population, $N$ determines the transition from sequential fixation to tunneling: if bubbles reach a maximum size that is of order $N$, they start to noticeably decrease the mean fitness, making fixation easier.
+% When $N\sigma \gg 1$, bubbles almost never reach a large enough size for this condition to obtain.
+% \fi
 
@@
 
-\iffalse
-First, we consider the case where $\rho = 0$ and focus on what \citet{weissman_2009} refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
-As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
-In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
-If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
-We refer to $w$ as the ``weight" of the bubble.
-The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
-Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
-The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
-This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.
+% \iffalse
+% First, we consider the case where $\rho = 0$ and focus on what \citet{weissman_2009} refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
+% As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
+% In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
+% If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
+% We refer to $w$ as the ``weight" of the bubble.
+% The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
+% Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
+% The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
+% This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.
 
-In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
-In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
-Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
-This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
-However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
-This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
-(Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
-Can start with equation 3 of \citet{neher_shraiman_2011} and drop the recombination term, which is what I'm doing now.
-This is tantamount to a ``mean field" approximation but might still work out okay.
-Other possible approach is to use something like the expression of  \citet{desai_fisher_2007} for the fixation probability and rescale it to a traveling wave model.
-Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)
+% In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
+% In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
+% Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
+% This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
+% However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
+% This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
+% (Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
+% Can start with equation 3 of \citet{neher_shraiman_2011} and drop the recombination term, which is what I'm doing now.
+% This is tantamount to a ``mean field" approximation but might still work out okay.
+% Other possible approach is to use something like the expression of  \citet{desai_fisher_2007} for the fixation probability and rescale it to a traveling wave model.
+% Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)
 
-One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
-That is to say, the total rate of valley crossing will be given by $\int P(x) p_{\mathrm{fix}}(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
-One cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
-This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
-We proceed to derive both $p_{\mathrm{fix}}(x,s)$ and $P(w)$ in the zero recombination limit.
+% One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+% That is to say, the total rate of valley crossing will be given by $\int P(x) p_{\mathrm{fix}}(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
+% One cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
+% This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
+% We proceed to derive both $p_{\mathrm{fix}}(x,s)$ and $P(w)$ in the zero recombination limit.
 
-The establishment probability $p(x,s)$ obeys the PDE
-\begin{equation}
-v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
-\end{equation}
-To see this, consider the generic expression for the fixation probability
-\begin{equation}
-p(X,t-dt) = p(X,t)-dt(D+B(X,t))p(X,t) + dt(B(X,t)(2p(X,t)-p(X,t)^2)
-\end{equation}
-\citep{barton_1995}. Taking the limit as $dt \to 0$ and inserting $D = 1, B = 1 + X -\bar{X}(t)+s$ yields a PDE for $p(X,t)$.
-Defining $x = X - vt$, we then have $\partial_t p(X-vt) = -v\partial_x p(x)$.
+% The establishment probability $p(x,s)$ obeys the PDE
+% \begin{equation}
+% v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
+% \end{equation}
+% To see this, consider the generic expression for the fixation probability
+% \begin{equation}
+% p(X,t-dt) = p(X,t)-dt(D+B(X,t))p(X,t) + dt(B(X,t)(2p(X,t)-p(X,t)^2)
+% \end{equation}
+% \citep{barton_1995}. Taking the limit as $dt \to 0$ and inserting $D = 1, B = 1 + X -\bar{X}(t)+s$ yields a PDE for $p(X,t)$.
+% Defining $x = X - vt$, we then have $\partial_t p(X-vt) = -v\partial_x p(x)$.
 
-This is a Bernoulli equation and is readily solved by substituting $u(x,s) = 1/p(x,s)$. We then have that $\partial_x u(x,s) = -\frac{\partial_x p(x,s)}{p(x,s)^2}$, so that
-\begin{equation}
-\partial_x u(x,s) = -\frac{(x+s)}{v}u(x,s) - \frac{1+x+s}{v}.
-\end{equation}
-The use of the integrating factor $e^{\int \frac{x+s}{v} dx} = Ce^{\frac{(x+s)^2}{2v}}$ yields
-\begin{equation}
-Ce^{\frac{(x+s)^2}{2v}} (\partial_x u + \frac{x+s}{v}u) = -Ce^{\frac{(x+s)^2}{2v}}\frac{1+x+s}{v}.
-\end{equation}
-The left side is simply $\partial_x(Ce^{\frac{(x+s)^2}{2v}}u)$, so (after dividing $C$ out) we can integrate both sides, then multiply by $e^{\frac{-(x+s)^2}{2v}}$ to obtain
-\begin{equation}
-u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}}) - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
-\end{equation}
-with $C$ a constant. Of course, $p = 1/u$.
+% This is a Bernoulli equation and is readily solved by substituting $u(x,s) = 1/p(x,s)$. We then have that $\partial_x u(x,s) = -\frac{\partial_x p(x,s)}{p(x,s)^2}$, so that
+% \begin{equation}
+% \partial_x u(x,s) = -\frac{(x+s)}{v}u(x,s) - \frac{1+x+s}{v}.
+% \end{equation}
+% The use of the integrating factor $e^{\int \frac{x+s}{v} dx} = Ce^{\frac{(x+s)^2}{2v}}$ yields
+% \begin{equation}
+% Ce^{\frac{(x+s)^2}{2v}} (\partial_x u + \frac{x+s}{v}u) = -Ce^{\frac{(x+s)^2}{2v}}\frac{1+x+s}{v}.
+% \end{equation}
+% The left side is simply $\partial_x(Ce^{\frac{(x+s)^2}{2v}}u)$, so (after dividing $C$ out) we can integrate both sides, then multiply by $e^{\frac{-(x+s)^2}{2v}}$ to obtain
+% \begin{equation}
+% u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}}) - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
+% \end{equation}
+% with $C$ a constant. Of course, $p = 1/u$.
 
-Unfortunately it is not clear how to get rid of $C$, as there is no obvious ``initial condition".
-The most we are likely to know about $p$ is that $p$ approaches $0$ in the limit of low $x$, but behaves $\sim s$ in the limit of high $x$ before saturating at $1$ (however, there are unlikely to be any individuals in the population at such high $x$).
-One possibility is to impose an \emph{ersatz} ``solvability condition": surely the fixation probability of a neutral mutation averaged over the entire population must equal $1/N$. We would then have
-\begin{equation}
-\int_{-\infty}^{\infty} P(x)p(x,0)dx = \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{e^{-x^2/2\sigma^2}}{-e^{-x^2/2\sigma^2} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - 1 + Ce^{-x^2/2\sigma^2}}dx = \frac{1}{N}.
-\end{equation}
-Multiplying the top and bottom of the integrand by $e^{x^2/2\sigma^2}$ would clean this up, yielding
-\begin{equation}
-\frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{1}{-\frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - e^{x^2/2\sigma^2} + C}dx = \frac{1}{N}.
-\end{equation}
-Unfortunately this integral has an imaginary error function in the denominator, so there is not likely to be a closed form expression: we could explore it numerically and see where it peaks, however.
+% Unfortunately it is not clear how to get rid of $C$, as there is no obvious ``initial condition".
+% The most we are likely to know about $p$ is that $p$ approaches $0$ in the limit of low $x$, but behaves $\sim s$ in the limit of high $x$ before saturating at $1$ (however, there are unlikely to be any individuals in the population at such high $x$).
+% One possibility is to impose an \emph{ersatz} ``solvability condition": surely the fixation probability of a neutral mutation averaged over the entire population must equal $1/N$. We would then have
+% \begin{equation}
+% \int_{-\infty}^{\infty} P(x)p(x,0)dx = \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{e^{-x^2/2\sigma^2}}{-e^{-x^2/2\sigma^2} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - 1 + Ce^{-x^2/2\sigma^2}}dx = \frac{1}{N}.
+% \end{equation}
+% Multiplying the top and bottom of the integrand by $e^{x^2/2\sigma^2}$ would clean this up, yielding
+% \begin{equation}
+% \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{1}{-\frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - e^{x^2/2\sigma^2} + C}dx = \frac{1}{N}.
+% \end{equation}
+% Unfortunately this integral has an imaginary error function in the denominator, so there is not likely to be a closed form expression: we could explore it numerically and see where it peaks, however.
 
-Another idea would be to consider an alternative way of solving the differential equation for $p$.
-Return to
-\begin{equation}
-v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
-\end{equation}
-Suppose we make the \emph{ansatz} that $x+s \ll 1$, and define $\chi = x/\sigma$, $\tilde{s} = s/\sigma$, and $\tilde{p}(\chi,\tilde{s}) = p(x,s)/\sigma$. Then
-\begin{equation}
-\partial_\chi \tilde{p}(\chi,\tilde{s}) = (\chi + \tilde{s})\tilde{p}(\chi,\tilde{s}) - \tilde{p}(\chi,\tilde{s})^2.
-\end{equation}
-Now define $q(\chi,\tilde{s})$ such that $\tilde{w}(\chi,\tilde{s}) = \partial_x \log q(\chi,\tilde{s})$.
-Then
-\begin{equation}
-\partial_\chi^2 q(\chi,\tilde{s}) - (\chi+\tilde{s})\partial_\chi q(\chi, \tilde{s}) = 0.
-\end{equation}
-Defining $\vartheta = \chi + \tilde{s}$ and $q(\vartheta) = e^{\vartheta^2/4}h(\vartheta)$ yields
-\begin{equation}
-\partial_\vartheta^2 h(\vartheta) - (\frac{1}{2} + \frac{\theta^2}{4})h(\vartheta) = 0,
-\end{equation}
-which is the parabolic cylinder equation.
-We then need to find the form of $q$ that has the correct asymptotics.
-This does have a solution:
-\begin{equation}
-q(\chi) = \int_0^\infty d\lambda e^{\vartheta \lambda - \lambda^2/2}/\lambda.
-\end{equation}
+% Another idea would be to consider an alternative way of solving the differential equation for $p$.
+% Return to
+% \begin{equation}
+% v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
+% \end{equation}
+% Suppose we make the \emph{ansatz} that $x+s \ll 1$, and define $\chi = x/\sigma$, $\tilde{s} = s/\sigma$, and $\tilde{p}(\chi,\tilde{s}) = p(x,s)/\sigma$. Then
+% \begin{equation}
+% \partial_\chi \tilde{p}(\chi,\tilde{s}) = (\chi + \tilde{s})\tilde{p}(\chi,\tilde{s}) - \tilde{p}(\chi,\tilde{s})^2.
+% \end{equation}
+% Now define $q(\chi,\tilde{s})$ such that $\tilde{w}(\chi,\tilde{s}) = \partial_x \log q(\chi,\tilde{s})$.
+% Then
+% \begin{equation}
+% \partial_\chi^2 q(\chi,\tilde{s}) - (\chi+\tilde{s})\partial_\chi q(\chi, \tilde{s}) = 0.
+% \end{equation}
+% Defining $\vartheta = \chi + \tilde{s}$ and $q(\vartheta) = e^{\vartheta^2/4}h(\vartheta)$ yields
+% \begin{equation}
+% \partial_\vartheta^2 h(\vartheta) - (\frac{1}{2} + \frac{\theta^2}{4})h(\vartheta) = 0,
+% \end{equation}
+% which is the parabolic cylinder equation.
+% We then need to find the form of $q$ that has the correct asymptotics.
+% This does have a solution:
+% \begin{equation}
+% q(\chi) = \int_0^\infty d\lambda e^{\vartheta \lambda - \lambda^2/2}/\lambda.
+% \end{equation}
 
-Next, we seek the weight distribution as a function of $x$.
-Let $p(w,T|k,t,x)$ be the probability that a bubble has attained a weight of $w$ at time $T$, given that there were $k$ individuals in the lineage at time $t$.
-Ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
-Here $w$ represents (as before) the weight of a mutant bubble and evolves according to
-\begin{align*}
--(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s)p(w,T|k,t,x) \\
-& + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
-& +kp(w,T|k-1,t,x). \\
-\end{align*}
-We have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$, by the assumption that the behavior of each of the $k$ copies of the mutant allele is independent of the other.
-Recall that, in the traveling wave approximation, $\bar{x} = vt = \sigma^2 t$.
-The notation here can be cleaned up by defining $\phi = 1 - \hat{p}$, as well as $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, $\tau = \sigma(T-t)$, and $\theta = \chi+\tau+\tilde{s}+z$.
-Rewriting the preceding in terms of $\phi$ (note that differentiating $p$ with respect to $w$ simply pulls out a factor of $z$) yields
-\begin{equation}
-\partial_\tau \phi(\tau,z,\chi) = z + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi).
-\end{equation}
-Unfortunately $\theta$ depends on $\tau$, so we will have to be careful here.
-This is a Riccati equation.
-Let $\phi_0(\tau,z,\chi)$ be a particular solution for $\phi$: then the general solution becomes
-\begin{equation}
-\phi(\tau,z,\chi) = \phi_0(\tau,z,\chi) + \frac{\Phi(\tau,z,\chi)}{C + \int \Phi(\tau,z,\chi) d\tau},
-\end{equation}
-where $\Phi(\tau,z,\chi) = e^{\int (-2\phi_0 + \theta ) d\tau}$.
-It remains to find a particular solution of $\phi$.
-We can ``guess" a steady state solution, in which case $0 = \phi^2 - \theta \phi - z = 0$: this corresponds to $\phi_0 = \frac{\theta \pm \sqrt{\theta^2 + 4z}}{2}$.
-Computing the exponent in the integral (and changing integration variables from $\tau$ to $\theta$) yields
-\begin{equation}
-\Phi (\tau, z, \chi) = e^{-\int \sqrt{\theta^2 + 4z} d\theta} = e^{\theta\sqrt{\theta^2+4z} + \frac{4z}{2}\log (\theta + \sqrt{\theta^2 + 4z})}.
-\end{equation}
-(Still need to solve this and carefully consider the integral boundary.)
+% Next, we seek the weight distribution as a function of $x$.
+% Let $p(w,T|k,t,x)$ be the probability that a bubble has attained a weight of $w$ at time $T$, given that there were $k$ individuals in the lineage at time $t$.
+% Ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
+% Here $w$ represents (as before) the weight of a mutant bubble and evolves according to
+% \begin{align*}
+% -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s)p(w,T|k,t,x) \\
+% & + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
+% & +kp(w,T|k-1,t,x). \\
+% \end{align*}
+% We have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$, by the assumption that the behavior of each of the $k$ copies of the mutant allele is independent of the other.
+% Recall that, in the traveling wave approximation, $\bar{x} = vt = \sigma^2 t$.
+% The notation here can be cleaned up by defining $\phi = 1 - \hat{p}$, as well as $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, $\tau = \sigma(T-t)$, and $\theta = \chi+\tau+\tilde{s}+z$.
+% Rewriting the preceding in terms of $\phi$ (note that differentiating $p$ with respect to $w$ simply pulls out a factor of $z$) yields
+% \begin{equation}
+% \partial_\tau \phi(\tau,z,\chi) = z + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi).
+% \end{equation}
+% Unfortunately $\theta$ depends on $\tau$, so we will have to be careful here.
+% This is a Riccati equation.
+% Let $\phi_0(\tau,z,\chi)$ be a particular solution for $\phi$: then the general solution becomes
+% \begin{equation}
+% \phi(\tau,z,\chi) = \phi_0(\tau,z,\chi) + \frac{\Phi(\tau,z,\chi)}{C + \int \Phi(\tau,z,\chi) d\tau},
+% \end{equation}
+% where $\Phi(\tau,z,\chi) = e^{\int (-2\phi_0 + \theta ) d\tau}$.
+% It remains to find a particular solution of $\phi$.
+% We can ``guess" a steady state solution, in which case $0 = \phi^2 - \theta \phi - z = 0$: this corresponds to $\phi_0 = \frac{\theta \pm \sqrt{\theta^2 + 4z}}{2}$.
+% Computing the exponent in the integral (and changing integration variables from $\tau$ to $\theta$) yields
+% \begin{equation}
+% \Phi (\tau, z, \chi) = e^{-\int \sqrt{\theta^2 + 4z} d\theta} = e^{\theta\sqrt{\theta^2+4z} + \frac{4z}{2}\log (\theta + \sqrt{\theta^2 + 4z})}.
+% \end{equation}
+% (Still need to solve this and carefully consider the integral boundary.)
 
-We now turn to the case where $\rho > 0$.
-Because we are explicitly dealing with crossovers, the results of \citet{neher_kessinger_2013} are useful here.
-We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
-The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
-The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
-If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
-If not, then the value of $N\sigma_b$ becomes the relevant factor.
-$N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
-On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
-\fi
+% We now turn to the case where $\rho > 0$.
+% Because we are explicitly dealing with crossovers, the results of \citet{neher_kessinger_2013} are useful here.
+% We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
+% The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
+% The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
+% If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
+% If not, then the value of $N\sigma_b$ becomes the relevant factor.
+% $N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
+% On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
+% \fi
 
-\bibliographystyle{plainnat}
+\bibliographystyle{genetics}
 \bibliography{bib}

### 2018-06-03 ms/valley_crossing.tex
edited intro

@@
-%\documentclass[rmp,twocolumn]{revtex4}
 \documentclass[rmp]{revtex4}
@@
 %\usepackage{subfig}
+
+\usepackage{fontspec}
 
@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+
+\usepackage{setspace}
 
@@
 
+\begin{spacing}{1.5}
+
 \section*{Introduction}
@@
 The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
-This appears to be how Darwin thought most complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies much of the field of adaptive dynamics.
-However, many adaptations exist that do not appear to have formed by this process.
-One class of such adaptations has been termed ``irreducibly complex" because, in their current forms, they cease to function when one or more of their parts is removed.
-\citet{muller_1918} described one method by which such adaptations could evolve: components that originally served independent functions might become dependent on each other as parts of a system.
-When any such component of a system is removed, the system ceases to function, giving the impression of ``irreducibility".
-An alternate method comes from the metaphor of a ``fitness landscape": a wild type individual must traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
-We seek to characterize this ``valley crossing" process, as there is reason to think it may in fact be a common mode of adaptation \citep{trotter_2014}.
+This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
+Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype where each mutant genotype in the sequence has a higher fitness than the previous one.
+If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
+However, many empirical fitness landscapes are not completely smooth and have multiple peaks \citep[reviewed in][]{Szendro:Schenk:2013,Visser:Krug:2014,Obolski:Ram:2018}.
+In such landscapes, wild type individuals may have to traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
+We seek to characterize this ``valley crossing" process in order to better understand the routes adaptation is likely to take on these rugged empirical landscapes \citep[e.g.,][]{Aguilar-Rodriguez:Payne:2017} and to assess the likelihood of valley crossing under more complex scenarios in which it may in fact be common \citep[e.g.][]{trotter_2014}.
 
 Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
-If the number of mutants produced every generation is high, the arisal of a complex adaptation that is destined to fix may be guaranteed.
-In small populations, wild type individuals may give rise to neutral or deleterious intermediate mutants, which drift to fixation.
-A third possibility is for wild type individuals to birth mutant ``bubbles", or transient subpopulations.
-These bubbles are ordinarily doomed to extinction due to drift, but a lucky bubble may give rise to a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
-Different modes of valley crossing were compared to simple sweep models by \citet{ochs_2015}, which found that valley crossing is disfavored at intermediate population sizes (where tunneling is the major mode of crossing) but can be favored in either large or small populations.
-Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
-Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
-High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.
+If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
+In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation.
+Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
+A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
+These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
+On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}.
+However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}.
+This suggests that tunneling may be the most difficult mode of valley crossing.
+% Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
+% Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
+% High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.
 
-Previous studies of valley crossing focused primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
+Previous studies of valley crossing focus primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
 This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
-In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles close to frequency zero or one.
-Such alleles can be effectively modeled by a diffusion approximation, and the ancestry of a population is described well by the Kingman coalescent.
-When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
-Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum total of hitchhiking effects due to the genetic background.
-Draft is fundamentally different from drift: for example, drafting populations exhibit genealogies that are better described by an alternative coalescent process such as the Bolthausen-Sznitman coalescent or another Beta coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}, and allele frequencies experience large jumps that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}.
-In addition, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency 1--or, equivalently, a depression in intermediate frequency alleles, which are swept out quickly \citep{kosheleva_2013, neher_shraiman_2011}.
-Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population, carrying linked alleles with them.
+In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare, close to frequency zero, or common, close to frequency one.
+Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982} where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
+When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by genetic drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
+Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
+Draft is fundamentally different from drift: for example, drafting populations experience large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}
+and have genealogies where more than two branches can merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
+Additionally under genetic draft, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency one--or, equivalently, a depression in intermediate frequency alleles, which are quickly swept out of the population \citep{neher_shraiman_2011,kosheleva_2013,neher_hallatschek_2013}.
+Only a handful of individuals near the ``nose" of the fitness distribution are likely to persist, and they give rise to the bulk of the future population and carry linked alleles with them.
 The parameter that determines whether drift or draft is more important is the product of the population size $N$ and the standard deviation in fitness $\sigma$ \citep{neher_hallatschek_2013}.
 
-In sexual populations, the accuracy of this description depends on the recombination model.
-When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
-Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
-When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
-If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.
+It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbial species, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller \citep{masel_2011}.
+The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
+Draft likewise may have profound effects on the evolution of complex adaptations. Preliminary evidence for this comes from
+\citet{neher_shraiman_2011} who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting population.
+They found that, compared to drift, draft generally makes small and intermediate sized mutant bubbles much rarer but very large bubbles are much more common.
+Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately clear.
+Here, we extend these previous results by studying in asexual populations how genetic draft affects the rate of crossing fitness valleys across a range of population sizes.
 
-It is increasingly realized that draft may play a critical role in shaping genetic diversity, especially in many microbes of interest, where population sizes can be large but ``effective population sizes" are many orders of magnitude smaller \citep{masel_2011}.
-The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
-Draft likewise has profoundly different effects on the evolution of complex adaptations.
-\citet{neher_shraiman_2011} explored the dynamics of stochastic tunneling in rapidly adapting sexual populations.
-The primary mode of recombination considered therein is ``free" recombination, in which parents essentially shuffle the entirety of their genomes during reproduction.
-We extend this to asexual populations and comment on alternate modes of recombination, such as crossovers.
-
-The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic without the smoothing effect provided by recombination.
-We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
+The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic.
+In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011}.
+We therefore must focus primarily on simulation approaches as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
 We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
 However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
-In addition, we confirm that in rapidly adapting population, adaptive events are more likely to be complex adaptations than they are in drifting, neutrally evolving populations: genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
-These observations add to the growing intuition that irreducible complexity may not be a surprising result of the evolutionary process, but rather an expected one.
+In addition, we confirm that in rapidly adapting populations, adaptive fixation of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
+These observations add to the growing intuition that complex adaptations that involve evolution across fitness valleys may not be a surprising result of the evolutionary process, but rather an expected one.
 
-\section*{Model}
+\section*{Mathematical background}
 
@@
 
-Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a particular beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
-A higher value of the ratio $\alpha$ means that valley crossing takes much longer than a simple beneficial sweep does.
-Though the precise value of this ratio will generally depend on $\delta$, as well as the selection coefficient $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend can nonetheless be observed.
+Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps.
+Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
+Though the precise value of this ratio depends on $\delta$ and the selection coefficients $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend nonetheless emerges.
 In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
@@
 If $\sigma$ is large, then draft dominates the crossing rate: if $\rho$ is large, then drift dominates it.
+
+%%% from into but more relevant here
+% In sexual populations, the accuracy of this description depends on the recombination model.
+% When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
+% Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
+% When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
+% If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.
 
@@
 
+\end{spacing}
+
 \end{document}

### 2018-06-04 ms/valley_crossing.tex
edited through Mathematical background

@@
 \usepackage{amssymb,amsfonts,amsmath}
-\usepackage{color}
+\usepackage{xcolor}
 \usepackage{breqn}
@@
 
-\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
-\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
+\newcommand{\EQ}[1]{Eq.~\eqref{eq:#1}}
+\newcommand{\EQS}[2]{Eqs.~\eqref{eq:#1} and \eqref{eq:#2}}
 \newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
@@
 
-\newcommand{\comment}[1]{{\color{red}#1}}
+\newcommand{\comment}[2][noinline]{\todo[color=red,#1]{#2}}
 
@@
 
+%%% math macros
 \DeclareMathOperator*{\argmin}{arg\,min}
+\newcommand{\pfix}{p_{\mathrm{fix}}}
 
 \usepackage{setspace}
+\usepackage[colorinlistoftodos]{todonotes}
 
@@
 If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
+This is sometimes called deterministic fixation of the double mutant \citep{weissman_2009}.
 In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation.
 Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
+This is the sequential fixation regime.
 A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
@@
 
+\begin{figure}
+\begin{center}
+  \includegraphics[scale=0.75]{Figures/valley.pdf}
+  \caption{Fitness landscape.}
+  \label{fig:landscape}
+\end{center}
+\end{figure}
+
 Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
 We focus on two loci, which are initially fixed for alleles $a$ and $b$.
-These alleles mutate to $A$ and $B$ at rates $\mu_a$ and $\mu_b$, respectively (we neglect back mutations).
-Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
-For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$.
-We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
-Let $x$ be the background fitness of a lineage.
-The distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
-Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.
+These alleles mutate to $A$ and $B$ at rates $\mu_A$ and $\mu_B$, respectively (we neglect back mutations).
+Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_A$, $1-\delta_B$, and $1+s$ respectively, with $\delta \geq 0$ (intermediate genotypes are deleterious) and $1/N < s < 1$ (double mutant $AB$ is strongly beneficial).
+For the remainder of our analysis, we will assume that $\mu_A = \mu_B = \mu$ and $\delta_A = \delta_B = \delta$.
+We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, sets the rate $v$ at which the mean fitness advances to $\sigma^{2}$.
+If $x$ is the background fitness of a lineage, then the distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
+Given this background fitness variance, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.
 
@@
 We begin by zeroing in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
-Once this is understood, extending our results to incorporate sequential fixation or the arisal of a complex adaptation in a single generation is straightforward.
+We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, tunneling, and deterministic fixation cases.
 
@@
 If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
-We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a ``bubble": the time integrated number of individuals $W$ is the ``weight".
-During the lifetime of a bubble, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
-Finally, the complex adaptation must sweep to fixation.
+We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a mutant ``bubble": the time integrated number of individuals $W$ is the ``weight" of the bubble.
+During the lifetime of a bubble, the complex adaptation must appear and establish; that is, it must rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu \pfix$ where $\pfix$ is the probability that the double mutant reaches fixation from a single initial individual.
+If it establishes, the complex adaptation will quickly sweep to fixation.
 The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
 In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
-We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.
+We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term
+\comment{JVC: justify this briefly}.
 
 Most of the time, a bubble will simply arise and go extinct.
-If mutation and fixation are sufficiently rare events, then a double mutant fixing can be modeled as a Poisson process: with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
-Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
-One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
-That is to say, it is not possible to simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
-Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
-As a result, $W$ will depend on $x$ and $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and $s$.
+If mutation and fixation are sufficiently rare events, then fixation of the double mutant can be modeled as a Poisson process: with probability $1-e^{-\mu \pfix W}$, the bubble with total size $W$ will give rise to a successful double mutant.
+Computing the expected value of this quantity over all bubble sizes, $\phi = \left< 1-e^{\mu \pfix W} \right>$, gives the probability that some bubble will lead to a successful valley crossing.
+One difficulty in computing the rate of tunneling in the high $N\sigma$ case where genetic draft is operating is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+That is to say, it is not possible to simply compute $\pfix$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
+Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals carry the complex adaptation, it will therefore likewise be more likely to fix.
+Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and the peak height $s$.
 These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
@@
 
-We begin by considering the dynamics of $p_{\mathrm{fix}}(x,s)$ before moving on to consider the distribution of bubble sizes.
+\comment[inline]{
+  \begin{itemize}      
+  \item  This section is still too technical.
+For example, the reader here doesn't need to know what  ``weak driver'' is, or if they do, it should be more fully explained and used more than just in what is effectively a background section.
+  \item I think having the PDE for the fixation probability is ok, but you don't need the details about the limits necessary for deriving it.
+The point of keeping it more informational for the reader.
+  \item Also, avoid introducing new variables, like $D$, and keep the existing notation, like $\pfix$, instead of changing to $p$ since the simplicity doesn't save much time (we only use the equation once) and the extra notation could be confusing.
+  \item Part of the motivation for introducing the PDE in the first place is to then say what the results of analyzing it, namely what the fixation probability is under draft with fitness background $x$.
+You explain this a little bit, but you can more fully describe how the $\pfix$ increases exponentially then linearly.
+        \item Also, I don't think you mention how the valley depth $\delta$ and peak height $s$ related to the background fitness $x$ and fitness variance $\sigma$.
+This should probably come up here, but also probably in the first paragraph of the ``Mathematical background''
+        \end{itemize}
+      }
+{\color{red}
+We begin by considering the dynamics of $\pfix(x,s)$ before moving on to consider the distribution of bubble sizes.
 We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
@@
 The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
+}
 
-In order to compute $\Phi$ we must next consider the behavior of mutant lineages.
-Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
+In order to compute the total valley crossing probability $\Phi$, the only remaining piece to compute is distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
+Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu \pfix$.
 \citet{neher_shraiman_2011} showed that $\phi$ follows the equation
@@
 but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
-Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu p_{\mathrm{fix}}(x)$), means that a simple closed form expression for $\Phi$ is not forthcoming.
-We therefore focus on simulation methods to study the dynamics of valley crossing.
+Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu \pfix(x)$), lead to difficulty in deriving a simple closed form expression for $\Phi$.
+Therefor, we focus on simulation methods to study the dynamics of valley crossing.
 
-% \iffalse
-% (Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
-% We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
-% To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
-% Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
-% Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
-% These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
-% Then the fixation probability turns out to be governed by the integro-differential equation
-% \begin{equation}
-% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
-% \end{equation}
-% where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.
-
-% We zero in on the ``infinitesimal" model, where mutations are of small effect compared to $\sigma$.
-% In that case, the fixation probability is governed by the equation
-
-% \begin{equation}
-% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
-% \end{equation}
-
-% The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
-% This should be intuitive, as lineages in a large population are doomed to extinction as the mean fitness inexorably advances, unless they an increase their fitness via recombination or mutation.
-% Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
-% \begin{equation}
-% p(X,s) =
-% \begin{cases}
-% 0 &\mathrm{~if~} X+s < 0, \\
-% X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
-% X+s &\mathrm{~if~} X+s > X_c.
-% \end{cases}
-% \end{equation}
-% (If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
-% Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
-% The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
-% Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
-% Then $X_c$ is determined by solving
-% \begin{equation}
-% 2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}.
-% \end{equation}
-
-% We now consider the behavior of single mutant bubbles.
-% Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
-% We proceed to derive an equation for the time evolution of the Laplace transform.
-
-% Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
-% We have
-% \begin{align*}
-% -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
-% & + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
-% & +kp(w,T|k-1,t,x).
-% \end{align*}
-% Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
-% Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
-% We therefore have
-% \begin{equation}
-% -\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
-% \end{equation}
-% Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
-% \begin{equation}
-% \partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
-% \end{equation}
-% where $\tilde{\phi}$ = $\phi/\sigma$.
-% We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
-% In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
-% However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
-% To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
-% Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
-% \begin{equation}
-% \partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
-% \end{equation}
-% Using matched asymptotics allows us to write
-% \begin{equation}
-% \tilde{\phi} = 
-% \begin{cases}
-% \tilde{z} &\mathrm{~if~} \theta < 0, \\
-% \theta &\mathrm{~if~} \theta > 0.
-% \end{cases}
-% \end{equation}
-% Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
-% Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
-% \fi
-
-%\begin{figure}
-\begin{figure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
-\end{subfigure}
-\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes.
-In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$.
-Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges.
-Each box represents about $100$ simulation runs.}
-\label{fig:deterministic}
-\end{figure}
-
-\section*{Methods}
+\section*{Simulation methods}
 
@@
 \section*{Results}
+
+\begin{figure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\end{subfigure}
+\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes.
+In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$.
+Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges.
+Each box represents about $100$ simulation runs.}
+\label{fig:deterministic}
+\end{figure}
 
@@
 At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
-This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
+This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
 
@@
 % One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
-% That is to say, the total rate of valley crossing will be given by $\int P(x) p_{\mathrm{fix}}(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
-% One cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
+% That is to say, the total rate of valley crossing will be given by $\int P(x) \pfix(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
+% One cannot simply compute $\pfix$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
 % This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
-% We proceed to derive both $p_{\mathrm{fix}}(x,s)$ and $P(w)$ in the zero recombination limit.
+% We proceed to derive both $\pfix(x,s)$ and $P(w)$ in the zero recombination limit.
 

### 2018-06-05 ms/valley_crossing.tex
edited through simulation methods

@@
 
-\newcommand{\comment}[2][noinline]{\todo[color=red,#1]{#2}}
+\newcommand{\jcomment}[2][noinline]{\todo[size=\tiny,author=JVC,color=red,#1,caption={},fancyline]{#2}}
 
@@
 \begin{abstract}
-
-Living systems are characterized by complex adaptations, at least some of which have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
-Such adaptations have been termed ``irreducibly complex", and the process by which they evolve is known as ``crossing a fitness valley".
-Previous efforts have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
-However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate of individual mutations.
-We investigate the rate at which irreducible complexity evolves in these rapidly adapting populations, focusing on asexual populations.
-We confirm that rapid adaptation overall increases the time required to cross a valley but can make it easier for deeper valleys to be crossed.
+Living systems are characterized by complex adaptations where multiple mutations are required for the function of the adaptation.
+Empirical studies of the fitness landscape that results from the many possible mutations in a gene region reveal multiple possible fitness peaks and valleys that connect them.
+Thus, it is possible that some complex adaptations have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
+Traversing these evolutionary paths when intermediates are deleterious is known as ``crossing a fitness valley".
+Previous efforts at studying this problem have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
+However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate of individual mutations due to the action of genetic draft.
+We investigate the rate at which complex adaptations evolve in these rapidly adapting populations in regions without recombination.
+We confirm that rapid adaptation overall increases the time required to cross a valley but can make it easier for deeper valleys to be crossed relative to the time required for single beneficial mutations to sweep to fixation.
 \end{abstract}
@@
 Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
-This is the sequential fixation regime.
+This is the sequential fixation regime \citep{weissman_2009}.
 A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
@@
 We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term
-\comment{JVC: justify this briefly}.
+\jcomment{justify this briefly}.
 
@@
 
-\comment[inline]{
+\jcomment[inline]{
   \begin{itemize}      
@@
 The mean offspring number is further adjusted to keep the population roughly at a pre-specified carrying capacity.
-We initialize a wild type population of $N$ individuals consisting of haploid genomes of length $L = 200$.
-Genetic loci are partitioned into two groups, the two ``focal" loci (at which the epistatic alleles will segregate) and ``background" loci (which are responsible for the underlying fitness distribution).
-For the background loci, we make use of a modified ``infinite sites" model: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
+We initialize a wild type population of $N$ individuals consisting of haploid genomes of with $L = 200$ loci.
+Genetic loci are partitioned into two groups, the two ``focal" loci where the epistatic alleles $A$ and $BB$ will segregate and the remaining ``background" loci that are responsible for the underlying fitness distribution.
+We set the focal loci to be at positions $L/4$ and $3L/4$.
+For the background loci, we make use of a modified infinite sites model \citep{Kimura:1971,Watterson:1975}: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
 In this way, the population experiences a constant influx of beneficial mutations at the background loci.
 These mutations' fitness effects are drawn from an exponential distribution; the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation (they are multipled by the current variance and divided by $\sigma^2$).
-In this way, the background mutation rate $U$ and average fitness effect $\bar{\epsilon}$ are not parameters of our simulation model but must be directly measured from simulations: they are constrained by $L$ and $\sigma$.
+In this way, the mutation rate $U$ \jcomment{Is this $U_{B}$ or $U$?} and average fitness effect $\bar{\epsilon}$ at the background loci \jcomment{this should be defined clearly either above in this section or in the math background section (it appears in that section but is not explicitly defined)} are not parameters of our simulation model but must be directly measured from simulations; i.e., they are constrained by $L$ and $\sigma$.
 We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.
 
-We set the ``focal" loci (the loci at which the epistatic alleles segregate) to be at positions $L/4$ and $3L/4$.
-Simulations are allowed to proceed for an ``equilibration" time of $N/10$ generations to allow sufficient genetic diversity to be introduced.
+Simulations are allowed to proceed for an equilibration time of $N/10$ generations to allow sufficient genetic diversity to be introduced.
 During this time, the mutation rate $\mu$ for the focal loci is set to zero.
-After equilibration, we set a per-site mutation rate $\mu$ for the focal loci (which controls both forward and backward mutations), with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$, consistent with our analytical model.
+After equilibration, we set a per-site mutation rate $\mu$ for the focal loci (allowing forward and backward mutations) with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$.
 When the double mutant reaches frequency $0.5$, we consider it to have fixed.
 We then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background by drawing anew from an exponential distribution, and allow the population to equilibrate for another $N/10$ generations.
-This offers some assurance that independent valley crossing trials are indeed independent despite occurring in the same simulation run.
+This ensures that independent valley crossing trials are independent despite occurring in the same simulation run.
 
 We partitioned our simulations as follows.
-We first studied the effect of increasing $\sigma$ in what \citet{weissman_2009} describes as the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes.
-In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
-In the drift limit, sequential fixation applies when $N \ll \mathrm{min}(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
-Next, we considered the stochastic tunneling regime.
+We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes \citep{weissman_2009}.
+\jcomment{commented out some text here; put back into the results if the pattern described is evident there}
+% In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
+For small populations, sequential fixation applies when $N \ll \min(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
 Tunneling becomes relevant when $N$ is too large for sequential fixation, i.e., the tunneling probability becomes more significant than the probability that the intermediate will fix: this mandates $N \gtrsim 1/\sqrt{\mu s}$.
-It also requires that $\ll 1/\mu$, meaning that only one mutant lineage is likely to be extant at any given time and, if they do co-occur, are unlikely to interfere with each other or substantially affect the mean fitness.
-This happens to be the necessary condition for a birth-death process to be an appropriate model.
-By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, reaching mutation-selection balance at low frequency and providing many opportunities for a lucky double mutant to appear (so-called ``semi-deterministic tunneling").
-At still higher population sizes ($\gtrsim s/\mu^2$), multiple double mutants are likely to appear every generation, and their dynamics can be modeled as deterministic.
+It also requires that $\ll 1/\mu$ \jcomment{$N$ missing here?}, meaning that only one mutant lineage is likely to be extant at any given time and, if lineages do co-occur, they are unlikely to interfere with each other or substantially affect the mean fitness.
+This happens to be the necessary condition for a birth-death process to be an appropriate model.\jcomment{is this previous sentence necessary?}
+By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, reaching mutation-selection balance at low frequency and providing many opportunities for a lucky double mutant to appear; this is semi-deterministic tunneling.
+At still higher population sizes ($N \gtrsim s/\mu^2$), multiple double mutants are likely to appear every generation, and their dynamics can be modeled as deterministic.
 
@@
 We recorded the time for a beneficial mutation to arise and reach frequency $0.5$, considering it to have fixed at that point.
-We then computed the ratio of the sweep time to the valley crossing time for a comparable complex adaptation, with valley depth $\delta$ and larger fitness advantage $s_{\mathrm{valley}}$.
-To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation favors valley crossing over simple sweeps.
+We then computed the ratio of the sweep time to the valley crossing time for a complex adaptation, with valley depth $\delta$ and larger fitness advantage $s_{\mathrm{valley}}$.
+To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation increases the likelihood of valley crossing relative to the speed of simple sweeps.
 
@@
 
+\jcomment{check out some text that was from the intro that I've commented out and see if it fits here}
 %%% from into but more relevant here
@@
 
-%In the case where $\rho = 0$ and the fitness variance $\sigma^2$ is small ($N\sigma \ll 1$), the rate of valley crossing agrees with \citet{weissman_2009}: when $\rho > 0$, the rate agrees with \citet{weissman_2010}.
-%This is because, provided that $N\sigma \ll 1$, genetic drift is the major evolutionary process affecting the frequency of deleterious intermediates, and the fixation probability of the complex adaptation simply scales with $s$.
-%We turn to the regime where $N\sigma \gg 1$, meaning that draft is more important than drift, but the fixation probability of the full complex adaptation is suppressed.
-%By the central limit theorem, the background loci create a background fitness wave whose shape is roughly Gaussian.
-%We assume that the constant influx of weak effect beneficial mutations keeps the fitness variance $\sigma^2$ roughly constant.
-%We sketch some heuristics in this limit before providing a more formal analysis.
-
-% \iffalse
-% First, we consider the case where $\rho = 0$ and focus on what \citet{weissman_2009} refer to as the process of ``stochastic tunneling", where $N\mu \ll 1$.
-% As we shall see, the results in the $\rho = 0$ and low $\rho$ limits can easily be extended to sexually recombining populations: moreover, understanding the rate of tunneling will allow us to easily examine the rate at which sequential fixation and deterministic fixation occur.
-% In stochastic tunneling, a deleterious intermediate arises (at rate $N\mu$), and its lineage persists over a short time $T$, forming a transient ``bubble".
-% If the number of individuals in the lineage is given by $n(t)$, then the total number of individuals in the ``bubble" is $w = \int_{t_0}^T n(t) dt$.
-% We refer to $w$ as the ``weight" of the bubble.
-% The overall rate of valley crossing becomes $1/N\mu p$, with $p$ the probability that a particular bubble will give rise to a successful double mutant.
-% Thus, given that a deleterious intermediate appears, it has $w$ ``chances" to generate a successful complex adaptation, which it does at rate $\sim \mu s$: double mutants arise at rate $\mu$ and have a probability $\sim s$ of fixing.
-% The probability of achieving a weight of at least $w$ scales as $1/\sqrt{w}$ (or, equivalently, $P(w) \sim w^{-3/2}$), meaning that with probability $\sqrt{\mu s}$, a bubble gives rise to a successful single mutant.
-% This probability drops off rapidly at $1/\delta^2$, so for higher values of $\delta$ (so called ``deleterious tunneling"), the rate at which new complex adaptations are generated starts to fall: the size of the mutant subpopulation is bounded from above by roughly $1/\delta$.
-
-% In the high $N\sigma$ regime, the argument is much the same, but the underlying dynamics are different.
-% In stark contrast to the drift case, the fate of a complex adaptation depends strongly on the fitness $x$ of the genetic background on which it appears.
-% Individuals with fitness lower than some critical fitness $x_c$ are almost guaranteed to go extinct.
-% This means the fixation probability of the complex adaptation no longer scales with $s$, and $P(w) \sim w^{-2}$ instead of $w^{-3/2}$, meaning that the probability of reaching weight $w$ is $\sim 1/w$.
-% However, individuals with higher fitnesses can be destined for fixation, even if they are carrying a significant load of deleterious mutations.
-% This means the dependence on $\delta$ is significantly lessened: even deep fitness valleys can be crossed by a rapid enough wave.
-% (Still needed: expression for fixation probability of beneficial mutant--this should be easy enough to obtain.
-% Can start with equation 3 of \citet{neher_shraiman_2011} and drop the recombination term, which is what I'm doing now.
-% This is tantamount to a ``mean field" approximation but might still work out okay.
-% Other possible approach is to use something like the expression of  \citet{desai_fisher_2007} for the fixation probability and rescale it to a traveling wave model.
-% Once we have the fixation probability, studying sequential and deterministic fixation will be easier than studying tunneling, since sequential fixation is essentially two different fixation events.)
-
-% One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
-% That is to say, the total rate of valley crossing will be given by $\int P(x) \pfix(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
-% One cannot simply compute $\pfix$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
-% This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
-% We proceed to derive both $\pfix(x,s)$ and $P(w)$ in the zero recombination limit.
-
-% The establishment probability $p(x,s)$ obeys the PDE
-% \begin{equation}
-% v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
-% \end{equation}
-% To see this, consider the generic expression for the fixation probability
-% \begin{equation}
-% p(X,t-dt) = p(X,t)-dt(D+B(X,t))p(X,t) + dt(B(X,t)(2p(X,t)-p(X,t)^2)
-% \end{equation}
-% \citep{barton_1995}. Taking the limit as $dt \to 0$ and inserting $D = 1, B = 1 + X -\bar{X}(t)+s$ yields a PDE for $p(X,t)$.
-% Defining $x = X - vt$, we then have $\partial_t p(X-vt) = -v\partial_x p(x)$.
-
-% This is a Bernoulli equation and is readily solved by substituting $u(x,s) = 1/p(x,s)$. We then have that $\partial_x u(x,s) = -\frac{\partial_x p(x,s)}{p(x,s)^2}$, so that
-% \begin{equation}
-% \partial_x u(x,s) = -\frac{(x+s)}{v}u(x,s) - \frac{1+x+s}{v}.
-% \end{equation}
-% The use of the integrating factor $e^{\int \frac{x+s}{v} dx} = Ce^{\frac{(x+s)^2}{2v}}$ yields
-% \begin{equation}
-% Ce^{\frac{(x+s)^2}{2v}} (\partial_x u + \frac{x+s}{v}u) = -Ce^{\frac{(x+s)^2}{2v}}\frac{1+x+s}{v}.
-% \end{equation}
-% The left side is simply $\partial_x(Ce^{\frac{(x+s)^2}{2v}}u)$, so (after dividing $C$ out) we can integrate both sides, then multiply by $e^{\frac{-(x+s)^2}{2v}}$ to obtain
-% \begin{equation}
-% u(x,s) =  -e^{\frac{-(x+s)^2}{2v}} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x+s}{\sqrt{2v}}}{\sqrt{v}}) - 1 + C(e^{\frac{-(x+s)^2}{2v}}),
-% \end{equation}
-% with $C$ a constant. Of course, $p = 1/u$.
-
-% Unfortunately it is not clear how to get rid of $C$, as there is no obvious ``initial condition".
-% The most we are likely to know about $p$ is that $p$ approaches $0$ in the limit of low $x$, but behaves $\sim s$ in the limit of high $x$ before saturating at $1$ (however, there are unlikely to be any individuals in the population at such high $x$).
-% One possibility is to impose an \emph{ersatz} ``solvability condition": surely the fixation probability of a neutral mutation averaged over the entire population must equal $1/N$. We would then have
-% \begin{equation}
-% \int_{-\infty}^{\infty} P(x)p(x,0)dx = \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{e^{-x^2/2\sigma^2}}{-e^{-x^2/2\sigma^2} \frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - 1 + Ce^{-x^2/2\sigma^2}}dx = \frac{1}{N}.
-% \end{equation}
-% Multiplying the top and bottom of the integrand by $e^{x^2/2\sigma^2}$ would clean this up, yielding
-% \begin{equation}
-% \frac{1}{\sqrt{2\pi}\sigma}\int_{-\infty}^{\infty} \frac{1}{-\frac{\sqrt{\pi /2} \textrm{erfi}(\frac{x}{\sqrt{2\sigma^2}})}{\sigma} - e^{x^2/2\sigma^2} + C}dx = \frac{1}{N}.
-% \end{equation}
-% Unfortunately this integral has an imaginary error function in the denominator, so there is not likely to be a closed form expression: we could explore it numerically and see where it peaks, however.
-
-% Another idea would be to consider an alternative way of solving the differential equation for $p$.
-% Return to
-% \begin{equation}
-% v \partial_x p(x,s) = (x+s)p(x,s) - (1+x+s)p(x,s)^2.
-% \end{equation}
-% Suppose we make the \emph{ansatz} that $x+s \ll 1$, and define $\chi = x/\sigma$, $\tilde{s} = s/\sigma$, and $\tilde{p}(\chi,\tilde{s}) = p(x,s)/\sigma$. Then
-% \begin{equation}
-% \partial_\chi \tilde{p}(\chi,\tilde{s}) = (\chi + \tilde{s})\tilde{p}(\chi,\tilde{s}) - \tilde{p}(\chi,\tilde{s})^2.
-% \end{equation}
-% Now define $q(\chi,\tilde{s})$ such that $\tilde{w}(\chi,\tilde{s}) = \partial_x \log q(\chi,\tilde{s})$.
-% Then
-% \begin{equation}
-% \partial_\chi^2 q(\chi,\tilde{s}) - (\chi+\tilde{s})\partial_\chi q(\chi, \tilde{s}) = 0.
-% \end{equation}
-% Defining $\vartheta = \chi + \tilde{s}$ and $q(\vartheta) = e^{\vartheta^2/4}h(\vartheta)$ yields
-% \begin{equation}
-% \partial_\vartheta^2 h(\vartheta) - (\frac{1}{2} + \frac{\theta^2}{4})h(\vartheta) = 0,
-% \end{equation}
-% which is the parabolic cylinder equation.
-% We then need to find the form of $q$ that has the correct asymptotics.
-% This does have a solution:
-% \begin{equation}
-% q(\chi) = \int_0^\infty d\lambda e^{\vartheta \lambda - \lambda^2/2}/\lambda.
-% \end{equation}
-
-% Next, we seek the weight distribution as a function of $x$.
-% Let $p(w,T|k,t,x)$ be the probability that a bubble has attained a weight of $w$ at time $T$, given that there were $k$ individuals in the lineage at time $t$.
-% Ultimately what we will be interested in is the Laplace transform $\hat{p} (z,T|k,t,x) = \int dz e^{-zw} p(w,T|k,t,x)$.
-% Here $w$ represents (as before) the weight of a mutant bubble and evolves according to
-% \begin{align*}
-% -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}+s)p(w,T|k,t,x) \\
-% & + k(1+x-\bar{x}+s)p(w,T|k+1,t,x) \\
-% & +kp(w,T|k-1,t,x). \\
-% \end{align*}
-% We have that $\hat{p} (z,T|k,t,x) = \hat{p}^k (z,T|1,t,x)$, by the assumption that the behavior of each of the $k$ copies of the mutant allele is independent of the other.
-% Recall that, in the traveling wave approximation, $\bar{x} = vt = \sigma^2 t$.
-% The notation here can be cleaned up by defining $\phi = 1 - \hat{p}$, as well as $\tilde{s} = s/\sigma$, $\chi = x/\sigma - \sigma T$, $\tau = \sigma(T-t)$, and $\theta = \chi+\tau+\tilde{s}+z$.
-% Rewriting the preceding in terms of $\phi$ (note that differentiating $p$ with respect to $w$ simply pulls out a factor of $z$) yields
-% \begin{equation}
-% \partial_\tau \phi(\tau,z,\chi) = z + \theta \phi(\tau,z,\chi) - \phi^2(\tau,z,\chi).
-% \end{equation}
-% Unfortunately $\theta$ depends on $\tau$, so we will have to be careful here.
-% This is a Riccati equation.
-% Let $\phi_0(\tau,z,\chi)$ be a particular solution for $\phi$: then the general solution becomes
-% \begin{equation}
-% \phi(\tau,z,\chi) = \phi_0(\tau,z,\chi) + \frac{\Phi(\tau,z,\chi)}{C + \int \Phi(\tau,z,\chi) d\tau},
-% \end{equation}
-% where $\Phi(\tau,z,\chi) = e^{\int (-2\phi_0 + \theta ) d\tau}$.
-% It remains to find a particular solution of $\phi$.
-% We can ``guess" a steady state solution, in which case $0 = \phi^2 - \theta \phi - z = 0$: this corresponds to $\phi_0 = \frac{\theta \pm \sqrt{\theta^2 + 4z}}{2}$.
-% Computing the exponent in the integral (and changing integration variables from $\tau$ to $\theta$) yields
-% \begin{equation}
-% \Phi (\tau, z, \chi) = e^{-\int \sqrt{\theta^2 + 4z} d\theta} = e^{\theta\sqrt{\theta^2+4z} + \frac{4z}{2}\log (\theta + \sqrt{\theta^2 + 4z})}.
-% \end{equation}
-% (Still need to solve this and carefully consider the integral boundary.)
-
-% We now turn to the case where $\rho > 0$.
-% Because we are explicitly dealing with crossovers, the results of \citet{neher_kessinger_2013} are useful here.
-% We consider the genome as divided into a series of effectively asexual ``blocks" in which the recombination rate is very low.
-% The size of each block is determined by a balance between recombination, which chops up the blocks, and the rate of adaptation, which amplifies them.
-% The size of such a block is not a parameter of our model but is given by $\xi_b = \frac{\sigma^2}{2L\rho^2 c \log N\sigma_b}$, with $L$ the number of loci, $c$ a constant of order $1$, and $\sigma_b$ the proportion of the fitness variance segregating within the block.
-% If $\xi_b$ is large enough that both focal loci fall within the same ``block", then an analysis similar to that of Neher and Shraiman (2009) may be appropriate: the two loci effectively segregate within the same traveling wave.
-% If not, then the value of $N\sigma_b$ becomes the relevant factor.
-% $N\sigma_b \gg 1$ implies that each locus effectively operates within its own traveling wave: in order for the full complex adaptation to arise, establish, and fix, both must be at high enough frequency at the same time for either mutation or recombination to produce a double mutant.
-% On the other hand, $N\sigma_b \ll 1$ implies that each locus effectively evolves on a neutral genetic background neutrally, and we recover the dynamics of genetic drift.
-% \fi
-
 \bibliographystyle{genetics}

### 2018-06-08 ms/valley_crossing.tex
done editing results.

@@
 
-Here we present the key results of our simulation runs.
-In general, we expect valley crossing in rapidly adapting populations to require more time, as bubbles are doomed to extinction more quickly and beneficial lineages are less likely to fix, especially when valleys are shallow.
+Here we present the key results of our simulation analysis.
+In general, we expect valley crossing to require more time in rapidly adapting populations (large $N \sigma$) experiencing genetic draft as bubbles are doomed to extinction more quickly \jcomment[inline]{I was going to add something like, ``(bubbles of intermediate size are less likely to occur under draft)'' to explain this, but is it more general (since you bring it up only under tunneling below)?
+Maybe add a short parenthetical explanation?} and beneficial lineages are less likely to fix, especially when valleys are shallow.
 Figures \ref{fig:deterministic} and \ref{fig:tunneling} confirm these expectations: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
 It is tempting to interpret this as a reduced $N_e$ due to a decrease in genetic diversity and the strength of selection.
-The underlying dynamics are more complicated, however.
-As $\sigma$ increases, the effect of the genetic background becomes more significant in determining the fate of an allele than the allele's fitness effect is.
-The frequency of alleles is buffeted by genetic draft, so alleles behave as though they are more neutral than they really are.
-This depresses the fixation probability of the complex adaptation, which appears to account for the majority of the change in fixation time.
-In the case of stochastic tunneling, the overall size distribution of bubble sizes is reduced, as well: the distribution $P(w)$ scales not as $w^{-3/2}$ but as $w^{-2}$ \citep{neher_shraiman_2011}, so there are fewer chances for the complex adaptation to arise and fix.
+The underlying dynamics are more complicated, however.\jcomment{Yes, but the following sentences don't highlight specifically enough I think why; e.g., what about a simple reduction in $N_{e}$ cannot explain what happens in this draft case?}
+As $\sigma$ increases, the effect of the genetic background in which an mutation finds itself becomes more significant in determining the fate of that mutation than the fitness effect of the mutation itself.
+This genetic draft buffets the frequency of mutations so that these mutations behave as though they are more neutral than they really are.
+This depresses the fixation probability of the double mutant and other complex adaptations, which appears to account for the majority of the change in fixation time.
+In the case of stochastic tunneling, the overall size distribution of bubble sizes is reduced as well: the distribution $P(w)$ scales not as $w^{-3/2}$ but as $w^{-2}$ \citep{neher_shraiman_2011}, so there are fewer chances for the complex adaptation to arise on a mutant background and fix.
 
@@
 
-On the other hand, the reduced effectiveness of selection should generally be helpful in crossing deeper valleys, by mitigating the deleterious effect of single mutants.
+On the other hand, the reduced effectiveness of selection should generally be helpful in crossing deeper valleys by mitigating the deleterious effect of single mutants.
 This can be seen in figure \ref{fig:sigma_delta}.
-At high values of $\sigma$, the fitness effect of the deleterious intermediate \emph{almost does not matter}, meaning that very deep valleys can be crossed.
-This supports the view that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations.
-
-% \iffalse
-% \begin{figure}
-% \includegraphics[width=0.4\textwidth]{Figures/bands.pdf}
-% \caption{Valley crossing times for fixed $N\sigma$, with $N$ allowed to vary up or down by a factor of $2$ (and $\sigma$ adjusted accordingly).
-Error bars are quartile ranges.
-Here, $\mu = 3 \times 10^{-6}$, $\delta = 10^{-3}$, and $s = 10^{-2}$.}
-% \label{fig:bands}
-% \end{figure}
-
-% In addition, we consider the effects of the compound parameter $N\sigma$, which has previously been shown to determine the coalescent behavior of populations \citep{neher_hallatschek_2013, neher_kessinger_2013}.
-% Figure \ref{fig:bands} suggests that, at high values of $\sigma$, the value of $N\sigma$ is more important than $N$ itself in determining the valley crossing rate.
-% The dependence on $N$ appears to be reduced, consistent with general features of very large, rapidly adapting populations.
-% In a slowly adapting population, $N$ determines the transition from sequential fixation to tunneling: if bubbles reach a maximum size that is of order $N$, they start to noticeably decrease the mean fitness, making fixation easier.
-% When $N\sigma \gg 1$, bubbles almost never reach a large enough size for this condition to obtain.
-% \fi
+At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect of the valley crossing time, which means that very deep valleys can be crossed.
+This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations.
 
 \begin{figure}
-\begin{subfigure}[b]{0.43\textwidth}
-\includegraphics[width=\textwidth]{Figures/tau_compare.pdf}
+\begin{subfigure}[b]{0.475\textwidth}
+  \includegraphics[width=\textwidth]{Figures/tau_compare_crop.pdf}
+  \caption{Valley crossing times on a $\log_{10}$ scale}
+  \label{fig:tau}
 \end{subfigure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/alpha_ratios.pdf}
+\begin{subfigure}[b]{0.475\textwidth}
+  \includegraphics[width=\textwidth]{Figures/alpha_ratios_crop.pdf}
+  \caption{Ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ on a $\log_{10}$ scale}
+  \label{fig:ratios}
 \end{subfigure}
@@
 In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
-Left two plots: valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
-Right two plots: the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.
-Deeper valleys in smaller populations are disfavored relative to simple sweeps.
-But at higher values of $\sigma$, they can be favored ($\alpha < 1$: note the log scaled colorbar).
-In both plots, $N = 10^5$ is a fairly hard cutoff beyond which tunneling becomes semi-deterministic, as $N\mu > 1$ means that, on average, more than one mutant lineage is seeded every generation.}
-\label{fig:ratios}
+Panel (a) shows valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
+Panel (b) shows the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.}
+\label{fig:tau-ratios}
 \end{figure}
@@
 
-Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+So far, we can see that valley crossing is slower in rapidly adapting populations but the crossing time is less sensitive to the depth of the valley.
+How do these two effects combine to shape the likelihood of valley crossing in rapidly adapting populations experiencing genetic draft compared to populations only experiencing drift?
+To explore this, we first introduce the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
 By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps.
 Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
-Though the precise value of this ratio depends on $\delta$ and the selection coefficients $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend nonetheless emerges.
-In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
-For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
-In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
-Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
-In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
-\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
-At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
-This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.
+Though the precise value of this ratio depends on $\delta$ and on the selection coefficients $s_{\mathrm{valley}}$ and $s_{\mathrm{sweep}}$ of the complex adaptation and the single beneficial mutation, respectively, a general trend nonetheless emerges.
+In Figure \ref{fig:tau-ratios}, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
+Figure \ref{fig:ratios} shows the expected pattern that deeper valleys in smaller populations take longer to cross.
+At a higher value of $\sigma$, valley depth has a weaker effect of the valley crossing time in both an absolute sense (Figure \ref{fig:tau}) and relative to the time for a single beneficial sweep (Figure \ref{fig:ratios}).
+Moreover, values of $\alpha$ are generally lower for the population with higher $\sigma$ experiencing genetic draft than for the population with lower $\sigma$ experiencing only drift.
+This implies that draft tends reduce the valley crossing time relative to the time a single beneficial mutation to sweep.
+
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by plotting the ratio $\alpha(10^{-2})/\alpha(10^{-6})$ in Figure \ref{fig:alpha_ratio}, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored for $N < 10^{5}$.
+In this regime, increasing $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
+For $N > 10^5$ in the low $\sigma$ case, tunneling becomes semi-deterministic where $N\mu > 1$ and, on average, more than one mutant lineage is generated every generation.\jcomment{Another figure presumably from simulations that you're running...}
+These mutants then give rise to double mutants at a low but constant rate that is not sensitive to the valley depth (Figure \ref{fig:tau-ratios} and equation 35 in \citealp{weissman_2009}) and that is much faster than in the stochastic tunneling case.
+In contrast, $N > 10^{5}$ for high $\sigma$ does not produce the same order of magnitude increase in valley crossing rates, which is likely due to no analogous qualitative change in valley crossing dynamics.
+Thus, for large enough population sizes, draft does not significantly improve the valley crossing rate relative to the rate of single beneficial sweeps.\jcomment{Double check my explanation here :-)}
+
+At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{Cvijovic:Good:2018}, but the average behavior is more or less the same.
+This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.\jcomment{I still don't get what the point of mentioning this is here; how does this connection to the behavior of the ratio of $\alpha$s?}
 

### 2018-06-12 ms/valley_crossing.tex
acknowledgments
edited discussion.

@@
 This suggests that tunneling may be the most difficult mode of valley crossing.
-% Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
-% Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
-% High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.
+Recombination also affects the rate of valley crossing where infrequent recombination decreases valley crossing times \citep{weissman_2010}.
+At high values of the recombination rate, however, double mutant individuals outcross with the wild type, and produce more deleterious intermediates that retard the growth of the double mutant.
+An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
 
@@
 \end{subfigure}
-\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes.
+\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes as a function of $\sigma$.
 In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$.
@@
 \end{subfigure}
-\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes.
+\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes as a function of $\sigma$.
 In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^4$ and $\mu = 3\times 10^{-5}$, and on the right, $N = 10^5$ and $\mu = 3\times 10^{-6}$.
@@
 
-We have used simulation methods to examine the rate at which rapidly adapting populations cross fitness valleys.
-Though the overall time required is generally higher than in slowly adapting (drifting) populations, there are two important caveats.
-First, the effect of the deleterious intermediate is much smaller: for $\delta \lesssim \sigma$, the precise value of $\delta$ almost does not affect the crossing rate.
-Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is likely to be much longer in rapidly adapting populations.
+We have used simulation methods to examine the rate of crossing fitness valleys in rapidly adapting populations that experience genetic draft.
+We find that although the overall valley crossing time is longer in rapidly adapting than in slowly adapting (drifting) populations, a higher fraction of the adaptation that occurs rapidly adapting populations will occur through valley crossing than in slowly adapting populations.
+This mean that the genetic draft that results from rapid adaptation accelerates the rate at which populations escape from local peaks in a fitness landscape and move towards global peaks.
+Thus, adaptation at loci unhindered by genetic draft occurs faster but is more likely to get stuck on a local peak in a rugged fitness landscape.
+
+The source of this difference in behavior between populations experiencing genetic draft and those experiencing drift comes from two effects.
+First, the effect of the deleterious intermediate in the rapidly adapting population is much smaller: for $\delta \lesssim \sigma$, the precise value of $\delta$ almost does not affect the crossing rate.
+Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is much longer in rapidly adapting populations.
 This means that the presence of a fitness valley is not as steep of an impediment to the evolution of a complex adaptation for a drafting population as it is for a drifting one.
-This confirms the intuition that traversing a rugged fitness landscape by valley crossing may be favored in rapidly adapting populations.
-These results compare favorably with those of \citet{ochs_2015}, which studied a similar problem in a slowly evolving population.
-They considered the scenario where a sweeping beneficial mutation \emph{competes} with a particular complex (valley crossing) adaptation, whereas we treat these two scenarios separately, but the trend is similar: in their model, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult especially for populations of intermediate size.
+
+Our results complement and build on those of \citet{ochs_2015} who compared valley crossing and a sweeping beneficial mutation in a slowly evolving population.
+They considered the scenario where the sweeping beneficial mutation \emph{competes} with a particular complex adaptation, whereas we treat these two scenarios separately.
+They find that large populations are more likely to cross valley than fix a simple beneficial mutation than small populations (for comparison, see the $\sigma=5 \times 10^{-6}$ panel in our Figure \ref{fig:ratios}).
+However, they also find that intermediate-sized populations in which stochastic tunneling occurs are the least likely to cross a fitness valley instead of fixing a beneficial mutation.
+In contrast, our results suggest that the stochastic tunneling regime is where rapid adaptation and genetic draft increase the likelihood of valley crossing relative to a simple beneficial mutation.
+Thus, it is possible that genetic draft improves the likelihood of valley crossing precisely where its least likely under genetic drift. 
+% in their model, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult especially for populations of intermediate size.
+\jcomment[inline]{I removed an argument had here about mean fitness that I wasn't sure of; how does the sweeping beneficial mutation increasing mean fitness affect intermediate population sizes more than large or small?}
 
@@
 If both $N\sigma$ and $N\sigma_b$ are large, and the focal loci segregate in the same block, then the analysis presented here should determine the crossing time.
-If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the distance between the focal loci is much larger than the block length $\xi_b$, then recombination between the focal loci can be modeled as ``free", with individuals effectively shuffling their entire genomes every generation.
+If, on the other hand, $N\sigma$ is large but $N\sigma_b$ is small, or the map distance between the focal loci is much larger than the block length, then recombination between the focal loci can be modeled as ``free", with individuals effectively shuffling their entire genomes every generation.
 In that case, the communal recombination analysis of \citet{neher_shraiman_2011} and \citet{neher_shraiman_2010} is likely to be important: the relationship between $\sigma$ and the recombination rate $\rho$ determines the crossing rate.
@@
 
-\jcomment{check out some text that was from the intro that I've commented out and see if it fits here}
-%%% from into but more relevant here
-% In sexual populations, the accuracy of this description depends on the recombination model.
-% When recombination is ``free", meaning that child genomes are assembled by randomly shuffling those of their parents, there is a chance for new alleles on weak fitness backgrounds to jump onto fitter ones near the nose of the distribution \citep{neher_shraiman_2011}.
-% Competition between recombination, which breaks up linkage, and adaptation, which creates it, determines whether draft or drift is more important.
-% When recombination occurs via crossovers, genomes can be partitioned into effectively asexual ``blocks", whose size depends on the per-site crossover rate and the total fitness variation segregating in the population.
-% If each block harbors some portion $\sigma_b$ of the total fitness standard deviation $\sigma$, then the conditions $N\sigma_b \gg 1$ and $N\sigma_b \ll 1$ correspond to draft and drift dominance, respectively \citep{neher_kessinger_2013}.
+One common feature of the effect of genetic draft and population subdivision on valley crossing is that they both increase the appearance of complex adaptations by reducing selection against deleterious intermediates (i.e., the valley genotypes).
+In contrast, high migration rates and high recombination rates \citep{neher_shraiman_2009} allow selection to more easily remove deleterious intermediates.
+This commonality has implications for Wright's ``shifting balance theory" \citep{Wright:1932}, which posits that population subdivision can enhance evolution on rugged fitness landscapes: deleterious intermediates accumulate due to genetic drift induced by subdivision, and subsequent beneficial mutations fix both in locally and are spread globally due to migration.
+However, it has long been known that the shifting balance theory has important limitations due to the fact that too much migration prevents the accumulation of deleterious intermediates and too little reduces the success of genotypes from the fitter peaks and prevents them from spreading throughout the metapopulation \citep{coyne_barton_turelli_2000,Van-Cleve:Weissman:2015}.
+Recent work by \citet{Bitbol:Schwab:2014} confirms the fact that intermediate migration rates yield the fastest valley crossing times; however, it also finds that valley crossing is faster in subdivided population with these intermediate migration rates than in panmictic populations.
 
-The dynamics by which rugged fitness landscapes are traversed has received significant attention in the literature in recent years.
-Valley crossing has previously been shown to be favored in small populations, where the effects of selection against intermediates are mitigated, and in large populations, where multiple mutants are much more likely to appear \citep{weissman_2009, ochs_2015}.
-Low amounts of recombination also appear to be helpful \citep{weissman_2010}: at high values of the recombination rate, double mutant individuals outcross with the wild type, producing more deleterious intermediates and retarding the growth of the double mutant.
-An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
-Like population subdivision, genetic draft favors the appearance of complex adaptations by forcing selection to act on entire genotypes, which may be carrying useful but provisionally harmful deleterious intermediate mutations.
-This can be contrasted with effects such as high migration rates and high recombination rates \citep{neher_shraiman_2009}, which force selection to act on individual alleles rather than genotypes.
+In light of our results that show that genetic draft can improve the likelihood of valley crossing, populations which experience both subdivision \emph{and} rapid adaptation are ideal candidates in which the shifting balance theory may apply.
+Such populations include HIV, in which host-specific adaptations evolve quickly \citep{zhang_1997, wain_2007, dapp_2017, theys_2018} and in which deleterious mutations are known to hitchhike to high frequency \citep{zanini_2013, zanini_2015}.
+Our work also shines light on the likely pathway through which multi-drug resistance evolves in HIV.
+In recent years of the HIV pandemic, resistance has generally been slower to evolve \citep{feder_2015} because more and stronger drugs are used.
+There are two possible explanations for this.
+One is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
+The other is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.
+Both factors are likely to play a role.
+However, since HIV is a population in which draft, not drift, is the dominant stochastic force and hence population size is less important in determining the evolutionary dynamics, our work suggests that the second factor is likely to be more significant in HIV compared to organisms in which draft is less important.
 
-Two further intuitions can be drawn from this argument.
-First, there is a sort of duality between the evolution of complex adaptations and the evolution of cooperation, which can be seen as a complex behavioral adaptation.
-In social evolution theory, high migration rates between demes, which cause organisms to be less likely to interact with close kin, disfavor the evolution of cooperation: likewise, when migration rates are low, cooperation can be favored \citep{van_cleve_2015}.
+Our results for the dynamics of valley crossing under genetic draft are likely to be important for other evolutionary problems in which stochastic forces are known to be important.
+For example, there is a sort of duality between the evolution of complex adaptations and the evolution of cooperation, which can be seen as a complex behavioral adaptation where the highest per capita fitness at equilibrium requires the combination of multiple cooperative individuals (i.e., the combination of genes among different individuals instead among different loci).
+In social evolution theory, high migration rates between demes, which cause organisms to be less likely to interact with close kin, disfavor the evolution of cooperation: likewise, when migration rates are low, cooperation can be favored \citep{Hamilton:1970,Rousset:2004,van_cleve_2015}.
 Evolutionary forces which strengthen existing associations between loci are more likely to lead to such complex phenotypes: evolutionary forces which weaken these associations hinder their evolution.
@@
 
-Second, the relationship between genetic draft and valley crossing has implications for Wright's ``shifting balance theory".
-Population subdivision followed by colonization of a new fitness peak may be an important aspect of the evolutionary process, but it is not obvious \emph{how} important \citep{coyne_barton_turelli_2000}.
-However, populations which experience both subdivision \emph{and} rapid adaptation are ideal candidates in which the shifting balance theory may apply.
-Such populations include HIV, in which host-specific adaptations evolve quickly \citep{zhang_1997, wain_2007, dapp_2017, theys_2018} and in which deleterious mutations are known to hitchhike to high frequency \citep{zanini_2013, zanini_2015}.
-Our work also shines light on the likely pathway through which multi-drug resistance evolves in HIV.
-In recent years of the HIV pandemic, resistance has generally been slower to evolve \citep{feder_2015}, in general because more and stronger drugs are used.
-There are two possible explanations for this.
-One is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
-The other is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.
-Both factors are likely to play a role.
-However, since HIV is a population in which draft, not drift, is the dominant factor and hence population size is less important in determining the evolutionary dynamics, our work suggests that the second is likely to be more significant compared to other populations.
+\section*{Acknowledgments}
+
+This work is supported by the National Science Foundation under Cooperative Agreement No. 1355438.
 

### 2018-06-14 ms/valley_crossing.tex
further edits and simplification

@@
 \begin{abstract}
-Living systems are characterized by complex adaptations where multiple mutations are required for the function of the adaptation.
-Empirical studies of the fitness landscape that results from the many possible mutations in a gene region reveal multiple possible fitness peaks and valleys that connect them.
+Living systems are characterized by complex adaptations which require multiple coordinated mutations in order to function.
+Empirical studies of fitness landscapes that result from the many possible mutations in a gene region reveal many fitness peaks and valleys that connect them.
 Thus, it is possible that some complex adaptations have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
-Traversing these evolutionary paths when intermediates are deleterious is known as ``crossing a fitness valley".
+When intermediates are deleterious, traversing such an evolutionary path is known as ``crossing a fitness valley".
 Previous efforts at studying this problem have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
-However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate of individual mutations due to the action of genetic draft.
+However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate and dynamics of individual mutations due to the action of genetic draft.
 We investigate the rate at which complex adaptations evolve in these rapidly adapting populations in regions without recombination.
-We confirm that rapid adaptation overall increases the time required to cross a valley but can make it easier for deeper valleys to be crossed relative to the time required for single beneficial mutations to sweep to fixation.
+We confirm that rapid adaptation overall increases the time required to cross a valley; however, rapid adaptation can make it easier for deeper valleys to be crossed relative to the time required for single beneficial mutations to sweep to fixation.
 \end{abstract}
@@
 This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
-Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype where each mutant genotype in the sequence has a higher fitness than the previous one.
+Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
 If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
@@
 In such landscapes, wild type individuals may have to traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
+This is a special case of ``sign epistasis".
 We seek to characterize this ``valley crossing" process in order to better understand the routes adaptation is likely to take on these rugged empirical landscapes \citep[e.g.,][]{Aguilar-Rodriguez:Payne:2017} and to assess the likelihood of valley crossing under more complex scenarios in which it may in fact be common \citep[e.g.][]{trotter_2014}.
@@
 Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
-If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
-This is sometimes called deterministic fixation of the double mutant \citep{weissman_2009}.
+If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
+This is sometimes called deterministic fixation of the multiple mutant\citep{weissman_2009}.
 In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation.
-Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
+Further mutations necessary for the adaptation can then arise on intermediate mutant backgrounds and fix due to positive selection.
 This is the sequential fixation regime \citep{weissman_2009}.
@@
 This suggests that tunneling may be the most difficult mode of valley crossing.
-Recombination also affects the rate of valley crossing where infrequent recombination decreases valley crossing times \citep{weissman_2010}.
-At high values of the recombination rate, however, double mutant individuals outcross with the wild type, and produce more deleterious intermediates that retard the growth of the double mutant.
+Recombination also affects the rate of valley crossing.
+Infrequent recombination decreases valley crossing times \citep{weissman_2010}.
+At high values of the recombination rate, however, individuals carrying the full complex adaptation outcross with the wild type and produce more deleterious intermediates that retard the growth of the multiple mutant.
 An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
@@
 This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
-In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare, close to frequency zero, or common, close to frequency one.
+In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare (close to frequency zero) or common (close to frequency one).
 Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982} where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
@@
 Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
-Draft is fundamentally different from drift: for example, drafting populations experience large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}
-and have genealogies where more than two branches can merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
+Draft is fundamentally different from drift: for example, drafting populations experience large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}.
+They commonly have genealogies in which more than two branches merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
 Additionally under genetic draft, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency one--or, equivalently, a depression in intermediate frequency alleles, which are quickly swept out of the population \citep{neher_shraiman_2011,kosheleva_2013,neher_hallatschek_2013}.
@@
 Draft likewise may have profound effects on the evolution of complex adaptations. Preliminary evidence for this comes from
-\citet{neher_shraiman_2011} who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting population.
-They found that, compared to drift, draft generally makes small and intermediate sized mutant bubbles much rarer but very large bubbles are much more common.
-Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately clear.
-Here, we extend these previous results by studying in asexual populations how genetic draft affects the rate of crossing fitness valleys across a range of population sizes.
+\citet{neher_shraiman_2011}, who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting population.
+They found that, compared to drift, draft generally causes the distribution of bubble sizes to drop off more quickly, meaning that small bubbles are more common but large bubbles are very rare.
+Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately obvious.
+Here, we extend these previous results by studying how genetic draft in asexual populations affects the rate of crossing fitness valleys across a range of population sizes.
 
 The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic.
-In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011}.
-We therefore must focus primarily on simulation approaches as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
+In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011, neher_shraiman_2009}.
+We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
 We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
 However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
-In addition, we confirm that in rapidly adapting populations, adaptive fixation of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
+In addition, we confirm that in rapidly adapting populations, adaptive fixations of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
 These observations add to the growing intuition that complex adaptations that involve evolution across fitness valleys may not be a surprising result of the evolutionary process, but rather an expected one.
@@
 For the remainder of our analysis, we will assume that $\mu_A = \mu_B = \mu$ and $\delta_A = \delta_B = \delta$.
-We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, sets the rate $v$ at which the mean fitness advances to $\sigma^{2}$.
+We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$.
+By Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, the rate $v$ at which the mean fitness advances due to selection is set to $\sigma^{2}$.
 If $x$ is the background fitness of a lineage, then the distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
 Given this background fitness variance, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.
+We emphasize that $\sigma^2$ is only a \emph{background} fitness variance and does not include the effects of the focal loci.
+For the most part, we assume that the focal loci are either at low enough frequency that they do not substantially affect the total fitness variance or that they are on the way to fixation and the population is destined to become roughly monomorphic (until new beneficial background mutations can be introduced).
 
@@
 At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
-We begin by zeroing in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
+We zero in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
 We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, tunneling, and deterministic fixation cases.
@@
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
-If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
-We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a mutant ``bubble": the time integrated number of individuals $W$ is the ``weight" of the bubble.
+If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = \int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
+As in the introduction, we refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a mutant ``bubble": the time integrated number of individuals $W$ is the ``weight" of the bubble.
 During the lifetime of a bubble, the complex adaptation must appear and establish; that is, it must rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu \pfix$ where $\pfix$ is the probability that the double mutant reaches fixation from a single initial individual.
@@
 In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
-We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term
-\jcomment{justify this briefly}.
+We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term: the time required for $T_1$ to establish will be of order $1/s$ \citep{desai_fisher_2007}, and the sweep time will be of order $1/s \log (Ns)$, but $T_0$ can be many orders of magnitude greater.
 
@@
 Computing the expected value of this quantity over all bubble sizes, $\phi = \left< 1-e^{\mu \pfix W} \right>$, gives the probability that some bubble will lead to a successful valley crossing.
-One difficulty in computing the rate of tunneling in the high $N\sigma$ case where genetic draft is operating is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+One difficulty in computing the rate of tunneling in the high $N\sigma$ case (where genetic draft is significant) is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
 That is to say, it is not possible to simply compute $\pfix$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
-Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals carry the complex adaptation, it will therefore likewise be more likely to fix.
+Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals give rise to the complex adaptation, it will therefore likewise be more likely to fix.
 Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and the peak height $s$.
@@
 
-\jcomment[inline]{
-  \begin{itemize}      
-  \item  This section is still too technical.
-For example, the reader here doesn't need to know what  ``weak driver'' is, or if they do, it should be more fully explained and used more than just in what is effectively a background section.
-  \item I think having the PDE for the fixation probability is ok, but you don't need the details about the limits necessary for deriving it.
-The point of keeping it more informational for the reader.
-  \item Also, avoid introducing new variables, like $D$, and keep the existing notation, like $\pfix$, instead of changing to $p$ since the simplicity doesn't save much time (we only use the equation once) and the extra notation could be confusing.
-  \item Part of the motivation for introducing the PDE in the first place is to then say what the results of analyzing it, namely what the fixation probability is under draft with fitness background $x$.
-You explain this a little bit, but you can more fully describe how the $\pfix$ increases exponentially then linearly.
-  \item Also, I don't think you mention how the valley depth $\delta$ and peak height $s$ related to the background fitness $x$ and fitness variance $\sigma$.
-This should probably come up here, but also probably in the first paragraph of the ``Mathematical background''
-  \end{itemize}
-}
-{\color{red}
 We begin by considering the dynamics of $\pfix(x,s)$ before moving on to consider the distribution of bubble sizes.
-We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
-In this case, the background beneficial mutation rate $U_b$ and mean fitness effect $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
-Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
+We focus on the case where background mutations are frequent but of weak effect--that is, the mean effect of background mutations $\bar{\epsilon}$ is small compared to $\sigma$, but background mutations occur at a high rate $U_b$.
+\iffalse
+In this case, the exact values of $U_b$ and $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
+\fi
 \citet{hallatschek_2011} and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
 \begin{equation}
-v\partial_x p(x,s) = x p(x,s) + D\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
+v\partial_x p(x,s) = x p(x,s) + \frac{U_b\bar{\epsilon}^2}{2}\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
 \end{equation}
-in which we have assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
-The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
-}
+in which we have further assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
+For a neutral mutation ($s = 0$), the fixation probability is effectively zero at low $x$ but increases exponentially up until $x_c$, at which it becomes roughly linear in $x$.
+In this regime, $x_c \approx 2\sigma^4/U_b\bar{\epsilon}^2$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
+The effect of introducing a nonzero $s$ is essentially to shift the value of $x$ and $x_c$: for large negative $s$, $x_c$ may be far enough in the nose that it is unlikely any individuals will actually be present, and for large positive $s$, even unfit individuals can give rise to a lineage that is destined to fix.
 
-In order to compute the total valley crossing probability $\Phi$, the only remaining piece to compute is distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
-Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu \pfix$.
+In order to obtain the total valley crossing probability $\Phi$, the only remaining element to compute is the distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
+Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu \pfix W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu \pfix$.
 \citet{neher_shraiman_2011} showed that $\phi$ follows the equation
@@
 \end{equation}
-but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
+but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation.
+They also avoided the need to explicitly send time to infinity, which is necessary to incorporate all possible bubble sizes and lifetimes.
 Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu \pfix(x)$), lead to difficulty in deriving a simple closed form expression for $\Phi$.
-Therefor, we focus on simulation methods to study the dynamics of valley crossing.
+Therefore, we focus on simulation methods to study the dynamics of valley crossing.
 
@@
 We initialize a wild type population of $N$ individuals consisting of haploid genomes of with $L = 200$ loci.
-Genetic loci are partitioned into two groups, the two ``focal" loci where the epistatic alleles $A$ and $BB$ will segregate and the remaining ``background" loci that are responsible for the underlying fitness distribution.
+Genetic loci are partitioned into two groups, the two ``focal" loci where the epistatic alleles $A$ and $B$ will segregate and the remaining ``background" loci that are responsible for the underlying fitness distribution.
 We set the focal loci to be at positions $L/4$ and $3L/4$.
@@
 These mutations' fitness effects are drawn from an exponential distribution; the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation (they are multipled by the current variance and divided by $\sigma^2$).
-In this way, the mutation rate $U$ \jcomment{Is this $U_{B}$ or $U$?} and average fitness effect $\bar{\epsilon}$ at the background loci \jcomment{this should be defined clearly either above in this section or in the math background section (it appears in that section but is not explicitly defined)} are not parameters of our simulation model but must be directly measured from simulations; i.e., they are constrained by $L$ and $\sigma$.
+In this way, the mutation rate $U_b$ and average fitness effect $\bar{\epsilon}$ at the background loci are not parameters of our simulation model but must be directly measured from simulations; i.e., they are constrained by $L$ and $\sigma$.
 We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.
@@
 Tunneling becomes relevant when $N$ is too large for sequential fixation, i.e., the tunneling probability becomes more significant than the probability that the intermediate will fix: this mandates $N \gtrsim 1/\sqrt{\mu s}$.
-It also requires that $\ll 1/\mu$ \jcomment{$N$ missing here?}, meaning that only one mutant lineage is likely to be extant at any given time and, if lineages do co-occur, they are unlikely to interfere with each other or substantially affect the mean fitness.
-This happens to be the necessary condition for a birth-death process to be an appropriate model.\jcomment{is this previous sentence necessary?}
+It also requires that $N \ll 1/\mu$, meaning that only one mutant lineage is likely to be extant at any given time and, if lineages do co-occur, they are unlikely to interfere with each other or substantially affect the mean fitness.
 By increasing $N$ far above the $1/\mu$ limit, multiple deleterious mutants are likely to occur every generation, reaching mutation-selection balance at low frequency and providing many opportunities for a lucky double mutant to appear; this is semi-deterministic tunneling.
@@
 Here we present the key results of our simulation analysis.
-In general, we expect valley crossing to require more time in rapidly adapting populations (large $N \sigma$) experiencing genetic draft as bubbles are doomed to extinction more quickly \jcomment[inline]{I was going to add something like, ``(bubbles of intermediate size are less likely to occur under draft)'' to explain this, but is it more general (since you bring it up only under tunneling below)?
-Maybe add a short parenthetical explanation?} and beneficial lineages are less likely to fix, especially when valleys are shallow.
+In general, we expect valley crossing to require more time in rapidly adapting populations (large $N \sigma$) experiencing genetic draft; bubbles are doomed to extinction more quickly (lineages experience larger fluctuations in size and decline in frequency as the mean fitness advances) and beneficial lineages are less likely to fix, especially when valleys are shallow.
 Figures \ref{fig:deterministic} and \ref{fig:tunneling} confirm these expectations: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
-It is tempting to interpret this as a reduced $N_e$ due to a decrease in genetic diversity and the strength of selection.
-The underlying dynamics are more complicated, however.\jcomment{Yes, but the following sentences don't highlight specifically enough I think why; e.g., what about a simple reduction in $N_{e}$ cannot explain what happens in this draft case?}
-As $\sigma$ increases, the effect of the genetic background in which an mutation finds itself becomes more significant in determining the fate of that mutation than the fitness effect of the mutation itself.
-This genetic draft buffets the frequency of mutations so that these mutations behave as though they are more neutral than they really are.
+At higher $\sigma$, the effect of the genetic background in which an mutation finds itself becomes more significant in determining the fate of that mutation than the fitness effect of the mutation itself.
+This genetic draft buffets the frequency of mutations so that these mutations behave as though they are more neutral than they really are: in effect, the fitness landscape is ``flattened".
 This depresses the fixation probability of the double mutant and other complex adaptations, which appears to account for the majority of the change in fixation time.
@@
 This can be seen in figure \ref{fig:sigma_delta}.
-At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect of the valley crossing time, which means that very deep valleys can be crossed.
-This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations.
+At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that very deep valleys can be crossed.
+This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations: as previously mentioned, they appear to be ``flattened".
 
@@
 \caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
-In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
-Panel (a) shows valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
+In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, and $\mu = 10^{-5}$.
+Panel (a) shows that valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
 Panel (b) shows the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.}
@@
 How do these two effects combine to shape the likelihood of valley crossing in rapidly adapting populations experiencing genetic draft compared to populations only experiencing drift?
-To explore this, we first introduce the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
+To explore this, we first introduce the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a single beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
 By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps.
@@
 Figure \ref{fig:ratios} shows the expected pattern that deeper valleys in smaller populations take longer to cross.
-At a higher value of $\sigma$, valley depth has a weaker effect of the valley crossing time in both an absolute sense (Figure \ref{fig:tau}) and relative to the time for a single beneficial sweep (Figure \ref{fig:ratios}).
+At a higher value of $\sigma$, valley depth has a weaker effect on the valley crossing time in both an absolute sense (Figure \ref{fig:tau}) and relative to the time for a single beneficial sweep (Figure \ref{fig:ratios}).
 Moreover, values of $\alpha$ are generally lower for the population with higher $\sigma$ experiencing genetic draft than for the population with lower $\sigma$ experiencing only drift.
-This implies that draft tends reduce the valley crossing time relative to the time a single beneficial mutation to sweep.
+This implies that draft tends reduce the valley crossing time relative to the time for a single beneficial mutation to sweep.
 
-We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by plotting the ratio $\alpha(10^{-2})/\alpha(10^{-6})$ in Figure \ref{fig:alpha_ratio}, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
-Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored for $N < 10^{5}$.
-In this regime, increasing $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
-For $N > 10^5$ in the low $\sigma$ case, tunneling becomes semi-deterministic where $N\mu > 1$ and, on average, more than one mutant lineage is generated every generation.\jcomment{Another figure presumably from simulations that you're running...}
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha(10^{-2})/\alpha(10^{-6})$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+The ratio can be seen in Figure \ref{fig:alpha_ratio}.
+Overall, we find that raising $\sigma$ increases the extent to which valley crossing is favored for lower population sizes: for this combination of parameters, the condition is $N < 10^{5}$.
+In this regime, raising $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
+For $N > 10^5$ in the low $\sigma$ case, the product $N\mu > 1$: tunneling becomes semi-deterministic and, on average, more than one mutant lineage is produced every generation.
 These mutants then give rise to double mutants at a low but constant rate that is not sensitive to the valley depth (Figure \ref{fig:tau-ratios} and equation 35 in \citealp{weissman_2009}) and that is much faster than in the stochastic tunneling case.
 In contrast, $N > 10^{5}$ for high $\sigma$ does not produce the same order of magnitude increase in valley crossing rates, which is likely due to no analogous qualitative change in valley crossing dynamics.
-Thus, for large enough population sizes, draft does not significantly improve the valley crossing rate relative to the rate of single beneficial sweeps.\jcomment{Double check my explanation here :-)}
-
+Thus, for large enough population sizes, draft does not significantly improve the valley crossing rate relative to the rate of single beneficial sweeps.
+\iffalse
+\jcomment{Double check my explanation here :-)}
 At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{Cvijovic:Good:2018}, but the average behavior is more or less the same.
 This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.\jcomment{I still don't get what the point of mentioning this is here; how does this connection to the behavior of the ratio of $\alpha$s?}
+\fi
 
@@
 We have used simulation methods to examine the rate of crossing fitness valleys in rapidly adapting populations that experience genetic draft.
-We find that although the overall valley crossing time is longer in rapidly adapting than in slowly adapting (drifting) populations, a higher fraction of the adaptation that occurs rapidly adapting populations will occur through valley crossing than in slowly adapting populations.
+We find that although the overall valley crossing time is longer in rapidly adapting than in slowly adapting (drifting) populations, a higher fraction of adaptive events in rapidly adapting populations will occur through valley crossing than in slowly adapting populations.
 This mean that the genetic draft that results from rapid adaptation accelerates the rate at which populations escape from local peaks in a fitness landscape and move towards global peaks.
@@
 
-Our results complement and build on those of \citet{ochs_2015} who compared valley crossing and a sweeping beneficial mutation in a slowly evolving population.
+Our results complement and build on those of \citet{ochs_2015}, who compared valley crossing and a sweeping beneficial mutation in a slowly evolving population.
 They considered the scenario where the sweeping beneficial mutation \emph{competes} with a particular complex adaptation, whereas we treat these two scenarios separately.
-They find that large populations are more likely to cross valley than fix a simple beneficial mutation than small populations (for comparison, see the $\sigma=5 \times 10^{-6}$ panel in our Figure \ref{fig:ratios}).
+They find that large populations are more likely to cross valleys than fix a simple beneficial mutation than small populations (for comparison, see the $\sigma=5 \times 10^{-6}$ panel in our Figure \ref{fig:ratios}).
 However, they also find that intermediate-sized populations in which stochastic tunneling occurs are the least likely to cross a fitness valley instead of fixing a beneficial mutation.
 In contrast, our results suggest that the stochastic tunneling regime is where rapid adaptation and genetic draft increase the likelihood of valley crossing relative to a simple beneficial mutation.
-Thus, it is possible that genetic draft improves the likelihood of valley crossing precisely where its least likely under genetic drift. 
+Thus, it is possible that genetic draft improves the likelihood of valley crossing precisely where it is least likely under genetic drift. 
 % in their model, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult especially for populations of intermediate size.
-\jcomment[inline]{I removed an argument had here about mean fitness that I wasn't sure of; how does the sweeping beneficial mutation increasing mean fitness affect intermediate population sizes more than large or small?}
 
@@
 However, it has long been known that the shifting balance theory has important limitations due to the fact that too much migration prevents the accumulation of deleterious intermediates and too little reduces the success of genotypes from the fitter peaks and prevents them from spreading throughout the metapopulation \citep{coyne_barton_turelli_2000,Van-Cleve:Weissman:2015}.
-Recent work by \citet{Bitbol:Schwab:2014} confirms the fact that intermediate migration rates yield the fastest valley crossing times; however, it also finds that valley crossing is faster in subdivided population with these intermediate migration rates than in panmictic populations.
+Recent work by \citet{Bitbol:Schwab:2014} confirms the fact that intermediate migration rates yield the fastest valley crossing times; however, it also finds that valley crossing is faster in subdivided populations with these intermediate migration rates than in panmictic populations.
 

### 2018-06-19 ms/valley_crossing.tex
edits of edits!

@@
 Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals give rise to the complex adaptation, it will therefore likewise be more likely to fix.
-Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and the peak height $s$.
+Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $\pfix$ will depend on $x$ and the peak height $s$.
 These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
@@
 We focus on the case where background mutations are frequent but of weak effect--that is, the mean effect of background mutations $\bar{\epsilon}$ is small compared to $\sigma$, but background mutations occur at a high rate $U_b$.
-\iffalse
-In this case, the exact values of $U_b$ and $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
-\fi
 \citet{hallatschek_2011} and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
 \begin{equation}
-v\partial_x p(x,s) = x p(x,s) + \frac{U_b\bar{\epsilon}^2}{2}\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
+v\partial_x \pfix(x,s) = x \pfix(x,s) + \frac{U_b\bar{\epsilon}^2}{2}\partial_x^2 \pfix(x,s) - \frac{\pfix(x,s)^2}{2},
 \end{equation}
@@
 For a neutral mutation ($s = 0$), the fixation probability is effectively zero at low $x$ but increases exponentially up until $x_c$, at which it becomes roughly linear in $x$.
-In this regime, $x_c \approx 2\sigma^4/U_b\bar{\epsilon}^2$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
-The effect of introducing a nonzero $s$ is essentially to shift the value of $x$ and $x_c$: for large negative $s$, $x_c$ may be far enough in the nose that it is unlikely any individuals will actually be present, and for large positive $s$, even unfit individuals can give rise to a lineage that is destined to fix.
+In this regime, $x_c \approx 2\sigma^4/U_b\bar{\epsilon}^2$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations.
+The effect of introducing a nonzero $s$ is essentially to shift the value of $x$ and $x_c$: for large negative $s$, $x_c$ may be far enough in the nose of the Gaussian distribution that it is unlikely any individuals will have a high enough background fitness to fix, and for large positive $s$, even individuals with an unfit genetic background can give rise to a lineage that is destined to fix.
 
@@
 We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes \citep{weissman_2009}.
-\jcomment{commented out some text here; put back into the results if the pattern described is evident there}
-% In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
 For small populations, sequential fixation applies when $N \ll \min(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
@@
 This can be seen in figure \ref{fig:sigma_delta}.
-At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that very deep valleys can be crossed.
-This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations: as previously mentioned, they appear to be ``flattened".
+At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that deep valleys can be more easily crossed.
+This suggests that the ``flattening'' of rugged fitness landscapes due to genetic draft may in fact yield some benefits to valley crossing in rapidly adapting populations easier.
 
@@
 In this regime, raising $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
-For $N > 10^5$ in the low $\sigma$ case, the product $N\mu > 1$: tunneling becomes semi-deterministic and, on average, more than one mutant lineage is produced every generation.
+For $N > 10^5$ in the low $\sigma$ case, tunneling becomes semi-deterministic where $N\mu > 1$ and, on average, more than one mutant lineage is produced every generation.
 These mutants then give rise to double mutants at a low but constant rate that is not sensitive to the valley depth (Figure \ref{fig:tau-ratios} and equation 35 in \citealp{weissman_2009}) and that is much faster than in the stochastic tunneling case.
@@
 Thus, for large enough population sizes, draft does not significantly improve the valley crossing rate relative to the rate of single beneficial sweeps.
-\iffalse
-\jcomment{Double check my explanation here :-)}
-At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{Cvijovic:Good:2018}, but the average behavior is more or less the same.
-This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.\jcomment{I still don't get what the point of mentioning this is here; how does this connection to the behavior of the ratio of $\alpha$s?}
-\fi
 
@@
 Thus, it is possible that genetic draft improves the likelihood of valley crossing precisely where it is least likely under genetic drift.
-% in their model, the complicating factor is that a partially sweeping beneficial mutation increases the mean fitness, making valley crossing more difficult especially for populations of intermediate size.
 
@@
 
+\clearpage
 \bibliographystyle{genetics}

### 2018-07-26 ms/valley_crossing.tex
updated explanation of alpha simulations

--- 2018-06-14 ms/valley_crossing.tex
@@
 We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
-However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down.
+However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down due to clonal interference and genetic background effects.
 In addition, we confirm that in rapidly adapting populations, adaptive fixations of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
@@
 These alleles mutate to $A$ and $B$ at rates $\mu_A$ and $\mu_B$, respectively (we neglect back mutations).
-Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_A$, $1-\delta_B$, and $1+s$ respectively, with $\delta \geq 0$ (intermediate genotypes are deleterious) and $1/N < s < 1$ (double mutant $AB$ is strongly beneficial).
+Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_A$, $1-\delta_B$, and $1+s$ respectively, with $\delta \geq 0$ (intermediate genotypes are deleterious) and $1/N < s < 1$ (double mutant $AB$ is strongly beneficial); see \ref{fig:landscape}.
 For the remainder of our analysis, we will assume that $\mu_A = \mu_B = \mu$ and $\delta_A = \delta_B = \delta$.
@@
 We zero in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
-We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, tunneling, and deterministic fixation cases.
+We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, stochastic tunneling, and semi-deterministic tunneling cases.
 
@@
 The expected valley crossing time then becomes $1/N\mu\Phi$.
+An additional difficulty is that technically, $\pfix$ will depend not on the fitness of the initial lineage in a bubble but on the fitness of the specific background on which the double mutant arose.
+These will not necessarily be the same; for example, individuals within the single mutant lineage may accumulate additional background mutations.
+Assuming that this does \emph{not} happen, that is, assuming that $\pfix$ depends on the fitness $x$ of the founding single mutant lineage, is tantamount to assuming that bubbles are too short lived to accumulate background mutations that significantly affect their fitness.
 
@@
 For a neutral mutation ($s = 0$), the fixation probability is effectively zero at low $x$ but increases exponentially up until $x_c$, at which it becomes roughly linear in $x$.
-In this regime, $x_c \approx 2\sigma^4/U_b\bar{\epsilon}^2$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
+In this regime, $x_c \approx 2\sigma^4/U_b\bar{\epsilon}^2$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by background mutations.
 The effect of introducing a nonzero $s$ is essentially to shift the value of $x$ and $x_c$: for large negative $s$, $x_c$ may be far enough in the nose that it is unlikely any individuals will actually be present, and for large positive $s$, even unfit individuals can give rise to a lineage that is destined to fix.
@@
 \begin{equation}
-\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - \sigma^2 t + \delta)\phi^2,
+\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi - (1 + x - \sigma^2 t + \delta)\phi^2,
 \end{equation}
@@
 During this time, the mutation rate $\mu$ for the focal loci is set to zero.
-After equilibration, we set a per-site mutation rate $\mu$ for the focal loci (allowing forward and backward mutations) with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$.
+After equilibration, we set the per-site mutation rate $\mu$ for the focal loci (allowing forward and backward mutations) with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$.
 When the double mutant reaches frequency $0.5$, we consider it to have fixed.
@@
 We partitioned our simulations as follows.
-We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and sequential fixation regimes \citep{weissman_2009}.
-\jcomment{commented out some text here; put back into the results if the pattern described is evident there}
+We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and deterministic fixation regimes \citep{weissman_2009}.
 % In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
@@
 \end{subfigure}
-\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes as a function of $\sigma$.
-In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$.
+\caption{Valley crossing times in the semi-deterministic tunneling regime as a function of $\sigma$.
+In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$; the right figure has a high enough value of $N\mu^2$ that fixation is almost deterministic.
 Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges.
@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{Figures/compare_ratios.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$.
-Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes.
-But when semi-deterministic tunneling becomes more important than stochastic tunneling, it can slow down the crossing rate.}
+\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$, for varied values of $\mu$.
+Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys: at higher values of $\mu$ this limitation disappears.}
 \label{fig:alpha_ratio}
@@
 
-We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha(10^{-2})/\alpha(10^{-6})$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha(10^{-2})/\alpha(10^{-6})$ (the arguments are values of $\sigma$), which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
 The ratio can be seen in Figure \ref{fig:alpha_ratio}.
@@
 In this regime, raising $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
-For $N > 10^5$ in the low $\sigma$ case, the product $N\mu > 1$: tunneling becomes semi-deterministic and, on average, more than one mutant lineage is produced every generation.
-These mutants then give rise to double mutants at a low but constant rate that is not sensitive to the valley depth (Figure \ref{fig:tau-ratios} and equation 35 in \citealp{weissman_2009}) and that is much faster than in the stochastic tunneling case.
-In contrast, $N > 10^{5}$ for high $\sigma$ does not produce the same order of magnitude increase in valley crossing rates, which is likely due to no analogous qualitative change in valley crossing dynamics.
-Thus, for large enough population sizes, draft does not significantly improve the valley crossing rate relative to the rate of single beneficial sweeps.
-\iffalse
-\jcomment{Double check my explanation here :-)}
-At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{Cvijovic:Good:2018}, but the average behavior is more or less the same.
-This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.\jcomment{I still don't get what the point of mentioning this is here; how does this connection to the behavior of the ratio of $\alpha$s?}
-\fi
+The exception is at very high values of $N$ and $\delta$ and low values of $\mu$, where draft turns out again not to be helpful; see the lower right part of the leftmost plot.
+In this regime, $\delta$ is on the order of $\sigma$ in our draft simulations, meaning that single mutant lineages are rapidly cut off by the advancing mean fitness and are likely almost never present in the nose of the fitness distribution.
+Essentially, the population size is large enough that intermediate mutants can be ``seen" by selection, but the fitness detriment is severe enough compared to $\sigma$ that draft is no longer helpful in allowing mutant lineages to reach high frequency.
+This problem disappears as $\mu$ increases (center and right plots) and the population is far beyond the semi-deterministic tunneling barrier.
 

### 2018-07-26 src/compare_alpha.py
updated versions of main scripts

@@ -1,0 +1,372 @@
+#!/usr/bin/env python
+'''
+plt_asex_crossings.py
+Author: Taylor Kessinger
+Date: July 24, 2018
+Description: Glob and plot valley crossing times as a function of input variables.
+'''
+import sys
+sys.path.append('/home/tkessinger/tools/FFPopSim/pkg/python/')
+sys.path.append('/home/tkessinger/Documents/draft_valley_crossing/FFPopSim/pkg/python/')
+import numpy as np
+import FFPopSim as h
+import random as rand
+import matplotlib.pyplot as plt
+from time import time
+import argparse
+import math
+import cPickle as pickle
+import glob
+import os
+
+import seaborn as sns
+sns.set_style('white')
+sns.set_context('paper')
+sns.set(font="CMU Sans Serif", font_scale=1.2,style='white')#, style='ticks')
+
+
+numruns = {}
+sigma_range = {}
+N_range = {}
+s_range = {}
+delta_range = {}
+mu_range = {}
+rho_range = {}
+
+run_time = {}
+
+N_range = [np.int(x) for x in np.logspace(3,6,7)]
+#    N_range[name] = [10**3,3*10**3, 10**4,10**5]#,3*10**4,10**5,3*10**5,10**6] #range of population size values
+sigma_range = [5*10**-6,5*10**-2]
+#delta_range = [0.0,10**-4,10**-2] #range of deleterious intermediate coefficients
+delta_range = [0.0, 10**-4.5,10**-4,10**-3.5,10**-3,10**-2.5,10**-2,10**-1.5]
+s_range_sweep = [10**-2] #range of selection coefficients for the adaptation
+s_range_valley = [10**-1]
+valley_name = 'compare_ratios_high_mu'
+if valley_name == 'compare_ratios_high_mu':
+    sweeps_name = 'compare_ratios_sweeps_high_mu'
+    mu_range = [5*10**-4]
+elif valley_name == 'compare_ratios_low_mu':
+    mu_range = [5*10**-6]
+    sweeps_name = 'compare_ratios_sweeps_low_mu'
+
+elif valley_name == 'compare_ratios_very_high_mu':
+    sweeps_name = 'compare_ratios_sweeps_very_high_mu'
+
+    mu_range = [10**-4]
+else:
+    sweeps_name= 'compare_ratios_sweeps_2'
+    mu_range = [5*10**-5]
+numruns = 3
[…]

### 2018-07-26 src/plt_asex_crossings.py

@@ -29,7 +29,7 @@ sns.set_style('ticks')
 #var_sigma_8: variable sigma, N = 10**5
 
 
-name = 'var_sigma_sequential_fix_N_4_delta_-5'
+name = 'var_sigma_tunneling_3'
 
 numruns = {}
 sigma_range = {}
@@ -348,6 +348,31 @@ if name == 'var_sigma_tunneling':
     sigma_range[name] = [10**x for x in range(-5,-1)] + [2*10**x for x in range(-5,-1)] + [8*10**x for x in range(-5,-1)] + [5*10**x for x in range(-5,-1)]    
     sigma_range[name] = np.sort(sigma_range[name])
     delta_range[name] = [10**-3] #range of deleterious intermediate coefficients
+    s_range[name] = [10**-2] #range of selection coefficients for the adaptation
+    mu_range[name] = [1*10**-6]
+    numruns[name] = 3
+    run_time[name] = 10000000
+    rho = 0.0
+    fixed_s = False
+    graphtype = 'sigma'
+
+if name == 'var_delta':
+    N_range[name] = [10**5]#,3*10**4,10**5,3*10**5,10**6] #range of population size values
+    sigma_range[name] = [5e-6,5e-2]
+    delta_range[name] = [10**-6,10**-5,10**-4,10**-3,10**-2] #range of deleterious intermediate coefficients
+    s_range[name] = [10**-1] #range of selection coefficients for the adaptation
+    mu_range[name] = [3*10**-6]
+    numruns[name] = 3
+    run_time[name] = 1000000
+    rho = 0.0
+    fixed_s = False
+    graphtype = 's_delta'
+
+if name == 'var_sigma_tunneling_3':
+    N_range[name] = [10**5]#,3*10**4,10**5,3*10**5,10**6] #range of population size values
+    sigma_range[name] = [10**x for x in range(-5,-1)] + [2*10**x for x in range(-5,-1)] + [8*10**x for x in range(-5,-1)] + [5*10**x for x in range(-5,-1)]    
+    sigma_range[name] = np.sort(sigma_range[name])
+    delta_range[name] = [10**-4] #range of deleterious intermediate coefficients
     s_range[name] = [10**-2] #range of selection coefficients for the adaptation
     mu_range[name] = [1*10**-6]
     numruns[name] = 3
@@ -681,11 +706,11 @@ def plt_s_delta_dependence():
                             if not params in inferred_Uvals.keys():
                                 inferred_Uvals[params] = []
                             tmp_Uvals = []
[…]

### 2018-07-26 src/plt_ratios.py

@@ -1,0 +1,376 @@
+#!/usr/bin/env python
+'''
+plt_asex_crossings.py
+Author: Taylor Kessinger
+Date: June 16, 2017
+Description: Glob and plot valley crossing times as a function of input variables.
+'''
+import sys
+sys.path.append('/home/tkessinger/tools/FFPopSim/pkg/python/')
+sys.path.append('/home/tkessinger/Documents/draft_valley_crossing/FFPopSim/pkg/python/')
+import numpy as np
+import FFPopSim as h
+import random as rand
+import matplotlib.pyplot as plt
+from time import time
+import argparse
+import math
+import cPickle as pickle
+import glob
+import os
+
+import seaborn as sns
+sns.set_style('white')
+sns.set_context('paper')
+sns.set(font="CMU Sans Serif", font_scale=1.2,style='white')#, style='ticks')
+
+
+numruns = {}
+sigma_range = {}
+N_range = {}
+s_range = {}
+delta_range = {}
+mu_range = {}
+rho_range = {}
+
+run_time = {}
+
+# N_range = [10**3,10**4,10**5]#,3*10**4,10**5,3*10**5,10**6] #range of population size values
+# sigma_range = [5*10**-6,5*10**-2]
+# delta_range = [0.0,10**-4,10**-2] #range of deleterious intermediate coefficients
+# s_range_valley = [10**-2] #range of selection coefficients for the adaptation
+# s_range_sweep = [10**-2]
+# mu_range = [5*10**-5]
+# numruns = 3
+# rho=0.0
+
+N_range = [np.int(x) for x in np.logspace(3,6,7)]
+#    N_range[name] = [10**3,3*10**3, 10**4,10**5]#,3*10**4,10**5,3*10**5,10**6] #range of population size values
+sigma_range = [5*10**-6,5*10**-2]
+#delta_range = [0.0,10**-4,10**-2] #range of deleterious intermediate coefficients
+delta_range = [0.0, 10**-4.5,10**-4,10**-3.5,10**-3,10**-2.5,10**-2,10**-1.5]
+s_range_sweep = [10**-2] #range of selection coefficients for the adaptation
+s_range_valley = [10**-1]
+valley_name = 'compare_ratios_high_mu'
+if valley_name == 'compare_ratios_high_mu':
+    sweeps_name = 'compare_ratios_sweeps_high_mu'
[…]

### 2018-07-26 src/simulate_population.py

@@ -34,6 +34,7 @@ parser.add_argument('--runno', type=int, default=0)
 parser.add_argument('--runtime', type=int, default=100000)
 parser.add_argument('--name', type=str, default='test')
 parser.add_argument('--fixed_s', type=bool, default=False)
+parser.add_argument('--expected_crossings', type=int, default=100)
 
 args=parser.parse_args()
 
@@ -44,6 +45,7 @@ sigma = args.sigma #fitness variance
 rho = args.rho #per generation outcrossing rate
 cx = 0 #per locus crossover rate
 #unfortunately, nonzero values of rho and cx increase the runtime somewhat
+expected_crossings = args.expected_crossings
 
 s = args.s
 delta = args.delta
@@ -128,7 +130,8 @@ if not os.path.exists('output/valley_crossing_sims_'+args.name):
 total_number_of_mutations = 0
 
 
-while pop.generation < run_time:
+#while pop.generation < run_time:
+while len(dm_successful_crossings) < expected_crossings:
     pfit = pop.get_trait_statistics(0)
     if pfit.variance > 0 and pop.generation > np.min([100, base_burn_time]) and not args.fixed_s:
         multiplier = sigma/np.sqrt(pfit.variance)
@@ -182,13 +185,13 @@ while pop.generation < run_time:
                 pop.set_allele_frequencies(allele_freqs,pop.N)
                 pop.carrying_capacity = N #reset the carrying capacity to its desired value
                 
-                #selection_coefficients = np.random.exponential(s_mean, size = L) #we might as well change up the background fitnesses at this point
-                #selection_coefficients[mutant_locs]=1e-10
+                selection_coefficients = np.random.exponential(s_mean, size = L) #we might as well change up the background fitnesses at this point
+                selection_coefficients[mutant_locs]=1e-10
                 
                 double_mut_in_play = False
                 
                 pop.mutation_rate = 0
-                burn_time = pop.generation + base_burn_time #we might as well let the population equilibrate again for a bit
+                burn_time = pop.generation + base_burn_time/10 #we might as well let the population equilibrate again for a bit
                 burned_in = False
                 pop.trait_weights += np.array([0,-1.0]) #"turn off" the focal loci: just a failsafe
 

### 2018-08-01 ms/valley_crossing.tex
minor edits

@@
 We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and deterministic fixation regimes \citep{weissman_2009}.
-% In the case of sequential fixation, the valley crossing time is dominated by the time until a deleterious mutant that is destined to fix arises: once it does, it does so in order $T_{\mathrm{MRCA}}$ generations, which is of order $N$ for low $N\sigma$ and scales weakly with $N$ for higher $N\sigma$ \citep{neher_hallatschek_2013}.
 For small populations, sequential fixation applies when $N \ll \min(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
@@
 This can be seen in figure \ref{fig:sigma_delta}.
-At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that very deep valleys can be crossed.
-This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations: as previously mentioned, they appear to be ``flattened".
+At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that deep valleys can be crossed.
+This suggests that traversing rugged fitness landscapes by valley crossing may be enhanced in rapidly adapting populations as landscapes under genetic draft appear to be ``flattened".
 
@@
 \includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$, for varied values of $\mu$.
-Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys: at higher values of $\mu$ this limitation disappears.}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$.
+Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
 \label{fig:alpha_ratio}
@@
 
-We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha(10^{-2})/\alpha(10^{-6})$ (the arguments are values of $\sigma$), which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha_{\sigma_2} / \alpha_{\sigma_1}$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
 The ratio can be seen in Figure \ref{fig:alpha_ratio}.
@@
 Essentially, the population size is large enough that intermediate mutants can be ``seen" by selection, but the fitness detriment is severe enough compared to $\sigma$ that draft is no longer helpful in allowing mutant lineages to reach high frequency.
-This problem disappears as $\mu$ increases (center and right plots) and the population is far beyond the semi-deterministic tunneling barrier.
+This problem disappears as $\mu$ increases (center and right plots) and the population is beyond the deleterious tunneling barrier where the valley crossing time depends on the depth of the valley $\delta$.
 

### 2018-08-01 ms/cover_letter.tex
cover letter draft

\documentclass[kentucky,10pt,foldmarks=off,backaddress=false,refline=dateleft,letterpaper]{scrlttr2}

\usepackage{xspace}
\usepackage[parfill]{parskip}

% Sender's information
\setkomavar{signature}{Jeremy Van Cleve\\Taylor Kessinger}
%\setkomavar{signature}{\includegraphics[scale=0.3]{hand_sig.pdf}\Jeremy Van Cleve} % use PDF image of signature
\setkomavar{date}{\today}
\setkomavar{subject}{}

\begin{document}

% Recipient's information
\begin{letter}{
    GENETICS Editorial Board\\
    6120 Executive Boulevard\\
    Suite 550\\
    Rockville, MD 20852}

\opening{Dear GENETICS Editorial Board}

We are excited to submit to GENTETICS our manuscript entitled ``Genetic draft and valley crossing''. Our work focuses on the intersection of two topoics: (i) how populations evolve when the mapping between genotype and fitness, the so-called ``fitness landscape'', is rugged with valleys between multiple peaks, and (ii) how rapid adaptation and the genetic draft that it creates profoundly alters evolutionary processes. Both topics have received a great deal of interest from geneticists and evolutionary biologists in the last few years with a number of important studies in GENETICS but no study has looked at how genetic draft affects evolution on rugged fitness landscapes.

Specifically, we look at how the time to cross a fitness valley is affected by the variation in fitness in a population that is adapting to a novel environment. When that variation is low, genetic drift is the primary stochastic force shaping evolutionary dynamics and classic valley crossing results, such as Weissman et al. (2009, GENETICS), hold. In contrast, genetic draft is more important when fitness variation is high, and we show that valley crossing is fundamentally different in this case.

We show that genetic draft generally increases the time to cross a fitness valley and hence the time for a complex adaptation to evolve on a rugged fitness landscape. This is due to the lower efficacy of selection on the complex adaptation in the presence of genetic draft. Crucially, we also find that complex adaptations are more likely relative to simple selective sweeps under genetic draft and rapid adaptation than under genetic drift. This means that populations evolving in novel environments or with pathogens may more easier construct complex adaptations than populations in constant environments. This result suggests that genetic draft may be an important and underappreciated mechanism in the evolution of complex traits.

We suggest the following reviewers for our manuscript:\\
Daniel Weissman (Emory University), Joshua Plotkin (University of Pennsylvania), Andreas Wagner (University of Zürich), and Ben Good (UC Berkeley).

We hope you find our work appropriate for GENETICS. Please let us know if you have any questions.

\closing{Best regards,}

\end{letter}

\end{document}

### 2018-08-01 ms/valley_crossing.tex
update to Genetics template
minor typos

@@
-\documentclass[rmp]{revtex4}
-\usepackage[english]{babel}
-\usepackage{amssymb,amsfonts,amsmath}
-\usepackage{xcolor}
+\documentclass[10pt,twocolumn,twoside]{gsajnl}
+\articletype{inv} % article type: {inv} Investigation
+
 \usepackage{breqn}
-\usepackage{graphicx}
-\usepackage{caption}
+
 \usepackage{subcaption}
-\captionsetup{compatibility=false}
-
-\usepackage{amsmath}
-\usepackage{natbib}
-\usepackage{hyperref}
-\usepackage{enumerate}
-%\usepackage{subfig}
-
-\usepackage{fontspec}
-
-% Code syntax highlighting
-\usepackage{listings}
-\lstloadlanguages{C++,Python}
 
@@
 
-\definecolor{dkgreen}{rgb}{0,0.6,0}
-\definecolor{gray}{rgb}{0.5,0.5,0.5}
-\definecolor{mauve}{rgb}{0.58,0,0.42}
-
 %%% math macros
@@
 \usepackage{setspace}
-\usepackage[colorinlistoftodos]{todonotes}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{Genetic draft and valley crossing}
-\author{Taylor~Kessinger}
-\author{Jeremy~Van~Cleve}
-%\affiliation{Department of Biology, University of Kentucky}
+\author[$\ast$]{Taylor Kessinger}
+\author[$\ast$,1]{Jeremy Van Cleve}
+\affil[$\ast$]{Department of Biology, University of Kentucky}
 
-\date{\today}
+\keywords{genetic drift; rapid adaptation; epistasis; fitness landscape}
+\runningtitle{Genetic draft and valley crossing}
+\runningauthor{Kessinger and Van Cleve}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{abstract}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+
 \maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+\correspondingauthoraffiliation{1}{Corresponding author: Department of Biology, University of Kentucky. E-mail: \href{emailto:jvancleve@uky.edu}{jvancleve@uky.edu}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
-This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
+This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
 Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
@@
 If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
-This is sometimes called deterministic fixation of the multiple mutant\citep{weissman_2009}.
+This is sometimes called deterministic fixation of the multiple mutant \citep{weissman_2009}.
 In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation.
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{center}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{subfigure}[b]{0.4\textwidth}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \includegraphics[width=0.4\textwidth]{Figures/var_sigma_delta.pdf}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{subfigure}[b]{0.475\textwidth}
@@
 
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$.
-Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
-\label{fig:alpha_ratio}
-\end{figure}
-
 So far, we can see that valley crossing is slower in rapidly adapting populations but the crossing time is less sensitive to the depth of the valley.
@@
 This implies that draft tends reduce the valley crossing time relative to the time for a single beneficial mutation to sweep.
+
+\begin{figure*}[t]
+\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$.
+Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
+\label{fig:alpha_ratio}
+\end{figure*}
 
@@
 
-\clearpage
-\bibliographystyle{genetics}
 \bibliography{bib}

### 2018-08-02 ms/valley_crossing.tex
format changes, fixed \Phi, and added data availability

@@
 
-\section*{Introduction}
+\section{Introduction}
 
 The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
-This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
+This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
 Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
@@
 If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
-This is sometimes called deterministic fixation of the multiple mutant \citep{weissman_2009}.
+This is sometimes called deterministic fixation of the multiple mutant\citep{weissman_2009}.
 In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation.
@@
 
-\section*{Mathematical background}
+\section{Mathematical background}
 
@@
 Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $\pfix$ will depend on $x$ and the peak height $s$.
-These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
-The expected valley crossing time then becomes $1/N\mu\Phi$.
+These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\upPhi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
+The expected valley crossing time then becomes $1/N\mu\upPhi$.
 An additional difficulty is that technically, $\pfix$ will depend not on the fitness of the initial lineage in a bubble but on the fitness of the specific background on which the double mutant arose.
@@
 
-In order to obtain the total valley crossing probability $\Phi$, the only remaining element to compute is the distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
+In order to obtain the total valley crossing probability $\upPhi$, the only remaining element to compute is the distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
 Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu \pfix W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu \pfix$.
@@
 They also avoided the need to explicitly send time to infinity, which is necessary to incorporate all possible bubble sizes and lifetimes.
-Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu \pfix(x)$), lead to difficulty in deriving a simple closed form expression for $\Phi$.
+Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu \pfix(x)$), lead to difficulty in deriving a simple closed form expression for $\upPhi$.
 Therefore, we focus on simulation methods to study the dynamics of valley crossing.
 
-\section*{Simulation methods}
+\section{Simulation methods}
 
@@
 
-\section*{Results}
+\subsection*{Data availability}
+
+Simulation and analysis scripts needed to replicate the results are available at \url{https://github.com/tkessinger/draft_valley_crossing} under the Creative Commons BY-NC-SA license (\url{https://creativecommons.org/licenses/by-nc-sa/4.0/}).
+
+\section{Results}
 
@@
 
-\section*{Discussion}
+\section{Discussion}
 

### 2018-11-29 ms/response_to_reviews.tex
begin revision

\documentclass[11pt]{article}

\usepackage{natbib}
%\setcitestyle{numbers}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm,amssymb}
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage[letterpaper]{geometry} % easy way to set margins and paper size
\geometry{hmargin={1in,1in},vmargin={1in,1in}}

\usepackage[hidelinks]{hyperref}

%%% Check for xetex or luatex
\usepackage{ifxetex,ifluatex}
\newif\ifxetexorluatex % a new conditional starts as false
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi>0
\xetexorluatextrue
\fi
%%%

%%% Font settings
\ifxetexorluatex
\usepackage{fontspec}
\defaultfontfeatures{Ligatures=TeX}
\setmainfont{TeX Gyre Pagella}
\setsansfont[Scale=MatchLowercase]{Myriad Pro}
\setmonofont[Scale=MatchLowercase]{Fira Code}

\usepackage{unicode-math}
\setmathfont[Scale=MatchLowercase]{TeX Gyre Pagella Math}
\else
\usepackage[T1]{fontenc}
\usepackage{tgpagella}
\fi
%%%

\usepackage{setspace}
\usepackage{parskip}

\usepackage[explicit]{titlesec}

% new math operators
\newcommand{\der}{\mathop{}\!\mathrm{d}}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}

\newenvironment{reviewerquote}{\begin{quote}\color{black}\itshape}{\end{quote}}
\newenvironment{response}{\color{DarkBlue}}{}

\begin{document}

\section*{Associate Editor}

\begin{reviewerquote}
As mentioned by all referees, the non-standard setup of the simulation scheme makes it very difficult to relate the results to previous work. Although your main conclusion may well be confirmed for a standard infinite-sites model without constraining the background variance, this is by no means guaranteed (see the remarks about the infinitesimal limit by referee 2). To address this point (in the absence of analytical results), there is probably no way around individual-based simulations.
\end{reviewerquote}

\begin{reviewerquote}
The first point by referee 2 is particularly important: Indeed, it is not clear that serial substitutions and tunneling will dominate in the same parameter regions as for drift (it also may depend on the depth of the fitness valley and on other parameters like the background mutation rate). In your simulations, this point is not addressed at all. However, in the absence of theoretical results, we need an even more comprehensive numerical study.
\end{reviewerquote}

\begin{reviewerquote}
As mentioned by both referees, also your parameter choices are too limited. Note that additional figures can always be part of an electronic supplement.
\end{reviewerquote}

\begin{reviewerquote}
Your ms (and in particular the mathematical background part) has clearly been written for specialists. This is ok, but still the comments of referee 3, who is an interested non-specialist, may help you to identify points where points can be made clearer to reach a broader audience.
\end{reviewerquote}

\section*{Reviewer \#2}

\begin{reviewerquote}
Primarily, I’m concerned that the simulations have been used to explore a very sparse set of parameters, and I’m not convinced they provide a comprehensive picture. Furthermore, the theoretical motivation for picking these exact parameters is also not obviously justified. In sum, these shortcomings make it difficult to evaluate the conclusions drawn from these results as well as the mechanisms put forward as explanations. The work would be substantially improved if the simulations were expanded to cover a broader range of parameters, so that known limiting forms at the ends of the parameter range can be recovered, and sanity checks are added. I give several suggestions below.
\end{reviewerquote}

\begin{reviewerquote}
1. The authors justify the choice of parameters and regimes to focus on by quoting the results from the analysis of valley crossing in non-drafting populations by Weissman et al. (2009). Yet, as the authors point out at several points, evolutionary dynamics and genealogies are very different in rapidly adapting populations. It is not clear to me that the regimes in rapidly adapting populations are likely to be the same as the regimes in the cases studied by Weissman et al. (2009), or that they occur at the same transition points in terms of the parameters.

For instance, the authors primarily focus on tunneling (where the beneficial mu- tation arises on the background of a polymorphic deleterious mutation that is not destined to fix), which in the limit studied by Weissman et al. (2009) requires that 1 􏰀 N 􏰀 1 . The lower limit of 1 comes from the calculation of bubble sizes in √μs μ √μs
the limit considered by Weissman et al. (2009). However, the distribution of bubble sizes will be very different in the limit that the authors consider. Therefore it is unclear why this is a relevant scale. Furthermore, it is unclear whether (depending on the population size, fitness effects, mutation rates, etc.) tunneling is likely to be an equally broad regime in rapidly adapting populations. Here, deleterious hitchhik- ers can have a substantial probability of fixation, which sensitively depends on the mentioned parameters and increases with the size of the population (see e.g. Good and Desai (2014)). Thus, I would expect that the regime of sequential fixation can be much broader in rapidly adapting populations, and that the precise population size at which it gives way to the tunneling regime would be a much more complicated function of the parameters.

In principle, it should be possible to increase the range of parameters explored by simulation and see whether, e.g. transitions in the quantities measured seem to occur at anticipated transition points, whether the behavior is monotonic (see next point), and at what parameter values known limiting forms are achieved (e.g. in Figure 2, how do the rates compare to the values anticipated in the limits of low σ calculated by Weissman et al. (2009)?).
\end{reviewerquote}

\begin{reviewerquote}
2. Furthermore, there seems to be substantial variance in the outcomes in drafting populations, so that though the average time to crossing the valley is increased, there are lucky occurrences where the time to crossing the valley can be substantially shorter than in non-drafting populations. This variance appears larger than the difference between the means, so that reporting the average behavior doesn’t seem to capture the complete picture. Is this variance in outcomes driven by differences in the fitness of the background that the mutation arises on, so that some valleys are guaranteed to be crossed if the process is initiated in a high-fitness individual? Is the variance reduced when the depth of the valley is reduced? In addition, in Figure 2, though the average appears to asymptote, the fastest time appears non-monotonic. Is this a result of the relatively small number of replicates of the simulations (∼ 100), or an phenomenon that arises from the dynamics? Answering these questions would be important to interpreting the results, and would make me more comfortable about the verbal interpretations of the figures.
\end{reviewerquote}

\begin{reviewerquote}
3. In addition to this, because there are no calculations to guide interpretation, it would be important to add additional simulations to measure these quantities. Specifically, I might be missing something in the design of the simulations, but it’s not entirely clear to me that what the authors are observing is true tunneling for all parameters considered. Might it be that the deleterious intermediates are fixing prior to the arising of the second “valley crossing mutation”? It might be a good idea to record the fixation times of both the intermediate and the double mutant to help interpret these results.
4. The way the simulations have been set up is non-standard and makes it hard to use previous results to interpret which regime the populations are in. For instance, the authors’ claim is that the population is in the infinitesimal limit, where the only relevant parameter is σ (or, more accurately, Nσ). Yet Good et al. (2014) showed that for moderate Nσ there is a whole manifold of behaviors depending on the mutation rates and selection coefficients, and that observables do not always agree with asymptotically large Nσ limits. The fact that fitness effects of “background” mutations in these simulations are adjusted every generation to keep σ constant makes it difficult to understand what the background selection coefficients and overall background mutation rates are. Furthermore, the choice of exponential distribution for the selection coefficients of the genetic background may put the population in a regime where the infinitesimal limit may not always apply (see Fisher (2013)). Finally, the equilibration time of N/10 generations seems potentially short (one would hope for minimally a timescale of coalescence, and it’s not possible to interpret whether this is satisfied here).
\end{reviewerquote}

\begin{reviewerquote}
1. Page 2, column 2: “Alleles whose dynamics are primarily governed by genetic drift can be effectively modeled by a diffusion approximation” and “drafting populations experience quite large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation”: I’m not sure this is correct: allele frequency trajectories can in fact be modeled by a diffusion equation, albeit a more complicated one (e.g. the reference cited, Neher and Shraiman (2011a), uses such an approach). I’m also not sure what is meant by ‘jumps’ here – allele frequency trajectories tend to be no less smooth in drafting populations than in drifting populations.
\end{reviewerquote}

\begin{reviewerquote}
2. Page 3, column 1: “small bubbles are more common, but large bubbles are very rare”: I don’t think the first half of this statement is correct. If mutations arise equally frequently in both cases, small bubbles must be equally common.
\end{reviewerquote}

\begin{reviewerquote}
3. Page 3, column 2: “In previous approaches, these stochastic effects were smoothed out due to the presence of recombination”: Perhaps I don’t understand what is meant here, but this strikes me as inaccurate. The effects of drift are in the nose of the dis- tribution are of critical importance in the case considered by Neher and Shraiman (2011a): these authors do take them into account to obtain the establishment prob- abilities and allele frequency trajectories (e.g. Eq. 4 in that work).
\end{reviewerquote}

\begin{reviewerquote}
4. Page 3, column 2: “This is consistent with the fact that all forms of adaptation are slowed down due to clonal interference and genetic background effects”: Does it mean to imply that the rate at which valleys are crossed is reduced by the same amount by the rate at which beneficial mutations of effect s are likely to fix compared to NUbs? This is a quantitative claim, and I don’t think it’s been shown. If I’m missing it, then perhaps a reference to a figure would be appropriate.
\end{reviewerquote}

\begin{reviewerquote}
5. Page 4, column 1: “By Fisher’s “fundamental theorem”, the rate v at which the mean fitness advances is set to σ2”: this implies that v = σ2, ignoring the μ⟨s⟩ term [in v = σ2 + μ⟨s⟩] that arises in the full moment expansion, e.g. see Neher and Shraiman (2011b) for a recent review. This term is sometimes subdominant, but not in all regimes (and it’s not obvious it’s subdominant in the simulations carried out by the authors).
\end{reviewerquote}

\begin{reviewerquote}
6. Page 4, column 2: The analysis of timescales seems inaccurate: the estimate that T0 is of order 1 log(Ns) is not appropriate for drafting populations.
\end{reviewerquote}

\section*{Reviewer \#3}

\begin{reviewerquote}
However, in my opinion the manuscript does not satisfy the criteria of quality and scientific substance required for publication in Genetics, for two main reasons. First, the modeling setup seems in several respects arbitrary and poorly motivated. Rather than using a true 'infinite sites' model for the background mutations as is standard in the fitness wave studies mentioned above, the authors choose a genome of 200 loci, two of which are designated as focal; the positioning of the focal loci is specified, though obviously this cannot matter in the absence of recombination. Although the authors do not explicitly say so, I assume that the background fitness is additive or multiplicative across loci. However, rather than simply letting background loci mutate to a new state with a randomly chosen selection coefficients, the authors impose a mutation scheme where a locus mutates whenever it becomes monomorphic. This seems like a highly unrealistic and awkward choice which makes it effectively impossible to externally control parameters such as the background mutation rate and selection coefficients. I find it hard to assess to what extent the results presented here generalize to a more reasonable modeling framework. 
\end{reviewerquote}

\begin{reviewerquote}
Compared to the extensive (and well-written) review of the state of the art on pages 1-5, the results section* of the paper is remarkably meager. Results are exclusively numerical, subject to large error bars, and largely qualitative. Again, arbitrary parameter choices make it hard to assess the generality of these results. For example, for the comparison between the times required for a valley 
crossing and the sweep of a single beneficial mutation the authors arbitrarily choose the 'valley' selection coefficient to be ten times that of the sweep. It would seem that choosing the selection coefficients to be same would be more informative.
\end{reviewerquote}

\begin{reviewerquote}
1. page 5, second column: It seems to me that in the discussion below Eq.(1), the background fitness x should be replaced by $x - \bar{x}$ (fitness relative to mean fitness). 
\end{reviewerquote}

\begin{reviewerquote}
2. page 6, second column, second paragraph: Here the authors say that backward mutations at the focal loci are allowed. This contradicts the statement in the first column on page 4.
\end{reviewerquote}

\begin{reviewerquote}
3. page 7, first paragraph of Results: Why is the null expectation that genetic draft increases the time for valley crossing? At least in their abstract, Neher \& Shraiman (2011) appear to claim the opposite trend (last sentence). Or is this only true in the presence of recombination?
\end{reviewerquote}

\begin{reviewerquote}
4. page 8, second column, line 14: $s_{sweep}$ -> $s_{valley}$
\end{reviewerquote}

\section*{Reviewer \#4}

\begin{reviewerquote}
1. Page 3: “The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like Drosophila, is one possible explanation for the ‘paradox of variation’, the fact that genetic diversity and population size often do not scale linearly”.
How do the authors reconcile this suggestion with the fact that natural populations of Drosophila show very small levels of LD? (e.g. Mackay et al. 2012)
\end{reviewerquote}

\begin{reviewerquote}
2. Pages 3-6: A long discussion of analytical math is provided in the introduction only to conclude that it is not applicable to the scenario investigated in this paper - which only uses simulations. The math section* should be shortened significantly or moved to the supplement.
\end{reviewerquote}

\begin{reviewerquote}
3. Page 6: It seems like only beneficial mutations are modeled in the background loci. The reality is that most background mutations are deleterious, so beneficial focal mutations are more likely to be “dragged” down by linked deleterious alleles than “drafted” up by beneficial ones. Is it realistic to completely ignore negative selection on linked sites?
\end{reviewerquote}

\begin{reviewerquote}
4. Page 6: “the population experiences a constant influx of beneficial mutations at the background loci. These mutations’ fitness effects are drawn from an exponential distribution”. In page 4 (mathematical description of the system): “If x is the background fitness of a lineage, then the distribution of background fitnesses f (x) is assumed to be roughly Gaussian”. How can sampling from an exponential distribution produce a Gaussian background? Any part of the mathematical discussion that does not pertain to the assumptions of this study should either be removed or explicitly mentioned to be not relevant to simulations done here.
\end{reviewerquote}

\begin{reviewerquote}
5. Page 6: “any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population. In this way, the population experiences a constant influx of beneficial mutations at the background loci.” This practice seems to me to mask one of the important effects of population size: the total number of mutations introduced every generation. So, it does not seem appropriate at least when the question is the effect of Ne on likelihood or time of valley crossing.
\end{reviewerquote}

\begin{reviewerquote}
6. Pages 7-9: Do all simulations always successfully cross the valley? If not, the rate of success should be reported in addition to the required time.
\end{reviewerquote}

\begin{reviewerquote}
7. Page 7-9: As the authors discuss, when 𝜎 is high, the effect of linked background loci might be more important in determining the fate of a lineage than the focal loci. A question then arises: what is the distribution of background fitness of the lineages that make the crossing? Do only focal mutations on favorable backgrounds cross the valley? If all of them cross, then a plot or summary statistics of dependence of 𝜏 on the lineages’ background fitness (averaged over loci and time to simplify) would be informative.
\end{reviewerquote}

\begin{reviewerquote}
8. Page 8, right column: “In Figure 5, we set ssweep = 0.01, ssweep = 0.1, and m = 10-5”. Should be “svalley = 0.1”.
\end{reviewerquote}

\begin{reviewerquote}
9. Page 8: “Figure 5b shows the expected pattern that deeper valleys in smaller populations take longer to cross”. This seems contrary to earlier arguments for easier crossing in smaller populations due to weaker selection on deleterious intermediates. Please clarify.
\end{reviewerquote}

\begin{reviewerquote}
10. Page 10: The authors mention the contrast of their results to those of Ochs and Desai. Can they offer any explanations as to the reason?
\end{reviewerquote}

\end{document}

### 2018-11-29 src/WaveFit.jl
recent submit and plotting scripts

@@ -1,0 +1,337 @@
+#!/usr/bin/env julia
+
+## WaveFit.jl
+##
+## Author: Taylor Kessinger <taylor.kessinger@uky.edu>
+## Simulation model for "traveling (fitness) wave" populations.
+
+
+module WaveFit
+export evolve!, selection!, mutation!, evolve_multi!
+export Clone
+export Population, Landscape
+export FitnessClass
+export get_mean_fitness, get_frequencies
+export copy
+
+using Distributions, Combinatorics, StatsBase
+
+mutable struct Landscape
+    σ::Float64
+    δ::Float64
+    s::Float64
+    β::Float64
+    UL::Float64
+    μ::Array{Float64,1}
+
+    Landscape() = new(1e-6, 0.001, 0.01, 1.0, 10, [1e-4, 1e-4])
+    Landscape(σ, δ, s, β, UL, μ) = new(σ, δ, s, β, UL, μ)
+end
+
+mutable struct FitnessClass
+    n::Int64
+    bg_mutations::Int64
+    loci::Array{Int64,1}
+    bg_fitness::Float64
+    loci_fitness::Float64
+
+    # constructors
+    FitnessClass(n, bg_mutations, loci) = new(n, bg_mutations, loci, 0.0, 0.0)
+    FitnessClass(n) = new(n, 0, [0, 0])
+    FitnessClass() = FitnessClass(1)
+end
+
+function copy(fc::FitnessClass)
+    return FitnessClass(fc.n, fc.bg_mutations, fc.loci)
+end
+
+# generate a key from a FitnessClass as an array
+function key(fc::FitnessClass)
+    return vcat([fc.bg_mutations], fc.loci)
+end
+
+mutable struct Population
+    K::Int64    # carrying capacity
+    N::Int64    # population size
+    landscape::Landscape
+    classes::Dict{Array{Int64,1}, FitnessClass}
+    mean_bg_fitness::Float64
+    mean_loci_fitness::Float64
+    mean_fitness::Float64
+    var_bg_fitness::Float64
+    generation::Int64
+
+    Population(K, landscape) = new(K, K, landscape, Dict(key(FitnessClass(K)) => FitnessClass(K)),
+                                   1.0, 1.0, 1.0, 1.0, 1)
+    Population(K) = Population(K, Landscape())
+end
+
+function loci_fitness(class::FitnessClass, pop::Population)
+    # returns the sign-epistatic fitness for the focal loci
[…]

### 2018-11-29 src/plt_grid_sims.jl

@@ -1,0 +1,26 @@
+#!/usr/bin/env julia
+
+## plt_grid_sims.jl
+##
+## Author: Taylor Kessinger <taylor.kessinger@uky.edu>
+## Make "grid" plots of crossing time versus δ and N.
+
+using PyPlot, JLD, Glob, WaveFit, LaTeXStrings
+#gr()
+
+function load_and_plot(namestring::AbstractString)
+
+    files = glob("output/vc_sims/grid_tunnel_*.jld")
+    # glob all the output files
+
+    # keys for this dict will be K value and s_u value
+
+    for file in files
+        f = load(file)
+        println(f["params"])
+    end
+end
+#sigmalist, UL_list, nt_results = load_and_plot("neut_tunnel")
+#UL_vals = unique(UL_list)
+#display(
+load_and_plot("grid_tunnel")

### 2018-11-29 src/sample_sim.jl

@@ -1,0 +1,62 @@
+#!/usr/bin/env julia
+
+## sample_sim.jl
+##
+## Author: Taylor Kessinger <taylor.kessinger@uky.edu>
+## Basic implementation of WaveFit.jl
+
+using Revise
+using WaveFit
+#using Profile
+using Distributions
+using Stats
+
+const K = 10000 # carrying capacity
+
+const sigma = .05
+const delta = .001
+const s = .01
+const UL = 10
+const beta = 0.01
+const mu = 0.0
+
+numgens = 100000
+
+landscape = Landscape(sigma, delta, s, UL, beta, [mu,mu])
+pop = Population(K,landscape)
+
+@time (
+#while get_frequencies(pop)[2] < 0.5
+for i in 1:numgens;
+    evolve!(pop)
+    if (pop.generation%(numgens/10) == 0)
+        #bar([c.bg_mutations for c in values(pop.classes)], [c.n for c in values(pop.classes)],yscale=:log10)
+        #bg_muts = [c.bg_mutations for c in values(pop.classes)]
+        #println(maximum(bg_muts) - minimum(bg_muts))
+        println(pop.generation)
+    end
+end
+)
+println("$K $numgens")
+
+landscape = Landscape(sigma, delta, s, UL, beta, [mu,mu])
+pop = Population(K,landscape)
+
+@time (
+#while get_frequencies(pop)[2] < 0.5
+for i in 1:numgens;
+    evolve_multi!(pop)
+    if (pop.generation%(numgens/10) == 0)
+        #bar([c.bg_mutations for c in values(pop.classes)], [c.n for c in values(pop.classes)],yscale=:log10)
+        #bg_muts = [c.bg_mutations for c in values(pop.classes)]
+        #println(maximum(bg_muts) - minimum(bg_muts))
+        println(pop.generation)
+    end
+end
+)
+println("$K $numgens")
+
+
+#using Plots
+# plot the distribution of background mutations
+#bar([c.bg_mutations for c in values(pop.classes)], [c.n for c in values(pop.classes)],yscale=:log10)

### 2019-01-24 ms/response_to_reviews.tex
began response to reviewers

@@
 
+We have redone most of the simulations in this paper using \texttt{WaveFit}, a custom simulation package written in Julia.
+Both the original \texttt{FFPopSim} simulations and the new ones are individual-based, though groups of individuals identical in fitness are abstracted away as "clones" during the natural selection step.
+We have made this clearer.
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+We have amended the results section to clarify that we are not claiming the parameter regimes in the cases of drift and draft are identical.
+In addition, we have explored a larger chunk of the relevant state space.
 
@@
 
+(I need to think about this.)
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+The fitness distribution drops off as $1/n^2$ in the draft case rather than $1/n$ in the drift case. This means most bubbles are short-lived, and long-lived bubbles are the exception.
 
@@
 
+We agree that the effects of drift on mutations in the nose are of critical importance, even in the presence of recombination.
+But high recombination rates have the effect of "flattening" the nose by limiting the fat-tailedness of the fitness distribution and making it more Gaussian: see the supplementary figures in that work.
+In asexual models, there is no equivalent.
+Fitness distributions tend to be even more fat-tailed and the nose dynamics are even more critical.
+(Technically, of course, even the concept of a Gaussian fitness distribution, especially in an asexual population, will be an abstraction: the "true" fitness distribution will be a series of delta functions which ensemble-averages out to an ideal form.
+That ideal form is less Gaussian in asexual populations than in sexual ones.) We have amended the text to be clearer on this front.
+
 \begin{reviewerquote}
@@
 
+We agree that this wording is confusing. We have fixed it to clarify that we are explicitly referring to selective sweeps acting on loci of strong effect ($Ns > 1$).
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+We agree.
+The fact that selection coefficients of background mutations are altered every generation so as to keep the fitness variance fixed means that adding an explicit mutation term, featuring a fixed distribution of selection coefficients, is difficult.
+In our \texttt{WaveFit} simulations, we explicitly force all background mutations to have an identical fitness effect $\beta$ (which is scaled up or down depending on the instantaneous fitness variance each generation).
+However, by construction, the value of $\beta$ is irrelevant, including whether it is positive or negative.
+So in a sense, the mutation term can be papered over: one can simply imagine that the distribution of fitness effects is symmetric around 0, so that different mutational contributions cancel out.
 
@@
 
+The choice of a 200-locus genome, with background mutations being injected randomly to keep each locus polymorphic, was made due to our use of \texttt{FFPopSim}.
+We have re-performed most of the relevant simulations with \texttt{WaveFit}, which allows for an explicit infinite-sites model and allows the background mutation rate to be controlled.
+Varying the selection coefficients is unfortunately necessary to keep the background fitness variance fixed.
+
 \begin{reviewerquote}
@@
 
+We fix $s$ to be larger in the case of the valley than in the sweep to model a situation wherein a population is forced to "choose" between an immediately available, easily accessible, but weak adaptation or a difficult-to-access but stronger adaptation that is hidden by a fitness valley.
+This parallels Ochs and Desai (2013).
+
 \begin{reviewerquote}
@@
 
+We agree. We have fixed this.
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+We have fixed this.
 
@@
 \end{reviewerquote}
+
+We agree that the null expectation is not really clear in this case. We have amended the text accordingly.
+
+Neher and Shraiman (2011) determined that the opposite trend applies in the case where $1/N < \delta \ltrsim sigma$.
+In this case, the stochastic effects of genetic draft can propel a deleterious mutation to high frequencies, allowing valley crossing to occur more quickly.
+The bubble size distribution, which determines the number of opportunities there are for a successful complex adaptation to appear and establish, drops off more quickly in the case of draft ($1/n^2$) than drift ($1/n$).
+So the overall rate of crossing should be higher in the draft case \emph{for neutral intermediates, or intermediates of small effect}.
+When the effect is large, the drift bubble size distribution is harshly cut off around $1/\delta^2$.
+The draft bubble size distribution is not.
+So the crossing rate is \emph{lower} in the case of neutral or weak intermediates.
+In addition, in a drafting population, clonal interference/multiple mutation effects retard the growth and fixation of the complex adaptation once it appears.
+So that, too, slows down the crossing rate in the draft case.
 
@@
 \end{reviewerquote}
+
+We agree.
 
@@
 \end{reviewerquote}
+
+By design, whether background mutations are mostly deleterious or mostly beneficial does not matter in our simulations, since the fitness coefficients are adjusted each generation so as to keep the fitness variance fixed.
+Similar results have been observed for wide-fitness-variance populations in which the background fitness is due primarily to beneficial or deleterious mutations.
 
@@
 
+We are not sure what the issue is here.
+Each selection coefficient is exponentially distributed, but the overall fitness distribution is the sum of the effects of a large number of mutations--by the central limit theorem, it turns out to be roughly Gaussian.
+We agree, however, that the mathematical background takes up too much of the paper.
+
 \begin{reviewerquote}
@@
 
+We have clarified this: that procedure is followed only at the \emph{background} loci, not the \emph{focal} loci which will ultimately contribute to the complex adaptation.
+At those loci, mutations are introduced at a fixed rate, so indeed N does matter.
+In addition, in our \texttt{WaveFit} simulations, this is no longer an issue.
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+We carry out our simulations until a pre-ordained number of valley crossings has occurred.
+But there is no in principle reason why we could not report the probability of crossing instead, with the caveat that the crossing probability is often low, meaning this is likely to be noisily estimated.
 
@@
 
+We have performed simulations to probe this.
+
 \begin{reviewerquote}
@@
 \end{reviewerquote}
+
+Fixed.
 
@@
 
+We have clarified this.
+
 \begin{reviewerquote}
@@
 
+We have clarified this.
+The differences between our results and those of Ochs and Desai are somewhat minor.
+One model difference is that theirs concerned explicit \emph{competition} between a simple sweep and a valley-crossing adaptation.
+Our model treats these cases separately.
+In addition, there was no draft in the Ochs-Desai simulations.
+
 \end{document}

### 2019-02-12 ms/valley_crossing.tex
methods section edits

@@
-\documentclass[10pt,twocolumn,twoside]{gsajnl}
+\documentclass[rmp]{revtex4}
 \articletype{inv} % article type: {inv} Investigation
@@
 This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
-Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
-If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
+Evolution will proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
+If the individual fitness of each genotype is visualized as a surface or landscape wherein the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
 However, many empirical fitness landscapes are not completely smooth and have multiple peaks \citep[reviewed in][]{Szendro:Schenk:2013,Visser:Krug:2014,Obolski:Ram:2018}.
@@
 This is the sequential fixation regime \citep{weissman_2009}.
-A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
+A third possibility occurs for intermediate population sizes wherein wild type individuals generate transient mutant subpopulations or ``bubbles".
 These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
 On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}.
-However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}.
+However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes in which tunneling occurs \citet{ochs_2015}.
 This suggests that tunneling may be the most difficult mode of valley crossing.
@@
 Infrequent recombination decreases valley crossing times \citep{weissman_2010}.
-At high values of the recombination rate, however, individuals carrying the full complex adaptation outcross with the wild type and produce more deleterious intermediates that retard the growth of the multiple mutant.
+At high values of the recombination rate, however, individuals carrying the full complex adaptation outcross with the wild type and produce deleterious intermediates that retard the growth of the multiple mutant.
 An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
@@
 In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare (close to frequency zero) or common (close to frequency one).
-Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982} where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
+Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982}, where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
 When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by genetic drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
 Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
-Draft is fundamentally different from drift: for example, drafting populations experience large jumps in allele frequencies that cannot be encapsulated by a diffusion approximation \citep{neher_shraiman_2011}.
-They commonly have genealogies in which more than two branches merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
+The standard technique of defining a reduced ``effective population size" fails to capture several important qualitative differences between drift and draft.
+Draft is fundamentally different from drift: for example, new mutations in a drafting population can rapidly be brought to high frequency, and the offspring number variance almost diverges \citep{neher_shraiman_2011}.
+Drafting populations commonly have genealogies in which more than two branches merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
 Additionally under genetic draft, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency one--or, equivalently, a depression in intermediate frequency alleles, which are quickly swept out of the population \citep{neher_shraiman_2011,kosheleva_2013,neher_hallatschek_2013}.
@@
 Draft likewise may have profound effects on the evolution of complex adaptations. Preliminary evidence for this comes from
-\citet{neher_shraiman_2011}, who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting population.
-They found that, compared to drift, draft generally causes the distribution of bubble sizes to drop off more quickly, meaning that small bubbles are more common but large bubbles are very rare.
+\citet{neher_shraiman_2011}, who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting facultatively sexual population.
+They found that, compared to drift, draft generally causes the distribution of bubble sizes to drop off more quickly, meaning that bubbles tend to be smaller overall: however, as the mutant lineage's fitness is determined not only by a mutation's intrinsic fitness effect but also the genetic background on which it appears, the fitness effect of a focal mutation plays less of a role in capping the size of the lineage.
 Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately obvious.
@@
 The complicating factor is that, in an asexual population, the dynamics are governed almost entirely by what happens in the nose of the fitness distribution, and these dynamics can be highly stochastic.
-In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011, neher_shraiman_2009}.
-We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are difficult to obtain.
+In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011, neher_shraiman_2009}: the critical nose dynamics are still important in a sexual population, but the fitness distribution is more Gaussian overall.
+We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are very difficult to obtain.
 We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
-However, this is consistent with the fact that \emph{all} forms of adaptation are slowed down due to clonal interference and genetic background effects.
+This is consistent with the fact that clonal interference and genetic background effects retard the growth of fit mutant lineages and make a given selective sweep less likely. (reword this later)
 In addition, we confirm that in rapidly adapting populations, adaptive fixations of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
@@
 By Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, the rate $v$ at which the mean fitness advances due to selection is set to $\sigma^{2}$.
-If $x$ is the background fitness of a lineage, then the distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
-Given this background fitness variance, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.
+Let $x$ be the background fitness of a lineage.
+The distribution of background fitnesses $f(x)$ will generally be a series of (Dirac) $\delta$ functions, but its ensemble average will be roughly Gaussian \citep{good_desai_2014}: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
+Given this background fitness distribution, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.
 We emphasize that $\sigma^2$ is only a \emph{background} fitness variance and does not include the effects of the focal loci.
@@
 
-To analyze our model, we perform forward simulations using a customized version of \texttt{FFPopSim} \citep{zanini_2012}, a discrete time forward simulation package that implements a modified Wright-Fisher model.
+To analyze our model, we perform forward simulations using the custom Julia module \texttt{WaveFit} that implements a modified Wright-Fisher model.
 The simulation code is available via \texttt{GitHub}.
-In \texttt{FFPopSim}, individuals are organized by ``clones" (sets of individuals with the same genome): each individual gives rise to a Poisson distributed number of offspring, with the mean dependent on the relative fitness.
-The mean offspring number is further adjusted to keep the population roughly at a pre-specified carrying capacity.
-We initialize a wild type population of $N$ individuals consisting of haploid genomes of with $L = 200$ loci.
-Genetic loci are partitioned into two groups, the two ``focal" loci where the epistatic alleles $A$ and $B$ will segregate and the remaining ``background" loci that are responsible for the underlying fitness distribution.
-We set the focal loci to be at positions $L/4$ and $3L/4$.
-For the background loci, we make use of a modified infinite sites model \citep{Kimura:1971,Watterson:1975}: any time a locus becomes monomorphic, a mutation at that locus is injected into a random individual in the population.
-In this way, the population experiences a constant influx of beneficial mutations at the background loci.
-These mutations' fitness effects are drawn from an exponential distribution; the background fitness variance is set to a constant value $\sigma^2$ by manually adjusting the selection coefficients every generation (they are multipled by the current variance and divided by $\sigma^2$).
-In this way, the mutation rate $U_b$ and average fitness effect $\bar{\epsilon}$ at the background loci are not parameters of our simulation model but must be directly measured from simulations; i.e., they are constrained by $L$ and $\sigma$.
-We estimate them by manually counting the number of injected mutations every generation and averaging over the fitness effect at each locus.
+In \texttt{WaveFit}, individuals are organized by ``clones" (sets of individuals with the same genome): each clone gives rise to a Poisson distributed number of offspring, with the mean dependent on the relative fitness $x-\bar{x}$.
+The mean offspring number is further adjusted to keep the population roughly at a pre-specified carrying capacity $K$.
+In addition, each individual receives a Poisson distributed number of mutations: if $L$ is the number of sites in the genome and $U$ is the (background) per-site mutation rate, then $UL$ is the mean number of mutations per individual per generation (and is a parameter in our simulations).
+In effect, this is an ``infinite sites" model \citep{Kimura:1971,Watterson:1975}.
+All background loci are assigned a common fitness effect $\epsilon$, which is scaled up or down each generation so as to keep the background fitness standard deviation $\sigma$ constant.
+This rescaling means that the assigned initial value of $\epsilon$ turns out not to matter, even if it is negative or even if both deleterious and beneficial background mutations are permitted.
+That feature of our model may seem counterintuitive, but previous work has found that populations experiencing strong background selection at many loci of weak effect (see, e.g., \citet{cvijovic_2015}) exhibit dynamics very similar to rapidly adapting populations \citep{desai_fisher_2007, neher_shraiman_2011}.
 
-Simulations are allowed to proceed for an equilibration time of $N/10$ generations to allow sufficient genetic diversity to be introduced.
-During this time, the mutation rate $\mu$ for the focal loci is set to zero.
-After equilibration, we set the per-site mutation rate $\mu$ for the focal loci (allowing forward and backward mutations) with single mutant fitness disadvantage $\delta$ and double mutant advantage $s$.
-When the double mutant reaches frequency $0.5$, we consider it to have fixed.
-We then set the focal loci allele frequencies back to zero, randomize the selection coefficients for the genetic background by drawing anew from an exponential distribution, and allow the population to equilibrate for another $N/10$ generations.
-This ensures that independent valley crossing trials are independent despite occurring in the same simulation run.
+In addition to the background loci, individuals are assigned values at each of two ``focal loci" which will ultimately contribute to the complex adaptation.
+Individuals that carry a mutation at only one focal locus have a fitness detriment $\delta$: individuals that carry mutations at \emph{both} focal loci have a fitness advantage $s$.
+$\delta$ and $s$ are \emph{not} rescaled as $\beta$ is.
+Forward mutations at the focal loci occur at rates $\mu_1$ and $\mu_2$: mirroring the convention of \citet{weissman_2009}, $\mu_1$ is the mutation rate from the wild type \emph{at the first locus}, and $\mu_2$ is the mutation rate \emph{at the second locus, for individuals who are already mutant at the first locus} (these mutations are assigned separately, so that it is possible for a double mutant to arise from the wild type in a single generation).
+We record that a crossing event has occurred when the complex adaptation (i.e., the frequency of a mutation at the second focal locus) reaches frequency $0.5$.
+
+For all simulations, we initialize a homogeneous wild type population (in which each individual has zero background mutations) of size $N = K$.
+The precise value of $UL$ seems not to affect the valley crossing dynamics provided it is large enough for the ensemble average of the fitness distribution to be roughly Gaussian.
+Accordingly, we set $UL = 0.1$, as a value less than $1$ cuts down on the number of operations needed for the mutation step.
+We permit the population to equilibrate for $N/10$ generations.
+While this is usually less than the coalescence time, it is long enough for the average number of clones to reach its approximate equilibrium value (at which the creation of new clones by mutation roughly cancels out the extinction of unfit clones).
+After the population has equilibrated, we set the mutation rates at the focal loci to their pre-selected values $\mu_1$ and $\mu_2$.
 

### 2019-02-25 ms/valley_crossing.tex
started rewrite of theory/methods

@@
 Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
-We focus on two loci, which are initially fixed for alleles $a$ and $b$.
-These alleles mutate to $A$ and $B$ at rates $\mu_A$ and $\mu_B$, respectively (we neglect back mutations).
-Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_A$, $1-\delta_B$, and $1+s$ respectively, with $\delta \geq 0$ (intermediate genotypes are deleterious) and $1/N < s < 1$ (double mutant $AB$ is strongly beneficial); see \ref{fig:landscape}.
-For the remainder of our analysis, we will assume that $\mu_A = \mu_B = \mu$ and $\delta_A = \delta_B = \delta$.
-We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$.
-By Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, the rate $v$ at which the mean fitness advances due to selection is set to $\sigma^{2}$.
+A complex adaptation can occur in the population via mutations at two ``focal" loci (which are treated separately from the background loci).
+The mutation rate is $\mu_1$ at the first focal locus and $\mu_2$ at the second locus.
+Individuals who are mutant at one locus and wild-type at the other have reduced fitness relative to the wild type, but the double mutant state corresponds to a ``complex" adaptation that increases fitness.
+For simplicity, we assume that individuals must mutate at the first locus \emph{before} they can mutate at the second locus, so that $\mu_1$ is the mutation rate into the deleterious intermediate state and $\mu_2$ is the mutation rate toward the full complex adaptation.
+The fitnesses of these states are $1-\delta$ and $1+s$, respectively, with $\delta \leq 0$ and $1/N < s \ll 1$ (so-called ``strong selection").
+This setup closely mirrors that of \citep{weissman_2009}.
+
+We further assume that the large number of background loci of weak effect contribute to a constant background fitness variance $\sigma^2$.
 Let $x$ be the background fitness of a lineage.
-The distribution of background fitnesses $f(x)$ will generally be a series of (Dirac) $\delta$ functions, but its ensemble average will be roughly Gaussian \citep{good_desai_2014}: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
-Given this background fitness distribution, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.
+If the number of background loci is large, the distribution of background fitnesses $f(x)$ will generally be a series of (Dirac) $\delta$ functions, but its ensemble average will be roughly Gaussian \citep{good_desai_2014}: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
+By Fisher's ``fundamental theorem", the mean fitness $\bar{x}$ will advance due to selection at a rate equal to $\sigma^2$: the total rate of advance will depend also on contributions from drift and the average effect of mutations.
+
+Given this background fitness distribution, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for a beneficial double mutant at the focal loci (a ``complex adaptation'') to arise and fix in the population.
 We emphasize that $\sigma^2$ is only a \emph{background} fitness variance and does not include the effects of the focal loci.
-For the most part, we assume that the focal loci are either at low enough frequency that they do not substantially affect the total fitness variance or that they are on the way to fixation and the population is destined to become roughly monomorphic (until new beneficial background mutations can be introduced).
+For the most part, we assume that mutations at the focal loci are either at low enough frequency, so that they do not substantially affect the total fitness variance, or that they are on the way to fixation, and hence the population is destined to become fixed at those loci.
 
-In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a double mutant to sweep through the population.
-At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
-We zero in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
-We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, stochastic tunneling, and semi-deterministic tunneling cases.
+\citet{weissman_2009} divided the possible modes of asexual valley crossing into several distinct regimes.
+We classify them, broadly, as \emph{sequential fixation}, \emph{deterministic fixation}, and \emph{tunneling}.
+What follows is a brief outline of the process of valley crossing in each of these regimes, as well as how these are likely to differ in our model.
+We then present results for valley crossing times using forward-time simulations across a range of parameter values that includes each of these major regimes.
+
+The \emph{sequential fixation} regime arises at small population sizes.
+The wait time $\mathbb{E}\left[ T\right]$ can be partitioned into two components: the time $\tau_1$ until an intermediate mutation arises \emph{that is destined to fix} and the time $\tau_2$ until the complex adaptation arises and is likewise destined to fix.
+$\tau_1$ is one divided by the rate $N\mu_1 p$ at which intermediate mutants arise that are destined to fix: on average, $N\mu_1$ mutants appear every generation, and the diffusion approximation yields $p = (e^\delta - 1)/(e^{N\delta} - 1)$.
+$\tau_2$, similarly, is approximately one divided by $N\mu_2 s$ and in general will be dwarfed by $\tau_1$.
+When genetic draft is more important than genetic drift, $p$ in $\tau_1$ takes on a much more complicated form similar to that in \citet{good_desai_2014}: in general, the dependence on $\delta$ is lessened provided that $\sigma \gtrsim \delta$, as deleterious mutations that arise on significantly fitter-than-average backgrounds can be destined to fix.
+$\tau_2$ likewise has lessened dependence on $s$, as a complex adaptation that does \emph{not} arise on a fitter-than-average background lineage may be doomed to extinction.
 
@@
 The precise value of $UL$ seems not to affect the valley crossing dynamics provided it is large enough for the ensemble average of the fitness distribution to be roughly Gaussian.
-Accordingly, we set $UL = 0.1$, as a value less than $1$ cuts down on the number of operations needed for the mutation step.
+Accordingly, we set $UL = 0.1$, as a value less than $1$ reduces the number of operations needed for the mutation step.
 We permit the population to equilibrate for $N/10$ generations.

### 2019-03-14 ms/valley_crossing.tex
stochastic tunneling rewrite

@@
-\documentclass[rmp]{revtex4}
+\documentclass[10pt,twocolumn,twoside]{gsajnl}
 \articletype{inv} % article type: {inv} Investigation
@@
 As in the introduction, we refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a mutant ``bubble": the time integrated number of individuals $W$ is the ``weight" of the bubble.
-During the lifetime of a bubble, the complex adaptation must appear and establish; that is, it must rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu \pfix$ where $\pfix$ is the probability that the double mutant reaches fixation from a single initial individual.
 If it establishes, the complex adaptation will quickly sweep to fixation.
@@
 In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
-We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term: the time required for $T_1$ to establish will be of order $1/s$ \citep{desai_fisher_2007}, and the sweep time will be of order $1/s \log (Ns)$, but $T_0$ can be many orders of magnitude greater.
+We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term: in the drift case, the time required for $T_1$ to establish will be of order $1/s$ \citep{desai_fisher_2007}, and the sweep time will be of order $1/s \log (Ns)$, but $T_0$ can be many orders of magnitude greater.
 
+This process is well understood in the drift case, and at first brush it seems possible to extend this understanding to the draft case.
 Most of the time, a bubble will simply arise and go extinct.
@@
 Computing the expected value of this quantity over all bubble sizes, $\phi = \left< 1-e^{\mu \pfix W} \right>$, gives the probability that some bubble will lead to a successful valley crossing.
-One difficulty in computing the rate of tunneling in the high $N\sigma$ case (where genetic draft is significant) is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
+In the draft case, the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
 That is to say, it is not possible to simply compute $\pfix$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
 Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals give rise to the complex adaptation, it will therefore likewise be more likely to fix.
-Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $\pfix$ will depend on $x$ and the peak height $s$.
-These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\upPhi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
-The expected valley crossing time then becomes $1/N\mu\upPhi$.
-An additional difficulty is that technically, $\pfix$ will depend not on the fitness of the initial lineage in a bubble but on the fitness of the specific background on which the double mutant arose.
-These will not necessarily be the same; for example, individuals within the single mutant lineage may accumulate additional background mutations.
-Assuming that this does \emph{not} happen, that is, assuming that $\pfix$ depends on the fitness $x$ of the founding single mutant lineage, is tantamount to assuming that bubbles are too short lived to accumulate background mutations that significantly affect their fitness.
+Thus, the distribution of $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $\pfix$ will depend on $x$ and the peak height $s$.
+Once $\phi$ is obtained, it will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\upPhi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
+The expected valley crossing time would then be $1/N\mu\upPhi$.
 
+Unfortunately, a number of complicating factors prevent a full analytical description of this process.
+First, $\pfix$ will not depend solely on the background fitness $x$ of a lineage.
+Individuals in each lineage will accumulate mutations, broadening the fitness distribution \emph{of that lineage}, so $\pfix$ will depend specifically on the background fitness of the precise individual within said lineage that gives rise to the complex adaptation.
+This issue can be swept aside by assuming that bubbles do not last long enough for their own distribution of fitnesses to become very broad.
+The second issue is somewhat more severe: there is no principled way to compute the Laplace transform of the bubble size distribution.
+Previous approaches such as \citep{weissman_2009} and \citep{neher_shraiman_2011} have relied on branching process approximations.
+\citep{neher_shraiman_2009} in particular were able to derive the asymptotics of the bubble size distribution in a facultatively sexual population: they began by writing down a master equation for the distribution of bubble sizes with a birth rate that decreases in time as the mean fitness advances.
+However, they relied on a recombination model that allowed them to directly obtain the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to explicitly solve the branching process model; our model being an asexual one, this option is not available to us.
+In addition, they evaded the need to send time to infinity in the model.
+In principle this is necessary to integrate over all possible bubble sizes: but in practice, because the birth rate of a lineage decreases with time, this would yield the absurd result that $\phi$ itself decreases with time.
+Since it seems unlikely that analytical approaches will yield a principled way of approximating the valley crossing time in the drift case, we focus on simulation methods to estimate the valley crossing time.
+
+\iffalse
 We begin by considering the dynamics of $\pfix(x,s)$ before moving on to consider the distribution of bubble sizes.
@@
 Therefore, we focus on simulation methods to study the dynamics of valley crossing.
+\fi
 

### 2019-03-15 ms/valley_crossing.tex
further rewrites, getting figures in place

@@
 Previous approaches such as \citep{weissman_2009} and \citep{neher_shraiman_2011} have relied on branching process approximations.
-\citep{neher_shraiman_2009} in particular were able to derive the asymptotics of the bubble size distribution in a facultatively sexual population: they began by writing down a master equation for the distribution of bubble sizes with a birth rate that decreases in time as the mean fitness advances.
+\citep{neher_shraiman_2009} in particular were able to derive the asymptotics of the bubble size distribution in a facultatively sexual population: they began by deriving a master equation for the distribution of bubble sizes with a birth rate that decreases in time as the mean fitness advances.
 However, they relied on a recombination model that allowed them to directly obtain the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to explicitly solve the branching process model; our model being an asexual one, this option is not available to us.
+\iffalse
 In addition, they evaded the need to send time to infinity in the model.
 In principle this is necessary to integrate over all possible bubble sizes: but in practice, because the birth rate of a lineage decreases with time, this would yield the absurd result that $\phi$ itself decreases with time.
+\fi
 Since it seems unlikely that analytical approaches will yield a principled way of approximating the valley crossing time in the drift case, we focus on simulation methods to estimate the valley crossing time.
@@
 Finally, we compared the dynamics of valley crossing versus sweeping mutations.
-We performed simulations with a sweeping focal beneficial mutation, with selection coefficient $s_{\mathrm{sweep}}$, at position $L/2$, with the background loci kept polymorphic as in the valley crossing simulations.
+We performed simulations with a sweeping focal beneficial mutation, with selection coefficient $s_{\mathrm{sweep}}$, and a constant influx of background beneficial mutations of weak effect as in the valley crossing simulations.
 We recorded the time for a beneficial mutation to arise and reach frequency $0.5$, considering it to have fixed at that point.
@@
 \section{Results}
+
+\begin{figure}[t]
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\end{subfigure}
+\begin{subfigure}[c]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\end{subfigure}
+\begin{subfigure}[d]{0.4\textwidth}
+\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\end{subfigure}
+\caption{Valley crossing times across a range of parameter values mirroring those of \citet{weissman_2009}.
+In figures (a) and (b), the population size is varied: increasing the size causes valley crossing to transition from tunneling to semi-deterministic fixation.
+In figure (c), the valley depth $\delta$ is varied: at negative $\delta$, our simulations closely match the expectation for beneficial mutations.
+In figure (d), the intermediate mutation rate $\mu_2$ is varied.}
+\label{fig:weissman}
+\end{figure}
 

### 2019-05-07 ms/valley_crossing.tex
further manuscript edits

@@
 Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
-If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
+If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated \emph{de novo}.
 This is sometimes called deterministic fixation of the multiple mutant \citep{weissman_2009}.
@@
 In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare (close to frequency zero) or common (close to frequency one).
-Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population is described well by the classic Kingman coalescent \citep{Kingman:1982}, where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
+Alleles whose dynamics are primarily determined by genetic drift can be effectively modeled by a diffusion approximation \citep{Wright:1945,Kimura:1955,Kimura:1957}, and the ancestry of a population in the absence of selection is described well by the classic Kingman coalescent \citep{Kingman:1982}, where pairs of branches in the genealogy merge back in time until the most recent common ancestor.
 When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by genetic drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
 Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
-The standard technique of defining a reduced ``effective population size" fails to capture several important qualitative differences between drift and draft.
-Draft is fundamentally different from drift: for example, new mutations in a drafting population can rapidly be brought to high frequency, and the offspring number variance almost diverges \citep{neher_shraiman_2011}.
-Drafting populations commonly have genealogies in which more than two branches merge at once and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
+The standard technique of defining a reduced ``effective population size" fails to capture several fundamental qualitative differences between drift and draft.
+For example, new mutations in a drafting population can rapidly be brought to high frequency, and the variance in offspring number almost diverges \citep{neher_shraiman_2011}.
+Drafting populations commonly have genealogies in which more than two branches merge at once (or almost at once) and that are better described by alternative coalescent processes such as the Bolthausen-Sznitman coalescent \citep{neher_hallatschek_2013, brunet_2007, schweinsberg_2017}.
 Additionally under genetic draft, the frequency spectrum of neutral alleles is non-monotonic, with a marked uptick near frequency one--or, equivalently, a depression in intermediate frequency alleles, which are quickly swept out of the population \citep{neher_shraiman_2011,kosheleva_2013,neher_hallatschek_2013}.
@@
 We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are very difficult to obtain.
-We show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
-This is consistent with the fact that clonal interference and genetic background effects retard the growth of fit mutant lineages and make a given selective sweep less likely. (reword this later)
+We expand on the work of \citet{neher_shraiman_2011} by considering not only stochastic tunneling but also sequential fixation; in fact, we find that, over a significant chunk of parameter space, stochastic tunneling effectively \emph{does not happen} in drafting populations, and sequential fixation appears to be the dominant mode of valley crossing.
+We also show that fitness valley crossing occurs at an overall lower rate in rapidly adapting populations.
+This is consistent with the fact that clonal interference and genetic background effects retard the growth of fit mutant lineages, slowing the expansion of a beneficial mutation.
 In addition, we confirm that in rapidly adapting populations, adaptive fixations of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
@@
 The second issue is somewhat more severe: there is no principled way to compute the Laplace transform of the bubble size distribution.
-Previous approaches such as \citep{weissman_2009} and \citep{neher_shraiman_2011} have relied on branching process approximations.
-\citep{neher_shraiman_2009} in particular were able to derive the asymptotics of the bubble size distribution in a facultatively sexual population: they began by deriving a master equation for the distribution of bubble sizes with a birth rate that decreases in time as the mean fitness advances.
+Previous approaches such as \citet{weissman_2009} and \citet{neher_shraiman_2011} have relied on branching process approximations.
+The latter in particular were able to derive the asymptotics of the bubble size distribution in a facultatively sexual population: they began by deriving a master equation for the distribution of bubble sizes with a birth rate that decreases in time as the mean fitness advances.
 However, they relied on a recombination model that allowed them to directly obtain the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to explicitly solve the branching process model; our model being an asexual one, this option is not available to us.
@@
 
-To analyze our model, we perform forward simulations using the custom Julia module \texttt{WaveFit} that implements a modified Wright-Fisher model.
+To analyze our model, we perform forward simulations using the custom Julia module \texttt{WaveFit}, which implements a modified Wright-Fisher model.
 The simulation code is available via \texttt{GitHub}.
@@
 
-For all simulations, we initialize a homogeneous wild type population (in which each individual has zero background mutations) of size $N = K$.
-The precise value of $UL$ seems not to affect the valley crossing dynamics provided it is large enough for the ensemble average of the fitness distribution to be roughly Gaussian.
-Accordingly, we set $UL = 0.1$, as a value less than $1$ reduces the number of operations needed for the mutation step.
-We permit the population to equilibrate for $N/10$ generations.
-While this is usually less than the coalescence time, it is long enough for the average number of clones to reach its approximate equilibrium value (at which the creation of new clones by mutation roughly cancels out the extinction of unfit clones).
+For all simulations, we initialize a homogeneous wild type population (in which each individual has zero background mutations) of size $N = K$, with $\mu_1 = \mu_2 = 0$.
+The precise value of $UL$ seems not to affect the valley crossing dynamics provided it is large enough for the ensemble average of the fitness distribution to be roughly Gaussian (see appendix).
+Accordingly, we set $UL = 1.0$.
+We then permit the population to equilibrate for $N/10$ generations.
+While this is usually less than the coalescence time, it is long enough for the average number of clones to reach its approximate equilibrium value (at which the creation of new clones by mutation roughly cancels out the extinction of unfit clones); see appendix.
 After the population has equilibrated, we set the mutation rates at the focal loci to their pre-selected values $\mu_1$ and $\mu_2$.
@@
 We partitioned our simulations as follows.
-We first studied the effect of increasing $\sigma$ in the sequential fixation, stochastic tunneling, semi-deterministic tunneling, and deterministic fixation regimes \citep{weissman_2009}.
+We first considered the parameter regimes modeled in \citet{weissman_2009}, which were previously studied solely in the presence of genetic drift, and introduced genetic draft by turning up $\sigma$.
+In the drift case, the dynamics are as follows: we comment on important differences for the draft case in the results section.
 For small populations, sequential fixation applies when $N \ll \min(1/\delta, 1/\sqrt{\mu s})$, so that a successful mutant is unlikely to arise until the intermediate has fixed, but the intermediate's fitness disadvantage is too weak for selection to inhibit it.
@@
 
+Next, we directly examined the effects of varying both $\sigma$ and $\delta$ on the valley crossing time.
+
 Finally, we compared the dynamics of valley crossing versus sweeping mutations.
@@
 We then computed the ratio of the sweep time to the valley crossing time for a complex adaptation, with valley depth $\delta$ and larger fitness advantage $s_{\mathrm{valley}}$.
-To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation increases the likelihood of valley crossing relative to the speed of simple sweeps.
+To determine the effect of $\sigma$ on this ratio, we performed simulations at both high and low $\sigma$ and computed the ratio of these ratios: the result is a measure of the extent to which rapid adaptation facilitates valley crossing rather than simple selective sweeps.
 
@@
 
-\begin{figure}[t]
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
+\begin{figure*}[t]
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{Figures/julia_weissman_5a.pdf}
 \end{subfigure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{Figures/julia_weissman_5b.pdf}
 \end{subfigure}
-\begin{subfigure}[c]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\begin{subfigure}[c]{0.5\textwidth}
+\includegraphics[width=\textwidth]{Figures/julia_weissman_5c.pdf}
 \end{subfigure}
-\begin{subfigure}[d]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
+\begin{subfigure}[d]{0.5\textwidth}
+\includegraphics[width=\textwidth]{Figures/julia_weissman_5d.pdf}
 \end{subfigure}
@@
 In figure (c), the valley depth $\delta$ is varied: at negative $\delta$, our simulations closely match the expectation for beneficial mutations.
-In figure (d), the intermediate mutation rate $\mu_2$ is varied.}
+In figure (d), the intermediate mutation rate $\mu_2$ is varied.
+Dotted lines represent the theoretical predictions from \citet{weissman_2009}.}
 \label{fig:weissman}
-\end{figure}
+\end{figure*}
 
+\iffalse
 \begin{figure}[t]
@@
 \end{figure}
+\fi
 
 Here we present the key results of our simulation analysis.
+Figure \ref{fig:weissman} summarizes the major differences between the drift (low $\sigma$) and draft (high $\sigma$) regimes in valley crossing.
+Note that the theoretical predictions differ slightly from those of \citet{weissman_2009} due to their use of a Moran model, compared to our Wright-Fisher model: this is comparable to an order of 2 difference in the population size.
+The general trend is that draft causes valley crossing to take more time but be less sensitive to the underlying parameter values.
+In figure (a), tunneling at low $\sigma$ occurs via neutral sequential fixation at low $N$ before transitioning into neutral tunneling and semi-deterministic tunneling at higher $N$; these are visible as minor kinks in the theoretical predictions (dotted line).
+At low $\sigma$, our results agree with those predictions, but as $\sigma$ increases, the dependence on $N$ is lessened: at higher $\sigma$, the population effectively behaves as though there is a very small $N_e$, resulting in no dependence on $N$.
+These trends are also visible in figure (c) where, in the drift case, the population transitions from neutral to deleterious tunneling, and figure (d), in which deleterious tunneling gives way to neutral tunneling as the mutation rate $\mu_2$ increases.
+
+One remarkable difference between drift and draft can be seen in figure (b).
+At $\sigma = 10^{-8}$, crossing proceeds via deleterious sequential fixation at low $N$, then transitions into deleterious tunneling around $N = 10^3$ and semi-deterministic tunneling at higher $N$.
+The increase in crossing time with $N$ is a hallmark of deleterious sequential fixation: higher $N$ means the deleterious intermediate is more ``visible" to selection, slowing its overall growth.
+Once the population becomes large enough that fixation of the intermediate is unlikely, raising $N$ increases the number of single mutant ``bubbles" that are generated, resulting in the $\tau \sim N^{-1}$ crossing time dependence.
+As $\sigma$ approaches even the moderate value $10^{-2}$, the crossing time curve flattens significantly; it still positively depends on $N$ but never approaches the peak crossing time $\approx 10^7$ generations in the $\sigma = 10^{-8}$ case.
+The behavior in these two cases is qualitatively different: the intermediate $\sigma$ curve's dependence on $N$ cannot be reduced to the drift case simply by defining a reduced $N_e$.
+In effect, from the point of view of the focal mutations, draft flattens the fitness landscape somewhat but not completely.
+At higher values of $\sigma$, $\tau$ is again mostly independent of $N$, an indicator that sequential fixation is the dominant mode of crossing.
+
+\iffalse
 In general, we expect valley crossing to require more time in rapidly adapting populations (large $N \sigma$) experiencing genetic draft; bubbles are doomed to extinction more quickly (lineages experience larger fluctuations in size and decline in frequency as the mean fitness advances) and beneficial lineages are less likely to fix, especially when valleys are shallow.
@@
 In the case of stochastic tunneling, the overall size distribution of bubble sizes is reduced as well: the distribution $P(w)$ scales not as $w^{-3/2}$ but as $w^{-2}$ \citep{neher_shraiman_2011}, so there are fewer chances for the complex adaptation to arise on a mutant background and fix.
+\fi
 
 \begin{figure}[t]
-\includegraphics[width=0.4\textwidth]{Figures/var_sigma_delta.pdf}
+\includegraphics[width=0.4\textwidth]{Figures/julia_sigma_delta.pdf}
 \caption{Valley crossing times for varied $\sigma$ and $\delta$.
-In these simulations, $N = 10^5$, $\mu = 3 \times 10^{-6}$, and $s = 10^{-2}$.
+In these simulations, $N = 10^5$, $\mu_1 = 3 \times 10^{-6}$, and $\mu_2 = 10^{-4}$.
 Note the lack of dependence on $\delta$ at high $\sigma$.}
@@
 
-\begin{figure}[t]
-\begin{subfigure}[b]{0.475\textwidth}
-  \includegraphics[width=\textwidth]{Figures/tau_compare_crop.pdf}
+\begin{figure*}[t]
+\begin{subfigure}[b]{0.5\textwidth}
+  \includegraphics[width=\textwidth]{Figures/tau_lowsig.pdf}
+\end{subfigure}
+\begin{subfigure}[b]{0.5\textwidth}
+  \includegraphics[width=\textwidth]{Figures/tau_highsig.pdf}
+\end{subfigure}
+\label{fig:tau}
+
   \caption{Valley crossing times on a $\log_{10}$ scale}
-  \label{fig:tau}
+\begin{subfigure}[b]{0.5\textwidth}
+  \includegraphics[width=\textwidth]{Figures/ratio_lowsig.pdf}
 \end{subfigure}
-\begin{subfigure}[b]{0.475\textwidth}
-  \includegraphics[width=\textwidth]{Figures/alpha_ratios_crop.pdf}
-  \caption{Ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ on a $\log_{10}$ scale}
-  \label{fig:ratios}
+\begin{subfigure}[b]{0.5\textwidth}
+  \includegraphics[width=\textwidth]{Figures/ratio_highsig.pdf}
 \end{subfigure}
 \caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
-In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, and $\mu = 10^{-5}$.
+In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, $\mu_1 = 3 \times 10^{-6}$, and $\mu_2 = 10^{-4}$.
 Panel (a) shows that valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
@@
 \label{fig:tau-ratios}
-\end{figure}
+\end{figure*}
 
@@
 \begin{figure*}[t]
-\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$.
+\includegraphics[width=0.95\textwidth]{Figures/julia_ratios.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2= 5 \times 10^{-2}$), for varied values of $\mu$.
 Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
@@
 
-We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha_{\sigma_2} / \alpha_{\sigma_1}$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha_{\sigma_2} / \alpha_{\sigma_1}$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2=5 \times 10^{-2}$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
 The ratio can be seen in Figure \ref{fig:alpha_ratio}.

### 2019-05-10 ms/valley_crossing.tex
further manuscript edits, added new figures

@@
 
-Next, we directly examined the effects of varying both $\sigma$ and $\delta$ on the valley crossing time.
+Next, we directly examined the effects of varying both $\sigma$ and $\delta$ on the valley crossing time by varying them independently for fixed $N$, $\mu_1$, and $\mu_2$.
 
@@
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5a.pdf}
+%\label{fig:weissmana}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5b.pdf}
+%\label{fig:weissmanb}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5c.pdf}
+%\label{fig:weissmanc}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5d.pdf}
+%\label{fig:weissmand}
 \end{subfigure}
@@
 
-\iffalse
-\begin{figure}[t]
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
-\end{subfigure}
-\caption{Valley crossing times in the semi-deterministic tunneling regime as a function of $\sigma$.
-In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$; the right figure has a high enough value of $N\mu^2$ that fixation is almost deterministic.
-Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges.
-Each box represents about $100$ simulation runs.}
-\label{fig:deterministic}
-\end{figure}
-
-\begin{figure}[t]
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/seq_fix.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.4\textwidth}
-\includegraphics[width=\textwidth]{Figures/neut_tunnel.pdf}
-\end{subfigure}
-\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes as a function of $\sigma$.
-In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^4$ and $\mu = 3\times 10^{-5}$, and on the right, $N = 10^5$ and $\mu = 3\times 10^{-6}$.
-Labeling is as in figure \ref{fig:deterministic}.}
-\label{fig:tunneling}
-\end{figure}
-\fi
-
 Here we present the key results of our simulation analysis.
@@
 The general trend is that draft causes valley crossing to take more time but be less sensitive to the underlying parameter values.
-In figure (a), tunneling at low $\sigma$ occurs via neutral sequential fixation at low $N$ before transitioning into neutral tunneling and semi-deterministic tunneling at higher $N$; these are visible as minor kinks in the theoretical predictions (dotted line).
+In figure \ref{fig:weissman}a, tunneling at low $\sigma$ occurs via neutral sequential fixation at low $N$ before transitioning into neutral tunneling and semi-deterministic tunneling at higher $N$; these are visible as minor kinks in the theoretical predictions (dotted line).
 At low $\sigma$, our results agree with those predictions, but as $\sigma$ increases, the dependence on $N$ is lessened: at higher $\sigma$, the population effectively behaves as though there is a very small $N_e$, resulting in no dependence on $N$.
-These trends are also visible in figure (c) where, in the drift case, the population transitions from neutral to deleterious tunneling, and figure (d), in which deleterious tunneling gives way to neutral tunneling as the mutation rate $\mu_2$ increases.
+These trends are also visible in figure \ref{fig:weissman}c where, in the drift case, the population transitions from neutral to deleterious tunneling, and figure \ref{fig:weissman}d, in which deleterious tunneling gives way to neutral tunneling as the mutation rate $\mu_2$ increases.
 
-One remarkable difference between drift and draft can be seen in figure (b).
+One remarkable difference between drift and draft can be seen in figure \ref{fig:weissman}b.
 At $\sigma = 10^{-8}$, crossing proceeds via deleterious sequential fixation at low $N$, then transitions into deleterious tunneling around $N = 10^3$ and semi-deterministic tunneling at higher $N$.
@@
 
-\iffalse
-In general, we expect valley crossing to require more time in rapidly adapting populations (large $N \sigma$) experiencing genetic draft; bubbles are doomed to extinction more quickly (lineages experience larger fluctuations in size and decline in frequency as the mean fitness advances) and beneficial lineages are less likely to fix, especially when valleys are shallow.
-Figures \ref{fig:deterministic} and \ref{fig:tunneling} confirm these expectations: the overall rate of valley crossing is generally slowed as $\sigma$ increases.
-At higher $\sigma$, the effect of the genetic background in which an mutation finds itself becomes more significant in determining the fate of that mutation than the fitness effect of the mutation itself.
-This genetic draft buffets the frequency of mutations so that these mutations behave as though they are more neutral than they really are: in effect, the fitness landscape is ``flattened".
-This depresses the fixation probability of the double mutant and other complex adaptations, which appears to account for the majority of the change in fixation time.
-In the case of stochastic tunneling, the overall size distribution of bubble sizes is reduced as well: the distribution $P(w)$ scales not as $w^{-3/2}$ but as $w^{-2}$ \citep{neher_shraiman_2011}, so there are fewer chances for the complex adaptation to arise on a mutant background and fix.
-\fi
-
 \begin{figure}[t]
-\includegraphics[width=0.4\textwidth]{Figures/julia_sigma_delta.pdf}
+\includegraphics[width=0.5\textwidth]{Figures/julia_sigma_delta.pdf}
 \caption{Valley crossing times for varied $\sigma$ and $\delta$.
 In these simulations, $N = 10^5$, $\mu_1 = 3 \times 10^{-6}$, and $\mu_2 = 10^{-4}$.
-Note the lack of dependence on $\delta$ at high $\sigma$.}
+Note the near lack of dependence on $\delta$ at high $\sigma$.}
 \label{fig:sigma_delta}
@@
 
+Overall, we find that $\tau$ is mostly elevated at high $\sigma$, as genetic draft dooms new lineages to extinction quickly and slows the growth of the complex adaptation.
 On the other hand, the reduced effectiveness of selection should generally be helpful in crossing deeper valleys by mitigating the deleterious effect of single mutants.
-This can be seen in figure \ref{fig:sigma_delta}.
+This can be seen at intermediate values of $N$ in Figure \ref{fig:weissman}b and in figure \ref{fig:sigma_delta}.
 At high values of $\sigma$, the fitness effect of the deleterious intermediate mutation has very little effect on the valley crossing time, which means that deep valleys can be crossed.
@@
 \begin{figure*}[t]
-\begin{subfigure}[b]{0.5\textwidth}
-  \includegraphics[width=\textwidth]{Figures/tau_lowsig.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.5\textwidth}
-  \includegraphics[width=\textwidth]{Figures/tau_highsig.pdf}
-\end{subfigure}
+  \includegraphics[width=\textwidth]{Figures/log_tau_s0-1.pdf}
+  \caption{Valley crossing times on a $\log_{10}$ scale as a function of $\sigma$, $N$, and $\delta$. In all three plots, $s = 0.1$, $\mu_1 = 10^{-5}$, and $\mu_2 = 10^{-4}$.}
 \label{fig:tau}
-
-  \caption{Valley crossing times on a $\log_{10}$ scale}
-\begin{subfigure}[b]{0.5\textwidth}
-  \includegraphics[width=\textwidth]{Figures/ratio_lowsig.pdf}
-\end{subfigure}
-\begin{subfigure}[b]{0.5\textwidth}
-  \includegraphics[width=\textwidth]{Figures/ratio_highsig.pdf}
-\end{subfigure}
-\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$.
-In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, $\mu_1 = 3 \times 10^{-6}$, and $\mu_2 = 10^{-4}$.
-Panel (a) shows that valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted.
-Panel (b) shows the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.}
+\end{figure*}
+\begin{figure*}[t]
+  \includegraphics[width=\textwidth]{Figures/log_alpha_s0-1.pdf}
+\caption{The ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\sigma$, $N$, and $\delta$.
+In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, $\mu_1 = 10^{-5}$, and $\mu_2 = 10^{-4}$.}
 \label{fig:tau-ratios}
@@
 To explore this, we first introduce the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a single beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
-By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps.
-Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
+By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple (``hard") selective sweeps.
+Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple sweep.
 Though the precise value of this ratio depends on $\delta$ and on the selection coefficients $s_{\mathrm{valley}}$ and $s_{\mathrm{sweep}}$ of the complex adaptation and the single beneficial mutation, respectively, a general trend nonetheless emerges.
-In Figure \ref{fig:tau-ratios}, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
-Figure \ref{fig:ratios} shows the expected pattern that deeper valleys in smaller populations take longer to cross.
-At a higher value of $\sigma$, valley depth has a weaker effect on the valley crossing time in both an absolute sense (Figure \ref{fig:tau}) and relative to the time for a single beneficial sweep (Figure \ref{fig:ratios}).
+In Figure \ref{fig:tau-ratios}, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, $\mu_1 = 10^{-5}$, and $\mu_2 = 10^{-4}$.
+Figure \ref{fig:tau-ratios} shows the expected pattern that deeper valleys in smaller populations take longer to cross.
+At a higher value of $\sigma$, valley depth has a weaker effect on the valley crossing time in both an absolute sense (figure \ref{fig:tau}) and relative to the time for a single beneficial sweep (figure \ref{fig:ratios}).
 Moreover, values of $\alpha$ are generally lower for the population with higher $\sigma$ experiencing genetic draft than for the population with lower $\sigma$ experiencing only drift.
@@
 \begin{figure*}[t]
-\includegraphics[width=0.95\textwidth]{Figures/julia_ratios.pdf}
+  \includegraphics[width=\textwidth]{Figures/log_ratio_s0-1.pdf}
+\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$), where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2$ values are on the right of the plot.
+$s_{\mathrm{sweep}}$ varies: other parameter values are as in figures \ref{fig:tau} and \ref{fig:tau-ratios}.
+A value lower than $0$ indicates that draft favors valley crossing.
+Genetic draft is helpful for crossing deep valleys, especially at low or intermediate sizes.}
+\label{fig:alpha-ratios}
+\end{figure*}
+
+\begin{figure*}[t]
+\includegraphics[width=\textwidth]{Figures/log_taumu.pdf}
 \caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2= 5 \times 10^{-2}$), for varied values of $\mu$.
-Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
+Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.
+(Dummy caption.
+Need to fix.)}
 \label{fig:alpha_ratio}
@@
 
-We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha_{\sigma_2} / \alpha_{\sigma_1}$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2=5 \times 10^{-2}$, which is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
-The ratio can be seen in Figure \ref{fig:alpha_ratio}.
-Overall, we find that raising $\sigma$ increases the extent to which valley crossing is favored for lower population sizes: for this combination of parameters, the condition is $N < 10^{5}$.
+We can get a clearer picture of how draft affects valley crossing times relative to single beneficial sweeps by considering the ratio $\alpha_{\sigma_2} / \alpha_{\sigma_1}$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2$ is allowed to vary.
+This is a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
+The ratio can be seen in Figure \ref{fig:alpha-ratios}.
+Overall, we find that raising $\sigma$ increases the extent to which valley crossing is favored for lower population sizes: for this combination of parameters, the condition is $N < 10^5$.
 In this regime, raising $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
@@
 We have used simulation methods to examine the rate of crossing fitness valleys in rapidly adapting populations that experience genetic draft.
-We find that although the overall valley crossing time is longer in rapidly adapting than in slowly adapting (drifting) populations, a higher fraction of adaptive events in rapidly adapting populations will occur through valley crossing than in slowly adapting populations.
+We find that although the overall valley crossing time is longer in rapidly adapting than in slowly adapting (drifting) populations, a higher fraction of adaptive events in rapidly adapting populations may occur through valley crossing than in slowly adapting populations.
 This mean that the genetic draft that results from rapid adaptation accelerates the rate at which populations escape from local peaks in a fitness landscape and move towards global peaks.
-Thus, adaptation at loci unhindered by genetic draft occurs faster but is more likely to get stuck on a local peak in a rugged fitness landscape.
+Thus, adaptation at loci unhindered by genetic draft occurs faster, but the population is more likely to get stuck on a local peak in a rugged fitness landscape.
 
 The source of this difference in behavior between populations experiencing genetic draft and those experiencing drift comes from two effects.
-First, the effect of the deleterious intermediate in the rapidly adapting population is much smaller: for $\delta \lesssim \sigma$, the precise value of $\delta$ almost does not affect the crossing rate.
+First, the effect of the deleterious intermediate in the rapidly adapting population is much smaller: for $\delta \ll \sigma$, the precise value of $\delta$ does not affect the crossing rate.
 Second, the time to the evolution of even \emph{simple} adaptations (i.e., ones requring substitution only at a single locus) is much longer in rapidly adapting populations.
@@
 They considered the scenario where the sweeping beneficial mutation \emph{competes} with a particular complex adaptation, whereas we treat these two scenarios separately.
-They find that large populations are more likely to cross valleys than fix a simple beneficial mutation than small populations (for comparison, see the $\sigma=5 \times 10^{-6}$ panel in our Figure \ref{fig:ratios}).
+They find that large populations are more likely to cross valleys than fix a simple beneficial mutation than small populations (for comparison, see the $\sigma=5 \times 10^{-8}$ panel in our Figure \ref{fig:tau-ratios}).
 However, they also find that intermediate-sized populations in which stochastic tunneling occurs are the least likely to cross a fitness valley instead of fixing a beneficial mutation.
-In contrast, our results suggest that the stochastic tunneling regime is where rapid adaptation and genetic draft increase the likelihood of valley crossing relative to a simple beneficial mutation.
+In contrast, our results suggest that the stochastic tunneling regime is fairly narrow in rapidly adapting populations, and draft can aid the crossing of fitness valleys via sequential fixation of mutations.
 Thus, it is possible that genetic draft improves the likelihood of valley crossing precisely where it is least likely under genetic drift.
@@
 In contrast, high migration rates and high recombination rates \citep{neher_shraiman_2009} allow selection to more easily remove deleterious intermediates.
-This commonality has implications for Wright's ``shifting balance theory" \citep{Wright:1932}, which posits that population subdivision can enhance evolution on rugged fitness landscapes: deleterious intermediates accumulate due to genetic drift induced by subdivision, and subsequent beneficial mutations fix both in locally and are spread globally due to migration.
+This commonality has implications for Wright's ``shifting balance theory" \citep{Wright:1932}, which posits that population subdivision can enhance evolution on rugged fitness landscapes: deleterious intermediates accumulate due to genetic drift induced by subdivision, and subsequent beneficial mutations fix locally and are spread globally due to migration.
 However, it has long been known that the shifting balance theory has important limitations due to the fact that too much migration prevents the accumulation of deleterious intermediates and too little reduces the success of genotypes from the fitter peaks and prevents them from spreading throughout the metapopulation \citep{coyne_barton_turelli_2000,Van-Cleve:Weissman:2015}.
@@
 In recent years of the HIV pandemic, resistance has generally been slower to evolve \citep{feder_2015} because more and stronger drugs are used.
-There are two possible explanations for this.
+There are several possible explanations for this.
 One is that the use of anti-HIV drugs decreases the number of virions segregating within an individual, thus lowering the number of possible chances for a resistant phenotype to appear.
-The other is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.
+Another is that such drugs, especially when used in concert, contort the fitness landscape so that evolution of a phenotype that is resistant to a drug cocktail is more difficult.
 Both factors are likely to play a role.
-However, since HIV is a population in which draft, not drift, is the dominant stochastic force and hence population size is less important in determining the evolutionary dynamics, our work suggests that the second factor is likely to be more significant in HIV compared to organisms in which draft is less important.
+However, since HIV is a population in which draft, not drift, is the dominant stochastic force and hence population size is less important in determining the evolutionary dynamics, our work suggests that the second factor is likely to be more significant in HIV compared to organisms in which draft is less important; indeed, the population size turns out to be less and less important as the effect of draft is ramped up.
 

### 2019-05-15 ms/valley_crossing.tex
revising discussion and figure captions

@@
 \begin{figure*}[t]
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[t]{0.48\textwidth}
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5a.pdf}
@@
 \end{subfigure}
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.48\textwidth}
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5b.pdf}
@@
 \end{subfigure}
-\begin{subfigure}[c]{0.5\textwidth}
+\begin{subfigure}[c]{0.48\textwidth}
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5c.pdf}
@@
 \end{subfigure}
-\begin{subfigure}[d]{0.5\textwidth}
+\begin{subfigure}[d]{0.48\textwidth}
 \includegraphics[width=\textwidth]{Figures/julia_weissman_5d.pdf}
@@
 \caption{Valley crossing times across a range of parameter values mirroring those of \citet{weissman_2009}.
+In each case, we set $s = 0.1$.
 In figures (a) and (b), the population size is varied: increasing the size causes valley crossing to transition from tunneling to semi-deterministic fixation.
-In figure (c), the valley depth $\delta$ is varied: at negative $\delta$, our simulations closely match the expectation for beneficial mutations.
-In figure (d), the intermediate mutation rate $\mu_2$ is varied.
+In (a), $\mu_1 = 10^{-5}$, $\mu_2 = 10^{-4}$, and $\delta = 2 \times 10^{-4}$, so that the valley depth is effectively neutral; in (b), $\mu_1 = 10^{-5}$, $\mu_2 = 10^{-6}$, and $\delta = 7 \times 10^{-3}$.
+In figure (c), we set $N = 10^5$, $\mu_1 = 10^{-6}$, and $\mu_2 = 4 \times 10^{-5}$.
+The valley depth $\delta$ is varied: at negative $\delta$, our simulations closely match the expectation for beneficial mutations.
+In figure (d), the intermediate mutation rate $\mu_2$ is varied, with $N = 10^{5}$, $\mu_1 = 10^{-6}$, and $\delta = 3 \times 10^{-3}$.
 Dotted lines represent the theoretical predictions from \citet{weissman_2009}.}
@@
 Figure \ref{fig:weissman} summarizes the major differences between the drift (low $\sigma$) and draft (high $\sigma$) regimes in valley crossing.
-Note that the theoretical predictions differ slightly from those of \citet{weissman_2009} due to their use of a Moran model, compared to our Wright-Fisher model: this is comparable to an order of 2 difference in the population size.
+Note that the theoretical predictions differ slightly from those of \citet{weissman_2009} due to their use of a Moran model, compared to our modified Wright-Fisher model: this is comparable to an order of 2 difference in the population size.
 The general trend is that draft causes valley crossing to take more time but be less sensitive to the underlying parameter values.
@@
 \includegraphics[width=\textwidth]{Figures/log_taumu.pdf}
-\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=5 \times 10^{-8}$ and $\sigma_2= 5 \times 10^{-2}$), for varied values of $\mu$.
-Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.
-(Dummy caption.
-Need to fix.)}
-\label{fig:alpha_ratio}
+\caption{The product of the crossing time $\tau$ and mutation rate $\mu_1$ at the first locus, as a function of $s$ and $\sigma$.
+The dotted line is $s = \sigma$.
+When $\sigma \gg s$, the genetic background dwarfs the fitness effect of the complex adaptation, and fixation is essentially neutral; when $\sigma \ll s$, the fixation time decreases with $s$.
+Note that, in the middle column, $s_{\mathrm{valley}} = s_{\mathrm{sweep}}$.}
+\label{fig:tau-mu}
 \end{figure*}
@@
 The ratio can be seen in Figure \ref{fig:alpha-ratios}.
-Overall, we find that raising $\sigma$ increases the extent to which valley crossing is favored for lower population sizes: for this combination of parameters, the condition is $N < 10^5$.
+Overall, we find that raising $\sigma$ increases the extent to which valley crossing is favored, especially for lower population sizes.
 In this regime, raising $\delta$ also increases the degree to which valley crossing is favored under draft, which means that the deeper the valley, the more that draft increases the likelihood of valley crossing relative to a single beneficial mutation.
-The exception is at very high values of $N$ and $\delta$ and low values of $\mu$, where draft turns out again not to be helpful; see the lower right part of the leftmost plot.
-In this regime, $\delta$ is on the order of $\sigma$ in our draft simulations, meaning that single mutant lineages are rapidly cut off by the advancing mean fitness and are likely almost never present in the nose of the fitness distribution.
-Essentially, the population size is large enough that intermediate mutants can be ``seen" by selection, but the fitness detriment is severe enough compared to $\sigma$ that draft is no longer helpful in allowing mutant lineages to reach high frequency.
-This problem disappears as $\mu$ increases (center and right plots) and the population is beyond the deleterious tunneling barrier where the valley crossing time depends on the depth of the valley $\delta$.
+The effect being exacerbated at low $N$ suggests that deleterious intermediates are surfing to very high frequency as the fitness wave advances.
+The middle column is of special interest, as $s_{\mathrm{valley}} = s_{\mathrm{sweep}}$.
+
+We further probe the effects of increasing $\sigma$ in Figure \ref{fig:tau-mu}.
+At low values of $\sigma$, a strongly selected complex adaptation (high $s$) is essentially far ahead of the fitness wave and behaves as though governed by genetic drift, not draft.
+This is consistent with the treatment of \citet{good_2012}, who found that, for beneficial lineages whose fitness $x + s$ is very far ahead of a critical fitness $x_c$, the fixation probability effectively scales with $x+s$.
+Raising $\sigma$ far above $s$ (i.e., above the dotted line) causes the effect of $s$ to be overwhelmed by the advancing mean fitness, and fixation appears to be effectively neutral.
 

### 2019-05-24 ms/valley_crossing.tex
adjust fig 2

@@
 An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.
+Intermediates can then drift to high frequency or even fixation, increasing the rate at which complex adaptations arise.
 
 Previous studies of valley crossing focus primarily on populations where all individuals are equal in fitness except for the focal loci, i.e., the loci at which the individual mutations comprising the complex adaptation segregate.
-This is tantamount to assuming that, if there is variation in the background fitness of the population, it is negligible.
+This is tantamount to assuming that variation in the background fitness of the population is negligible.
 In such populations, genetic drift governs the fate of neutral alleles, as well as the behavior of deleterious or beneficial alleles where they are rare (close to frequency zero) or common (close to frequency one).
@@
 When population sizes are very large or the fitness variation in the population is substantial, however, the behavior of genetic variation is governed more by genetic \emph{draft} than by genetic drift \citep{gillespie_2000, gillespie_2001, masel_2011, neher_shraiman_2011}.
-Whereas drift is the effect of imperfect sampling from generation to generation, draft is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
+Broadly speaking, drift results from the fact that each generation is imperfectly sampled from the previous generation.
+Draft, on the other hand, is the sum of hitchhiking effects due to selection on sites linked to the focal loci.
 The standard technique of defining a reduced ``effective population size" fails to capture several fundamental qualitative differences between drift and draft.
@@
 The possibility that neutral variants are affected more strongly by selection at linked sites than by genetic drift, even in organisms like \emph{Drosophila}, is one possible explanation for the ``paradox of variation", the fact that genetic diversity and population size often do not scale linearly \citep{gillespie_2000, gillespie_2001, neher_kessinger_2013, corbett-detig_2015}.
-Draft likewise may have profound effects on the evolution of complex adaptations. Preliminary evidence for this comes from
-\citet{neher_shraiman_2011}, who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting facultatively sexual population.
+Draft likewise may have profound effects on the evolution of complex adaptations.
+Preliminary evidence for this comes from \citet{neher_shraiman_2011}, who explored how draft affects valley crossing via stochastic tunneling by calculating the mutant bubble size distribution in a rapidly adapting facultatively sexual population.
 They found that, compared to drift, draft generally causes the distribution of bubble sizes to drop off more quickly, meaning that bubbles tend to be smaller overall: however, as the mutant lineage's fitness is determined not only by a mutation's intrinsic fitness effect but also the genetic background on which it appears, the fitness effect of a focal mutation plays less of a role in capping the size of the lineage.
-Thus, while draft may make valley crossing via tunneling more difficult, the total effect isn't immediately obvious.
-Here, we extend these previous results by studying how genetic draft in asexual populations affects the rate of crossing fitness valleys across a range of population sizes.
+Thus, while draft may make valley crossing via tunneling more difficult, the total effect is not immediately obvious.
+Here, we extend these results by studying how genetic draft in asexual populations affects the rate of crossing fitness valleys across a range of population sizes, including ones in which tunneling is not the dominant mode of valley crossing.
 
@@
 In previous approaches, these stochastic effects were smoothed due to the presence of recombination \citep{neher_shraiman_2011, neher_shraiman_2009}: the critical nose dynamics are still important in a sexual population, but the fitness distribution is more Gaussian overall.
+Moreover, a full understanding of the behavior of complex adaptations requires that one consider the dynamics of individual lineages.
+The equations that describe said dynamics, however, are very difficult to solve; \citet{neher_shraiman_2011} were able to circumvent this issue by carefully choosing a recombination model that allowed them to avoid the need to directly solve these equations.
 We therefore must focus primarily on simulation approaches, as analytical solutions for the behavior of a mutation in the nose of an asexual population are very difficult to obtain.
@@
 In addition, we confirm that in rapidly adapting populations, adaptive fixations of alleles are more likely to be complex adaptations involving a fitness valley than they are in populations where genetic drift is the primary force shaping genetic variation at linked sites; essentially, genetic draft maintains the linkage that makes it possible to leap across fitness landscapes rather than adapt primarily by climbing to local peaks.
-These observations add to the growing intuition that complex adaptations that involve evolution across fitness valleys may not be a surprising result of the evolutionary process, but rather an expected one.
+These observations add to the growing intuition that complex adaptations that involve evolution across fitness valleys may not be a surprising result of the evolutionary process but rather an expected one.
 
@@
 Let $x$ be the background fitness of a lineage.
-If the number of background loci is large, the distribution of background fitnesses $f(x)$ will generally be a series of (Dirac) $\delta$ functions, but its ensemble average will be roughly Gaussian \citep{good_desai_2014}: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
+If the number of background loci is large, the distribution of background fitnesses $f(x)$ will generally be a series of (Dirac) $\delta$ functions, but its ensemble average will be roughly Gaussian \citep{good_2014}: $f(x) \approx \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
 By Fisher's ``fundamental theorem", the mean fitness $\bar{x}$ will advance due to selection at a rate equal to $\sigma^2$: the total rate of advance will depend also on contributions from drift and the average effect of mutations.
@@
 \citet{weissman_2009} divided the possible modes of asexual valley crossing into several distinct regimes.
-We classify them, broadly, as \emph{sequential fixation}, \emph{deterministic fixation}, and \emph{tunneling}.
+We classify them, broadly, as \emph{sequential fixation}, \emph{deterministic fixation}, and \emph{stochastic tunneling}.
 What follows is a brief outline of the process of valley crossing in each of these regimes, as well as how these are likely to differ in our model.
@@
 $\tau_2$, similarly, is approximately one divided by $N\mu_2 s$ and in general will be dwarfed by $\tau_1$.
-When genetic draft is more important than genetic drift, $p$ in $\tau_1$ takes on a much more complicated form similar to that in \citet{good_desai_2014}: in general, the dependence on $\delta$ is lessened provided that $\sigma \gtrsim \delta$, as deleterious mutations that arise on significantly fitter-than-average backgrounds can be destined to fix.
+When genetic draft is more important than genetic drift, $p$ in $\tau_1$ takes on a much more complicated form similar to that in \citet{good_2014}: in general, the dependence on $\delta$ is lessened provided that $\sigma \gtrsim \delta$, as deleterious mutations that arise on significantly fitter than average backgrounds can be destined to fix.
 $\tau_2$ likewise has lessened dependence on $s$, as a complex adaptation that does \emph{not} arise on a fitter-than-average background lineage may be doomed to extinction.
 
-There are three components to the process of stochastic tunneling.
+The process of \emph{deterministic fixation} occurs when the number of single mutants introduced every generation is sufficiently large: $N\mu_1 \gg 1$.
+One possibility is that the intermediate exists in what is roughly mutation-selection balance, so that it persists at a low frequency.
+Lucky complex adaptations can then arise on this background of roughly constant intermediate influx, then sweep to fixation.
+This is known as \emph{semi-deterministic tunneling}; we class it under the ``deterministic" umbrella because it is not necessary to consider the stochastic dynamics of individual deleterious mutant lineages.
+Switching from drift to draft will affect the likelihood that a particular complex adaptation will sweep to fixation, but it seems unlikely to affect the mutation-selection balance dynamics for intermediates.
+The strictest ``deterministic fixation" case occurs when $N \gg 2s/\pi \mu_1 \mu_2$.
+In this case, multiple complex adaptations are likely to arise on independent backgrounds, and there is no guarantee that the first one to reach significant frequency will sweep through the population on its own.
+In the draft case, we expect the fixation time to be increased due to the blunting effect of the genetic background.
+However, note that the $N \gg 2s/\pi \mu_1 \mu_2$ cutoff is generally very large: for $\mu_1 = \mu_2 = 10^{-4}$ and $s = 0.1$, we have $N \approx 10^7$, a size that will prove difficult to probe with our simulation model.
+
+Finally, we consider the intermediate case: \emph{stochastic tunneling}.
+This process has three major components.
 First, a single mutant lineage must appear, which occurs with rate $N\mu$.
@@
 This rescaling means that the assigned initial value of $\epsilon$ turns out not to matter, even if it is negative or even if both deleterious and beneficial background mutations are permitted.
-That feature of our model may seem counterintuitive, but previous work has found that populations experiencing strong background selection at many loci of weak effect (see, e.g., \citet{cvijovic_2015}) exhibit dynamics very similar to rapidly adapting populations \citep{desai_fisher_2007, neher_shraiman_2011}.
+That feature of our model may seem counterintuitive, but previous work has found that populations experiencing strong background selection at many loci of weak effect (see, e.g., \citet{cvijovic_2018}) exhibit dynamics very similar to rapidly adapting populations \citep{desai_fisher_2007, neher_shraiman_2011}.
 
@@
 
-\begin{figure*}[t]
-\begin{subfigure}[t]{0.48\textwidth}
-\includegraphics[width=\textwidth]{Figures/julia_weissman_5a.pdf}
+\begin{figure*}[!t]
+  \begin{tabular}{r@{\hspace{-1ex}}r}
+\includegraphics[width=0.48\textwidth]{Figures/julia_weissman_5a.pdf} &
 %\label{fig:weissmana}
-\end{subfigure}
-\begin{subfigure}[b]{0.48\textwidth}
-\includegraphics[width=\textwidth]{Figures/julia_weissman_5b.pdf}
+\includegraphics[width=0.48\textwidth]{Figures/julia_weissman_5b.pdf}\\[-3ex]
 %\label{fig:weissmanb}
-\end{subfigure}
-\begin{subfigure}[c]{0.48\textwidth}
-\includegraphics[width=\textwidth]{Figures/julia_weissman_5c.pdf}
+\includegraphics[width=0.48\textwidth]{Figures/julia_weissman_5c.pdf} &
 %\label{fig:weissmanc}
-\end{subfigure}
-\begin{subfigure}[d]{0.48\textwidth}
-\includegraphics[width=\textwidth]{Figures/julia_weissman_5d.pdf}
-%\label{fig:weissmand}
-\end{subfigure}
+\includegraphics[width=0.48\textwidth]{Figures/julia_weissman_5d.pdf}
+% \label{fig:weissmand}
+  \end{tabular}
   \caption{Valley crossing times across a range of parameter values mirroring those of \citet{weissman_2009}.

### 2023-02-27 manuscript/response_to_reviews.tex
rename folder

--- 2019-01-24 ms/response_to_reviews.tex
@@
 
-We have redone most of the simulations in this paper using \texttt{WaveFit}, a custom simulation package written in Julia.
-Both the original \texttt{FFPopSim} simulations and the new ones are individual-based, though groups of individuals identical in fitness are abstracted away as "clones" during the natural selection step.
-We have made this clearer.
+We have redone most of the simulations in this paper using \texttt{WaveFit}, a custom simulation package written in Julia for running individually-based simulations using an infinite sites model for the background fitness linked to two focal loci.
+Though the original \texttt{FFPopSim} simulations were individual-based, we have replaced our results completely with results from \texttt{WaveFit}.
+%though groups of individuals identical in fitness are abstracted away as "clones" during the natural selection step. We have made this clearer.
 
@@
 
-Neher and Shraiman (2011) determined that the opposite trend applies in the case where $1/N < \delta \ltrsim sigma$.
+Neher and Shraiman (2011) determined that the opposite trend applies in the case where $1/N < \delta \lesssim sigma$.
 In this case, the stochastic effects of genetic draft can propel a deleterious mutation to high frequencies, allowing valley crossing to occur more quickly.
